### 2017-12-06 manuscript/introduction_notes.tex
added current draft of the manuscript

\subsection*{Molecular clock dating \textit{underestimates} divergence dates for many different viruses.}

Underestimation might be extreme in RNA  viruses because of their mutation rate~\citep{holmes2003molecular}.

\textbf{\textit{Purifying Selection Can Obscure the Ancient Age of Viral Lineages}}\citep{wertheim2011purifying}
This paper uses simulations as well as sequences from Ebola, Measles, and Influenza to show how codon-based substitution models (specifically with site-to-site and lineage-to-lineage variation of non-synonymous to synonymous substitution rates.)
\begin{quote}
Statistical methods for molecular dating of viral origins have been used extensively to infer the time of most common recent ancestor for many rapidly evolving pathogens. However, there are a number of cases, in which epidemiological, historical, or genomic evidence suggests much older viral origins than those obtained via molecular dating. We demonstrate how pervasive purifying selection can mask the ancient origins of recently sampled pathogens, in part due to the inability of nucleotide-based substitution models to properly account for complex patterns of spatial and temporal variability in selective pressures. We use codon-based substitution models to infer the length of branches in viral phylogenies; these models produce estimates that are often considerably longer than those obtained with traditional nucleotide-based substitution models. Correcting the apparent underestimation of branch lengths suggests substantially older origins for measles, Ebola, and avian influenza viruses. This work helps to reconcile some of the inconsistencies between molecular dating and other types of evidence concerning the age of viral lineages.
\end{quote}

\subsubsection*{Actual examples}

There are many examples with real viruses sequences which show either a) that the current molecular clock method seems to drastically underestimate divergence times or b) substitution models which incorporate rate variation or purifying selection produces longer branches and therefore early divergence estimates.

\textbf{HIV}

HIV appears to be the best studied example of the molecular clock underestimation. 
While molecular clock methods estimate the common ancestor of HIV-1 and HIV-2 to be several thousand of years, the co-divergence theory of SIV and non-human primates would be the estimates around several million years ago.

It seems that there is more and more evidence that the co-divergence hypothesis is probably not completely right but molecular clock methods are probably still underestimating the dates.

\comment{There are more references for this section but I was working by way through the Wertheim bib.}
\comment{I also need to find the biogeography paper}

\textbf{\textit{A Challenge to the Ancient Origin of SIVagm
Based on African Green Monkey
Mitochondrial Genomes}}\citep{wertheim2007challenge}
This paper used more genomes of AGM to test the hypothesis that AGM and SIVagm co-diverged. They found that that mitochondrial AGM tree did not match the SIV tree. 
\begin{quote}
While the circumstances surrounding the origin and spread of HIV are becoming clearer, the particulars of the origin of simian immunodeficiency virus (SIV) are still unknown. Specifically, the age of SIV, whether it is an ancient or recent infection, has not been resolved. Although many instances of cross-species transmission of SIV have been documented, the similarity between the African green monkey (AGM) and SIVagm phylogenies has long been held as suggestive of ancient codivergence between SIVs and their primate hosts. Here, we present well-resolved phylogenies based on full- length AGM mitochondrial genomes and seven previously published SIVagm genomes; these allowed us to perform the first rigorous phylogenetic test to our knowledge of the hypothesis that SIVagm codiverged with the AGMs. Using the Shimodaira?H	asegawa test, we show that the AGM mitochondrial genomes and SIVagm did not evolve along the same topology. Furthermore, we demonstrate that the SIVagm topology can be explained by a pattern of west-to-east transmission of the virus across existing AGM geographic ranges. Using a relaxed molecular clock, we also provide a date for the most recent common ancestor of the AGMs at approximately 3 million years ago. This study substantially weakens the theory of ancient SIV infection followed by codivergence with its primate hosts.
\end{quote}

\textbf{\textit{Dating the Age of the SIV Lineages That Gave Rise to HIV-1 and HIV-2}}\citep{wertheim2009dating}
This paper showed that SIVsmm had an indistinguishable rate estimate to HIV-2 and that either SIV is very young or that both HIV and SIV dates are severally underestimated. 
\begin{quote}
Great strides have been made in understanding the evolutionary history of simian immunodeficiency virus (SIV) and the zoonoses that gave rise to HIV-1 and HIV-2. What remains unknown is how long these SIVs had been circulating in non- human primates before the transmissions to humans. Here, we use relaxed molecular clock dating techniques to estimate the time of most recent common ancestor for the SIVs infecting chimpanzees and sooty mangabeys, the reservoirs of HIV-1 and HIV-2, respectively. The date of the most recent common ancestor of SIV in chimpanzees is estimated to be 1492 (1266? 1685), and the date in sooty mangabeys is estimated to be 1809 (1729?1875). Notably, we demonstrate that SIV sequences sampled from sooty mangabeys possess sufficient clock-like signal to calibrate a molecular clock; despite the differences in host biology and viral dynamics, the rate of evolution of SIV in sooty mangabeys is indistinguishable from that of its human counterpart, HIV-2. We also estimate the ages of the HIV-2 human-to-human transmissible lineages and provide the first age estimate for HIV-1 group N at 1963 (1948?1977). Comparisons between the SIV most recent common ancestor dates and those of the HIV lineages suggest a difference on the order of only hundreds of years. Our results suggest either that SIV is a surprisingly young lentiviral lineage or that SIV and, perhaps, HIV dating estimates are seriously compromised by unaccounted-for biases.
\end{quote}

\textbf{\textit{Island Biogeography Reveals the Deep History of SIV}}\citep{worobey2010island}
This paper sequenced SIV from non-human primates which have been geographically isolated from humans for a long time. 
The SIV formed a monophyletic group on the tree with the SIV from the mainland which suggests that the virus had infected the most recent common ancestor of the species. 
This disagrees with molecular clock estimates.

\textbf{Flu}

\comment{I know there are more flu papers, I just have not added them.}

\citep{wertheim2011purifying} looks at Eboloa, Measles, and IAV sequences. 
They are more concerned with \textit{dating} Ebola and Measles but show that the general patterns of purifying selection masking substitutions along long branches exists in IAV NA as well.

\textbf{EBOV}
\citep{wertheim2011purifying} looks at Eboloa, Measles, and IAV sequences.

\textbf{MeV}
\citep{wertheim2011purifying} looks at Eboloa, Measles, and IAV sequences.

\textbf{Coronaviruses}

\textbf{\textit{A Case for the Ancient Origin of Coronaviruses}}\citep{wertheim2013case}
\begin{quote}
Coronaviruses are found in a diverse array of bat and bird species, which are believed to act as natural hosts. Molecular clock dating analyses of coronaviruses suggest that the most recent common ancestor of these viruses existed around 10,000 years ago. This relatively young age is in sharp contrast to the ancient evolutionary history of their putative natural hosts, which began diversifying tens of millions of years ago. Here, we attempted to resolve this discrepancy by applying more realistic evolutionary models that have previously revealed the ancient evolutionary history of other RNA viruses. By explicitly modeling variation in the strength of natural selection over time and thereby improving the modeling of substitution saturation, we found that the time to the most recent ancestor common for all coronaviruses is likely far greater (millions of years) than the previously inferred range.
\end{quote}

\textbf{HSV}

\textbf{\textit{Evolutionary Origins of Human Herpes Simplex Viruses 1 and 2}}\citep{wertheim2014evolutionary}
\begin{quote}
Herpesviruses have been infecting and co-diverging with their vertebrate hosts for hundreds of millions of years. The
primate simplex viruses exemplify this pattern of virus?host co-divergence, at a minimum, as far back as the most recent
common ancestor of New World monkeys, Old World monkeys, and apes. Humans are the only primate species known to
be infected with two distinct herpes simplex viruses: HSV-1 and HSV-2. Human herpes simplex viruses are ubiquitous,
with over two-thirds of the human population infected by at least one virus. Here, we investigated whether the additional
human simplex virus is the result of ancient viral lineage duplication or cross-species transmission. We found that
standard phylogenetic models of nucleotide substitution are inadequate for distinguishing among these competing
hypotheses; the extent of synonymous substitutions causes a substantial underestimation of the lengths of some of
the branches in the phylogeny, consistent with observations in other viruses (e.g., avian influenza, Ebola, and coronaviruses).
To more accurately estimate ancient viral divergence times, we applied a branch-site random effects likelihood
model of molecular evolution that allows the strength of natural selection to vary across both the viral phylogeny and the
gene alignment. This selection-informed model favored a scenario in which HSV-1 is the result of ancient co-divergence
and HSV-2 arose from a cross-species transmission event from the ancestor of modern chimpanzees to an extinct Homo
precursor of modern humans, around 1.6 Ma. These results provide a new framework for understanding human herpes
simplex virus evolution and demonstrate the importance of using selection-informed models of sequence evolution when
investigating viral origin hypotheses.
\end{quote}

\textbf{Herpes}

\textbf{Foamy virus}

\subsubsection*{TDRP - generally}

\subsection*{Substitution rate misspecification - generally}

\subsubsection*{Accounting for site-to-site variation, purifying selection}

\citep{halpern1998evolutionary}
\citep{brown1982mitochondrial}

\subsection*{other reasons why the divergence timing might be off.}

### 2017-12-06 manuscript/manuscript.tex
added current draft of the manuscript

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[11pt]{article}

\usepackage[round,semicolon]{natbib}
\usepackage{etoolbox}
\AtBeginEnvironment{quote}{\singlespacing\tiny}
% Use times if you have the font installed; otherwise, comment out the
% following line.

% added by SKH
%\usepackage{lineno}
%\linenumbers

\usepackage{times}
\usepackage{amssymb}
\usepackage{amsmath}

\usepackage[export]{adjustbox}

\usepackage{graphicx}
\graphicspath{ {images/} }

% for adjustwidth
\usepackage{changepage}

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

\usepackage{newfloat}
\usepackage{amsmath}
\usepackage[labelfont=bf]{caption}
\usepackage{nameref}
\usepackage{rotating}
\usepackage{color}
\usepackage{float}
\renewcommand{\figurename}{{}}
\renewcommand{\thefigure}{{Figure \arabic{figure}}}

\renewcommand{\tablename}{{}}
\renewcommand{\thetable}{{Table \arabic{table}}}

\newfloat{suppfile}{thp}{losuppfile}
\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
\floatname{suppfile}{}

\newfloat{suppfig}{thp}{losuppfig}
\renewcommand{\thesuppfig}{Supplementary figure \arabic{suppfig}}
\floatname{suppfig}{}

%
\newfloat{supptable}{thp}{losupptable}
\renewcommand{\thesupptable}{Supplementary table \arabic{supptable}}
\floatname{supptable}{}
%

\renewcommand{\theequation}{Equation \arabic{equation}}

\newcommand{\mutDNA}{\textbf{mutDNA}}
\newcommand{\mutvirus}{\textbf{mutvirus}}
\newcommand{\DNA}{\textbf{DNA}}
\newcommand{\virus}{\textbf{virus}}

\newcommand\comment[1]{{\color{magenta}#1}}

\usepackage{hyperref}
\hypersetup{colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue}
\hypersetup{colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue}

\usepackage{seqsplit}

\usepackage{array}
\newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}

\title{Experimentally Informed Site-Specific Substitution Models Substantially Deepen Divergence Estimates}

\author
{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2,}$\\
\\
\normalsize{$^1$Division of Basic Sciences and Computational Biology Program,}\\
\normalsize{Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}\\
\normalsize{$^2$Department of Genome Sciences, University of Washington, Seattle, WA}\\
\normalsize{E-mail:  jbloom@fredhutch.org.}\\
}

% Include the date command, but leave its argument blank.

\date{}

\usepackage{setspace}
\onehalfspacing

\begin{document}

% Double-space the manuscript.

% Make the title.

\maketitle

\begin{abstract}
\textit{$\leq$ 250 words, currently 159}

Molecular dating techniques have been used to estimate the divergence date of many viruses. 
However, these estimates are consistently substantially younger than estimates from methods which are independent of branch length estimation. 
This discrepancy is caused, in part, by inadequate modeling of purifying selection leading to branch length underestimation. 
Here, we show that substitution models informed by empirical measurements of mutational constraint better than traditional models and extend branch lengths. 
We used models informed by deep mutational scanning experiments performed in two, highly diverged influenza virus hemagglutinin homologs to optimize the branch lengths of a phylogenetic tree. 
For each experimentally informed model, we observed extension in branch length from the experiment's focal sequence. 
This extension in branch length due to explicit modeling of site-specific purifying selection is observed in the presence and absence of standard methods for modeling site-to-site variation. 
Overall, this study underscores the importance of modeling purifying selection when estimating branch lengths and, by extension, divergence dates.

\end{abstract}

\clearpage

\section*{Introduction}

\textit{Introduction outline}    
\textit{1. Estimating the divergence date of viruses is important and common.}  
\textit{2. There are numerous examples of molecular dating techniques contradicting other dating methods for viruses.}  
\textit{3. Why purifying selection could cause branch length underestimation.}  
\textit{4. Discussion of the attempts/methods made thus far to model site-to-site rate variation.}  
\textit{5. We use empirical measurements to model site-specific purifying selection. }  
\textit{6. In this study we compare branch lengths of trees optimized by GY94 and ExpCM models and see an extension in branch lengths with ExCM and with $\Gamma\omega$. }

\textbf{Estimating the age of viruses is important to understanding their evolutionary history, including co-evolution with their host species.}
Molecular dating techniques are commonly used to estimate the dates of viruses. 
However, these dates are consistently younger than the dates obtained by other methods which are independent of phylogenetic trees, often by orders of magnitude. 
\textit{example HIV/SIV, foamy virus, or measles}

\textbf{Failing to account for site-specific purifying selection results in an underestimation of branch lengths.} 
It has long been observed that protein-coding genes have a non-uniform distribution of amino-acid frequencies across the sites in the protein. 
Multiple sequence alignments of homologs show some sites which are conserved and appeared to be mutationally constrained and other sites which are variable and appear to be mutationally tolerant. 
This site-specific purifying selection dictates the expectation of change at a given site which is translated into branch length. 
\textit{The strength of purifying selection at a given site gives the expectation of change which directly relates to branch length. 
Specifically, failing to account for constraint will lead to an underestimation in branch length because your expectation of change will be too high.}

\textbf{Different substitution models have been developed to address the effect of site-specific purifying selection.}
Most commonly, a rate parameter is described by some statistical distribution fit across the entire protein, such as the $\Gamma$-distributed $\omega$ in the YNGKP M5 \citep{yang2000codon}. 
Other approaches have tried to explicitly model the site-specific amino-acid frequencies, such as mutation-selection models \citep{halpern1998evolutionary}. 
While perhaps the most biologically-relevant, mutation-selection models are heavily parametrized and run the risk of overfitting in most practical applications. 
However, we have previously shown that we can model this site-specific purifying selection using empirical measurements of mutational constraint from a high-throughput assay called deep mutational scanning. 
We hypothesize that using these experimentally informed models will estimate longer branches than traditional models.

\textbf{Here, we address this hypothesis by comparing the branch lengths on phylogenetic trees describing influenza virus hemagglutinin optimized by different substitution models.}
By comparing the trees optimized with and without the experimental descriptions of purifying selection, we show that failing to account for site-specific constraints results in shorter estimations of branch length. 
We used two different experimentally informed models, defined by experimental measurements from one of two diverged homologs. 
For each model, the branchs from the experimental focal sequence are extended compared to traditional models.  
This branch length extension is seen even in the presence of model parameters which are traditionally used to describe site-to-site rate variation.  
These results underscore the importance of modeling purifying selection when estimating divergence dates and suggest that experimental measurement may be used to account for this selection while avoiding overparameterization.

\section*{Results and Discussion}

\subsection*{Substitution models}
\subsubsection*{GY94 models}
\subsubsection*{ExpCMs}
We recap the \textbf{Exp}erimentally Informed \textbf{C}odon \textbf{M}odel (ExpCM) \citep{bloom2014experimentally,bloom2014informed,bloom2017identification,hilton2017phydms} to introduce nomenclature.

In an ExpCM, rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
\begin{equation}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form:
\begin{equation}
Q_{xy} = 
\begin{cases}
\phi_w & \mbox{if $x$ and $y$ differ by a transversion to nucleotide $w$} \\
\kappa \phi_w & \mbox{if $x$ and $y$ differ by a transition to nucleotide $w$} \\
0 & \mbox{if $x$ and $y$ differ by $>1$ nucleotide.}
\end{cases}
\end{equation}
The $\kappa$ parameter represents the transition-transversion ratio, and the $\phi_w$ values give the expected frequency of nucleotide $w$ in the absence of selection on amino-acid substitutions, and are constrained by $1 = \sum_w \phi_w$.

The deep mutational scanning data are incorporated into the ExpCM via the $F_{r,xy}$ terms.
The experiments measure the preference $\pi_{r,a}$ of every site $r$ for every amino-acid $a$.
The $F_{r,xy}$ terms are defined in terms of these experimentally measured amino-acid preferences as
\begin{equation}
\label{eq:Frxy}
F_{r,xy} = 
\begin{cases}
   1 & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$} \\
   \omega \times \frac{\ln\left[\left(\pi_{r,\mathcal{A}\left(y\right)} / \pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta}\right]}{1 - \left(\pi_{r,\mathcal{A}\left(x\right)} / \pi_{r,\mathcal{A}\left(y\right)}\right)^{\beta}} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$}
   \end{cases}
\end{equation}
where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\beta$ is the stringency parameter, and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
The ExpCM has six free parameters (three $\phi_w$ values, $\kappa$, $\beta$, and $\omega$).
The preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined by an experiment independent of the sequence alignment being analyzed.

The ExpCM stationary state frequency $p_{r,x}$ of codon $x$ at site $r$ is~\citep{bloom2017identification} 
\begin{equation}
\label{eq:p_rx}
p_{r,x} = \frac{\left(\pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta} \phi_{x_0} \phi_{x_1} \phi_{x_2}}{\sum_z \left(\pi_{r,\mathcal{A}\left(z\right)}\right)^{\beta} \phi_{z_0} \phi_{z_1} \phi_{z_2}},
\end{equation}
\subsection*{Theoretical effect of model choice on branch length}
\subsection*{Effect of model choice on natural sequences}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature}}
\caption{\label{model_feature}
\textbf{Comparison of substitution model features.}
Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
Blue indicates presence and white indicates absence of a feature. 
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/decay}}
\caption{\label{decay}
\textbf{ExpCMs infer longer branches than YNGKP models for sites with mutational constraint.}
The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:decay_params}.
The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller.  
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{simulations}
\textbf{Branch lengths simulated under an ExpCM are underestimated by YNGKP models and long branches are disproportionally affected.} 
Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
The randomized ExpCMs have amino-acid profiles shuffled among the sites 
These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
\caption{\label{fig:tree_doud}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
\caption{\label{fig:tree_lee}
\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
\caption{\label{fig:tree_average}
\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
}
\end{figure}

\section*{Conclusion}

\newpage
\section*{Materials and Methods}

\subsubsection*{Experimentally informed codon model (ExpCM)}

\subsubsection*{YNGKP M0}

\subsubsection*{ExpCM + $\Gamma\omega$ and YNGKP M5}

\subsubsection*{Spielman $\omega_{r}$ values inferred from the ExpCM} 
We inferred the average nonsynonymous fixation rate from the ExpCM following~\citet{spielman2015relationship} as 
\begin{equation}
\label{eq:w_r}
\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times P_{r,xy}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
\end{equation}
where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.

\subsubsection*{Expected pairwise amino-acid identity}
The expected pairwise amino-acid identity at a site $r$ over time $t$ for a given model is 
\begin{equation}
\label{eq:f}
\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
\end{equation}
where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.

\newpage
\section*{Supplemental Information}

\subsection*{Model Parameters for the simulations}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_doud}}
\caption{\label{fig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.21$)} 
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_lee}}
\caption{\label{fig:prefs_lee}
\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)} 
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_average}}
\caption{\label{fig:prefs_average}
\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.82$)}}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
\caption{\label{fig:experiment_doud}
\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
The colors denote the alignment and the black circle denotes the focal H3 sequence. 
\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
The branch lengths are in average number of codon substitutions per site. 
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_lee}}
\caption{\label{fig:experiment_lee}
\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
\textbf{(B)} The phylogenetic tree of the "high" alignment. 
The colors denote the alignment and the black circle denotes the focal H3 sequence. 
\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
The branch lengths are in average number of codon substitutions per site. 
}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/dms}}
\caption{\label{fig:dms}
\textbf{Schematic of deep mutational scanning.} 
\textbf{(A)} All single codon mutations are introduced into the wildtype HA gene. 
\textbf{(B)} Each virus in the mutant virus library contains one HA variant. 
\textbf{(C)} The mutant virus library is passaged in cell culture to select for functional variants. 
\textbf{(D)} Deep sequencing quantifies the frequency of each variant before and after selection. 
The preference for each amino acid at each site (as quantified by the deep sequencing) is represented as a logoplot.
}
\end{figure}

\begin{table}[t!]
\caption{\label{tab:simulation_params}
ExpCM parameters used to simulate sequences in Fig.~\ref{fig:simulation}.}
      \begin{tabular}{ccccc}
        \hline
          Parameter & Value\\ \hline
       	$\beta$ & $1.54$\\
	$\kappa$ & $3.60$\\
	$\omega$ & $0.20$\\
	$\phi_A$, $\phi_C$, $\phi_G$& $0.38$, $0.17$, $0.23$\\
      \end{tabular}
\end{table}

\begin{table}[t!]
\caption{\label{tab:decay_params}
Model parameters used in  Fig.~\ref{fig:decay}.}
      \begin{tabular}{ccccc}
        \hline
          Model & Parameters\\ \hline
          ExpCM & $\beta=2.0$, $\kappa=1.0$, $\omega=1.0$, $\phi_A=\phi_C=\phi_T=0.25$, $\pi_{r, A\left(X\right)}$:\cite{Doud2016accurate}\\
          YNGKP M0 & $\kappa=1.0$, $\omega=1.0$, $\phi_{rw}=0.25$\\
          YNGKP M5 & $\kappa=1.0$, $\alpha_\omega=0.36$, $\beta_\omega=1.9$, $\phi_{rw}=0.25$\\\
      \end{tabular}
\end{table}

\clearpage 
\bibliographystyle{mbe}
\bibliography{references.bib}

\end{document}

### 2017-12-07 analysis/decay_to_stationary/decay_to_stationary_plot.py
added decay_to_stationary analysis

@@ -1,0 +1,241 @@
+"""
+The purpose of this script is to perform the calculations to make the
+"decay to stationary state" plot.
+
+This includes calculating the "spielman w_r" rates of non-syn change from the
+`ExpCM` applying these site-specific values to the `YNGKP_M0`.
+
+To run doctest, type:
+`python3 -m doctest -v test_doctest.py`
+
+SKH 20170910
+"""
+
+import pandas as pd
+from phydmslib.constants import *
+import phydmslib.models
+import phydmslib.file_io
+
+
+## Define the models
+def read_phydms_model_params(param):
+    params = pd.read_csv(param, engine="python", sep= " = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    return params
+
+def create_model_YNGKP_M0(param, nsites):
+    """
+    This is the basic `YNGKP_M0` from `phydms`
+    """
+    params = read_phydms_model_params(param)
+    e_pw = [[] for x in range(3)]
+    for i in range(3):
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys() if x.startswith("phi{0}".format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in range(N_NT)]
+    e_pw = scipy.array(e_pw)
+    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], omega=params["omega"], mu=0.3,
+            freeparams=['mu'])
+
+def create_model_YNGKP_M5(param, nsites):
+    """
+    This is the basic `YNGKP_M5` from `phydms` with 4 categories.
+    """
+    params = read_phydms_model_params(param)
+    e_pw = [[] for x in range(3)]
+    for i in range(3):
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys() if x.startswith("phi{0}".format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in range(N_NT)]
+    e_pw = scipy.array(e_pw)
+    m =  phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], mu=0.3,
+            freeparams=['mu', 'omega'])
+    return phydmslib.models.GammaDistributedOmegaModel(m, ncats=4, alpha_lambda=params["alpha_omega"], beta_lambda=params["beta_omega"])
+
[…]

### 2017-12-19 analysis/HA/data/references/make_hybrid_prefs.py
final version of preferences

@@ -1,0 +1,57 @@
+"""
+The purpose of this script is to make the hybrid preference set between Mike
+and Juhye's preferences.
+
+SKH 20171219
+"""
+
+import pandas as pd
+import numpy as np
+
+def average_df(df1, df2):
+    """
+    >>> pd.DataFrame({"A":[0.45,1.1,3.5],"B":[-.5,1,4]}).equals(average_df(pd.DataFrame({"A":[1,2,3],"B":[-2,0,5]}), pd.DataFrame({"A":[-0.1,0.2,4],"B":[1,2,3]})))
+    True
+    """
+    p = pd.Panel({n: df for n, df in enumerate([df1, df2])})
+    average = p.mean(axis=0)
+    return average
+
+def main():
+    # set up the file names
+    doud_fname = "HA_Doud_prefs.csv"
+    lee_fname = "HA_Lee_prefs.csv"
+    hybrid_numbering = "hybrid_numbering.csv"
+    outname = "HA_average_prefs.csv"
+
+    # read in the files
+    doud = pd.read_csv(doud_fname)
+    lee = pd.read_csv(lee_fname)
+    hybrid_numbers = pd.read_csv(hybrid_numbering)
+
+    # subset to only the shared sites
+    doud = doud[doud["site"].isin(hybrid_numbers["WSN"].tolist())]
+    lee = lee[lee["site"].isin(hybrid_numbers["Perth"].tolist())]
+    assert len(doud) == len(lee)
+
+    # get rid of the sites column
+    doud = doud.drop("site", axis=1)
+    lee = lee.drop("site", axis=1)
+
+    # average the values
+    average = average_df(doud, lee)
+    assert np.allclose(average.sum(axis=1), [1] * len(average))
+
+    # add back in the site column
+    average["site"] = [x+1 for x in range(len(average))]
+    cols = ["site"] + [x for x in average.columns.values if x not in ["site", "index"]]
+    average = average[cols]
+
+    # output the file
+    average.to_csv(outname, index=False)
+
+
+if __name__ == '__main__':
+    import doctest
+    doctest.testmod()
+    main()

### 2017-12-20 manuscript/manuscript.tex
beginning of the introduction

@@
 
-\newcommand\comment[1]{{\color{magenta}#1}}
+\newcommand\skhcomment[1]{{\color{magenta}#1}}
 
@@
 
-\title{Experimentally Informed Site-Specific Substitution Models Substantially Deepen Divergence Estimates}
+\title{Experimentally Informed Site-Specific Substitution Models Substantially Deepen Viral Divergence Estimates}
 
@@
 \begin{document}
-
-% Double-space the manuscript.
 
@@
 \section*{Introduction} 
+Estimating the age of a virus is essential to understanding its evolutionary history, including its emergence, spread, and past zoonoses. 
+This estimation is commonly done using the concept a ``molecular clock" to transform the branch lengths of the viral phylogenetic tree into age in years. 
+However, this molecular dating technique often underestimates the age of many viruses, including measles, foamy virus, and ebola \skhcomment{(citations)}, compared to other methods which are independent of the viral phylogeny.
+For example, SIV (the original source of HIV) is estimated to be less than a million years old based on the viral phylogeny \citep{sharp2000origins,wertheim2009dating,worobey2010island} but estimated to be several million years old based on the host tree or endogenous retroviral elements  \citep{compton2013host} \skhcomment{(other citations)}.
+Overall, there is a systematic and substantially large underestimation of of branch length on viral phylogenies. 
+\skhcomment{long branches}
 
-\textit{Introduction outline}    
-\textit{1. Estimating the divergence date of viruses is important and common.}  
-\textit{2. There are numerous examples of molecular dating techniques contradicting other dating methods for viruses.}  
-\textit{3. Why purifying selection could cause branch length underestimation.}  
-\textit{4. Discussion of the attempts/methods made thus far to model site-to-site rate variation.}  
-\textit{5. We use empirical measurements to model site-specific purifying selection. }  
-\textit{6. In this study we compare branch lengths of trees optimized by GY94 and ExpCM models and see an extension in branch lengths with ExCM and with $\Gamma\omega$. }
+Branch length underestimation is due, in part, to strong purifying selection masking the evolutionary signal in the observed sequences. 
+Purifying selection can lead to mutational saturation, where multiple unobserved, substitutions occur at a single site along a long branch and erase the divergence signal \citep{holmes2003molecular}.
+Furthermore, proteins do not have equal preference for all amino acids at all sites, this evident by a simple visual inspection of a multiple sequence alignment. 
+How many and which amino acids tolerated at each site of the protein generate a site-specific expected rate of change. 
+Failing to account for these site-specific constraints will lead to branch length underestimation. 
+\skhcomment{you will have mutational saturation no matter what - this is a separate, addressable issue?}
+\skhcomment{talk about the high mutation rate in viruses?}
 
-\textbf{Estimating the age of viruses is important to understanding their evolutionary history, including co-evolution with their host species.}
-Molecular dating techniques are commonly used to estimate the dates of viruses. 
-However, these dates are consistently younger than the dates obtained by other methods which are independent of phylogenetic trees, often by orders of magnitude. 
-\textit{example HIV/SIV, foamy virus, or measles}
+Substitution models which address site-to-site rate variation have been developed to decrease the bias in long branch estimation. 
+The most common strategy is to allow a single rate-controlling parameter to vary according to some statistical distribution, such as a $\Gamma$-distributed $\omega$ (~dN/dS) \citep{yang2000codon}. 
+This flexibility in the value of $\omega$ accounts for the site-to-site rate variation by allow some sites to have a higher dN/dS value than others. 
+While this modification is simple and only requires the addition of one extra parameter, it does not describe site-specificity in its stationary state. 
+That is, at evolutionary equilibrium, this model still assumes that each site in the protein evolves identically.
 
-\textbf{Failing to account for site-specific purifying selection results in an underestimation of branch lengths.} 
-It has long been observed that protein-coding genes have a non-uniform distribution of amino-acid frequencies across the sites in the protein. 
-Multiple sequence alignments of homologs show some sites which are conserved and appeared to be mutationally constrained and other sites which are variable and appear to be mutationally tolerant. 
-This site-specific purifying selection dictates the expectation of change at a given site which is translated into branch length. 
-\textit{The strength of purifying selection at a given site gives the expectation of change which directly relates to branch length. 
-Specifically, failing to account for constraint will lead to an underestimation in branch length because your expectation of change will be too high.}
+An alternative approach is to model the site-specific amino-acid frequencies explicitly, such as those models in the mutation-selection family \citep{halpern1998evolutionary}. 
+In these models, each amino-acid at each site in the protein is described by its own parameter and these differences are reflected in the stationary state of the model. 
+The rate of change at a given site is controlled by these amino acid profiles and can now vary from site to site, as expected based on observations in nature. 
+Importantly, these rate variations are not constrained to an arbitrary statistical distribution but by parameters with a direct biological interpretation.
 
-\textbf{Different substitution models have been developed to address the effect of site-specific purifying selection.}
-Most commonly, a rate parameter is described by some statistical distribution fit across the entire protein, such as the $\Gamma$-distributed $\omega$ in the YNGKP M5 \citep{yang2000codon}. 
-Other approaches have tried to explicitly model the site-specific amino-acid frequencies, such as mutation-selection models \citep{halpern1998evolutionary}. 
-While perhaps the most biologically-relevant, mutation-selection models are heavily parametrized and run the risk of overfitting in most practical applications. 
-However, we have previously shown that we can model this site-specific purifying selection using empirical measurements of mutational constraint from a high-throughput assay called deep mutational scanning.
-We hypothesize that using these experimentally informed models will estimate longer branches than traditional models.
+Mutation-selection models are assumably more biologically relevant but pose more practical challenges than the $\Gamma\omega$ models. 
+These models are highly parametrized with 19 free parameters (the 20 amino acid preferences are constrained to sum to one) per site leading to thousands of parameters for the length of a normal protein.
+One way to avoid overfitting is to implement the model as a mixture model in either a bayesian \citep{lartillot2004bayesian} or maximum likelihood framework \citep{si2008empirical}. 
+Alternatively, you can reduce the parameter space by defining the amino-acid frequencies \textit{a priori}. 
+We have shown previously that we can define an Experimentally Informed Codon Model (ExpCM) \citep{bloom2014experimentally,bloom2014informed} from the mutation-selection family using measurements from deep mutational scanning \citep{fowler2014deep}, a high-throughput functional assay.
+ExpCM are therefore defined by amino-acid preferences measured in a \textit{single} genetic background and do not reflect any epistatic changes which may have occurred over the virus's evolutionary history.
+But they do contain as a few parameters as the traditional codon models while maintaining a site-specific stationary state. 
+We hypothesize that the ExpCM will estimate longer branches than the traditional models due to the protein-specific description of purifying selection. 
+\skhcomment{CAT model has been shown to work well (better) on saturated data.}
 
-\textbf{Here, we address this hypothesis by comparing the branch lengths on phylogenetic trees describing influenza virus hemagglutinin optimized by different substitution models.}
-By comparing the trees optimized with and without the experimental descriptions of purifying selection, we show that failing to account for site-specific constraints results in shorter estimations of branch length.
-We used two different experimentally informed models, defined by experimental measurements from one of two diverged homologs. 
-For each model, the branchs from the experimental focal sequence are extended compared to traditional models.  
-This branch length extension is seen even in the presence of model parameters which are traditionally used to describe site-to-site rate variation.  
-These results underscore the importance of modeling purifying selection when estimating divergence dates and suggest that experimental measurement may be used to account for this selection while avoiding overparameterization.
+In order to test this hypothesis, we compared the branch lengths of a influenza virus HA phylogenetic trees optimized by different substitution models. 
+We found that the ExpCM did extend the length of branches from the focal sequence on the tree \skhcomment{define focal} and that this extension was seen even in the context of $\Gamma$-distributed rate variation.
+Furthermore, we found this extension occurred even in the presence of $\Gamma$-distributed $\omega$, indicating that they are both important for modeling purifying selection. 
+This supports the conclusion that modeling purifying selection, especially in a model with a non-uniform stationary state, is important to estimating the branch lengths on phylogenetic trees.
 
 \section*{Results and Discussion}
+
+\subsection*{Effect of stationary state and rate variation on long branch estimation}
+
+What are the points for this section? 
+\begin{enumerate}
+  \item Different substitution models have different definitions of rate variation and this allows us to compare the effects of different attributes. 
+  \item GY94 has a uniform stationary state and no rate variation - it estimates the shortest branches and reaches equilibrium right around 0.05. 
+  \item The addition of rate variation for a model with a uniform stationary state does increase the branch length. 
+  \item Using site-specific models allows you to model site-specific constraint. The more constrained the site the bigger the difference between non-uniform and uniform substitution models 
+  \item You can infer an dN/dS value from mutation-selection models. When you add this value to the base model you see the line come up but not as much as the true ExpCM. 
+  \item 
+\end{enumerate}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature}}
+\caption{\label{model_feature}
+\textbf{Comparison of substitution model features.}
+Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
+The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
+Blue indicates presence and white indicates absence of a feature. 
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/decay2.pdf}}
+\caption{\label{decay}
+\textbf{Effect of stationary state and rate variation on long branch estimation.}
+The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:decay_params}.
+The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
+The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
+The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller.
+}
+\end{figure}
+
+\subsection*{Model performance under simulated, site-specific data.}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
+\caption{\label{simulations}
+\textbf{Model performance under simulated, site-specific data.} 
+Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
+The randomized ExpCMs have amino-acid profiles shuffled among the sites 
+These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
+Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
+The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
+}
+\end{figure}
+
+\subsection*{Actual Data}
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
+\caption{\label{fig:tree_doud}
+\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
+\caption{\label{fig:tree_lee}
+\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
+\caption{\label{fig:tree_average}
+\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
+The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
+}
+\end{figure}
+
+\subsection{Competing effects of shifting preferences and long branches.}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
+\caption{\label{fig:experiment_doud}
+\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
+\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
+\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
+The colors denote the alignment and the black circle denotes the focal H3 sequence. 
+\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
+\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
+Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
+The branch lengths are in average number of codon substitutions per site. 
+}
+\end{figure}
+
+\section*{Conclusion}
+
+\begin{enumerate}
+  \item We don't allow any of the models to vary by lineage. 
+\end{enumerate}
+
+\newpage
+\section*{Materials and Methods}
 
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature}}
-\caption{\label{model_feature}
-\textbf{Comparison of substitution model features.}
-Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
-The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
-Blue indicates presence and white indicates absence of a feature. 
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/decay}}
-\caption{\label{decay}
-\textbf{ExpCMs infer longer branches than YNGKP models for sites with mutational constraint.}
-The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:decay_params}.
-The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
-The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
-The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller.
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
-\caption{\label{simulations}
-\textbf{Branch lengths simulated under an ExpCM are underestimated by YNGKP models and long branches are disproportionally affected.} 
-Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
-The randomized ExpCMs have amino-acid profiles shuffled among the sites 
-These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
-Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
-The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
-\caption{\label{fig:tree_doud}
-\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
-\caption{\label{fig:tree_lee}
-\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
-\caption{\label{fig:tree_average}
-\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
-The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
-}
-\end{figure}
-
-\section*{Conclusion}
-
-\newpage
-\section*{Materials and Methods}
-
-\subsubsection*{Experimentally informed codon model (ExpCM)}
-
-\subsubsection*{YNGKP M0}
-
 \subsubsection*{ExpCM + $\Gamma\omega$ and YNGKP M5}
@@
 \subsubsection*{Expected pairwise amino-acid identity}
+\textit{Do I need to talk about the branchScale scaling I used?}
 The expected pairwise amino-acid identity at a site $r$ over time $t$ for a given model is 
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_doud}}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_lee}}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_average}}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
-\caption{\label{fig:experiment_doud}
-\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
-\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
-\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
-The colors denote the alignment and the black circle denotes the focal H3 sequence. 
-\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
-\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
-Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
-The branch lengths are in average number of codon substitutions per site. 
-}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[H]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_lee}}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/dms}}

### 2018-01-16 manuscript/manuscript.tex
updates to manuscript

@@
 \begin{abstract}
-\textit{$\leq$ 250 words, currently 159}
+\textit{$\leq$ 250 words}
 
-Molecular dating techniques have been used to estimate the divergence date of many viruses. 
-However, these estimates are consistently substantially younger than estimates from methods which are independent of branch length estimation. 
-This discrepancy is caused, in part, by inadequate modeling of purifying selection leading to branch length underestimation. 
-Here, we show that substitution models informed by empirical measurements of mutational constraint better than traditional models and extend branch lengths. 
+Molecular dating techniques is often used to estimate the divergence time of many viruses. 
+However, these estimates are substantially younger than estimates from methods which do not rely on viral phylogenies. 
+This discrepancy is thought to be caused in part by inadequate modeling of purifying selection leading to branch length underestimation. 
+Here, we show that substitution models informed by empirical measurements of mutational model constraint better than traditional models and extend branch lengths. 
 We used models informed by deep mutational scanning experiments performed in two, highly diverged influenza virus hemagglutinin homologs to optimize the branch lengths of a phylogenetic tree. 
@@
 Overall, this study underscores the importance of modeling purifying selection when estimating branch lengths and, by extension, divergence dates. 
+\skhcomment{from JDB: also shows a way to actually do this in addition to showing importance.}
 
@@
 \section*{Introduction} 
-Estimating the age of a virus is essential to understanding its evolutionary history, including its emergence, spread, and past zoonoses. 
+\skhcomment{from JDB: what is the "age" of a virus? Maybe "divergence time of viral lineages"}
+skhcomment{from JDB: what is the less than a million actually? "Old" is not the right phrase.}
+Estimating the divergence time of viral lineages of a virus is essential to understanding its evolutionary history, including its emergence, spread, and past zoonoses. 
 This estimation is commonly done using the concept a ``molecular clock" to transform the branch lengths of the viral phylogenetic tree into age in years. 
@@
 
-Substitution models which address site-to-site rate variation have been developed to decrease the bias in long branch estimation. 
+Substitution models that incorportate site-to-site rate variation have been developed to decrease the bias in long branch estimation. 
 The most common strategy is to allow a single rate-controlling parameter to vary according to some statistical distribution, such as a $\Gamma$-distributed $\omega$ (~dN/dS) \citep{yang2000codon}. 
@@
 
-Mutation-selection models are assumably more biologically relevant but pose more practical challenges than the $\Gamma\omega$ models. 
+Mutation-selection models are presumably more biologically relevant but pose more practical challenges than the $\Gamma\omega$ models. 
 These models are highly parametrized with 19 free parameters (the 20 amino acid preferences are constrained to sum to one) per site leading to thousands of parameters for the length of a normal protein.
 One way to avoid overfitting is to implement the model as a mixture model in either a bayesian \citep{lartillot2004bayesian} or maximum likelihood framework \citep{si2008empirical}.
+
 Alternatively, you can reduce the parameter space by defining the amino-acid frequencies \textit{a priori}. 
@@
 ExpCM are therefore defined by amino-acid preferences measured in a \textit{single} genetic background and do not reflect any epistatic changes which may have occurred over the virus's evolutionary history.
-But they do contain as a few parameters as the traditional codon models while maintaining a site-specific stationary state. 
+But they contain no more parameters than the traditional codon models while maintaining a site-specific stationary state. 
 We hypothesize that the ExpCM will estimate longer branches than the traditional models due to the protein-specific description of purifying selection. 
@@
 
-\subsection*{Effect of stationary state and rate variation on long branch estimation}
-
-What are the points for this section? 
-\begin{enumerate}
-  \item Different substitution models have different definitions of rate variation and this allows us to compare the effects of different attributes. 
-  \item GY94 has a uniform stationary state and no rate variation - it estimates the shortest branches and reaches equilibrium right around 0.05. 
-  \item The addition of rate variation for a model with a uniform stationary state does increase the branch length. 
-  \item Using site-specific models allows you to model site-specific constraint. The more constrained the site the bigger the difference between non-uniform and uniform substitution models 
-  \item You can infer an dN/dS value from mutation-selection models. When you add this value to the base model you see the line come up but not as much as the true ExpCM. 
-  \item 
-\end{enumerate}
+\subsection*{Substitution models}
 
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature}}
-\caption{\label{model_feature}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature.pdf}}
+\caption{\label{fig:model_feature}
 \textbf{Comparison of substitution model features.}
@@
 Blue indicates presence and white indicates absence of a feature. 
+from \ref{tab:wsn_low_params}
 }
@@
 
+\skhcomment{opening sentence or two about modeling site-to-site rate variation and/or purifying selection}
+
+Codon substitution models commonly include a parameter ($\omega$) which describes the relative rate of non-synonymous to synonymous change to account for purifying selection or site-to-site rate variation more generally.
+This parameter can be constrained to one value, representing a gene-wide average rate, or allowed to vary among sites following some statistical distribution, commonly a $\Gamma$-distribution (\ref{fig:model_feature}B).
+By allowing $\omega$ to vary among the sites, this models acknowledges that some sites evolve faster or slower than others. 
+However, in practical implementation the distribution is discretized into some number of categories (commonly four) and $\omega$ takes on the mean of each category.
+
+Another strategy to account for purifying selection and site-to-site rate variation is to use a model with a non-uniform stationary state (\ref{fig:model_feature}A). 
+In this framework, differences in rate between sites is modeled explicitly by site-specific amino-acid frequency parameters. 
+These parameters describe the ''preference" for a given amino acid at a given site. 
+According to this model, the rate of change at a "constrained" site, which prefers very few amino acids, is much slower than the rate of change at a tolerant site, which prefers many amino acids equally.
+
+The challenge of a model with a non-uniform stationary state is that all 19 * L (sites in the protein) amino acid preference parameters have to be defined. 
+In this study, we chose to use non-uniform stationary state model called ExpCM which defines these parameters using the measurements from a high-throughput functional assay called Deep Mutational Scanning (DMS).
+DMS measures the effect of every single amino acid change on some aspect of protein function. 
+We used two deep mutational scans of influenza virus surface protein hemagglutinin (HA) which measured the effect of mutations on viral growth.
+
+The models we chose to use in this study represent all four combinations of these two features (\ref{fig:model_feature}C). 
+The GY94 model has a uniform stationary state while the ExpCM has a non-uniform stationary state and each model can be implemented with a $\Gamma$-distributed $\omega$. 
+Therefore, we can example the effect of one feature (stationary state or $\Gamma\omega$) while controlling for the presence of the other feature.
+
+\subsection*{Effect of stationary state and rate variation on branch length estimation}
+
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/decay2.pdf}}
-\caption{\label{decay}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/decay.pdf}}
+\caption{\label{fig:decay}
 \textbf{Effect of stationary state and rate variation on long branch estimation.}
-The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:decay_params}.
+The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
 The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
@@
 
-\subsection*{Model performance under simulated, site-specific data.}
+To explore the effect of substitution model choice on branch length estimation, we first examined how the relationship between sequence divergence and time is described by different models. 
+While substitution models can only accurately estimate branch length when there is a clear transformation from sequence divergence to time (branch length), every model has a saturation point at which the temporal signal has effectively decayed.
+This effect can been seen in the long, almost horizontal tails of \ref{fig:decay}, where the expectation of sequence divergence stays constant as time passes.
+
+While this saturation phenomenon is inevitable, the amount of time it takes to reach the saturation point and the sequence divergence at the point are dependent on the specific features of the substitution model.
+The addition of $\Gamma\omega$ to the GY94 model modulates the exact sequence divergence of the saturation point but the overall effect is minimal. 
+Furthermore, neither the GY94 nor the GY+$\Gamma\omega$ are site-specific and consequently the saturation points for each model are exactly same for every site. 
+In comparison, the saturation point of the ExpCM differs from site to site. 
+The ExpCM saturation point is very similar to the saturation point for the GY94 model at sites with a relatively high tolerance for mutations (\ref{fig:decay} sites 90 and 51). 
+But the constrained sites show saturation points with much higher expected sequence conservation (\ref{fig:decay} sites 413, 300, and 262) and/or a longer period of time until the saturation point is reached (\ref{fig:decay} sites 413 and 300).
+
+Finally, we can see that the difference between the ExpCM and the GY94 is due to the non-uniform stationary state and not just because the ExpCM "takes advantage" of more information (the DMS). 
+We can infer a site-specific dN/dS values from the ExpCM \cite{spielman2015relationship} and apply them to the GY94 model, thereby incorporating information from the DMS into a uniform stationary state model.
+This model takes a longer time to reach the saturation point but does the saturation point does have the same sequence divergence as the GY94 model. 
+The underestimation of long branches will be an issue for all models but especially models with uniform stationary states, even if great care is taken to model rate variation.
+
+\skhcomment{be more clear about how this plot shows branch lengths }
+
+\subsection*{Failure to account for site-specificity leads to branch length underestimation.}
 
@@
 
-\subsection*{Actual Data}
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
-\caption{\label{fig:tree_doud}
-\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
-}
-\end{figure}
+Next, we explored how accurately these models infer branch length when site-specificity is assumed, as expected in natural sequences.  
+We simulated sequences under an ExpCM defined by the amino-acid preferences measured by \cite{doud2016accurate} along an influenza tree. 
+We then re-inferred the branch lengths using the models described in \ref{fig:model_feature}.
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
-\caption{\label{fig:tree_lee}
-\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
-}
-\end{figure}
+The two models with uniform stationary states, GY94 and GY94+$\Gamma\omega$, both underestimate long branches (> ~5 codon substitutions per site). 
+The degree of underestimation becomes more pronounced as the branch length increases, as shown by the curve of the correlation. 
+The addition of $\Gamma\omega$ rate variation does not mitigate the effect as both models show roughly the same degree of underestimation.
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
-\caption{\label{fig:tree_average}
-\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
-The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
-}
-\end{figure}
+Conversely, the ExpCM and ExpCM+$\Gamma\omega$ both avoid the underestimation bias, even at the longest branch lengths measured. 
+As the branch length increases, the variance of the branch length estimation also increases (grey points). 
+But, unlike the GY94 and the GY94+$\Gamma\omega$, the variance is unbiased in regards to under- or overestimation of the true branch length.
+
+Finally, the randomized control ExpCM and ExpCM+$\Gamma\omega$ show that the importance is truly in site-\textit{specificity}. 
+If the preferences are randomized along sites in the protein, the model exhibits the same degree of underestimation as the GY94 and GY+$\Gamma\omega$ models. 
+This underscores the point that the non-uniform stationary state models are site-specific and that mismatches are just as much of a problem as not accounting for these differences at all.
+
+Overall, the simulations show that if site-specificity is true, the addition of $\Gamma\omega$ will still have problems at extremely long branches.
+
+\subsection*{empirical Data}
+
+Next we compared the branch lengths inferred by the different models on empirical data. 
+We looked at the influenza virus surface protein hemagglutinin (HA). 
+HA proteins cluster into tight phylogenetic groups with long branches in between them. 
+Furthermore, HAs are extremely diverse. 
+The HAs which commonly circulate in human viruses (H1 and H3) are only about 40\% identical on the amino acid level.
+
+We took a representative sample of HA from 14 of the 18 groups and optimized the branch lengths of the tree using the models described in \ref{fig:model_feature}. 
+First, we used an ExpCM defined by the site-specific amino-acid preferences measured by deep mutational scanning from \cite{doud2016accurate}. 
+These amino-acid preferences were measured in the genetic background of the H1 strain A/Wilson-Smith/1993. 
+We designate this strain (and more broadly the H1 subtype) as the focal sequence of the model.
+
+\ref{fig:tree_doud} shows the results. 
+Comparing GY94 to GY94+$\Gamma\omega$ or ExpCM to ExpCM+$\Gamma\omega$ shows the addition of $\Gamma\omega$ causes branch length extension independent of the presence of preferences. 
+In addition, comparing comparing GY94 to ExpCM or GY94+$\Gamma\omega$ to ExpCM+$\Gamma\omega$ shows that the addition of preferences causes branch length extension from the \textit{focal} sequence independent of the presence of $\Gamma\omega$.
+Together this suggests that site-specificity of the model is important because branch lengths are shorter in the absence of preferences. 
+However, the preferences used here to define the ExpCM are clearly not equally relevant across all parts of the tree. 
+Specifically, they are most relevant near the strain that they were measured in. 
+This not at all unexpected because epistasis, where the effect of a mutation at one site is dependent on the identity of another site is well documented. 
+It would surprising if epistasis \textit{was not} a factor at these large divergences across the HA tree.
+
+This becomes even more apparent when the performance of the ExpCM defined by H1 preferences is compared to the ExpCM with H3 preferences. 
+Namely, the exact same pattern is observed. 
+There is extension across the whole tree when $\Gamma\omega$ is added to the model, independent of whether or not preferences are included in the model, and there is an extension of branch lengths from the focal sequence when preferences are added, independent of whether or not $\Gamma\omega$ is included in the model.
+Finally, we defined an ExpCM with the average of the H1 and H3 preferences, which can be seen as a rough proxy for the model sampling both preference sets. 
+In this model, there is an extension in branch length from \textit{both} focal sequences.
+
+These three sets of trees imply that site-specificity is important and not constant. 
+This observation is not surprising. 
+Epistasis, where the effect of a mutation at a site is dependent on the identity of another site in the protein, is well documented. 
+It is not all together surprising that there are epistatic interactions between H1 and H3 due to the large sequence divergence between the two strains. 
+In contrast, the deep mutational scanning preferences are measured in a single genetic background and therefore only measure the effect of mutations in one, specific genetic context.  
+The difference between the trees using the H1 preferences and the H3 preferences is most likely that the preferences "shifted" due to the genetic context. 
+The model with the average preferences could be seen as a proxy for a model which is able to sample both preference sets.
 
 \subsection{Competing effects of shifting preferences and long branches.}
+We explored these competing effects of shifting preferences and long branches by comparing the effects of the model across trees which vary in overall divergence levels. 
+We used the $\beta$ parameter of the ExpCM model as a proxy for the degree to which the preferences have shifted. 
+In each run of the model, the parameter $\beta$ is fit and effectively rescales the preferences. 
+When $\beta = 1$, we infer that selection in nature and selection in the lab prefers the same amino acids and with the same level of stringency. 
+When $\beta > 1$, we infer that selection in nature prefers the same amino acids as selection in lab but with much higher stringency and when $\beta < 1$ we infer that selection in lab and selection in nature prefer different amino acids or at a \textit{much} lower stringency.
+An alternative way of saying this is that the value of $\beta$ and the perceived "relevance" of the preferences are inversely correlated.
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
-\caption{\label{fig:experiment_doud}
-\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
-\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
-\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
-The colors denote the alignment and the black circle denotes the focal H3 sequence. 
-\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
-\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
-Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
-The branch lengths are in average number of codon substitutions per site. 
-}
-\end{figure}
+For each preference set, we took a subset the full HA tree shown in \ref{fig:tree_doud} or \ref{fig:tree_lee} to create a tree with either low divergence from the focal sequence (low) or intermediate divergence from the focal sequence (intermediate) and estimated the branch lengths on these trees using the same models as above.
+As we hypothesized above, as the overall divergence of the tree increases the $\beta$ value decreases. 
+That is as the tree begins to include more and more sequences which are highly diverged from the focal sequence, the less relevant the preferences appear to be overall. 
+As divergence increases the relevance of the preferences decreases.
 
+On the other hand, the effect of having a site-specific stationary state is more apparent when there are long branches on the tree.
+There is no real bias in estimation in any of the models for the low or intermediate trees. 
+This is not surprising as \ref{fig:decay} shows that the differences in these models really only matters at the long branches. 
+The underestimation is only apparent for the "high divergence" alignment (aka the full tree). 
+Here it becomes clear that you have these competing effects. 
+It is most important to have the site-specificity at these long branches but it is also where the site-specific preferences are the least relevant.
+
+For set of preferences, 
 \section*{Conclusion}
@@
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/dms}}
-\caption{\label{fig:dms}
-\textbf{Schematic of deep mutational scanning.} 
-\textbf{(A)} All single codon mutations are introduced into the wildtype HA gene. 
-\textbf{(B)} Each virus in the mutant virus library contains one HA variant. 
-\textbf{(C)} The mutant virus library is passaged in cell culture to select for functional variants. 
-\textbf{(D)} Deep sequencing quantifies the frequency of each variant before and after selection. 
-The preference for each amino acid at each site (as quantified by the deep sequencing) is represented as a logoplot.
-}
-\end{figure}
-
 \begin{table}[t!]
@@
 \begin{table}[t!]
-\caption{\label{tab:decay_params}
+\caption{\label{tab:wsn_low_params}
 Model parameters used in  Fig.~\ref{fig:decay}.}
@@
           Model & Parameters\\ \hline
-          ExpCM & $\beta=2.0$, $\kappa=1.0$, $\omega=1.0$, $\phi_A=\phi_C=\phi_T=0.25$, $\pi_{r, A\left(X\right)}$:\cite{Doud2016accurate}\\
-          YNGKP M0 & $\kappa=1.0$, $\omega=1.0$, $\phi_{rw}=0.25$\\
-          YNGKP M5 & $\kappa=1.0$, $\alpha_\omega=0.36$, $\beta_\omega=1.9$, $\phi_{rw}=0.25$\\\
+          ExpCM & $\beta=1.54196$\\
+           & $\kappa=3.47184$\\
+           & $\omega=0.219225$\\ 
+           & $\phi_A$=0.386529\\ 
+           & $\phi_C$=0.17187\\
+           & $\phi_T$=0.230319\\
+           & $\pi_{r, A\left(X\right)}$:Doud\\
+          YNGKP M0 & $\kappa=2.9984$\\
+          & $\omega=0.09076$\\
+          & $\phi_{1A}=0.326007$\\
+          & $\phi_{1C}=0.136307$\\ 
+          &$\phi_{1G}=0.278377$\\
+          &$\phi_{2A}=0.382017$\\ 
+          &$\phi_{2C}=0.179763$\\ 
+          &$\phi_{2G}=0.194067$\\
+          &$\phi_{3A}=0.354364$\\
+          &$\phi_{3C}=0.190112$\\
+          &$\phi_{3G}=0.214605$\\
+          YNGKP M5 & $\kappa=2.9984$\\
+          & $\omega=0.09076$\\
+          & $\phi_{1A}=0.326007$\\
+          & $\phi_{1C}=0.136307$\\ 
+          &$\phi_{1G}=0.278377$\\
+          &$\phi_{2A}=0.382017$\\ 
+          &$\phi_{2C}=0.179763$\\ 
+          &$\phi_{2G}=0.194067$\\
+          &$\phi_{3A}=0.354364$\\
+          &$\phi_{3C}=0.190112$\\
+          &$\phi_{3G}=0.214605$\\
       \end{tabular}
 \end{table}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
+\caption{\label{fig:tree_doud}
+\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
+\caption{\label{fig:tree_lee}
+\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
+\caption{\label{fig:tree_average}
+\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
+The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
+\caption{\label{fig:experiment_doud}
+\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
+\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
+\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
+The colors denote the alignment and the black circle denotes the focal H3 sequence. 
+\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
+\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
+Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
+The branch lengths are in average number of codon substitutions per site. 
+}
+\end{figure}
 

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.13.sh
small tweak to simulations pipeline

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_0_simulatedalignment.fasta"], "threads": 1, "jobid": 13, "rule": "run_phydms", "output": ["phydms/WSN_high_0_0/modelcomparison.md", "phydms/WSN_high_0_0/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_0/modelcomparison.md phydms/WSN_high_0_0/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_0_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/13.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/13.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.15.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_2_simulatedalignment.fasta"], "threads": 1, "jobid": 15, "rule": "run_phydms", "output": ["phydms/WSN_high_0_2/modelcomparison.md", "phydms/WSN_high_0_2/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_2/modelcomparison.md phydms/WSN_high_0_2/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_2_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/15.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/15.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.17.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_1_simulatedalignment.fasta"], "threads": 1, "jobid": 17, "rule": "run_phydms", "output": ["phydms/WSN_high_0_1/modelcomparison.md", "phydms/WSN_high_0_1/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_1/modelcomparison.md phydms/WSN_high_0_1/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_1_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/17.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/17.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.2.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_8_simulatedalignment.fasta"], "threads": 1, "jobid": 2, "rule": "run_phydms", "output": ["phydms/WSN_high_0_8/modelcomparison.md", "phydms/WSN_high_0_8/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_8/modelcomparison.md phydms/WSN_high_0_8/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_8_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/2.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/2.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.20.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_9_simulatedalignment.fasta"], "threads": 1, "jobid": 20, "rule": "run_phydms", "output": ["phydms/WSN_high_0_9/modelcomparison.md", "phydms/WSN_high_0_9/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_9/modelcomparison.md phydms/WSN_high_0_9/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_9_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/20.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/20.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.3.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_6_simulatedalignment.fasta"], "threads": 1, "jobid": 3, "rule": "run_phydms", "output": ["phydms/WSN_high_0_6/modelcomparison.md", "phydms/WSN_high_0_6/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_6/modelcomparison.md phydms/WSN_high_0_6/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_6_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/3.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/3.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.4.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_3_simulatedalignment.fasta"], "threads": 1, "jobid": 4, "rule": "run_phydms", "output": ["phydms/WSN_high_0_3/modelcomparison.md", "phydms/WSN_high_0_3/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_3/modelcomparison.md phydms/WSN_high_0_3/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_3_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/4.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/4.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.5.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_7_simulatedalignment.fasta"], "threads": 1, "jobid": 5, "rule": "run_phydms", "output": ["phydms/WSN_high_0_7/modelcomparison.md", "phydms/WSN_high_0_7/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_7/modelcomparison.md phydms/WSN_high_0_7/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_7_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/5.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/5.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.6.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_4_simulatedalignment.fasta"], "threads": 1, "jobid": 6, "rule": "run_phydms", "output": ["phydms/WSN_high_0_4/modelcomparison.md", "phydms/WSN_high_0_4/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_4/modelcomparison.md phydms/WSN_high_0_4/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_4_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/6.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/6.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.9.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_5_simulatedalignment.fasta"], "threads": 1, "jobid": 9, "rule": "run_phydms", "output": ["phydms/WSN_high_0_5/modelcomparison.md", "phydms/WSN_high_0_5/"], "cluster": {}, "local": false}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_5/modelcomparison.md phydms/WSN_high_0_5/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_5_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/9.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/9.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.11.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_2_modelcomparison.md_simulatedalignment.fasta"], "jobid": 11, "params": {}, "output": ["phydms/WSN_high_0_2_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_2_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_2_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_2_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_2_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/11.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/11.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.14.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_5_modelcomparison.md_simulatedalignment.fasta"], "jobid": 14, "params": {}, "output": ["phydms/WSN_high_0_5_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_5_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_5_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_5_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_5_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/14.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/14.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.15.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_7_modelcomparison.md_simulatedalignment.fasta"], "jobid": 15, "params": {}, "output": ["phydms/WSN_high_0_7_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_7_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_7_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_7_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_7_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/15.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/15.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.18.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_9_modelcomparison.md_simulatedalignment.fasta"], "jobid": 18, "params": {}, "output": ["phydms/WSN_high_0_9_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_9_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_9_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_9_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_9_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/18.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/18.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.20.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_6_modelcomparison.md_simulatedalignment.fasta"], "jobid": 20, "params": {}, "output": ["phydms/WSN_high_0_6_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_6_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_6_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_6_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_6_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/20.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/20.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.3.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_0_modelcomparison.md_simulatedalignment.fasta"], "jobid": 3, "params": {}, "output": ["phydms/WSN_high_0_0_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_0_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_0_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_0_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_0_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/3.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/3.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.4.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_3_modelcomparison.md_simulatedalignment.fasta"], "jobid": 4, "params": {}, "output": ["phydms/WSN_high_0_3_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_3_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_3_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_3_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_3_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/4.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/4.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.6.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_1_modelcomparison.md_simulatedalignment.fasta"], "jobid": 6, "params": {}, "output": ["phydms/WSN_high_0_1_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_1_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_1_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_1_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_1_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/6.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/6.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.7.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_8_modelcomparison.md_simulatedalignment.fasta"], "jobid": 7, "params": {}, "output": ["phydms/WSN_high_0_8_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_8_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_8_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_8_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_8_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/7.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/7.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.9.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_4_modelcomparison.md_simulatedalignment.fasta"], "jobid": 9, "params": {}, "output": ["phydms/WSN_high_0_4_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_4_modelcomparison.md"]}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_4_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_4_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_4_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/9.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/9.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.13.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {}, "input": ["_WSN_high_0_5_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_5/modelcomparison.md", "phydms/WSN_high_0_5/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 13}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_5/modelcomparison.md phydms/WSN_high_0_5/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_5_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/13.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/13.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.16.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {}, "input": ["_WSN_high_0_6_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_6/modelcomparison.md", "phydms/WSN_high_0_6/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 16}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_6/modelcomparison.md phydms/WSN_high_0_6/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_6_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/16.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/16.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.17.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {}, "input": ["_WSN_high_0_1_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_1/modelcomparison.md", "phydms/WSN_high_0_1/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 17}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_1/modelcomparison.md phydms/WSN_high_0_1/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_1_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/17.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/17.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.3.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {}, "input": ["_WSN_high_0_4_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_4/modelcomparison.md", "phydms/WSN_high_0_4/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 3}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_4/modelcomparison.md phydms/WSN_high_0_4/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_4_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/3.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/3.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.4.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {}, "input": ["_WSN_high_0_0_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_0/modelcomparison.md", "phydms/WSN_high_0_0/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 4}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_0/modelcomparison.md phydms/WSN_high_0_0/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_0_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/4.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/4.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.5.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {}, "input": ["_WSN_high_0_9_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_9/modelcomparison.md", "phydms/WSN_high_0_9/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 5}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_9/modelcomparison.md phydms/WSN_high_0_9/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files _WSN_high_0_9_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/5.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/5.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.14.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {"r": "8", "prefix": "_WSN_high_0_8", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_8_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 14}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_8_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/14.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/14.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.18.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {"r": "3", "prefix": "_WSN_high_0_3", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_3_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 18}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_3_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/18.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/18.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.19.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {"r": "7", "prefix": "_WSN_high_0_7", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_7_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 19}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_7_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/19.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/19.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.20.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+# properties = {"params": {"r": "2", "prefix": "_WSN_high_0_2", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_2_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 20}
+cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
+/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_2_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
+--force -j --keep-target-files --keep-shadow --keep-remote \
+--wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
+--benchmark-repeats 1 \
+--force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
+   --nocolor \
+--notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/20.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/20.jobfailed"; exit 1)
+

### 2018-01-16 analysis/simulations/pymodules/utils.py

@@ -1,0 +1,100 @@
+"""
+Python utilities for simulations analysis
+Written by Sarah Hilton.
+"""
+from phydmslib.constants import *
+import phydmslib.file_io
+from ete3 import Tree
+import os
+import itertools
+from scipy.special import comb
+import pandas as pd
+import scipy
+
+
+def createModel(prefs, params):
+    """
+    This function creates the ExpCM for the `pyvolve` simulation
+    input:
+        `str`, path to the preferences file
+        `str`, path to file with the preference parameters
+    """
+    prefs = phydmslib.file_io.readPrefs(prefs)
+    sites = sorted(prefs.keys())
+    prefs = [prefs[r] for r in sites]
+    params = pd.read_csv(params, engine="python", sep=" = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    params["phiT"] = 1 - sum([params[x] for x in params.keys()
+                             if x.startswith("phi")])
+    phi = scipy.array([params["phi{0}".format(INDEX_TO_NT[x])]
+                       for x in range(N_NT)])
+    return phydmslib.models.ExpCM(prefs, kappa=params["kappa"],
+                                  omega=params["omega"], beta=params["beta"],
+                                  mu=0.3, phi=phi, freeparams=['mu'])
+
+
+def extract_model_name(model_string):
+    """
+    This function takes a tree output file from phydms and extracts the name of
+    the model.
+    input:
+        `str`, tree file name
+    output:
+        `str`, model name
+        `int`, 1 if "control" model, 0 otherwise
+    """
+    if "YNGKP" in model_string:
+        if "YNGKP_M0" in model_string:
+            return "GY94", 0
+        elif "YNGKP_M5" in model_string:
+            return "GY94_gammaomega", 0
+        else:
+            raise ValueError("Cannot parse model string {0}"
+                             .format(model_string))
+    elif "ExpCM" in model_string:
+        if "gammaomega" in model_string:
+            if "randomized_ExpCM" in model_string:
+                return "ExpCM_gammaomega_randomized", 1
+            elif "averaged_ExpCM" in model_string:
+                return "ExpCM_gammaomega_averaged", 1
+            else:
+                return "ExpCM_gammaomega", 0
+        else:
+            if "randomized_ExpCM" in model_string:
+                return "ExpCM_randomized", 1
+            elif "averaged_ExpCM" in model_string:
[…]

### 2018-01-16 analysis/simulations/snakemake_cleanup.py

@@ -1,0 +1,64 @@
+#!/python
+
+"""
+This script cleans up the out files from the Halern-Bruno-like simulations `Snakefile`.
+
+This script is greedy and not robust.
+
+SKH 20170420
+"""
+
+import glob
+import os
+import sys
+import argparse
+
+class ArgumentParserNoArgHelp(argparse.ArgumentParser):
+    """Like *argparse.ArgumentParser*, but prints help when no arguments."""
+    def error(self, message):
+        sys.stderr.write('error: %s\n\n' % message)
+        self.print_help()
+        sys.exit(2)
+
+
+def ParseArguments():
+    """Argument parser for script."""
+    parser = ArgumentParserNoArgHelp(
+            description='Clean up files from simulation',
+            formatter_class=argparse.ArgumentDefaultsHelpFormatter,
+            )
+    parser.set_defaults(tree=False)
+    parser.add_argument('--tree', dest='tree', action=\
+            'store_true', help="Remove temp tree files.")
+    parser.set_defaults(alignment=False)
+    parser.add_argument('--alignment', dest='alignment', action=\
+            'store_true', help="Remove alignment files.")
+    parser.set_defaults(phydms=False)
+    parser.add_argument('--phydms', dest='phydms', action=\
+            'store_true', help="Remove phydms files.")
+    parser.set_defaults(slurm=False)
+    parser.add_argument('--slurm', dest='slurm', action=\
+            'store_true', help="Remove slurm files.")
+    return parser
+
+def delete_files(keyword):
+    for f in glob.glob("{0}*".format(keyword)):
+        if os.path.isfile(f):
+            os.remove(f)
+
+def main():
+    keyword_dictionary = {"tree":"_tempTree_", "alignment":"*simulatedalignment.fasta", "phydms":"phydms_", "slurm":"slurm-"}
+    args = vars(ParseArguments().parse_args())
+    print("Read the following command line arguments:")
+    print("\t{0}".format("\n\t".join(["{0} = {1}".format(key, value) for (key, value) in args.items()])))
+
+    if True in args.values():
+        print("Cleaning up files from simulations")
+        for arg in args.keys():
+            if args[arg] == True:
+                print("Removing {0}".format(arg))
+                delete_files(keyword_dictionary[arg])
+    print("done.")
+
+if __name__ == '__main__':
+ 	main() # run the script

### 2018-01-16 manuscript/manuscript.tex

@@
 \newcommand\skhcomment[1]{{\color{magenta}#1}}
+\newcommand\jdbcomment[1]{{\color{red}#1}}
 
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature.pdf}}
+\centerline{\includegraphics[width=0.55\textwidth]{figures/model_feature.pdf}}
 \caption{\label{fig:model_feature}
 \textbf{Comparison of substitution model features.}
+\jdbcomment{This figure is too big. Maybe put the three panels side-by-side, increase the space between panels, and increase the font sizes. The logo stacks are kind of tall relative to their width.}
 Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/decay.pdf}}
+\centerline{\includegraphics[width=0.75\textwidth]{figures/decay.pdf}}
 \caption{\label{fig:decay}
+\jdbcomment{I still think text could be bigger. Logo stacks still show trimming, maybe a bit tall? The Model key could probably be to the right or left.}
 \textbf{Effect of stationary state and rate variation on long branch estimation.}
@@
 \textbf{Model performance under simulated, site-specific data.} 
+\jdbcomment{Fonts are small, and tick labels definitely too small. I would show averaged.}
 Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
@@
 The HAs which commonly circulate in human viruses (H1 and H3) are only about 40\% identical on the amino acid level.
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.5\textwidth]{figures/tree_doud}}
+\caption{\label{fig:tree_doud}
+\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
+\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
+}
+\end{figure}
 
@@
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_doud}}
-\caption{\label{fig:prefs_doud}
+\begin{suppfig}[H]
+\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
+\caption{\label{suppfig:prefs_doud}
 \textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.21$)} 
 }
-\end{figure}
+\end{suppfig}
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_lee}}
-\caption{\label{fig:prefs_lee}
+\begin{suppfig}[H]
+\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
+\caption{\label{suppfig:prefs_lee}
 \textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)} 
 }
-\end{figure}
+\end{suppfig}
 
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
-\caption{\label{fig:tree_doud}
-\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
-}
-\end{figure}
-
-\begin{figure}[H]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.13.sh
more directory cleanup

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_0_simulatedalignment.fasta"], "threads": 1, "jobid": 13, "rule": "run_phydms", "output": ["phydms/WSN_high_0_0/modelcomparison.md", "phydms/WSN_high_0_0/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_0/modelcomparison.md phydms/WSN_high_0_0/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_0_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/13.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/13.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.15.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_2_simulatedalignment.fasta"], "threads": 1, "jobid": 15, "rule": "run_phydms", "output": ["phydms/WSN_high_0_2/modelcomparison.md", "phydms/WSN_high_0_2/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_2/modelcomparison.md phydms/WSN_high_0_2/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_2_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/15.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/15.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.17.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_1_simulatedalignment.fasta"], "threads": 1, "jobid": 17, "rule": "run_phydms", "output": ["phydms/WSN_high_0_1/modelcomparison.md", "phydms/WSN_high_0_1/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_1/modelcomparison.md phydms/WSN_high_0_1/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_1_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/17.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/17.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.2.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_8_simulatedalignment.fasta"], "threads": 1, "jobid": 2, "rule": "run_phydms", "output": ["phydms/WSN_high_0_8/modelcomparison.md", "phydms/WSN_high_0_8/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_8/modelcomparison.md phydms/WSN_high_0_8/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_8_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/2.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/2.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.20.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_9_simulatedalignment.fasta"], "threads": 1, "jobid": 20, "rule": "run_phydms", "output": ["phydms/WSN_high_0_9/modelcomparison.md", "phydms/WSN_high_0_9/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_9/modelcomparison.md phydms/WSN_high_0_9/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_9_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/20.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/20.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.3.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_6_simulatedalignment.fasta"], "threads": 1, "jobid": 3, "rule": "run_phydms", "output": ["phydms/WSN_high_0_6/modelcomparison.md", "phydms/WSN_high_0_6/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_6/modelcomparison.md phydms/WSN_high_0_6/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_6_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/3.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/3.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.4.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_3_simulatedalignment.fasta"], "threads": 1, "jobid": 4, "rule": "run_phydms", "output": ["phydms/WSN_high_0_3/modelcomparison.md", "phydms/WSN_high_0_3/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_3/modelcomparison.md phydms/WSN_high_0_3/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_3_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/4.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/4.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.5.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_7_simulatedalignment.fasta"], "threads": 1, "jobid": 5, "rule": "run_phydms", "output": ["phydms/WSN_high_0_7/modelcomparison.md", "phydms/WSN_high_0_7/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_7/modelcomparison.md phydms/WSN_high_0_7/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_7_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/5.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/5.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.6.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_4_simulatedalignment.fasta"], "threads": 1, "jobid": 6, "rule": "run_phydms", "output": ["phydms/WSN_high_0_4/modelcomparison.md", "phydms/WSN_high_0_4/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_4/modelcomparison.md phydms/WSN_high_0_4/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_4_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/6.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/6.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.7bbfhsx8/snakejob.run_phydms.9.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"resources": {}, "params": {}, "input": ["_WSN_high_0_5_simulatedalignment.fasta"], "threads": 1, "jobid": 9, "rule": "run_phydms", "output": ["phydms/WSN_high_0_5/modelcomparison.md", "phydms/WSN_high_0_5/"], "cluster": {}, "local": false}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_5/modelcomparison.md phydms/WSN_high_0_5/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_5_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8 --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/9.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.7bbfhsx8/9.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.11.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_2_modelcomparison.md_simulatedalignment.fasta"], "jobid": 11, "params": {}, "output": ["phydms/WSN_high_0_2_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_2_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_2_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_2_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_2_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/11.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/11.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.14.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_5_modelcomparison.md_simulatedalignment.fasta"], "jobid": 14, "params": {}, "output": ["phydms/WSN_high_0_5_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_5_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_5_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_5_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_5_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/14.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/14.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.15.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_7_modelcomparison.md_simulatedalignment.fasta"], "jobid": 15, "params": {}, "output": ["phydms/WSN_high_0_7_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_7_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_7_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_7_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_7_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/15.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/15.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.18.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_9_modelcomparison.md_simulatedalignment.fasta"], "jobid": 18, "params": {}, "output": ["phydms/WSN_high_0_9_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_9_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_9_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_9_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_9_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/18.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/18.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.20.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_6_modelcomparison.md_simulatedalignment.fasta"], "jobid": 20, "params": {}, "output": ["phydms/WSN_high_0_6_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_6_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_6_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_6_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_6_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/20.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/20.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.3.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_0_modelcomparison.md_simulatedalignment.fasta"], "jobid": 3, "params": {}, "output": ["phydms/WSN_high_0_0_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_0_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_0_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_0_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_0_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/3.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/3.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.4.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_3_modelcomparison.md_simulatedalignment.fasta"], "jobid": 4, "params": {}, "output": ["phydms/WSN_high_0_3_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_3_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_3_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_3_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_3_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/4.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/4.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.6.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_1_modelcomparison.md_simulatedalignment.fasta"], "jobid": 6, "params": {}, "output": ["phydms/WSN_high_0_1_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_1_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_1_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_1_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_1_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/6.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/6.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.7.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_8_modelcomparison.md_simulatedalignment.fasta"], "jobid": 7, "params": {}, "output": ["phydms/WSN_high_0_8_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_8_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_8_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_8_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_8_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/7.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/7.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.ih1tumzp/snakejob.run_phydms.9.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"local": false, "rule": "run_phydms", "resources": {}, "threads": 1, "cluster": {}, "input": ["_WSN_high_0_4_modelcomparison.md_simulatedalignment.fasta"], "jobid": 9, "params": {}, "output": ["phydms/WSN_high_0_4_modelcomparison.md/modelcomparison.md", "phydms/WSN_high_0_4_modelcomparison.md"]}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_4_modelcomparison.md/modelcomparison.md phydms/WSN_high_0_4_modelcomparison.md --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_4_modelcomparison.md_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/9.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.ih1tumzp/9.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.13.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {}, "input": ["_WSN_high_0_5_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_5/modelcomparison.md", "phydms/WSN_high_0_5/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 13}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_5/modelcomparison.md phydms/WSN_high_0_5/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_5_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/13.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/13.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.16.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {}, "input": ["_WSN_high_0_6_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_6/modelcomparison.md", "phydms/WSN_high_0_6/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 16}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_6/modelcomparison.md phydms/WSN_high_0_6/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_6_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/16.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/16.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.17.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {}, "input": ["_WSN_high_0_1_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_1/modelcomparison.md", "phydms/WSN_high_0_1/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 17}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_1/modelcomparison.md phydms/WSN_high_0_1/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_1_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/17.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/17.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.3.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {}, "input": ["_WSN_high_0_4_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_4/modelcomparison.md", "phydms/WSN_high_0_4/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 3}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_4/modelcomparison.md phydms/WSN_high_0_4/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_4_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/3.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/3.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.4.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {}, "input": ["_WSN_high_0_0_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_0/modelcomparison.md", "phydms/WSN_high_0_0/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 4}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_0/modelcomparison.md phydms/WSN_high_0_0/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_0_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/4.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/4.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.run_phydms.5.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {}, "input": ["_WSN_high_0_9_simulatedalignment.fasta"], "output": ["phydms/WSN_high_0_9/modelcomparison.md", "phydms/WSN_high_0_9/"], "rule": "run_phydms", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 5}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake phydms/WSN_high_0_9/modelcomparison.md phydms/WSN_high_0_9/ --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files _WSN_high_0_9_simulatedalignment.fasta /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules run_phydms  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/5.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/5.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.14.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {"r": "8", "prefix": "_WSN_high_0_8", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_8_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 14}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_8_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/14.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/14.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.18.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {"r": "3", "prefix": "_WSN_high_0_3", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_3_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 18}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_3_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/18.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/18.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.19.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {"r": "7", "prefix": "_WSN_high_0_7", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_7_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 19}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_7_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/19.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/19.jobfailed"; exit 1)
-

### 2018-01-16 analysis/simulations/.snakemake/tmp.zxlmuba_/snakejob.simulate_alignment.20.sh

@@ -1,11 +1,0 @@
-#!/bin/sh
-# properties = {"params": {"r": "2", "prefix": "_WSN_high_0_2", "tree": "../HA/branch_lengths/phydms/WSN_high_0_ExpCM_HA_Doud_prefs_tree.newick"}, "input": [], "output": ["_WSN_high_0_2_simulatedalignment.fasta"], "rule": "simulate_alignment", "local": false, "threads": 1, "cluster": {}, "resources": {}, "jobid": 20}
-cd /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations && \
-/app/python3/3.4.1/bin/python3.4 -m snakemake _WSN_high_0_2_simulatedalignment.fasta --snakefile /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/Snakefile \
---force -j --keep-target-files --keep-shadow --keep-remote \
---wait-for-files /fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_ --latency-wait 180 \
---benchmark-repeats 1 \
---force-use-threads --wrapper-prefix https://bitbucket.org/snakemake/snakemake-wrappers/raw/ \
-   --nocolor \
---notemp --quiet --no-hooks --nolock --force-use-threads  --allowed-rules simulate_alignment  && touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/20.jobfinished" || (touch "/fh/fast/bloom_j/computational_notebooks/skhilton/2017/divergence_timing_manuscript/analysis/simulations/.snakemake/tmp.zxlmuba_/20.jobfailed"; exit 1)
-

### 2018-01-16 analysis/HA/branch_lengths/pymodules/utils.py
added small housecleaning scripts for branch lengths

@@ -1,0 +1,87 @@
+"""
+Python utilities for branch_lengths analysis
+Written by Sarah Hilton.
+"""
+from phydmslib import constants
+from ete3 import Tree
+import os
+import itertools
+from scipy.special import comb
+import pandas as pd
+
+def extract_model_name(model_string):
+    """
+    This function takes a tree output file from phydms and extracts the name of the model. 
+    input: 
+        `str`, tree file name
+    output:
+        `str`, model name
+        `int`, 1 if "control" model, 0 otherwise
+    """
+    if "YNGKP" in model_string:
+        if "YNGKP_M0" in model_string:
+            return "GY94", 0
+        elif "YNGKP_M5" in model_string:
+            return "GY94_gammaomega", 0
+        else:
+            raise ValueError("Cannot parse model string {0}".format(model_string))
+    elif "ExpCM" in model_string:
+        if "gammaomega" in model_string:
+            if "randomized_ExpCM" in model_string:
+                return "ExpCM_gammaomega_randomized", 1
+            elif "averaged_ExpCM" in model_string:
+                return "ExpCM_gammaomega_averaged", 1
+            else:
+                return "ExpCM_gammaomega", 0
+        else:
+            if "randomized_ExpCM" in model_string:
+                return "ExpCM_randomized", 1
+            elif "averaged_ExpCM" in model_string:
+                return "ExpCM_averaged", 1
+            else:
+                return "ExpCM", 0
+    elif "RAxML" in model_string:
+        return "GTR", 0
+    else:
+        raise ValueError("Cannot parse model string {0}".format(model_string))
+
+def translate_with_gaps(seq):
+    """
+    This function uses `phydmslib.constants` to translate a nucleotide sequence.
+    input: `str`
+    output: `str`
+    """
+    prot_seq = []
+    for i in range(0, len(seq),3):
+        codon = seq[i:i+3]
+        codon = constants.CODON_TO_INDEX[codon]
+        AA = constants.CODON_TO_AA[codon]
+        AA = constants.INDEX_TO_AA[AA]
+        prot_seq.append(AA)
+    return "".join(prot_seq)
+
+def calc_distances(treePath):
+    '''
+    This function calculates the distance between all pairs of tips on a given tree.
+    input:
+        treePath: `str`, file path to tree
+    output:
+        `pandas dataframe`, dataframe summarizing the tree. Columns include the
[…]

### 2018-01-16 analysis/HA/branch_lengths/pymodules/utils.py
small code tweaks

@@ -9,10 +9,12 @@ import itertools
 from scipy.special import comb
 import pandas as pd
 
+
 def extract_model_name(model_string):
     """
-    This function takes a tree output file from phydms and extracts the name of the model. 
-    input: 
+    This function takes a tree output file from phydms and extracts the name
+    f the model.
+    input:
         `str`, tree file name
     output:
         `str`, model name
@@ -24,7 +26,8 @@ def extract_model_name(model_string):
         elif "YNGKP_M5" in model_string:
             return "GY94_gammaomega", 0
         else:
-            raise ValueError("Cannot parse model string {0}".format(model_string))
+            raise ValueError("Cannot parse model string {0}"
+                             .format(model_string))
     elif "ExpCM" in model_string:
         if "gammaomega" in model_string:
             if "randomized_ExpCM" in model_string:
@@ -45,14 +48,16 @@ def extract_model_name(model_string):
     else:
         raise ValueError("Cannot parse model string {0}".format(model_string))
 
+
 def translate_with_gaps(seq):
     """
-    This function uses `phydmslib.constants` to translate a nucleotide sequence.
+    This function uses `phydmslib.constants` to translate a nucleotide
+    sequence.
     input: `str`
     output: `str`
     """
     prot_seq = []
-    for i in range(0, len(seq),3):
+    for i in range(0, len(seq), 3):
         codon = seq[i:i+3]
         codon = constants.CODON_TO_INDEX[codon]
         AA = constants.CODON_TO_AA[codon]
@@ -60,28 +65,31 @@ def translate_with_gaps(seq):
         prot_seq.append(AA)
     return "".join(prot_seq)
 
+
 def calc_distances(treePath):
     '''
-    This function calculates the distance between all pairs of tips on a given tree.
+    This function calculates the distance between all pairs of tips on a
+    given tree.
     input:
         treePath: `str`, file path to tree
     output:
-        `pandas dataframe`, dataframe summarizing the tree. Columns include the
-        names of the two sequences, the identifier for the branch (`seq_id`),
-        the two HA groups the sequences come from and the branch length.
+        `pandas dataframe`, dataframe summarizing the tree. Columns include
+        the names of the two sequences, the identifier for the branch
+        (`seq_id`), the two HA groups the sequences come from and the branch
+        length.
[…]

### 2018-01-16 analysis/HA/data/pymodules/utils.py

@@ -10,8 +10,8 @@ def createDateDictionary(sequences):
     """
     This function organizes the sequences by year of collection.
 
-    This function assumes that for a sequence collected in 2000 there is a string
-            '2000_2000' somewhere in the header.
+    This function assumes that for a sequence collected in 2000 there is a
+    string '2000_2000' somewhere in the header.
 
     A list of headers which cannot be parsed will be printed at the end.
 
@@ -21,29 +21,30 @@ def createDateDictionary(sequences):
     """
     datesDictionary = {}
 
-    for seq in sequences: #go through each sequence
-        regex = r"(/\d+_+\d+/)" #look for dates
+    for seq in sequences:  # go through each sequence
+        regex = r"(/\d+_+\d+/)"  # look for dates
         match = re.search(regex, seq)
-        if match: # if a date is found, extract the exact date
+        if match:  # if a date is found, extract the exact date
             match = re.search(r"(\d\d\d\d)", (match.group(0)))
             if match:
-                year =  int(match.group(0))
-                if year not in range(1918,2018): #make sure the extracted date makes sense
+                year = int(match.group(0))
+                if year not in range(1918, 2018):  # does the date makse sense?
                     pass
-                elif year in datesDictionary.keys(): #add sequence to the dictionary
+                elif year in datesDictionary.keys():  # add seq to dictionary
                     datesDictionary[year].extend([seq])
                 else:
                     datesDictionary[year] = [seq]
     return datesDictionary
 
+
 def translate_with_gaps(seq):
     """
-    This function uses `phydmslib.constants` to translate a nucleotide sequence.
+    This function uses `phydmslib.constants` to translate a nucleotide seq.
     input: `str`
     output: `str`
     """
     prot_seq = []
-    for i in range(0, len(seq),3):
+    for i in range(0, len(seq), 3):
         codon = seq[i:i+3]
         codon = constants.CODON_TO_INDEX[codon]
         AA = constants.CODON_TO_AA[codon]

### 2018-01-18 manuscript/manuscript.tex
tweaks to figures

@@
 
-\newcommand\skhcomment[1]{{\color{magenta}#1}}
+\newcommand\skhcomment[1]{{\color{cyan}#1}}
 \newcommand\jdbcomment[1]{{\color{red}#1}}
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.55\textwidth]{figures/model_feature.pdf}}
+\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
 \caption{\label{fig:model_feature}
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.75\textwidth]{figures/decay.pdf}}
+\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
 \caption{\label{fig:decay}
-\jdbcomment{I still think text could be bigger. Logo stacks still show trimming, maybe a bit tall? The Model key could probably be to the right or left.}
+\jdbcomment{I still think text could be bigger.}
+\skhcomment{Should I remove the second from the left site?}
 \textbf{Effect of stationary state and rate variation on long branch estimation.}
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.5\textwidth]{figures/tree_doud}}
-\caption{\label{fig:tree_doud}
+\centerline{\includegraphics[width=0.90\textwidth]{figures/empirical_trees}}
+\caption{\label{fig:empirical_trees}
+\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
+\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
+The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
+The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
+}
+\end{figure}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.70\textwidth]{figures/empirical_trees_2}}
+\caption{\label{fig:empirical_trees2}
 \textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
@@
 An alternative way of saying this is that the value of $\beta$ and the perceived "relevance" of the preferences are inversely correlated.
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/doud_compete}}
+\caption{\label{fig:doud_compete}
+\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
+\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
+\textbf{(B)} The phylogenetic tree of the "high" alignment. 
+The colors denote the alignment and the black circle denotes the focal H3 sequence. 
+\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
+\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
+Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
+The branch lengths are in average number of codon substitutions per site. 
+}
+\end{figure}
 
@@
 \caption{\label{suppfig:prefs_doud}
-\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.21$)} 
+\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.54$)} 
 }
@@
 \caption{\label{suppfig:prefs_lee}
-\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)} 
+\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.76$)} 
 }
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_average}}
+\centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
 \caption{\label{fig:prefs_average}
-\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.82$)}}
-\end{figure}
-
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_lee}}
-\caption{\label{fig:experiment_lee}
-\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
-\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
-\textbf{(B)} The phylogenetic tree of the "high" alignment. 
-The colors denote the alignment and the black circle denotes the focal H3 sequence. 
-\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
-\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
-Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
-The branch lengths are in average number of codon substitutions per site. 
-}
+\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.77$)}}
 \end{figure}
@@
            & $\omega=0.219225$\\ 
-           & $\phi_A$=0.386529\\ 
-           & $\phi_C$=0.17187\\
-           & $\phi_T$=0.230319\\
-           & $\pi_{r, A\left(X\right)}$:Doud\\
           YNGKP M0 & $\kappa=2.9984$\\
           & $\omega=0.09076$\\
-          & $\phi_{1A}=0.326007$\\
-          & $\phi_{1C}=0.136307$\\ 
-          &$\phi_{1G}=0.278377$\\
-          &$\phi_{2A}=0.382017$\\ 
-          &$\phi_{2C}=0.179763$\\ 
-          &$\phi_{2G}=0.194067$\\
-          &$\phi_{3A}=0.354364$\\
-          &$\phi_{3C}=0.190112$\\
-          &$\phi_{3G}=0.214605$\\
           YNGKP M5 & $\kappa=2.9984$\\
           & $\omega=0.09076$\\
-          & $\phi_{1A}=0.326007$\\
-          & $\phi_{1C}=0.136307$\\ 
-          &$\phi_{1G}=0.278377$\\
-          &$\phi_{2A}=0.382017$\\ 
-          &$\phi_{2C}=0.179763$\\ 
-          &$\phi_{2G}=0.194067$\\
-          &$\phi_{3A}=0.354364$\\
-          &$\phi_{3C}=0.190112$\\
-          &$\phi_{3G}=0.214605$\\
       \end{tabular}
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
-\caption{\label{fig:tree_lee}
-\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
-}
-\end{figure}
-
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
-\caption{\label{fig:tree_average}
-\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
-The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
-}
-\end{figure}
-
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
-\caption{\label{fig:experiment_doud}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/lee_competing}}
+\caption{\label{fig:lee_compete}
 \textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
-\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
-\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
+\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
+\textbf{(B)} The phylogenetic tree of the "high" alignment. 
 The colors denote the alignment and the black circle denotes the focal H3 sequence. 
-\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
+\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
 \textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 

### 2018-01-18 manuscript/manuscript.tex
outline first two results sections

@@
 
-\skhcomment{opening sentence or two about modeling site-to-site rate variation and/or purifying selection}
+\skhcomment{Section Outline:}
 
-Codon substitution models commonly include a parameter ($\omega$) which describes the relative rate of non-synonymous to synonymous change to account for purifying selection or site-to-site rate variation more generally.
-This parameter can be constrained to one value, representing a gene-wide average rate, or allowed to vary among sites following some statistical distribution, commonly a $\Gamma$-distribution (\ref{fig:model_feature}B).
-By allowing $\omega$ to vary among the sites, this models acknowledges that some sites evolve faster or slower than others. 
-However, in practical implementation the distribution is discretized into some number of categories (commonly four) and $\omega$ takes on the mean of each category.
-
-Another strategy to account for purifying selection and site-to-site rate variation is to use a model with a non-uniform stationary state (\ref{fig:model_feature}A). 
-In this framework, differences in rate between sites is modeled explicitly by site-specific amino-acid frequency parameters. 
-These parameters describe the ''preference" for a given amino acid at a given site. 
-According to this model, the rate of change at a "constrained" site, which prefers very few amino acids, is much slower than the rate of change at a tolerant site, which prefers many amino acids equally.
-
-The challenge of a model with a non-uniform stationary state is that all 19 * L (sites in the protein) amino acid preference parameters have to be defined. 
-In this study, we chose to use non-uniform stationary state model called ExpCM which defines these parameters using the measurements from a high-throughput functional assay called Deep Mutational Scanning (DMS).
-DMS measures the effect of every single amino acid change on some aspect of protein function. 
-We used two deep mutational scans of influenza virus surface protein hemagglutinin (HA) which measured the effect of mutations on viral growth.
-
-The models we chose to use in this study represent all four combinations of these two features (\ref{fig:model_feature}C). 
-The GY94 model has a uniform stationary state while the ExpCM has a non-uniform stationary state and each model can be implemented with a $\Gamma$-distributed $\omega$. 
-Therefore, we can example the effect of one feature (stationary state or $\Gamma\omega$) while controlling for the presence of the other feature.
+\begin{itemize}
+\item There are different strategies to account for purifying selection and site-to-site rate variation. 
+\begin{itemize}
+\item $\Gamma\omega$ rate variation 
+\begin{itemize}
+\item single parameter controlling relative rate of non-synonymous to synonymous substitutions can vary across sites following some statistical distribution (usually $\Gamma$)
+\item very common 
+\item practical implementation - discretized distribution with the mean of $K$ bins. 
+\end{itemize}
+\item stationary state
+\begin{itemize}
+\item Explicitly model each amino-acid frequency at each site in the protein 
+\item a more mechanistic description, no arbitrary statistical distribution 
+\item practical limitation - how do you estimate these parameters without overfitting the data? We use experimental measurements to define the parameters \textit{a priori}
+\end{itemize}
+\end{itemize}
+\item We can compare the effect of a model feature ($\Gamma\omega$ rate variation or non-uniform stationary state) while controlling the presence/absence of the other feature using models from the ExpCM and GY94 families.
+\end{itemize}
 
@@
 
-To explore the effect of substitution model choice on branch length estimation, we first examined how the relationship between sequence divergence and time is described by different models. 
-While substitution models can only accurately estimate branch length when there is a clear transformation from sequence divergence to time (branch length), every model has a saturation point at which the temporal signal has effectively decayed.
-This effect can been seen in the long, almost horizontal tails of \ref{fig:decay}, where the expectation of sequence divergence stays constant as time passes.
+\skhcomment{Section Outline:}
 
-While this saturation phenomenon is inevitable, the amount of time it takes to reach the saturation point and the sequence divergence at the point are dependent on the specific features of the substitution model.
-The addition of $\Gamma\omega$ to the GY94 model modulates the exact sequence divergence of the saturation point but the overall effect is minimal. 
-Furthermore, neither the GY94 nor the GY+$\Gamma\omega$ are site-specific and consequently the saturation points for each model are exactly same for every site. 
-In comparison, the saturation point of the ExpCM differs from site to site. 
-The ExpCM saturation point is very similar to the saturation point for the GY94 model at sites with a relatively high tolerance for mutations (\ref{fig:decay} sites 90 and 51). 
-But the constrained sites show saturation points with much higher expected sequence conservation (\ref{fig:decay} sites 413, 300, and 262) and/or a longer period of time until the saturation point is reached (\ref{fig:decay} sites 413 and 300).
-
-Finally, we can see that the difference between the ExpCM and the GY94 is due to the non-uniform stationary state and not just because the ExpCM "takes advantage" of more information (the DMS). 
-We can infer a site-specific dN/dS values from the ExpCM \cite{spielman2015relationship} and apply them to the GY94 model, thereby incorporating information from the DMS into a uniform stationary state model.
-This model takes a longer time to reach the saturation point but does the saturation point does have the same sequence divergence as the GY94 model. 
-The underestimation of long branches will be an issue for all models but especially models with uniform stationary states, even if great care is taken to model rate variation.
-
-\skhcomment{be more clear about how this plot shows branch lengths }
+\begin{itemize}
+\item Substitution models are only useful for inferring branch lengths when they can accurately transform sequence divergence to time (branch length)
+\item There is a saturation point in every substitution model where the temporal signal is lost (long tails in figure). 
+\begin{itemize}
+\item In the long tails the expected sequence divergence remains constant as time increases. 
+\item The lost of temporal signal will affect long branches specifically.
+\end{itemize}
+\item Different model features (\ref{fig:model_feature}) can affect \textit{how long} it takes to reach the saturation point and the \textit{sequence divergence} at the saturation point 
+\begin{itemize}
+\item Adding $\Gamma\omega$ rate variation affects the ``path" to the saturation point but affects the sequence divergence at the saturation point minimally.
+Importantly, the saturation point is identically for every site in the protein.
+\skhcomment{These $\Gamma\omega$ parameters might not the best to show the effect of $\Gamma\omega$ on long branch estimation.}
+\item Adding a non-uniform stationary state creates a unique saturation point for each site in the protein.
+Sites which can tolerate many mutations have saturation points which are quite similar to the GY94 saturation point.
+However, constrained sites show a dramatic difference.
+\item This difference cannot be recapitulated by more complex modeling of the $\omega$ parameter.
+We can transform the GY94 model into a completely site-specific value by inferring a unique $\omega$ value from the ExpCM for each site in the protein.
+This model shows a substantial difference but does not recover the full effect of the ExpCM.
+\end{itemize}
+\item No matter how ``well" you model site-to-site rate variation, you will underestimate \textit{very} long branches with a uniform stationary state. 
+\end{itemize}
 
@@
 
-Next, we explored how accurately these models infer branch length when site-specificity is assumed, as expected in natural sequences.  
-We simulated sequences under an ExpCM defined by the amino-acid preferences measured by \cite{doud2016accurate} along an influenza tree. 
-We then re-inferred the branch lengths using the models described in \ref{fig:model_feature}.
-
-The two models with uniform stationary states, GY94 and GY94+$\Gamma\omega$, both underestimate long branches (> ~5 codon substitutions per site). 
-The degree of underestimation becomes more pronounced as the branch length increases, as shown by the curve of the correlation. 
-The addition of $\Gamma\omega$ rate variation does not mitigate the effect as both models show roughly the same degree of underestimation.
-
-Conversely, the ExpCM and ExpCM+$\Gamma\omega$ both avoid the underestimation bias, even at the longest branch lengths measured. 
-As the branch length increases, the variance of the branch length estimation also increases (grey points). 
-But, unlike the GY94 and the GY94+$\Gamma\omega$, the variance is unbiased in regards to under- or overestimation of the true branch length.
-
-Finally, the randomized control ExpCM and ExpCM+$\Gamma\omega$ show that the importance is truly in site-\textit{specificity}. 
-If the preferences are randomized along sites in the protein, the model exhibits the same degree of underestimation as the GY94 and GY+$\Gamma\omega$ models. 
-This underscores the point that the non-uniform stationary state models are site-specific and that mismatches are just as much of a problem as not accounting for these differences at all.
-
-Overall, the simulations show that if site-specificity is true, the addition of $\Gamma\omega$ will still have problems at extremely long branches.
-
 \subsection*{empirical Data}
-
-Next we compared the branch lengths inferred by the different models on empirical data. 
-We looked at the influenza virus surface protein hemagglutinin (HA). 
-HA proteins cluster into tight phylogenetic groups with long branches in between them. 
-Furthermore, HAs are extremely diverse. 
-The HAs which commonly circulate in human viruses (H1 and H3) are only about 40\% identical on the amino acid level.
 
@@
 
-We took a representative sample of HA from 14 of the 18 groups and optimized the branch lengths of the tree using the models described in \ref{fig:model_feature}. 
-First, we used an ExpCM defined by the site-specific amino-acid preferences measured by deep mutational scanning from \cite{doud2016accurate}. 
-These amino-acid preferences were measured in the genetic background of the H1 strain A/Wilson-Smith/1993. 
-We designate this strain (and more broadly the H1 subtype) as the focal sequence of the model.
-
-\ref{fig:tree_doud} shows the results. 
-Comparing GY94 to GY94+$\Gamma\omega$ or ExpCM to ExpCM+$\Gamma\omega$ shows the addition of $\Gamma\omega$ causes branch length extension independent of the presence of preferences. 
-In addition, comparing comparing GY94 to ExpCM or GY94+$\Gamma\omega$ to ExpCM+$\Gamma\omega$ shows that the addition of preferences causes branch length extension from the \textit{focal} sequence independent of the presence of $\Gamma\omega$.
-Together this suggests that site-specificity of the model is important because branch lengths are shorter in the absence of preferences. 
-However, the preferences used here to define the ExpCM are clearly not equally relevant across all parts of the tree. 
-Specifically, they are most relevant near the strain that they were measured in. 
-This not at all unexpected because epistasis, where the effect of a mutation at one site is dependent on the identity of another site is well documented. 
-It would surprising if epistasis \textit{was not} a factor at these large divergences across the HA tree.
-
-This becomes even more apparent when the performance of the ExpCM defined by H1 preferences is compared to the ExpCM with H3 preferences. 
-Namely, the exact same pattern is observed. 
-There is extension across the whole tree when $\Gamma\omega$ is added to the model, independent of whether or not preferences are included in the model, and there is an extension of branch lengths from the focal sequence when preferences are added, independent of whether or not $\Gamma\omega$ is included in the model.
-Finally, we defined an ExpCM with the average of the H1 and H3 preferences, which can be seen as a rough proxy for the model sampling both preference sets. 
-In this model, there is an extension in branch length from \textit{both} focal sequences.
-
-These three sets of trees imply that site-specificity is important and not constant. 
-This observation is not surprising. 
-Epistasis, where the effect of a mutation at a site is dependent on the identity of another site in the protein, is well documented. 
-It is not all together surprising that there are epistatic interactions between H1 and H3 due to the large sequence divergence between the two strains. 
-In contrast, the deep mutational scanning preferences are measured in a single genetic background and therefore only measure the effect of mutations in one, specific genetic context.  
-The difference between the trees using the H1 preferences and the H3 preferences is most likely that the preferences "shifted" due to the genetic context. 
-The model with the average preferences could be seen as a proxy for a model which is able to sample both preference sets.
-
 \subsection{Competing effects of shifting preferences and long branches.}
-We explored these competing effects of shifting preferences and long branches by comparing the effects of the model across trees which vary in overall divergence levels. 
-We used the $\beta$ parameter of the ExpCM model as a proxy for the degree to which the preferences have shifted. 
-In each run of the model, the parameter $\beta$ is fit and effectively rescales the preferences. 
-When $\beta = 1$, we infer that selection in nature and selection in the lab prefers the same amino acids and with the same level of stringency. 
-When $\beta > 1$, we infer that selection in nature prefers the same amino acids as selection in lab but with much higher stringency and when $\beta < 1$ we infer that selection in lab and selection in nature prefer different amino acids or at a \textit{much} lower stringency.
-An alternative way of saying this is that the value of $\beta$ and the perceived "relevance" of the preferences are inversely correlated.
 
@@
 
-For each preference set, we took a subset the full HA tree shown in \ref{fig:tree_doud} or \ref{fig:tree_lee} to create a tree with either low divergence from the focal sequence (low) or intermediate divergence from the focal sequence (intermediate) and estimated the branch lengths on these trees using the same models as above.
-As we hypothesized above, as the overall divergence of the tree increases the $\beta$ value decreases. 
-That is as the tree begins to include more and more sequences which are highly diverged from the focal sequence, the less relevant the preferences appear to be overall. 
-As divergence increases the relevance of the preferences decreases.
-
-On the other hand, the effect of having a site-specific stationary state is more apparent when there are long branches on the tree.
-There is no real bias in estimation in any of the models for the low or intermediate trees. 
-This is not surprising as \ref{fig:decay} shows that the differences in these models really only matters at the long branches. 
-The underestimation is only apparent for the "high divergence" alignment (aka the full tree). 
-Here it becomes clear that you have these competing effects. 
-It is most important to have the site-specificity at these long branches but it is also where the site-specific preferences are the least relevant.
-
-For set of preferences, 
 \section*{Conclusion}

### 2018-01-18 manuscript/manuscript.tex
outlined the third results section

@@
 
+\skhcomment{Section Outline:}
+
+\begin{itemize}
+\item To test the effect of substitution model on data with site-specific amino acid frequencies, we simulated sequences under an ExpCM and re-inferred the branch lengths using different substitution models.
+\item Uniform stationary state models underestimate long branches
+\begin{itemize}
+\item GY94 and GY94 + $\Gamma\omega$ both underestimate long branches 
+\item The \textit{averaged} ExpCMs also underestimate long branches. These control models have uniform stationary states but still derive some information from the experiments. 
+\item Non-uniform stationary state models accurately estimate long branches
+\end{itemize}
+\item If sequences have site-specific amino acid frequencies then non-uniform stationary state models, even with $\Gamma\omega$, will underestimate long branches.
+\end{itemize}
+
 \subsection*{empirical Data}

### 2018-01-18 manuscript/manuscript.tex
rough outlines of last two sections

@@
 
-\subsection{Competing effects of shifting preferences and long branches.}
+\skhcomment{Section Outline:}
+
+\begin{itemize}
+\item We tested the effect substitution model choice on branch length estimation using influenza HA sequences. 
+\begin{itemize}
+\item These sequences presumably evolve with site-specific frequencies but unlike the simulations we don't \textit{know} the model
+\item These sequences are from fairly diverged proteins. Some sequences share only 40\% identity on the amino acid level 
+\item HA has small, closely related subtypes with long branches between subtypes and groups. 
+\end{itemize}
+\item Both $\Gamma\omega$ and non-uniform stationary states result in an extension of branch lengths. 
+\begin{itemize}
+\item The addition of $\Gamma\omega$ extends branches independent of preferences. 
+\item The addition of preferences extends branches \textit{from the focal sequence} independent of $\Gamma\omega$. 
+\end{itemize}
+\item The ExpCM does not have a uniform effect across the whole tree. 
+\begin{itemize}
+\item This is not too surprising. DMS makes accurate measurements in a given genetic background. 
+\item You would expect epistatic effects over such a large divergence distance. ``Shifting preferences"
+\end{itemize}
+\end{itemize}
+
+\subsection*{Competing effects of shifting preferences and long branches.}
 
@@
 \end{figure}
+
+\begin{itemize}
+\item We investigated the competing effects of shifting preferences and long branch estimation.
+That is, our preferences are most relevant to sequences they are close to but the effect of a non-uniform stationary is only observed on long branches.
+\begin{itemize}
+\item We broke the large tree into small, sub trees with differing maximum divergence levels. 
+\item We estimated the branch lengths using the same models we used for empirical data above. 
+\end{itemize}
+\item The ``relevance" of the preferences is inversely correlated with sequence divergence. 
+\begin{itemize}
+\item We used the ExpCM stringency parameter as the measure of preference ``relevance". The larger the $\beta$ value the more ``relevant" the preferences. 
+\item As the maximum divergence of the tree increases the $\beta$ value decreases. 
+\end{itemize}
+\item All models estimate short branches relatively equally.
+
+\end{itemize}
 

### 2018-01-19 manuscript/manuscript.tex
small note on supp figure
small tweaks to empirical trees and decay

@@
 \textbf{Comparison of substitution model features.}
-\jdbcomment{This figure is too big. Maybe put the three panels side-by-side, increase the space between panels, and increase the font sizes. The logo stacks are kind of tall relative to their width.}
 Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
@@
 \caption{\label{fig:decay}
-\jdbcomment{I still think text could be bigger.}
-\skhcomment{Should I remove the second from the left site?}
 \textbf{Effect of stationary state and rate variation on long branch estimation.}
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.90\textwidth]{figures/empirical_trees}}
+\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees}}
 \caption{\label{fig:empirical_trees}
-\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
-\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
-}
-\end{figure}
-
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.70\textwidth]{figures/empirical_trees_2}}
-\caption{\label{fig:empirical_trees2}
 \textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
@@
 \caption{\label{fig:lee_compete}
+\skhcomment{something weird is going on with the low alignment. I am going to go back and look at this.}
 \textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 

### 2018-01-19 analysis/decay_to_stationary/decay_to_stationary_plot.py
re-ran after bug fix in phydms

@@ -18,11 +18,12 @@ import phydmslib.file_io
 import os
 
 
-## Define the models
+# Define the models
 def read_phydms_model_params(param):
-    params = pd.read_csv(param, engine="python", sep= " = ", header=None)
+    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
     params = dict(zip(params[0], params[1]))
     return params
+
 
 def create_model_YNGKP_M0(param, nsites):
     """
@@ -31,11 +32,16 @@ def create_model_YNGKP_M0(param, nsites):
     params = read_phydms_model_params(param)
     e_pw = [[] for x in range(3)]
     for i in range(3):
-        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys() if x.startswith("phi{0}".format(i))])
-        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in range(N_NT)]
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
+                                               if x.startswith("phi{0}"
+                                               .format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
+                   range(N_NT)]
     e_pw = scipy.array(e_pw)
-    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], omega=params["omega"], mu=0.3,
-            freeparams=['mu'])
+    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"],
+                                     omega=params["omega"], mu=0.3,
+                                     freeparams=['mu'])
+
 
 def create_model_YNGKP_M5(param, nsites):
     """
@@ -44,12 +50,18 @@ def create_model_YNGKP_M5(param, nsites):
     params = read_phydms_model_params(param)
     e_pw = [[] for x in range(3)]
     for i in range(3):
-        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys() if x.startswith("phi{0}".format(i))])
-        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in range(N_NT)]
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
+                                               if x.startswith("phi{0}"
+                                               .format(i))])
[…]

### 2018-01-22 analysis/HA/data/references/make_hybrid_prefs.py
added doud and lee preferences with only the shared sites

@@ -32,7 +32,11 @@ def main():
     # subset to only the shared sites
     doud = doud[doud["site"].isin(hybrid_numbers["WSN"].tolist())]
     lee = lee[lee["site"].isin(hybrid_numbers["Perth"].tolist())]
+    doud["site"] = [x+1 for x in range(len(doud))]
+    lee["site"] = [x+1 for x in range(len(lee))]
     assert len(doud) == len(lee)
+    doud.to_csv("HA_hybridDoud_prefs.csv", index=False)
+    lee.to_csv("HA_hybridLee_prefs.csv", index=False)
 
     # get rid of the sites column and reset the index
     doud = doud.reset_index()

### 2018-01-22 manuscript/manuscript.tex
tweaks to figures
updated manuscript

@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/doud_compete}}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/doud_compete_2}}
 \caption{\label{fig:doud_compete}
@@
 \caption{\label{suppfig:prefs_doud}
-\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.54$)} 
+\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.19$)} 
+\skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.}
 }
@@
 \caption{\label{suppfig:prefs_lee}
-\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.76$)} 
+\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)}
+\skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.} 
 }
@@
 \textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.77$)}}
+\skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.}
 \end{figure}
@@
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/lee_competing}}
+\begin{suppfig}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/lee_compete}}
 \caption{\label{fig:lee_compete}
-\skhcomment{something weird is going on with the low alignment. I am going to go back and look at this.}
 \textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
@@
 }
-\end{figure}
+\end{suppfig}
 

### 2018-01-23 manuscript/manuscript.tex

@@
 \begin{abstract}
-\textit{$\leq$ 250 words}
+\textit{$\leq$ 250 words, currently 164}
 
-Molecular dating techniques is often used to estimate the divergence time of many viruses. 
-However, these estimates are substantially younger than estimates from methods which do not rely on viral phylogenies. 
+Molecular dating techniques are often used to estimate the origin date of modern, circulating viruses. 
+These estimates are often substantially younger than estimates from methods which do not rely on viral phylogenies. 
 This discrepancy is thought to be caused in part by inadequate modeling of purifying selection leading to branch length underestimation. 
-Here, we show that substitution models informed by empirical measurements of mutational model constraint better than traditional models and extend branch lengths. 
+Here, we show that substitution models informed by empirical measurements of mutational tolerance model constraint better than traditional models and extend branch lengths. 
 We used models informed by deep mutational scanning experiments performed in two, highly diverged influenza virus hemagglutinin homologs to optimize the branch lengths of a phylogenetic tree. 
@@
 Overall, this study underscores the importance of modeling purifying selection when estimating branch lengths and, by extension, divergence dates.
-\skhcomment{from JDB: also shows a way to actually do this in addition to showing importance.}
 
@@
 \begin{itemize}
-\item Substitution models are only useful for inferring branch lengths when they can accurately transform sequence divergence to time (branch length)
-\item There is a saturation point in every substitution model where the temporal signal is lost (long tails in figure). 
+\item Substitution models are only useful for inferring branch lengths when they can accurately transform sequence divergence to time (branch lengths)
+\item Every substitution model has a saturation point where this temporal signal is lost.  
 \begin{itemize}
 \item In the long tails the expected sequence divergence remains constant as time increases. 
-\item The lost of temporal signal will affect long branches specifically.
+\item The saturation point will affect long branches more than short branches.
 \end{itemize}
@@
 Importantly, the saturation point is identically for every site in the protein.
-\skhcomment{These $\Gamma\omega$ parameters might not the best to show the effect of $\Gamma\omega$ on long branch estimation.}
-\item Adding a non-uniform stationary state creates a unique saturation point for each site in the protein.
+(green and blue)
+\item Adding a non-uniform stationary state creates a unique saturation point for each site in the protein (red).
 Sites which can tolerate many mutations have saturation points which are quite similar to the GY94 saturation point.
-However, constrained sites show a dramatic difference.
+Constrained sites show a dramatic difference.
+At a constrained site, ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at the site is expected to be small.
 \item This difference cannot be recapitulated by more complex modeling of the $\omega$ parameter.
-We can transform the GY94 model into a completely site-specific value by inferring a unique $\omega$ value from the ExpCM for each site in the protein.
-This model shows a substantial difference but does not recover the full effect of the ExpCM.
+We can transform the GY94 model into a  site-specific model by inferring a unique $\omega$ value from the ExpCM for each site in the protein.
+This model (yellow) takes longer to reach the saturation point than the GY94 model but ultimately has the same divergence level at the saturation point.
 \end{itemize}
@@
 \begin{itemize}
-\item GY94 and GY94 + $\Gamma\omega$ both underestimate long branches 
-\item The \textit{averaged} ExpCMs also underestimate long branches. These control models have uniform stationary states but still derive some information from the experiments. 
-\item Non-uniform stationary state models accurately estimate long branches
+\item Both the GY94 and the GY94+$\Gamma\omega$ underestimate long branches. The $\Gamma\omega$ rate variation scheme is not enough to prevent this underestimation at every long branches. 
+\item Another uniform stationary state model, the averaged ExpCM, also underestimates long branches. This model derives some information from the experiments but lacks site-specificity. 
+\item Non-uniform stationary state models accurately estimate long branches (both the ExpCM and ExpCM+$\Gamma\omega$
 \end{itemize}
@@
 \begin{itemize}
-\item These sequences presumably evolve with site-specific frequencies but unlike the simulations we don't \textit{know} the model
+\item These sequences presumably evolve with site-specific frequencies but unlike the simulations we do not \textit{know} the model
 \item These sequences are from fairly diverged proteins. Some sequences share only 40\% identity on the amino acid level 
-\item HA has small, closely related subtypes with long branches between subtypes and groups. 
+\item The HA tree has closely related subtypes with long branches between subtypes.
 \end{itemize}
@@
 \end{itemize}
-\item The ExpCM does not have a uniform effect across the whole tree. 
+\item The extension of branch length from the focal sequence shows that the ExpCM does not have equal "relevance" across the whole tree. 
 \begin{itemize}
-\item This is not too surprising. DMS makes accurate measurements in a given genetic background. 
-\item You would expect epistatic effects over such a large divergence distance. ``Shifting preferences"
+\item This result is not entirely surprising. DMS measure the effects of single amino-acid changes in the context of a single genetic background. 
+\item It is expected that there would be epistatic changes across such a highly diverged phylogenetic tree. ``Shifting preferences".
 \end{itemize}
@@
 
+\skhcomment{Section Outline:}
+
 \begin{itemize}
-\item We investigated the competing effects of shifting preferences and long branch estimation.
-That is, our preferences are most relevant to sequences they are close to but the effect of a non-uniform stationary is only observed on long branches.
+\item We investigated the competing effects of ``shifting preferences" and long branches by comparing ExpCMs with different "relevances"
+\item We use the fit stringency parameter $\beta$ as a measure of relevance. 
 \begin{itemize}
-\item We broke the large tree into small, sub trees with differing maximum divergence levels. 
-\item We estimated the branch lengths using the same models we used for empirical data above. 
+\item The stringency parameter rescales the preferences.
+A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength.
+\item There is an inverse relationship between the stringency parameter and overall sequence divergence. As the overall divergence of the tree increases, the stringency parameter decreases. 
 \end{itemize}
-\item The ``relevance" of the preferences is inversely correlated with sequence divergence. 
+\item \skhcomment{hypothetical data} There is also a positive relationship between the stringency parameter and branch lengths 
 \begin{itemize}
-\item We used the ExpCM stringency parameter as the measure of preference ``relevance". The larger the $\beta$ value the more ``relevant" the preferences. 
-\item As the maximum divergence of the tree increases the $\beta$ value decreases. 
+\item We optimized the same tree with ExpCMs with different stringency parameters. 
+\item The ExpCMs with higher stringency values estimate longer branches than the ExpCMs with lower stringency values. 
 \end{itemize}
-\item All models estimate short branches relatively equally.
+\end{itemize}
 
-\end{itemize}
+\subsection*{texttt{phylobayes}}
 

### 2018-01-23 manuscript/manuscript.tex
edits to abstract

@@
 
-\newcommand\skhcomment[1]{{\color{cyan}#1}}
-\newcommand\jdbcomment[1]{{\color{red}#1}}
+\newcommand\skhcomment[1]{{\color{cyan}[#1]}}
+\newcommand\jdbcomment[1]{{\color{red}[#1]}}
 
@@
 
-\title{Experimentally Informed Site-Specific Substitution Models Substantially Deepen Viral Divergence Estimates}
+\title{Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages}
 
 \author
-{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2,}$\\
+{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2}$\\
 \\
@@
 \begin{abstract}
-\textit{$\leq$ 250 words, currently 164}
-
-Molecular dating techniques are often used to estimate the origin date of modern, circulating viruses. 
-These estimates are often substantially younger than estimates from methods which do not rely on viral phylogenies. 
-This discrepancy is thought to be caused in part by inadequate modeling of purifying selection leading to branch length underestimation. 
-Here, we show that substitution models informed by empirical measurements of mutational tolerance model constraint better than traditional models and extend branch lengths. 
-We used models informed by deep mutational scanning experiments performed in two, highly diverged influenza virus hemagglutinin homologs to optimize the branch lengths of a phylogenetic tree. 
-For each experimentally informed model, we observed extension in branch length from the experiment's focal sequence. 
-This extension in branch length due to explicit modeling of site-specific purifying selection is observed in the presence and absence of standard methods for modeling site-to-site variation. 
-Overall, this study underscores the importance of modeling purifying selection when estimating branch lengths and, by extension, divergence dates.
-
+\noindent \skhcomment{$\leq$ 250 words, currently 164}   
+Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
+Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
+For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
+This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
+Here we show that substitution models informed by experimental measurements of site-specific purifying selection lengthen the branches on phylogenies of influenza virus hemagglutinin.
+This deepening of branch lengths is due to better modeling how site-specific amino-acid preferences affect the stationary state of the substitution models, and is independent of the branch-length-extension due to modeling site-to-site variation in substitution rate.
+We show that the deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
+However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
+Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state as well as the substitution rate when estimating deep divergence times. 
 \end{abstract}

### 2018-01-26 manuscript/manuscript.tex

@@
 \begin{itemize}
+\item It is well known that different protein sites evolve under constraints
 \item There are different strategies to account for purifying selection and site-to-site rate variation. 
@@
 \end{itemize}
+\item Differences in stationary state are manifested in rate differences (e.g., Spielman work), but the resulting differences in rate don't fully capture the site-specific stationary state
+\item These are not mutually exclusive, you can imagine why there would be both
 \end{itemize}
-\item We can compare the effect of a model feature ($\Gamma\omega$ rate variation or non-uniform stationary state) while controlling the presence/absence of the other feature using models from the ExpCM and GY94 families.
 \end{itemize}
@@
 \begin{itemize}
-\item Substitution models are only useful for inferring branch lengths when they can accurately transform sequence divergence to time (branch lengths)
-\item Every substitution model has a saturation point where this temporal signal is lost.  
+\item Along a single branch, a substitution model transforms sequence divergence to branch lengths, which under a molecular clock is assumed proportional time
+\item Every substitution model has a stationary state, which is the expected sequence divergence at very long time.  
 \begin{itemize}
@@
 \end{itemize}
-\item Different model features (\ref{fig:model_feature}) can affect \textit{how long} it takes to reach the saturation point and the \textit{sequence divergence} at the saturation point 
+\item Some model features (\ref{fig:model_feature}) affect \textit{how long} it takes to reach the stationary state, whereas others affect the stationary state itself: 
 \begin{itemize}
-\item Adding $\Gamma\omega$ rate variation affects the ``path" to the saturation point but affects the sequence divergence at the saturation point minimally.
-Importantly, the saturation point is identically for every site in the protein.
+\item Adding $\Gamma\omega$ rate variation affects the ``path" to the stationary state but not the stationary state itself.
+Importantly, the stationary state is identically for every site in the protein if there isn't site-specific amino-acid preferences.
 (green and blue)
-\item Adding a non-uniform stationary state creates a unique saturation point for each site in the protein (red).
+\item Adding a non-uniform stationary state creates a unique stationary state for each site in the protein (red).
 Sites which can tolerate many mutations have saturation points which are quite similar to the GY94 saturation point.

### 2018-01-30 manuscript/figures/temp_compete.py
added debugging scripts

@@ -1,0 +1,132 @@
+"""
+This script is the scratch script for optimizing a model with different betas.
+
+The script is a modified (simplified) verison of the `phydms` script
+
+SKH 20180122
+"""
+
+import sys
+import os
+import re
+import logging
+import random
+import time
+import math
+import operator
+import multiprocessing
+import warnings
+warnings.simplefilter('always')
+import scipy
+import scipy.stats
+import statsmodels.sandbox.stats.multicomp
+import Bio.Phylo
+import phydmslib
+from phydmslib.constants import *
+import phydmslib.file_io
+import phydmslib.parsearguments
+import phydmslib.models
+import phydmslib.treelikelihood
+import pandas as pd
+
+def read_phydms_model_params(param):
+    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    return params
+
+def create_model_ExpCM(param, prefs, beta):
+    """
+    This is the basic `ExpCM` from `phydms`
+
+    `omega` is set to 1.0
+    """
+    params = read_phydms_model_params(param)
+    prefs = phydmslib.file_io.readPrefs(prefs)
+    sites = sorted(prefs.keys())
+    prefs = [prefs[r] for r in sites]
+    params["phiT"] = 1 - sum([params[x] for x in params.keys() if
+                              x.startswith("phi")])
+    phi = scipy.array([params["phi{0}".format(INDEX_TO_NT[x])] for x in
+                       range(N_NT)])
+    return phydmslib.models.ExpCM(prefs, kappa=params["kappa"],
+                                  omega=params["omega"], beta=beta,
+                                  mu=0.3, phi=phi, freeparams=['mu'])
+
+def main():
+    """Main body of script."""
+
+    alignment = "../../analysis/HA/data/subsample/HA_hybrid_lowH1_0.fasta"
+    prefs_fname = "../../analysis/HA/data/references/HA_hybridDoud_prefs.csv"
+    params_fname = "../../analysis/HA/branch_lengths/phydms/hybrid_lowH1_0_ExpCM_HA_hybridDoud_prefs_modelparams.txt"
+    tree_fname = "../../analysis/HA/branch_lengths/phydms/hybrid_lowH1_0_RAxML_tree.newick"
+    # betas = [1.56, 1.34, 1.22]
+    betas = [1.00]
+
+    for beta in betas:
+        outprefix = "_temp_{0}/".format(beta)
[…]

### 2018-01-30 analysis/decay_to_stationary/decay_to_stationary_plot.py
reran decay analysis with bug fix

@@ -97,7 +97,7 @@ def create_model_YNGKP_wr(param, spielman_wr):
                    range(N_NT)]
     e_pw = scipy.array(e_pw)
     return phydmslib.models.YNGKP_M0(e_pw, 1, kappa=params["kappa"],
-                                     omega=spielman_wr,
+                                     omega=spielman_wr * params["omega"],
                                      mu=0.3, freeparams=['mu'])
 
 
@@ -190,13 +190,13 @@ def main():
     # Set up the parameter files
     phydms_dir = "../HA/branch_lengths/phydms/"
     prefs_dir = "../HA/data/references/"
-    ExpCM_modelparams_fname = ("{0}WSN_low_0_ExpCM_HA_Doud_prefs_modelparams"
+    ExpCM_modelparams_fname = ("{0}hybrid_lowH1_0_ExpCM_HA_hybridDoud_prefs_modelparams"
                                ".txt".format(phydms_dir))
-    YNGKP_M0_modelparams_fname = ("{0}WSN_low_0_YNGKP_M0_modelparams"
+    YNGKP_M0_modelparams_fname = ("{0}hybrid_lowH1_0_YNGKP_M0_modelparams"
                                   ".txt".format(phydms_dir))
-    YNGKP_M5_modelparams_fname = ("{0}WSN_low_0_YNGKP_M5_modelparams"
+    YNGKP_M5_modelparams_fname = ("{0}hybrid_lowH1_0_YNGKP_M5_modelparams"
                                   ".txt".format(phydms_dir))
-    prefs_fname = "{0}HA_Doud_prefs.csv".format(prefs_dir)
+    prefs_fname = "{0}HA_hybridDoud_prefs.csv".format(prefs_dir)
     if not os.path.isdir("outputs"):
         os.makedirs("outputs")
     # define the maximum amount of time

### 2018-01-30 manuscript/figures/graft_tree.py
removed debugging scripts

@@ -1,57 +1,0 @@
-"""
-The purpose of this script is to graft a small tree ("cion") onto a larger
-tree ("stock").
-
-SKH 20180120
-"""
-
-from ete3 import Tree
-import argparse
-import sys
-
-
-class ArgumentParserNoArgHelp(argparse.ArgumentParser):
-    """Like *argparse.ArgumentParser*, but prints help when no arguments."""
-    def error(self, message):
-        sys.stderr.write('error: {0}\n\n'.format(message))
-        self.print_help()
-        sys.exit(2)
-
-
-def ParseArguments():
-    """Argument parser for script."""
-    parser = ArgumentParserNoArgHelp(
-            description='Graft trees',
-            formatter_class=argparse.ArgumentDefaultsHelpFormatter,
-            )
-    parser.add_argument('stock', help='File name of stock tree.')
-    parser.add_argument('cion', help='File name of cion tree.')
-    parser.add_argument('out_prefix', help='Prefix for the final tree file')
-    return parser
-
-
-def tree_graft(stock, cion):
-    names = []
-    for node in cion.traverse("postorder"):
-        if len(node.name) >= 1:
-            names.append(node.name)
-    target_node = stock.get_common_ancestor(names)
-    for child in target_node.get_children():
-        child.detach()
-    cion_root = cion.get_tree_root()
-    for child in cion_root.get_children():
-        target_node.add_child(child)
-    return stock
-
-
-def main():
-    args = vars(ParseArguments().parse_args())
-    stock = args["stock"]
-    cion = args["cion"]
-
-    catch = tree_graft(Tree(stock), Tree(cion))
-    catch.write(format=1, outfile="{0}.newick".format(args["out_prefix"]))
-
-
-if __name__ == '__main__':
-    main()

### 2018-02-01 analysis/phylobayes/plot.R
first pass analysis with phylobayes

@@ -1,0 +1,22 @@
+library(ggplot2)
+library(cowplot)
+
+black = "black"
+blue = "#439FC3"
+green = "#76A88D"
+orange = "#F9842A"
+red = "#B52A24"
+pink = "#CC79A7"
+purple = "#8E7790"
+
+setwd("/Users/sarah/Google Drive/GS/lab/divergence_timing_manuscript/analysis/phylobayes")
+doud = read.csv("outputs/model_by_model_norm.csv")
+m <- ceiling(max(max(doud$distance1), max(doud$distance2)))
+doud_branch_lengths = ggplot(doud, aes(distance1, distance2)) + geom_abline(intercept=0, slope=1, color = "grey", linetype="dashed") + geom_point(size=1.0, color="grey") 
+doud_branch_lengths = doud_branch_lengths + facet_grid(divergence_level~ model_pair)
+doud_branch_lengths = doud_branch_lengths + theme(strip.background = element_blank(), strip.text = element_text(size=8), legend.position="bottom", panel.spacing = unit(2, "lines"), panel.border = element_rect(colour = "black"))
+doud_branch_lengths = doud_branch_lengths + xlim(c(0,m)) + ylim(c(0,m)) + coord_fixed() 
+doud_branch_lengths = doud_branch_lengths + xlab("model 1") + ylab("model 2") 
+doud_branch_lengths = doud_branch_lengths + geom_point(size = 0.5, data = subset(doud, from_ref == 'WSN'), color=blue) + geom_point(size = 0.5, data = subset(doud, from_ref == 'Perth'), color=orange)
+doud_branch_lengths
+ggsave("outputs/branch_lengths.png", width =12)

### 2018-02-01 analysis/phylobayes/pymodules/utils.py

@@ -1,0 +1,95 @@
+"""
+Python utilities for branch_lengths analysis
+Written by Sarah Hilton.
+"""
+from phydmslib import constants
+from ete3 import Tree
+import os
+import itertools
+from scipy.special import comb
+import pandas as pd
+
+
+def extract_model_name(model_string):
+    """
+    This function takes a tree output file from phydms and extracts the name
+    f the model.
+    input:
+        `str`, tree file name
+    output:
+        `str`, model name
+        `int`, 1 if "control" model, 0 otherwise
+    """
+    if "YNGKP" in model_string:
+        if "YNGKP_M0" in model_string:
+            return "GY94", 0
+        elif "YNGKP_M5" in model_string:
+            return "GY94_gammaomega", 0
+        else:
+            raise ValueError("Cannot parse model string {0}"
+                             .format(model_string))
+    elif "ExpCM" in model_string:
+        if "gammaomega" in model_string:
+            if "randomized_ExpCM" in model_string:
+                return "ExpCM_gammaomega_randomized", 1
+            elif "averaged_ExpCM" in model_string:
+                return "ExpCM_gammaomega_averaged", 1
+            else:
+                return "ExpCM_gammaomega", 0
+        else:
+            if "randomized_ExpCM" in model_string:
+                return "ExpCM_randomized", 1
+            elif "averaged_ExpCM" in model_string:
+                return "ExpCM_averaged", 1
+            else:
+                return "ExpCM", 0
+    elif "RAxML" in model_string:
+        return "GTR", 0
+    else:
+        raise ValueError("Cannot parse model string {0}".format(model_string))
+
+
+def translate_with_gaps(seq):
+    """
+    This function uses `phydmslib.constants` to translate a nucleotide
+    sequence.
+    input: `str`
+    output: `str`
+    """
+    prot_seq = []
+    for i in range(0, len(seq), 3):
+        codon = seq[i:i+3]
+        codon = constants.CODON_TO_INDEX[codon]
+        AA = constants.CODON_TO_AA[codon]
+        AA = constants.INDEX_TO_AA[AA]
+        prot_seq.append(AA)
+    return "".join(prot_seq)
+
+
+def calc_distances(treePath):
+    '''
+    This function calculates the distance between all pairs of tips on a
+    given tree.
+    input:
+        treePath: `str`, file path to tree
+    output:
[…]

### 2018-02-01 analysis/phylobayes/plot.R
cleaned up phylobayes script

@@ -1,22 +1,0 @@
-library(ggplot2)
-library(cowplot)
-
-black = "black"
-blue = "#439FC3"
-green = "#76A88D"
-orange = "#F9842A"
-red = "#B52A24"
-pink = "#CC79A7"
-purple = "#8E7790"
-
-setwd("/Users/sarah/Google Drive/GS/lab/divergence_timing_manuscript/analysis/phylobayes")
-doud = read.csv("outputs/model_by_model_norm.csv")
-m <- ceiling(max(max(doud$distance1), max(doud$distance2)))
-doud_branch_lengths = ggplot(doud, aes(distance1, distance2)) + geom_abline(intercept=0, slope=1, color = "grey", linetype="dashed") + geom_point(size=1.0, color="grey") 
-doud_branch_lengths = doud_branch_lengths + facet_grid(divergence_level~ model_pair)
-doud_branch_lengths = doud_branch_lengths + theme(strip.background = element_blank(), strip.text = element_text(size=8), legend.position="bottom", panel.spacing = unit(2, "lines"), panel.border = element_rect(colour = "black"))
-doud_branch_lengths = doud_branch_lengths + xlim(c(0,m)) + ylim(c(0,m)) + coord_fixed() 
-doud_branch_lengths = doud_branch_lengths + xlab("model 1") + ylab("model 2") 
-doud_branch_lengths = doud_branch_lengths + geom_point(size = 0.5, data = subset(doud, from_ref == 'WSN'), color=blue) + geom_point(size = 0.5, data = subset(doud, from_ref == 'Perth'), color=orange)
-doud_branch_lengths
-ggsave("outputs/branch_lengths.png", width =12)

### 2018-02-01 manuscript/manuscript.tex
added figure for phylobayes

--- 2018-01-23 manuscript/manuscript.tex
@@
 \begin{abstract}
-\noindent \skhcomment{$\leq$ 250 words, currently 164}   
+\noindent \skhcomment{$\leq$ 250 words, currently 217}   
 Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
@@
 \textbf{Comparison of substitution model features.}
-Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
-The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
-Blue indicates presence and white indicates absence of a feature. 
-from \ref{tab:wsn_low_params}
+(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
+(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution. 
+To ensure computational tractability, $\omega$ takes on the mean value of $K$ bins ($K=4$ shown). 
+(C) Substitution models can both, one, or neither of these features. 
+We used models from the GY94 and the ExpCM families. 
 }
@@
 
-\skhcomment{Section Outline:}
+It is well known that protein sites evolve under different pressures. 
+Some sites appear to tolerate many amino acids while other sites appear to tolerate only a select few. 
+Therefore, it is important to use substitution models which are able to describe this purifying selection, and the resulting site-to-site rate variation, when modeling protein evolution.
 
-\begin{itemize}
-\item There are different strategies to account for purifying selection and site-to-site rate variation. 
-\begin{itemize}
-\item $\Gamma\omega$ rate variation 
-\begin{itemize}
-\item single parameter controlling relative rate of non-synonymous to synonymous substitutions can vary across sites following some statistical distribution (usually $\Gamma$)
-\item very common 
-\item practical implementation - discretized distribution with the mean of $K$ bins. 
-\end{itemize}
-\item stationary state
-\begin{itemize}
-\item Explicitly model each amino-acid frequency at each site in the protein 
-\item a more mechanistic description, no arbitrary statistical distribution 
-\item practical limitation - how do you estimate these parameters without overfitting the data? We use experimental measurements to define the parameters \textit{a priori}
-\end{itemize}
-\end{itemize}
-\item We can compare the effect of a model feature ($\Gamma\omega$ rate variation or non-uniform stationary state) while controlling the presence/absence of the other feature using models from the ExpCM and GY94 families.
-\end{itemize}
+The standard strategy to account for purifying selection is to allow the rate of non-synonymous change to vary among sites following some statistical distribution. 
+This variation is commonly implemented by allowing the rate parameter, $\omega$, to follow a $\Gamma$-distribution ($\Gamma\omega$).
+In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is assigned to the mean of each bin and the site likelihood is the average likelihood given each of the $\omega$ values.
+Under this model, some sites would have a higher rate of non-synonymous change than others but this rate would be agnostic to the amino-acid identities themselves. 
+All non-synonymous changes would be treated as equally.
+
+Another strategy to account for purifying selection is to define a substitution model with a site-specific stationary state. 
+Such models describe the evolution at each site uniquely and independently of the other sites. 
+For example, models from the Halpern-Bruno mutation-selection family explicitly describe the frequency of each amino acid at each site. 
+However, these more mechanistic models, without an arbitrary statistical distribution, come at a cost in the form of an increased number of parameters. 
+When using a mutation-selection model is crucial to invoke a strategy to avoiding overfitting the data. 
+We chose to define the amino-acid frequency parameters \textit{a priori} using measurements from a high-throughput functional assay called deep mutational scanning, Experimentally Informed Codon Models (ExpCMs).
+
+These two strategies, $\Gamma\omega$ rate variation and non-uniform stationary states, are not truly distinct. 
+Differences in stationary state manifest as differences in non-synonymous rates across sites \citep{spielman2015relationship}. 
+Moreover, these two strategies are not mutually exclusive. 
+Technically, substitution models can have a non-uniform stationary state and $\Gamma\omega$ rate variation and biologically, these models can account for both strong purifying and strong positive selection.
 
@@
 \textbf{Effect of stationary state and rate variation on long branch estimation.}
-The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
-The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
-The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
-The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller.
+The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models. 
+The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
+The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
 }
@@
 
-\skhcomment{Section Outline:}
+Along a single branch, a substitution model transforms sequence divergence to branch length, which under a molecular clock assumption is proportional to time. 
+However, every substitution model has a stationary state, the expected sequence divergence given a very long time, at which the relationship between sequence divergence and time is lost. 
+The stationary state is represented by the long ``tails" in \ref{fig:decay}, where the expected sequence divergence remains constant as time increases.
 
-\begin{itemize}
-\item Substitution models are only useful for inferring branch lengths when they can accurately transform sequence divergence to time (branch lengths)
-\item Every substitution model has a saturation point where this temporal signal is lost.  
-\begin{itemize}
-\item In the long tails the expected sequence divergence remains constant as time increases. 
-\item The saturation point will affect long branches more than short branches.
-\end{itemize}
-\item Different model features (\ref{fig:model_feature}) can affect \textit{how long} it takes to reach the saturation point and the \textit{sequence divergence} at the saturation point 
-\begin{itemize}
-\item Adding $\Gamma\omega$ rate variation affects the ``path" to the saturation point but affects the sequence divergence at the saturation point minimally.
-Importantly, the saturation point is identically for every site in the protein.
-(green and blue)
-\item Adding a non-uniform stationary state creates a unique saturation point for each site in the protein (red).
-Sites which can tolerate many mutations have saturation points which are quite similar to the GY94 saturation point.
-Constrained sites show a dramatic difference.
-At a constrained site, ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at the site is expected to be small.
-\item This difference cannot be recapitulated by more complex modeling of the $\omega$ parameter.
-We can transform the GY94 model into a  site-specific model by inferring a unique $\omega$ value from the ExpCM for each site in the protein.
-This model (yellow) takes longer to reach the saturation point than the GY94 model but ultimately has the same divergence level at the saturation point.
-\end{itemize}
-\item No matter how ``well" you model site-to-site rate variation, you will underestimate \textit{very} long branches with a uniform stationary state. 
-\end{itemize}
+A substitution model's stationary state is defined by the parameters in the model and can vary from model to model. 
+Some model features affect how long it takes a model to reach its stationary state, whereas other features affect the stationary state itself. 
+Adding $\Gamma\omega$ rate variation affects the ``path" to the stationary state but does not affect the stationary state itself. 
+In addition, while $\Gamma\omega$ accounts for site-to-site rate variation, the resulting stationary state is identical at each site in the protein. 
+Under these models with uniform stationary states, a mutationally constrained site and a site which tolerates many mutations will have the sequence divergence given a very long time. 
+Models with non-uniform stationary states, by definition, acknowledge that differences in sites will persist even at a very long period of time. 
+The difference in branch length estimation will be most dramatic at constrained sites in the protein. 
+At a constrained site, the ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at that site is expected to be small.
+
+This difference in branch length estimation cannot be recapitulated by more complex modeling of the $\omega$ parameter. 
+We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM, following the work of \cite{spielman2015relationship}, for each site in the protein. 
+This model takes longer to reach stationary state than the GY94 model ultimately has the same divergence level at stationary state. 
+As this model shows, no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state model.
 
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees}}
+\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees.pdf}}
 \caption{\label{fig:empirical_trees}
@@
 \end{figure}
+
+\begin{table}[t!]
+\caption{\label{tab:empirical_data}
+Model comparison for Fig.~\ref{fig:empirical_trees}.}
+     \begin{tabular}{ccccc}
+        \hline
+          Model & $\Delta$AIC & Log Likelihood & Parameters\\ \hline
+       	ExpCM + $\Gamma\omega$ (average) & 0.00 & -48751.16 & $\omega$ = (0.19,  0.50,  0.90,  1.86), $\beta=1.70$, $\kappa=3.73$\\
+	ExpCM (average) & 949.70 & -49227.01 & $\beta=1.78$, $\kappa=3.40$, $\omega=0.15$\\
+	ExpCM + $\Gamma\omega$ (H1) & 1306.44 & -49404.38  & $\omega$ = (0.13 ,  0.44,  0.91,  2.16), $\beta=1.12$, $\kappa=3.64$\\
+	ExpCM + $\Gamma\omega$ (H3) & 1737.40 & -49619.86 & $\omega$ = (0.09,  0.33,  0.72,  1.77), $\beta=1.28$, $\kappa=3.71$\\
+	ExpCM (H1) & 2555.58 & -50029.95 & $\beta=1.22$, $\kappa=3.24$, $\omega=0.13$\\
+	ExpCM (H3) & 3196.50 & -50350.41 & $\beta=1.45$, $\kappa=3.38$, $\omega=0.12$\\
+	GY94 + $\Gamma\omega$ & 4719.34 & -51105.83 & $\omega$ = (0.00,  0.03,  0.08,  0.26), $\kappa=3.18$\\
+	GY94 & 7624.64 & -52559.48  & $\kappa=2.89$, $\omega=0.07$\\
+      \end{tabular}
+\end{table}
 
@@
 
-\subsection*{texttt{phylobayes}}
+\subsection*{\texttt{phylobayes}}
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.5\textwidth]{figures/phylobayes.pdf}}
+\caption{\label{fig:phylobayes}
+\textbf{Comparison of ExpCM and phylobayes}
+}
+\end{figure}
 

### 2018-02-01 manuscript/figures/_temp_compete.py
added temp figures for compete

@@ -1,0 +1,170 @@
+"""
+This script is the scratch script for optimizing a model with different betas.
+
+The script is a modified (simplified) verison of the `phydms` script
+
+SKH 20180122
+"""
+
+import sys
+import os
+import re
+import logging
+import random
+import time
+import math
+import operator
+import multiprocessing
+import warnings
+warnings.simplefilter('always')
+import scipy
+import scipy.stats
+import statsmodels.sandbox.stats.multicomp
+import Bio.Phylo
+import phydmslib
+from phydmslib.constants import *
+import phydmslib.file_io
+import phydmslib.parsearguments
+import phydmslib.models
+import phydmslib.treelikelihood
+import pandas as pd
+import phydmslib.simulate
+from ete3 import Tree
+
+
+def read_phydms_model_params(param):
+    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    return params
+
+def create_model_ExpCM(param, prefs, beta):
+    """
+    This is the basic `ExpCM` from `phydms`
+
+    `omega` is set to 1.0
+    """
+    params = read_phydms_model_params(param)
+    prefs = phydmslib.file_io.readPrefs(prefs)
+    sites = sorted(prefs.keys())
+    prefs = [prefs[r] for r in sites]
+    params["phiT"] = 1 - sum([params[x] for x in params.keys() if
+                              x.startswith("phi")])
+    phi = scipy.array([params["phi{0}".format(INDEX_TO_NT[x])] for x in
+                       range(N_NT)])
+    return phydmslib.models.ExpCM(prefs, kappa=params["kappa"],
+                                  omega=params["omega"], beta=beta,
+                                  mu=0.3, phi=phi, freeparams=['mu'])
+
+def run_phydms(alignment, prefs_fname, params_fname, tree_fname, beta, outprefix):
+    # create output directory if needed
+    outdir = os.path.dirname(outprefix)
+    if outdir:
+        if not os.path.isdir(outdir):
+            if os.path.isfile(outdir):
+                os.remove(outdir)
+            os.mkdir(outdir)
+
+    # output files, remove if they already exist
+    underscore = '' if outprefix == '/' else '_'
[…]

### 2018-02-01 manuscript/figures/_temp_empirical.py

@@ -1,0 +1,26 @@
+"""
+The purpose of this script is to manipulate tree figures to make the
+alternative tree figures.
+
+SKH 20180131
+"""
+
+from ete3 import Tree
+import glob
+import os
+
+def main():
+    tree_list = glob.glob("../../analysis/HA/branch_lengths/phydms/hybrid_high_0_*_tree.newick")
+    tree_list = [x for x in tree_list if "randomized" not in x]
+    tree_list = [x for x in tree_list if "averaged" not in x]
+    tree_list = [x for x in tree_list if "RAxML" not in x]
+    assert len(tree_list) == 8
+    for tree in tree_list:
+        outName = "_temp_{0}_{1}.newick".format(os.path.basename(tree).split(".")[0], "out")
+        t = Tree(tree)
+        R = t.get_midpoint_outgroup()
+        t.set_outgroup(R)
+        t.write(format=1, outfile=outName)
+
+if __name__ == '__main__':
+    main()

### 2018-02-01 manuscript/figures/temp_compete.py

@@ -1,132 +1,0 @@
-"""
-This script is the scratch script for optimizing a model with different betas.
-
-The script is a modified (simplified) verison of the `phydms` script
-
-SKH 20180122
-"""
-
-import sys
-import os
-import re
-import logging
-import random
-import time
-import math
-import operator
-import multiprocessing
-import warnings
-warnings.simplefilter('always')
-import scipy
-import scipy.stats
-import statsmodels.sandbox.stats.multicomp
-import Bio.Phylo
-import phydmslib
-from phydmslib.constants import *
-import phydmslib.file_io
-import phydmslib.parsearguments
-import phydmslib.models
-import phydmslib.treelikelihood
-import pandas as pd
-
-def read_phydms_model_params(param):
-    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
-    params = dict(zip(params[0], params[1]))
-    return params
-
-def create_model_ExpCM(param, prefs, beta):
-    """
-    This is the basic `ExpCM` from `phydms`
-
-    `omega` is set to 1.0
-    """
-    params = read_phydms_model_params(param)
-    prefs = phydmslib.file_io.readPrefs(prefs)
-    sites = sorted(prefs.keys())
-    prefs = [prefs[r] for r in sites]
-    params["phiT"] = 1 - sum([params[x] for x in params.keys() if
-                              x.startswith("phi")])
-    phi = scipy.array([params["phi{0}".format(INDEX_TO_NT[x])] for x in
-                       range(N_NT)])
-    return phydmslib.models.ExpCM(prefs, kappa=params["kappa"],
-                                  omega=params["omega"], beta=beta,
-                                  mu=0.3, phi=phi, freeparams=['mu'])
-
-def main():
-    """Main body of script."""
-
-    alignment = "../../analysis/HA/data/subsample/HA_hybrid_lowH1_0.fasta"
-    prefs_fname = "../../analysis/HA/data/references/HA_hybridDoud_prefs.csv"
-    params_fname = "../../analysis/HA/branch_lengths/phydms/hybrid_lowH1_0_ExpCM_HA_hybridDoud_prefs_modelparams.txt"
-    tree_fname = "../../analysis/HA/branch_lengths/phydms/hybrid_lowH1_0_RAxML_tree.newick"
-    # betas = [1.56, 1.34, 1.22]
-    betas = [1.00]
-
-    for beta in betas:
-        outprefix = "_temp_{0}/".format(beta)
[…]

### 2018-02-02 manuscript/figures/_temp_compete.py
.gitignore

@@ -1,170 +1,0 @@
-"""
-This script is the scratch script for optimizing a model with different betas.
-
-The script is a modified (simplified) verison of the `phydms` script
-
-SKH 20180122
-"""
-
-import sys
-import os
-import re
-import logging
-import random
-import time
-import math
-import operator
-import multiprocessing
-import warnings
-warnings.simplefilter('always')
-import scipy
-import scipy.stats
-import statsmodels.sandbox.stats.multicomp
-import Bio.Phylo
-import phydmslib
-from phydmslib.constants import *
-import phydmslib.file_io
-import phydmslib.parsearguments
-import phydmslib.models
-import phydmslib.treelikelihood
-import pandas as pd
-import phydmslib.simulate
-from ete3 import Tree
-
-
-def read_phydms_model_params(param):
-    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
-    params = dict(zip(params[0], params[1]))
-    return params
-
-def create_model_ExpCM(param, prefs, beta):
-    """
-    This is the basic `ExpCM` from `phydms`
-
-    `omega` is set to 1.0
-    """
-    params = read_phydms_model_params(param)
-    prefs = phydmslib.file_io.readPrefs(prefs)
-    sites = sorted(prefs.keys())
-    prefs = [prefs[r] for r in sites]
-    params["phiT"] = 1 - sum([params[x] for x in params.keys() if
-                              x.startswith("phi")])
-    phi = scipy.array([params["phi{0}".format(INDEX_TO_NT[x])] for x in
-                       range(N_NT)])
-    return phydmslib.models.ExpCM(prefs, kappa=params["kappa"],
-                                  omega=params["omega"], beta=beta,
-                                  mu=0.3, phi=phi, freeparams=['mu'])
-
-def run_phydms(alignment, prefs_fname, params_fname, tree_fname, beta, outprefix):
-    # create output directory if needed
-    outdir = os.path.dirname(outprefix)
-    if outdir:
-        if not os.path.isdir(outdir):
-            if os.path.isfile(outdir):
-                os.remove(outdir)
-            os.mkdir(outdir)
-
-    # output files, remove if they already exist
-    underscore = '' if outprefix == '/' else '_'
[…]

### 2018-02-02 manuscript/figures/_temp_empirical.py
cleaned up files in figures

@@ -1,26 +1,0 @@
-"""
-The purpose of this script is to manipulate tree figures to make the
-alternative tree figures.
-
-SKH 20180131
-"""
-
-from ete3 import Tree
-import glob
-import os
-
-def main():
-    tree_list = glob.glob("../../analysis/HA/branch_lengths/phydms/hybrid_high_0_*_tree.newick")
-    tree_list = [x for x in tree_list if "randomized" not in x]
-    tree_list = [x for x in tree_list if "averaged" not in x]
-    tree_list = [x for x in tree_list if "RAxML" not in x]
-    assert len(tree_list) == 8
-    for tree in tree_list:
-        outName = "_temp_{0}_{1}.newick".format(os.path.basename(tree).split(".")[0], "out")
-        t = Tree(tree)
-        R = t.get_midpoint_outgroup()
-        t.set_outgroup(R)
-        t.write(format=1, outfile=outName)
-
-if __name__ == '__main__':
-    main()

### 2018-02-05 manuscript/manuscript.tex
small figure tweaks

@@
 
-Substitution models that incorportate site-to-site rate variation have been developed to decrease the bias in long branch estimation. 
+Substitution models that incorporate site-to-site rate variation have been developed to decrease the bias in long branch estimation. 
 The most common strategy is to allow a single rate-controlling parameter to vary according to some statistical distribution, such as a $\Gamma$-distributed $\omega$ (~dN/dS) \citep{yang2000codon}. 
@@
 \textbf{Model performance under simulated, site-specific data.} 
-\jdbcomment{Fonts are small, and tick labels definitely too small. I would show averaged.}
 Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
@@
 
-\skhcomment{Section Outline:}
+First, we wanted to test the effect of substitution models on branch length estimation given sequences with site-specific amino-acid frequencies. 
+To this end, we simulated sequences under and ExpCM and re-inferred the branch lengths using the different substitution models described in \ref{fig:model_feature}C. 
+This allowed us to test whether or not a model with a uniform stationary state could accurately estimate branch lengths and, if not, whether or not the addition of $\Gamma\omega$ could prevent branch length underestimation.
 
-\begin{itemize}
-\item To test the effect of substitution model on data with site-specific amino acid frequencies, we simulated sequences under an ExpCM and re-inferred the branch lengths using different substitution models.
-\item Uniform stationary state models underestimate long branches
-\begin{itemize}
-\item Both the GY94 and the GY94+$\Gamma\omega$ underestimate long branches. The $\Gamma\omega$ rate variation scheme is not enough to prevent this underestimation at every long branches. 
-\item Another uniform stationary state model, the averaged ExpCM, also underestimates long branches. This model derives some information from the experiments but lacks site-specificity. 
-\item Non-uniform stationary state models accurately estimate long branches (both the ExpCM and ExpCM+$\Gamma\omega$
-\end{itemize}
-\item If sequences have site-specific amino acid frequencies then non-uniform stationary state models, even with $\Gamma\omega$, will underestimate long branches.
-\end{itemize}
+As expected, the ExpCM and ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
+As the branch lengths increase, the variance in the estimation of an given branch increases between the simulations but this error is not biased towards over- or underestimation.
+
+The models with a uniform stationary state consistently underestimate the long branches. 
+This underestimation is seen with both the GY94 and the GY94+$\Gamma\omega$, indicating that accounting for rate variation via the rate parameter cannot prevent branch length underestimation at these long branches.
+The ExpCM can also be transformed into a uniform stationary state model by setting the amino-acid at every site to the average value for that amino acid across all sites. 
+In contrast to the GY94 models, this average control does not have preference for amino acid and does extract some information from the experiments, but like the GY94 models the stationary state is uniform across sites.
+These models performed similarly to the GY94 and GY94+$\Gamma\omega$ and underestimate long branches.
+
+These simulations show that when the sequences have site-specific amino-acid frequencies, models with uniform stationary states cannot accurately estimate branch lengths and, will in fact, underestimate long branches.
 

### 2018-02-05 manuscript/manuscript.tex
change 'avg' to 'H1+H3 average'

@@
           Model & $\Delta$AIC & Log Likelihood & Parameters\\ \hline
-       	ExpCM + $\Gamma\omega$ (average) & 0.00 & -48751.16 & $\omega$ = (0.19,  0.50,  0.90,  1.86), $\beta=1.70$, $\kappa=3.73$\\
-	ExpCM (average) & 949.70 & -49227.01 & $\beta=1.78$, $\kappa=3.40$, $\omega=0.15$\\
+       	ExpCM + $\Gamma\omega$ (H1+H3 average) & 0.00 & -48751.16 & $\omega$ = (0.19,  0.50,  0.90,  1.86), $\beta=1.70$, $\kappa=3.73$\\
+	ExpCM (H1+H3 average) & 949.70 & -49227.01 & $\beta=1.78$, $\kappa=3.40$, $\omega=0.15$\\
 	ExpCM + $\Gamma\omega$ (H1) & 1306.44 & -49404.38  & $\omega$ = (0.13 ,  0.44,  0.91,  2.16), $\beta=1.12$, $\kappa=3.64$\\
@@
 \textbf{Comparison of ExpCM and phylobayes}
+\skhcomment{y=x line too faint?}
 }
@@
 
-\begin{figure}[H]
+\begin{suppfig}[H]
 \centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
@@
 \skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.}
-\end{figure}
+\end{suppfig}
 

### 2018-02-06 manuscript/manuscript.tex
small tweaks to first three results sections

@@
 (B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
-To ensure computational tractability, $\omega$ takes on the mean value of $K$ bins ($K=4$ shown). 
+In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
+A $\Gamma$ distribution with $K=4$ bins is shown here. 
 (C) Substitution models can both, one, or neither of these features. 
@@
 
-It is well known that protein sites evolve under different pressures. 
-Some sites appear to tolerate many amino acids while other sites appear to tolerate only a select few. 
-Therefore, it is important to use substitution models which are able to describe this purifying selection, and the resulting site-to-site rate variation, when modeling protein evolution.
+\skhcomment{Should this all be in the introduction?}
 
-The standard strategy to account for purifying selection is to allow the rate of non-synonymous change to vary among sites following some statistical distribution. 
-This variation is commonly implemented by allowing the rate parameter, $\omega$, to follow a $\Gamma$-distribution ($\Gamma\omega$).
-In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is assigned to the mean of each bin and the site likelihood is the average likelihood given each of the $\omega$ values.
-Under this model, some sites would have a higher rate of non-synonymous change than others but this rate would be agnostic to the amino-acid identities themselves. 
-All non-synonymous changes would be treated as equally.
+In nature, viral proteins evolve under both purifying selection, to maintain structure and function, and diversifying selection, to escape immune pressure. 
+However, neither one of these constraints is homogenous across the entire protein. 
+Some protein sites appear to be tolerant of many mutations, and sample many amino acids, while others appear to be mutationally tolerant, and sample only a select few. 
+When modeling evolution it is important to account for this heterogeneity and phylogenetic substitution models do this in a variety of ways.
 
-Another strategy to account for purifying selection is to define a substitution model with a site-specific stationary state. 
-Such models describe the evolution at each site uniquely and independently of the other sites. 
-For example, models from the Halpern-Bruno mutation-selection family explicitly describe the frequency of each amino acid at each site. 
-However, these more mechanistic models, without an arbitrary statistical distribution, come at a cost in the form of an increased number of parameters. 
-When using a mutation-selection model is crucial to invoke a strategy to avoiding overfitting the data. 
-We chose to define the amino-acid frequency parameters \textit{a priori} using measurements from a high-throughput functional assay called deep mutational scanning, Experimentally Informed Codon Models (ExpCMs).
+The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution (\cite{yang1994maximum, yang2000codon}, \ref{fig:model_feature}B).
+Under this model, some sites would have a higher rate of non-synonymous change than others. 
+However, this rate is agnostic to the amino-acid identities themselves and all non-synonymous changes at at given site are treated equally. 
+In contrast, models from the mutation-selection family \cite{halpern1998evolutionary} account for purifying selection by explicitly defining the frequency of each amino acid at each site in the protein.
+This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A).
 
-These two strategies, $\Gamma\omega$ rate variation and non-uniform stationary states, are not truly distinct. 
-Differences in stationary state manifest as differences in non-synonymous rates across sites \citep{spielman2015relationship}. 
-Moreover, these two strategies are not mutually exclusive. 
-Technically, substitution models can have a non-uniform stationary state and $\Gamma\omega$ rate variation and biologically, these models can account for both strong purifying and strong positive selection.
+The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
+While most codon models have $<$ 20 parameters, mutation-selection models will have more than 10,000 parameters for an averaged size protein. 
+As with all parameter-rich models, it is important to invoke some strategy to avoid overfitting when using a mutation-selection model.
+We chose to define the amino-acid frequency parameters \textit{a priori} from a high-throughput assay called deep mutational scanning. 
+The resulting model, an Experimentally Informed Codon Model (ExpCM) \citep{bloom2014experimentally, hilton2017phydms}, has a parameter space comparable to the other codon models.
+
+These two strategies to account for site-to-site rate variation are not mutually exclusive. 
+While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.
+Proteins which are under strong purifying and diversifying selection may be better modeled by a combination of the two. 
+For this study, we used four models from the GY94 and ExpCM families, which together represent all possible combinations of these two strategies (\ref{fig:model_feature}C).
 
@@
 
-Along a single branch, a substitution model transforms sequence divergence to branch length, which under a molecular clock assumption is proportional to time. 
-However, every substitution model has a stationary state, the expected sequence divergence given a very long time, at which the relationship between sequence divergence and time is lost. 
-The stationary state is represented by the long ``tails" in \ref{fig:decay}, where the expected sequence divergence remains constant as time increases.
+Along a single branch, a substitution model transforms sequence divergence into branch length. 
+Under the molecular clock assumption, this branch length is proportional to time.  
+However, every substitution model has a stationary state, defining the expected sequence divergence given a very long time, at which this relationship between sequence divergence and time is lost. 
+In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.
 
-A substitution model's stationary state is defined by the parameters in the model and can vary from model to model. 
-Some model features affect how long it takes a model to reach its stationary state, whereas other features affect the stationary state itself. 
-Adding $\Gamma\omega$ rate variation affects the ``path" to the stationary state but does not affect the stationary state itself. 
-In addition, while $\Gamma\omega$ accounts for site-to-site rate variation, the resulting stationary state is identical at each site in the protein. 
-Under these models with uniform stationary states, a mutationally constrained site and a site which tolerates many mutations will have the sequence divergence given a very long time. 
-Models with non-uniform stationary states, by definition, acknowledge that differences in sites will persist even at a very long period of time. 
-The difference in branch length estimation will be most dramatic at constrained sites in the protein. 
-At a constrained site, the ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at that site is expected to be small.
+The specific parametrization scheme of a model will control features of its stationary state. 
+Comparing GY94 to GY94+$\Gamma\omega$ will change how long it takes for the model to reach the stationary state but it does not change the expected sequence divergence at the stationary state. 
+Furthermore, the GY94 and GY94+$\Gamma\omega$ stationary state is exactly the same at each site in the protein. 
+Site-specific stationary states modulate both the amount of time to reach and the overall divergence level at the stationary state. 
+These differences arise from whether or not the site is modeled as constrained or mutationally tolerant.  
+The stationary state is important to branch length estimation because differences in expected divergence will lead to differences in branch length estimation and branch lengths cannot be estimated with any accuracy when two sequences have reached the divergence level of the stationary state.
+
+The difference in branch length estimation between uniform and site-specific stationary state models will be most dramatic at constrained sites. 
+Most uniform stationary state models define roughly equal frequency of each amino acid, in essence they assume that each site is mutationally tolerant. 
+If the sequence divergence at constrained site is quite small, a site-uniform model will expect a small amount of time to have passed, a short branch, and a site-specific model will expect that a large amount of time has passed, long branch.
+At a mutationally tolerant site, both models will estimate roughly the same branch length for any given divergence.
 
 This difference in branch length estimation cannot be recapitulated by more complex modeling of the $\omega$ parameter. 
-We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM, following the work of \cite{spielman2015relationship}, for each site in the protein. 
-This model takes longer to reach stationary state than the GY94 model ultimately has the same divergence level at stationary state. 
+We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM for each site in the protein, following the work of \cite{spielman2015relationship}. 
+This model takes longer to reach stationary state than the GY94 model but, ultimately, has the same divergence level at stationary state. 
 As this model shows, no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state model.
@@
 \textbf{Model performance under simulated, site-specific data.} 
-Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
-The randomized ExpCMs have amino-acid profiles shuffled among the sites 
-These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
-Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
+Alignments were simulated under an ExpCM along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
+The averaged ExpCMs amino-acid frequencies defined by the average preference of that amino acid across all sites in the protein. 
+While these models extract information from the deep mutational scanning experiment, they are not site-specific. 
+Grey points represent the length of one branch and the black points are the mean branch lengths over ten simulations. 
 The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
@@
 
-First, we wanted to test the effect of substitution models on branch length estimation given sequences with site-specific amino-acid frequencies. 
-To this end, we simulated sequences under and ExpCM and re-inferred the branch lengths using the different substitution models described in \ref{fig:model_feature}C. 
-This allowed us to test whether or not a model with a uniform stationary state could accurately estimate branch lengths and, if not, whether or not the addition of $\Gamma\omega$ could prevent branch length underestimation.
-
+To test the effect of substitution models on branch length estimation given sequences with site-specific amino-acid frequencies we simulated sequences under an ExpCM and re-inferred the branch lengths using the different substitution models described in \ref{fig:model_feature}C.
 As expected, the ExpCM and ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
-As the branch lengths increase, the variance in the estimation of an given branch increases between the simulations but this error is not biased towards over- or underestimation.
-
+The variance in the estimation of an given branch increases between the simulations increased as branch length increased but this error is not biased towards over- or underestimation.
 The models with a uniform stationary state consistently underestimate the long branches. 
@@
 In contrast to the GY94 models, this average control does not have preference for amino acid and does extract some information from the experiments, but like the GY94 models the stationary state is uniform across sites.
-These models performed similarly to the GY94 and GY94+$\Gamma\omega$ and underestimate long branches.
+These models performed similarly to the GY94 and GY94+$\Gamma\omega$ and underestimated long branches.
 
@@
 \caption{\label{fig:empirical_trees}
-\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
-\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
-The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
-The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
+\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to GY94 models.} ).
 }
@@
 \item These sequences are from fairly diverged proteins. Some sequences share only 40\% identity on the amino acid level 
-\item The HA tree has closely related subtypes with long branches between subtypes.
+\item The HA tree has closely related subtypes with long branches between subtypes. This is good because we are looking for the effect on long branches. 
 \end{itemize}
@@
 \item It is expected that there would be epistatic changes across such a highly diverged phylogenetic tree. ``Shifting preferences".
+\item "Sampling" from both preference sets via preference averaging extends the branch length from both focal sequences.
 \end{itemize}
@@
 \begin{itemize}
-\item We investigated the competing effects of ``shifting preferences" and long branches by comparing ExpCMs with different "relevances"
+\item We investigated the competing effects of ``shifting preferences" and long branches by comparing ExpCMs with trees of increasing overall divergence from the focal sequence. 
 \item We use the fit stringency parameter $\beta$ as a measure of relevance. 
@@
 \subsection*{\texttt{phylobayes}}
+
+\skhcomment{Section Outline}
+
+\begin{itemize}
+\item \texttt{phylobayes} implements the mutation-selection model in a bayesian framework as a way to avoid overfitting. 
+\item \texttt{phylobayes} and ExpCM (H1+H3 average) estimate branches of roughly equal length 
+\begin{itemize}
+\item the ExpCM estimate longer branches from the focal sequences 
+\item \texttt{phylobayes} estimates longer branches from the non-focal sequences 
+\end{itemize}
+\end{itemize}
 

### 2018-02-06 manuscript/manuscript.tex
some tweaks to abstract

@@
 \begin{abstract}
-\noindent \skhcomment{$\leq$ 250 words, currently 217}   
+\noindent  
 Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
@@
 This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
-Here we show that substitution models informed by experimental measurements of site-specific purifying selection lengthen the branches on phylogenies of influenza virus hemagglutinin.
-This deepening of branch lengths is due to better modeling how site-specific amino-acid preferences affect the stationary state of the substitution models, and is independent of the branch-length-extension due to modeling site-to-site variation in substitution rate.
-We show that the deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
+Here we show that substitution models informed by experimental measurements of the purifying selection due to site-specific amino-acid preferences lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
+This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
+The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
 However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
-Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state as well as the substitution rate when estimating deep divergence times. 
+Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times. 
 \end{abstract}

### 2018-02-06 manuscript/manuscript.tex
revisions to first section of Results

@@
 
-\subsection*{Substitution models}
+\subsection*{Different ways hat substitution models account for purifying selection}
+
+\jdbcomment{Some other comments on this section, which I think in general is pretty good: $\omega$ shows up in the figure, but is never mentioned in the text.
+I feel like GY94 models need to be explained at least in terms of what they and $\Gamma$ distribution stand for.
+Just introducing nomenclature.
+I feel like the last paragraph should tie back to panel C of the figure.}
 
@@
 \caption{\label{fig:model_feature}
-\textbf{Comparison of substitution model features.}
+\textbf{Different ways of modeling site-to-site variation in purifying selection.}
+\jdbcomment{Panels A and B need to be switched to match order in text. Then text updated too}
 (A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
@@
 
-\skhcomment{Should this all be in the introduction?}
+Proteins evolve under both purifying selection to maintain their structure and function. 
+This purifying selection is not homogenous across sites in a protein.
+It is also not homogenous across the different amino acids at a given site.
+For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
+In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.
 
-In nature, viral proteins evolve under both purifying selection, to maintain structure and function, and diversifying selection, to escape immune pressure. 
-However, neither one of these constraints is homogenous across the entire protein. 
-Some protein sites appear to be tolerant of many mutations, and sample many amino acids, while others appear to be mutationally tolerant, and sample only a select few. 
-When modeling evolution it is important to account for this heterogeneity and phylogenetic substitution models do this in a variety of ways.
+The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}B;][]{yang1994maximum, yang2000codon}.
+Under such models, some sites have a higher rate of non-synonymous change than others. 
+However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.
 
-The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution (\cite{yang1994maximum, yang2000codon}, \ref{fig:model_feature}B).
-Under this model, some sites would have a higher rate of non-synonymous change than others. 
-However, this rate is agnostic to the amino-acid identities themselves and all non-synonymous changes at at given site are treated equally. 
-In contrast, models from the mutation-selection family \cite{halpern1998evolutionary} account for purifying selection by explicitly defining the frequency of each amino acid at each site in the protein.
+In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein.
 This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A). 
+These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\jdbcomment{cite Rodrigue / Lartillot papers, our papers, etc}.
 
 The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
-While most codon models have $<$ 20 parameters, mutation-selection models will have more than 10,000 parameters for an averaged size protein. 
-As with all parameter-rich models, it is important to invoke some strategy to avoid overfitting when using a mutation-selection model.
-We chose to define the amino-acid frequency parameters \textit{a priori} from a high-throughput assay called deep mutational scanning. 
-The resulting model, an Experimentally Informed Codon Model (ExpCM) \citep{bloom2014experimentally, hilton2017phydms}, has a parameter space comparable to the other codon models.
+While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
+This corresponds to $19\times L$ parameters for a protein of length $L$, which leads to $\sim 10^4$ parameters for a typical size protein.
+As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
+Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data \jdbcomment{include citations}.
+Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
+Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian or maximum-likelihood estimation are discussed in the last section of the Results.
 
-These two strategies to account for site-to-site rate variation are not mutually exclusive. 
+These two strategies to account for site-to-site variation in purifying selection are not mutually exclusive. 
+\jdbcomment{I'm not sure if I love the rest of this paragraph.
+I feel like GY94 needs to be introduce more directly---perhaps in the paragraph beginning the ``The most common strategy...'' above.
+Also, it's a bit unclear what ``most appropriate'' means.}
 While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.

### 2018-02-07 manuscript/manuscript.tex
tweaks to model comparison table

@@
 \caption{\label{tab:empirical_data}
-Model comparison for Fig.~\ref{fig:empirical_trees}.}
-     \begin{tabular}{ccccc}
+Model comparison for Fig.~\ref{fig:empirical_trees}.\skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}} 
+     \begin{tabular}{cccccccccc}
         \hline
-          Model & $\Delta$AIC & Log Likelihood & Parameters\\ \hline
-       	ExpCM + $\Gamma\omega$ (H1+H3 average) & 0.00 & -48751.16 & $\omega$ = (0.19,  0.50,  0.90,  1.86), $\beta=1.70$, $\kappa=3.73$\\
-	ExpCM (H1+H3 average) & 949.70 & -49227.01 & $\beta=1.78$, $\kappa=3.40$, $\omega=0.15$\\
-	ExpCM + $\Gamma\omega$ (H1) & 1306.44 & -49404.38  & $\omega$ = (0.13 ,  0.44,  0.91,  2.16), $\beta=1.12$, $\kappa=3.64$\\
-	ExpCM + $\Gamma\omega$ (H3) & 1737.40 & -49619.86 & $\omega$ = (0.09,  0.33,  0.72,  1.77), $\beta=1.28$, $\kappa=3.71$\\
-	ExpCM (H1) & 2555.58 & -50029.95 & $\beta=1.22$, $\kappa=3.24$, $\omega=0.13$\\
-	ExpCM (H3) & 3196.50 & -50350.41 & $\beta=1.45$, $\kappa=3.38$, $\omega=0.12$\\
-	GY94 + $\Gamma\omega$ & 4719.34 & -51105.83 & $\omega$ = (0.00,  0.03,  0.08,  0.26), $\kappa=3.18$\\
-	GY94 & 7624.64 & -52559.48  & $\kappa=2.89$, $\omega=0.07$\\
+         Model & {\shortstack{Stationary\\ State}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
+       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & yes & yes & 0 & -487510 & 0.19,  0.50,  0.90,  1.86 &  1.70, \\
+	ExpCM (H1+H3 avg) & yes & no &  950 & -492270 & 0.15 & 1.78\\
+	ExpCM + $\Gamma\omega$ (H1) & yes & yes & 13060 & -494040  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
+	ExpCM + $\Gamma\omega$ (H3) & yes & yes & 17370 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
+	ExpCM (H1) & yes & no & 2556 & -50030 &  0.13 & 1.22\\
+	ExpCM (H3) & yes & no &  3197 & -50350 & 0.12 & 1.45\\
+	GY94 + $\Gamma\omega$ & no & yes & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & N/A \\
+	GY94 & no & no & 7625 & -52560  & 0.07 & N/A\\
       \end{tabular}

### 2018-02-07 manuscript/manuscript.tex
changed model feature figure and add citations

@@
 \textbf{Different ways of modeling site-to-site variation in purifying selection.}
-\jdbcomment{Panels A and B need to be switched to match order in text. Then text updated too}
 (A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
@@
 
-The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}B;][]{yang1994maximum, yang2000codon}.
+The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}A;][]{yang1994maximum, yang2000codon}.
 Under such models, some sites have a higher rate of non-synonymous change than others. 
@@
 In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein.
-This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A). 
-These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\jdbcomment{cite Rodrigue / Lartillot papers, our papers, etc}.
-
+This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
+These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, lartillot2007suppression, le2008phylogenetic, rodrigue2010mutation,si2008empirical,rodrigue2014site,bloom2017identification,hilton2017phydms,bloom2014experimentally}.
 The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
@@
 As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
-Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data \jdbcomment{include citations}.
+Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
 Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
-Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian or maximum-likelihood estimation are discussed in the last section of the Results.
+Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.
 
@@
 The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
+\skhcomment{calculated using equation. Maybe change long branch length to expected asymptotic sequence divergence.}
 }
 \end{figure}
+
+Given a single branch, a substitution model transforms sequence divergence into branch length. 
+Under the molecular clock assumption, this branch length is proportional to time. 
+However, every substitution model has a stationary state, defining the expected sequence composition after a long evolutionary time. 
+This stationary state can be used to calculate the expected amino-acid divergence a very long amount of time has elapsed. 
+(Rate only affects how fast asymptotic divergence approached, whereas). 
+In \ref{fig:decay}, the stationary state is represented by the ``long" tails where the expected sequence divergence remains constant as time increases.
+
+This is a fundamental feature. 
+The stationary state is principle eigenvector of substitution matrix. 
+Rate variation only multiples by a number, and stationary state of a stochastic matrix is invariant to multiplication.
 
@@
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees.pdf}}
+\centerline{\includegraphics[width=0.8\textwidth]{figures/empirical_trees.pdf}}
 \caption{\label{fig:empirical_trees}

### 2018-02-09 analysis/HA/branch_lengths/pymodules/utils.py
summarize parameter values from phydms run

@@ -93,3 +93,25 @@ def calc_distances(treePath):
         df["distance"].append(t.get_distance(seqs[0], seqs[1]))
     df = pd.DataFrame(df)
     return df
+
+def extract_pref_name(model_string):
+    """
+    This function takes a tree output file from phydms and extracts the name
+    preference set.
+    input:
+        `str`, tree file name
+    output:
+        `str`, pref name
+    """
+    if "HA_average_prefs" in model_string:
+        return "average"
+    elif "Doud" in model_string:
+        assert ("HA_Doud_prefs" in model_string) or ("HA_hybridDoud_prefs" in model_string), "Unexpected pref set {0}".format(model_string)
+        return "Doud"
+    elif "Lee" in model_string:
+        assert ("HA_Lee_prefs" in model_string) or ("HA_hybridLee_prefs" in model_string), "Unexpected pref set {0}".format(model_string)
+        return "Lee"
+    elif "YNGKP" in model_string:
+        return "N/A"
+    else:
+        raise ValueError("Cannot parse model string {0}".format(model_string))

### 2018-02-09 manuscript/manuscript.tex
small tweaks to model feature results section

@@
 \textbf{Different ways of modeling site-to-site variation in purifying selection.}
-(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
-(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
+(A) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
 In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
 A $\Gamma$ distribution with $K=4$ bins is shown here. 
-(C) Substitution models can both, one, or neither of these features. 
-We used models from the GY94 and the ExpCM families. 
+(B) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
+(C) Substitution models can both, one, or neither of these features and we use models from the GY94 and ExpCM families to represent all possible combinations. 
 }
@@
 
-The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}A;][]{yang1994maximum, yang2000codon}.
-Under such models, some sites have a higher rate of non-synonymous change than others. 
+The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
+For example, in the M5 of \cite{yang2000codon} the implied dN/dS parameter, $\omega$, follows a gamma distribution (GY94+$\Gamma\omega$, \ref{fig:model_feature}A).
+Under the GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
 However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.
@@
 This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
-These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, lartillot2007suppression, le2008phylogenetic, rodrigue2010mutation,si2008empirical,rodrigue2014site,bloom2017identification,hilton2017phydms,bloom2014experimentally}.
+These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
 The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
 While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
-This corresponds to $19\times L$ parameters for a protein of length $L$, which leads to $\sim 10^4$ parameters for a typical size protein.
-As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
-Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
-Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
+This corresponds to $19\times L$ parameters for a protein of length $L$, or $\sim 10^4$ parameters for a typical size protein.
+As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting \skhcomment{cite rodrigue paper}.
+Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
+Therefore, the number of remaining free parameters that are fit from the phylogenetic data for an ExpCM are similar to a non-site-specific substitution model.
 Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.
 
-These two strategies to account for site-to-site variation in purifying selection are not mutually exclusive. 
-\jdbcomment{I'm not sure if I love the rest of this paragraph.
-I feel like GY94 needs to be introduce more directly---perhaps in the paragraph beginning the ``The most common strategy...'' above.
-Also, it's a bit unclear what ``most appropriate'' means.}
-While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.
-Proteins which are under strong purifying and diversifying selection may be better modeled by a combination of the two. 
-For this study, we used four models from the GY94 and ExpCM families, which together represent all possible combinations of these two strategies (\ref{fig:model_feature}C).
+Finally, these two strategies to account for purifying selection, $\Gamma\omega$ rate variation and site-specific stationary states, are not mutually exclusive. 
+Some proteins may be better modeled as a combination of the two. 
+Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to examine the effects of each strategy on branch length estimation.
 
@@
 \textbf{Effect of stationary state and rate variation on long branch estimation.}
-The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models. 
+The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models calculated using \ref{eq:f}. 
 The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
-The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
-\skhcomment{calculated using equation. Maybe change long branch length to expected asymptotic sequence divergence.}
+The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship} (\ref{eq:w_r}).
+\skhcomment{ Maybe change long branch length to expected asymptotic sequence divergence.}
 }
@@
 
+\skhcomment{shorten the following sections}
 The specific parametrization scheme of a model will control features of its stationary state. 

### 2018-02-12 manuscript/manuscript.tex
first draft of empirical trees section

@@
 
-Given a single branch, a substitution model transforms sequence divergence into branch length. 
-Under the molecular clock assumption, this branch length is proportional to time. 
+Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
 However, every substitution model has a stationary state, defining the expected sequence composition after a long evolutionary time. 
-This stationary state can be used to calculate the expected amino-acid divergence a very long amount of time has elapsed. 
-(Rate only affects how fast asymptotic divergence approached, whereas). 
-In \ref{fig:decay}, the stationary state is represented by the ``long" tails where the expected sequence divergence remains constant as time increases.
-
-This is a fundamental feature. 
-The stationary state is principle eigenvector of substitution matrix. 
-Rate variation only multiples by a number, and stationary state of a stochastic matrix is invariant to multiplication.
-
-Along a single branch, a substitution model transforms sequence divergence into branch length. 
-Under the molecular clock assumption, this branch length is proportional to time.  
-However, every substitution model has a stationary state, defining the expected sequence divergence given a very long time, at which this relationship between sequence divergence and time is lost. 
 In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.
 
-\skhcomment{shorten the following sections}
-The specific parametrization scheme of a model will control features of its stationary state. 
-Comparing GY94 to GY94+$\Gamma\omega$ will change how long it takes for the model to reach the stationary state but it does not change the expected sequence divergence at the stationary state. 
-Furthermore, the GY94 and GY94+$\Gamma\omega$ stationary state is exactly the same at each site in the protein. 
-Site-specific stationary states modulate both the amount of time to reach and the overall divergence level at the stationary state. 
-These differences arise from whether or not the site is modeled as constrained or mutationally tolerant.  
-The stationary state is important to branch length estimation because differences in expected divergence will lead to differences in branch length estimation and branch lengths cannot be estimated with any accuracy when two sequences have reached the divergence level of the stationary state.
+By definition, the expected sequence composition varies from site to site for the site-specific stationary state models (\ref{fig:decay} ExpCM, red) but is constant across sites for the uniform stationary state models (\ref{fig:decay} GY94, blue).
+The biggest difference in stationary state between the models is at constrained sites, such as site 305 in \ref{fig:decay}. 
+At such a site, the ExpCM would estimate a much longer branch than the GY94 given the observed divergence of a set of sequences.
 
-The difference in branch length estimation between uniform and site-specific stationary state models will be most dramatic at constrained sites. 
-Most uniform stationary state models define roughly equal frequency of each amino acid, in essence they assume that each site is mutationally tolerant. 
-If the sequence divergence at constrained site is quite small, a site-uniform model will expect a small amount of time to have passed, a short branch, and a site-specific model will expect that a large amount of time has passed, long branch.
-At a mutationally tolerant site, both models will estimate roughly the same branch length for any given divergence.
+Rate variation, as modeled by $\Gamma\omega$, affect the rate at which stationary state is approached by not the sequence composition at stationary state itself. 
+Stationary state is a fundamental feature of stochastic matrices and is invariant to multiplicative terms, such as $\omega$. 
+In \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical.
 
-This difference in branch length estimation cannot be recapitulated by more complex modeling of the $\omega$ parameter. 
-We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM for each site in the protein, following the work of \cite{spielman2015relationship}. 
-This model takes longer to reach stationary state than the GY94 model but, ultimately, has the same divergence level at stationary state. 
-As this model shows, no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state model.
+This means even complex modeling of the $\omega$ to account for purifying selection will not affect the expectation at evolutionary equilibrium. 
+We inferred a site-specific $\omega$ value from the ExpCM \citep{spielman2015relationship} and applied these values to the GY94 model, GY94+$\Gamma\omega$ (\ref{fig:decay} yellow). 
+At constrained sites, this model takes even longer to reach stationary state than the GY94+$\Gamma\omega$ but the eventual stationary state is identical to both the GY94 and GY94+$\Gamma\omega$ stationary state.
+
+No matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.
 
@@
 
-To test the effect of substitution models on branch length estimation given sequences with site-specific amino-acid frequencies we simulated sequences under an ExpCM and re-inferred the branch lengths using the different substitution models described in \ref{fig:model_feature}C.
-As expected, the ExpCM and ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
-The variance in the estimation of an given branch increases between the simulations increased as branch length increased but this error is not biased towards over- or underestimation.
+Next, we wanted to test the effect of substitution model stationary state on branch length estimation given sequences with site-specific amino-acid frequencies. 
+We simulated sequences under an ExpCM and re-inferred the branch lengths using different substitution models described in \ref{fig:model_feature}C.
+
+The ExpCM and the ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
+This accuracy is expected as this was the model the sequences were simulated under. 
+The variance in the estimation of a given branch increases between the simulations increased as branch length increased. 
+However, this error is not biased towards over- or underestimation.
+
 The models with a uniform stationary state consistently underestimate the long branches. 
@@
 
-\skhcomment{Section Outline:}
+Next, we tested the effect of substitution model on branch length estimation using influenza HA sequences. 
+Like the simulated sequences, we assume the influenza HA sequences have site-specific frequencies but unlike the simulations we do not know the underlying model.
 
-\begin{itemize}
-\item We tested the effect substitution model choice on branch length estimation using influenza HA sequences. 
-\begin{itemize}
-\item These sequences presumably evolve with site-specific frequencies but unlike the simulations we do not \textit{know} the model
-\item These sequences are from fairly diverged proteins. Some sequences share only 40\% identity on the amino acid level 
-\item The HA tree has closely related subtypes with long branches between subtypes. This is good because we are looking for the effect on long branches. 
-\end{itemize}
-\item Both $\Gamma\omega$ and non-uniform stationary states result in an extension of branch lengths. 
-\begin{itemize}
-\item The addition of $\Gamma\omega$ extends branches independent of preferences. 
-\item The addition of preferences extends branches \textit{from the focal sequence} independent of $\Gamma\omega$. 
-\end{itemize}
-\item The extension of branch length from the focal sequence shows that the ExpCM does not have equal "relevance" across the whole tree. 
-\begin{itemize}
-\item This result is not entirely surprising. DMS measure the effects of single amino-acid changes in the context of a single genetic background. 
-\item It is expected that there would be epistatic changes across such a highly diverged phylogenetic tree. ``Shifting preferences".
-\item "Sampling" from both preference sets via preference averaging extends the branch length from both focal sequences.
-\end{itemize}
-\end{itemize}
+Influenza HA proteins are group into two groups, based on phylogenetic and antigenic properties. 
+Each HA subtype within the larger groups contains closely related sequences and is connected to the other HA subtype by long branches. 
+The most diverged homologs on this tree are upwards of 60\% diverged on the amino acid level. 
+Consequently, the HA tree is a good model to look at long branch estimation as the there are numerous long branches spread more or less ``evenly" across the entire tree.
+
+We built an HA tree sequences from 14 of the 18 subtype groups, excluding subtype groups with very few sequences. 
+Then we optimized the branch lengths on the tree using models from the GY94 and ExpCM families. 
+We compared ExpCMs defined by two different deep mutational scans measured in two different genetic background. 
+One deep mutational scan was measured in an H1 background and the other was measured in an H3 background. 
+The two focal sequences of these deep mutational scan are only 42\% identical. 
+Comparison of these two models allowed us to examine the effect of ExpCM focal sequence on the branch length estimation. 
+Additionally, because the models we used represent all possible combinations of $\Gamma\omega$ rate variation and site-specific stationary state, we were able to examine the effect of each of these features while controlling for the presence of the other.
+
+The addition of $\Gamma\omega$ rate variation or site-specific stationary state increased the branch lengths estimated. 
+The addition of site-specific stationary state and $\Gamma\omega$ rate variation each increased branch length compared to the model without the feature. 
+The longest branches were estimated with the ExpCM+$\Gamma\omega$, which included both features.
+
+However, the branch length extension for the ExpCM was not uniform across the entire tree. 
+The longest branch length extension is on branches near the focal sequence of the deep mutational scan. 
+The greatest difference between the GY94 and the ExpCM (H1 preferences) is on branches near the H1 clade (black circle) while the greatest difference between the GY94 and the ExpCM (H3 preferences) is on the branches near the H3 clade (black triangle).
+This indicates that the stationary state described by the DMS preferences is much more accurate near the focal sequence of the scan.
+
+This result is not entirely surprising. 
+DMS measures the effect of single amino-acid changes in the context of a single genetic background. 
+However, it is expected that there would be epistatic changes and interactions with such highly diverged homologs. 
+That is, the effect of an amino acid change at one site in the protein might be affected by the amino acid identity at another site in the protein. 
+This would cause the preferences, and consequently the stationary state, to ``shift" across the tree and would explain why there is some locality to our ExpCMs. 
+To test the effect of ``sampling" multiple preference sets, inferred the branch lengths using an ExpCM defined by the average of the H1 and H3 preferences. 
+This model estimated longer branches from both the H1 and H3 clades compared to the GY94 model, indicating that there are these shifts across the entire tree.
 

### 2018-02-12 manuscript/manuscript.tex
first draft of competing

@@
 
-\skhcomment{Section Outline:}
+We further investigated the effect of ``shifting" preferences by comparing behavior of ExpCM across trees with increasing overall divergence from the focal sequence of the deep mutational scans. 
+Based on our results with the large HA trees, we expected the preferences to be more "relevant" on trees which have a low overall sequence divergence from the DMS focal sequence.
 
-\begin{itemize}
-\item We investigated the competing effects of ``shifting preferences" and long branches by comparing ExpCMs with trees of increasing overall divergence from the focal sequence. 
-\item We use the fit stringency parameter $\beta$ as a measure of relevance. 
-\begin{itemize}
-\item The stringency parameter rescales the preferences.
-A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength.
-\item There is an inverse relationship between the stringency parameter and overall sequence divergence. As the overall divergence of the tree increases, the stringency parameter decreases. 
-\end{itemize}
-\item \skhcomment{hypothetical data} There is also a positive relationship between the stringency parameter and branch lengths 
-\begin{itemize}
-\item We optimized the same tree with ExpCMs with different stringency parameters. 
-\item The ExpCMs with higher stringency values estimate longer branches than the ExpCMs with lower stringency values. 
-\end{itemize}
-\end{itemize}
+We use the value of the ExpCM stringency parameter to measure the ``relevance" of a given DMS preference set and tree. 
+A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength, or that the preferences are ``relevant" to evolution in nature.
+Conversely, when the stringency parameter is fit to a value close to zero, the preferences are not ``relevant" and de-emphasize the measurements from the lab so that the preferences are flat. 
+Using the ExpCM (H1 preferences), we observe an inverse relationship between overall divergence and the stringency. 
+As the tree includes more and more sequences which are further from the focal sequence, the relevance of the preferences declines. 
+This indicates that the preferences truly are most relevant on trees with sequences closely related to the focal sequence (short branches).
+
+While the preferences are most relevant for the trees with short branches, due to the low divergence from the focal sequence, the effect of site-specific stationary states on branch length estimation is most important on long branches.
+On the low divergence tree, with the most ``relevant" preferences, roughly the same branch length estimated by the GY94+$\Gamma\omega$  and the ExpCM+$\Gamma\omega$. 
+In fact, it is not until the high divergence tree that a true effect is seen, in the branches from the focal sequences. 
+The effect from site-specific substitution models doesn't really have an effect until the branch lengths are a particular length.
+
+This leads to a regime with competing effects. 
+As the branch lengths increase, the ``importance" of having a site-specific stationary state increases. 
+At the same time, the ``relevance" of an given site-specific stationary decreases as the divergence increases. 
+Current models which have one stationary state across the entire tree, site-specific or not, will not be able to address this tension.
 
@@
 
-\skhcomment{Section Outline}
-
-\begin{itemize}
-\item \texttt{phylobayes} implements the mutation-selection model in a bayesian framework as a way to avoid overfitting. 
-\item \texttt{phylobayes} and ExpCM (H1+H3 average) estimate branches of roughly equal length 
-\begin{itemize}
-\item the ExpCM estimate longer branches from the focal sequences 
-\item \texttt{phylobayes} estimates longer branches from the non-focal sequences 
-\end{itemize}
-\end{itemize}
+Finally, we compared the branch lengths estimated by ExpCM to branch lengths estimated by another site-specific stationary state model. 
+We inferred the branch lengths on the ``high" divergence tree using the mutation-selection model implement in \texttt{phylobayes}. 
+Unlike the ExpCM, this mutation-selection model in \texttt{phylobayes}
 

### 2018-02-12 manuscript/manuscript.tex
edits to the simulations section

@@
 \centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
-\caption{\label{simulations}
+\caption{\label{fig:simulations}
 \textbf{Model performance under simulated, site-specific data.} 
@@
 
-Next, we wanted to test the effect of substitution model stationary state on branch length estimation given sequences with site-specific amino-acid frequencies. 
-We simulated sequences under an ExpCM and re-inferred the branch lengths using different substitution models described in \ref{fig:model_feature}C.
+Next, we wanted to test the effect of substitution model on branch length estimation given sequences with site-specific amino-acid frequencies. 
+To this end, we simulated sequences under an ExpCM and re-inferred the branch lengths using the models  in \ref{fig:model_feature}C.
 
-The ExpCM and the ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
-This accuracy is expected as this was the model the sequences were simulated under. 
-The variance in the estimation of a given branch increases between the simulations increased as branch length increased. 
-However, this error is not biased towards over- or underestimation.
+Models with uniform stationary state underestimated the branch lengths of the simulated sequences. 
+This underestimation is most pronounced on the long branches, with the GY94 estimating branches which are only $\sim \frac{2}{3}$ of the true branch length (\ref{fig:simulations}).
+Accounting for the site-to-site rate variation by $\Gamma\omega$ does not prevent this underestimation as the GY94+$\Gamma\omega$ estimates roughly the same branch length as the GY94. 
+We also estimated branch lengths using an average ExpCM. 
+These models have a uniform stationary state defined by the average preference of amino acid across all sites. 
+Even these models, which share some information with the model used to simulate the sequences, underestimate the long branches.
 
-The models with a uniform stationary state consistently underestimate the long branches. 
-This underestimation is seen with both the GY94 and the GY94+$\Gamma\omega$, indicating that accounting for rate variation via the rate parameter cannot prevent branch length underestimation at these long branches.
-The ExpCM can also be transformed into a uniform stationary state model by setting the amino-acid at every site to the average value for that amino acid across all sites. 
-In contrast to the GY94 models, this average control does not have preference for amino acid and does extract some information from the experiments, but like the GY94 models the stationary state is uniform across sites.
-These models performed similarly to the GY94 and GY94+$\Gamma\omega$ and underestimated long branches.
-
-These simulations show that when the sequences have site-specific amino-acid frequencies, models with uniform stationary states cannot accurately estimate branch lengths and, will in fact, underestimate long branches.
+As expected neither the ExpCM nor the ExpCM+$\Gamma\omega$ underestimate the long branches. 
+However, the variance between simulations increases as the branch length increases, though this error is not biased towards over- or underestimation. 
+This increase in error despite a perfect model match simply underscores the difficulty of estimating extremely long branches. 
+Despite this universal difficulty, it is clear that models with uniform stationary states will always drastically underestimate the long branch lengths.
 
@@
 We inferred the branch lengths on the ``high" divergence tree using the mutation-selection model implement in \texttt{phylobayes}. 
-Unlike the ExpCM, this mutation-selection model in \texttt{phylobayes}
+The amino-acid composition parameters of the mutation-selection model are estimated in a Bayesian framework in \texttt{phylobayes}, rather than defined \textit{a priori} as in the ExpCM.
+
+Both site-specific stationary state models estimate approximately equivalent branch lengths. 
+For a given branch, the two estimates are not exactly the same but each model estimate roughly the same number of branches as longer and to a very similar effect size. 
+However, the locality of the ExpCM can still be seen. 
+Branch lengths from the focal sequences are enriched in the branches which lengths are estimated to be longer by the ExpCM model. 
+That is, the preferences are still more accurate directly around the focal sequence than the amino-acid composition parameters estimated by \texttt{phylobayes} across the entire tree. 
+Altogether, this shows that site-specific models, even those which define the stationary state by different strategies, estimate longer branch lengths than uniform stationary state models but that epistasis plays a non-trivial role in how accurate a single stationary state can be.
 

### 2018-02-13 manuscript/manuscript.tex
edits to the empirical trees section

@@
 
-Next, we tested the effect of substitution model on branch length estimation using influenza HA sequences. 
-Like the simulated sequences, we assume the influenza HA sequences have site-specific frequencies but unlike the simulations we do not know the underlying model.
+Next, we want to examine the effect of substitution model on branch length estimation for actual protein coding sequences. 
+We assume protein coding sequences have site-specific amino-acid preferences, like the simulations, but unlike the simulations, we do not know the underlying model.
 
-Influenza HA proteins are group into two groups, based on phylogenetic and antigenic properties. 
-Each HA subtype within the larger groups contains closely related sequences and is connected to the other HA subtype by long branches. 
-The most diverged homologs on this tree are upwards of 60\% diverged on the amino acid level. 
-Consequently, the HA tree is a good model to look at long branch estimation as the there are numerous long branches spread more or less ``evenly" across the entire tree.
+To this end, we used sequences from the influenza virus surface protein hemagglutinin (HA). 
+HA is a good model for long branch estimation because the 18 HA subtypes are all separated by long branches. 
+Furthermore, HA homologs are incredibly diverse and the most diverged homologs on the tree are only 42\% identical on the amino-acid level. 
+We used a tree with 87 sequences from 14 of the 18 subtypes, excluding subtypes with very few sequences.
 
-We built an HA tree sequences from 14 of the 18 subtype groups, excluding subtype groups with very few sequences. 
-Then we optimized the branch lengths on the tree using models from the GY94 and ExpCM families. 
-We compared ExpCMs defined by two different deep mutational scans measured in two different genetic background. 
-One deep mutational scan was measured in an H1 background and the other was measured in an H3 background. 
-The two focal sequences of these deep mutational scan are only 42\% identical. 
-Comparison of these two models allowed us to examine the effect of ExpCM focal sequence on the branch length estimation. 
-Additionally, because the models we used represent all possible combinations of $\Gamma\omega$ rate variation and site-specific stationary state, we were able to examine the effect of each of these features while controlling for the presence of the other.
+We optimized the branch lengths on the HA tree using the models in \ref{fig:model_feature}C. 
+We used two different ExpCMs defined by DMS preference sets measured independently in two different HA genetic backgrounds. 
+One DMS set was measured in an H1 background (\ref{fig:empirical_trees}, black circle) and the other DMS preference set was measured in an H3 background (\ref{fig:empirical_trees}, black triangle). 
+The two focal sequences of the DMS are only 42\% identical on the amino-acid level. 
+By comparing the two ExpCMs, we can examine the effect of the ExpCM focal sequence on branch length estimation. 
+Additionally, because the models we use represent all possible combinations of $\Gamma\omega$ rate variation and site-specific stationary state, we were able to examine the effect of each of these features while controlling for the presence of the other.
 
-The addition of $\Gamma\omega$ rate variation or site-specific stationary state increased the branch lengths estimated. 
-The addition of site-specific stationary state and $\Gamma\omega$ rate variation each increased branch length compared to the model without the feature. 
-The longest branches were estimated with the ExpCM+$\Gamma\omega$, which included both features.
+Both site-specific stationary states and $\Gamma\omega$ rate variation increase branch length estimates (\ref{fig:empirical_trees}). 
+The addition of either site-specific stationary state or $\Gamma\omega$ rate variation leads to an increase but the longest branches were estimated with the ExpCM+$\Gamma\omega$, which includes both.
 
-However, the branch length extension for the ExpCM was not uniform across the entire tree. 
-The longest branch length extension is on branches near the focal sequence of the deep mutational scan. 
-The greatest difference between the GY94 and the ExpCM (H1 preferences) is on branches near the H1 clade (black circle) while the greatest difference between the GY94 and the ExpCM (H3 preferences) is on the branches near the H3 clade (black triangle).
-This indicates that the stationary state described by the DMS preferences is much more accurate near the focal sequence of the scan.
+However, the branch length extension for the ExpCM compared to the GY94 (or ExpCM+$\Gamma\omega$ compared to the GY94+$\Gamma\omega$) was not uniform across the entire tree. 
+The branches with the largest difference in branch length are near the focal sequence of the deep mutational scans (\ref{fig:empirical_trees}, circle for H1 and triangle for H3). 
+This local effect of the ExpCM indicates that the stationary state described by the DMS preferences is most accurate near the focal sequences of the scan.
 
 This result is not entirely surprising. 
-DMS measures the effect of single amino-acid changes in the context of a single genetic background. 
-However, it is expected that there would be epistatic changes and interactions with such highly diverged homologs. 
-That is, the effect of an amino acid change at one site in the protein might be affected by the amino acid identity at another site in the protein. 
-This would cause the preferences, and consequently the stationary state, to ``shift" across the tree and would explain why there is some locality to our ExpCMs. 
-To test the effect of ``sampling" multiple preference sets, inferred the branch lengths using an ExpCM defined by the average of the H1 and H3 preferences. 
-This model estimated longer branches from both the H1 and H3 clades compared to the GY94 model, indicating that there are these shifts across the entire tree.
+It is expected that protein homologs as diverged as HA would be affected by epistatic interactions, when the effect of a mutation at one site is dependent on the amino-acid identity at another site. 
+One of the strengths of deep mutational scanning is the ability to accurately measure the effect of a single amino-acid mutation. 
+However, these measurements are in the context of a single genetic background and are therefore completely blind to the effect of epistatic interactions. 
+Epistatic interactions would cause the DMS measurements to be different in different genetic backgrounds, resulting in ``shifting" preferences across the tree, and would explain the strong local effect we see with the ExpCM.
+When we average the H1 and H3 preferences, or balanced sampling the two different preference sets, we see estimation of longer branches from both the H1 and H3 clades compared to the GY94 model.  
+This is further evidence of shifts across the tree.
 

### 2018-02-14 manuscript/manuscript.tex
rewrite of phylobayes section

@@
 Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
-However, every substitution model has a stationary state, defining the expected sequence composition after a long evolutionary time. 
+All phylogenetic substitution models \skhcomment{all?} are stochastic matrices and, as such, reach stationary state. 
+Stationary state describes the expected sequence composition after a very long evolutionary time and this stationary state sequence composition is independent to time. 
 In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.
@@
 
-We further investigated the effect of ``shifting" preferences by comparing behavior of ExpCM across trees with increasing overall divergence from the focal sequence of the deep mutational scans. 
-Based on our results with the large HA trees, we expected the preferences to be more "relevant" on trees which have a low overall sequence divergence from the DMS focal sequence.
+We looked closer at the effect of shifting preferences on branch length estimation. 
+We examined the behavior of the ExpCM on trees with varying levels of overall sequence divergence from the ExpCM focal sequence. 
+For each tree, we asked ``does the site-specific stationary state model estimate longer branches than the uniform stationary state model?" and ``is the preference set defining the ExpCM relevant?"
 
-We use the value of the ExpCM stringency parameter to measure the ``relevance" of a given DMS preference set and tree. 
-A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength, or that the preferences are ``relevant" to evolution in nature.
-Conversely, when the stringency parameter is fit to a value close to zero, the preferences are not ``relevant" and de-emphasize the measurements from the lab so that the preferences are flat. 
-Using the ExpCM (H1 preferences), we observe an inverse relationship between overall divergence and the stringency. 
-As the tree includes more and more sequences which are further from the focal sequence, the relevance of the preferences declines. 
-This indicates that the preferences truly are most relevant on trees with sequences closely related to the focal sequence (short branches).
+The effect of site-specific stationary state on branch length estimation is seen most strongly on long branches. 
+As the full HA tree represents ``high" divergence from the DMS focal sequence, we took subsets the tree to create trees with either ``intermediate" or ``low" divergence from the DMS focal sequence. 
+The majority of the branch length estimates are very similar between the GY94+$\Gamma\omega$ and the ExpCM+$\Gamma\omega$ for the ``low" and ``intermediate" alignments. 
+Only when the tree has a high overall divergence from the DMS focal sequence, and therefore the longest branches, is there a difference between the two models. 
+This result is not unsurprising. 
+One of the original motivations for the ``mutation-selection" models was the effect the model would have on long branches specifically.
 
-While the preferences are most relevant for the trees with short branches, due to the low divergence from the focal sequence, the effect of site-specific stationary states on branch length estimation is most important on long branches.
-On the low divergence tree, with the most ``relevant" preferences, roughly the same branch length estimated by the GY94+$\Gamma\omega$  and the ExpCM+$\Gamma\omega$. 
-In fact, it is not until the high divergence tree that a true effect is seen, in the branches from the focal sequences. 
-The effect from site-specific substitution models doesn't really have an effect until the branch lengths are a particular length.
+We wanted to determine the accuracy or relevance of the ExpCM on these subtrees. 
+Due to the local branch length extension in \ref{fig:empirical_trees}, we expect that the preferences are most accurate on the low divergence tree.
+We use the ExpCM stringency parameters as a proxy for accuracy of the preference set.
+The stringency parameter rescales the raw preferences from the deep mutational scan. 
+Large stringency parameter values ($>1$) indicate that selection in nature prefers the same amino acids as selection in lab but with a higher stringency.
+Or that the preferences measured in lab are relevant for describing the evolution of that protein. 
+The stringency parameter for ExpCM(H1 prefs) has the highest value for the low divergence alignment and declines as the overall divergence increases. 
+The ExpCM(H3 prefs) show the same pattern. 
+The inverse relationship between stringency parameter and divergence supports the hypothesis that epistatic interactions degrade the ExpCM stationary state as defined by DMS preferences. 
+\skhcomment{All epistatic? Could there just be a change?}
 
-This leads to a regime with competing effects. 
-As the branch lengths increase, the ``importance" of having a site-specific stationary state increases. 
-At the same time, the ``relevance" of an given site-specific stationary decreases as the divergence increases. 
 Current models which have one stationary state across the entire tree, site-specific or not, will not be able to address this tension.
@@
 
-Finally, we compared the branch lengths estimated by ExpCM to branch lengths estimated by another site-specific stationary state model. 
-We inferred the branch lengths on the ``high" divergence tree using the mutation-selection model implement in \texttt{phylobayes}. 
-The amino-acid composition parameters of the mutation-selection model are estimated in a Bayesian framework in \texttt{phylobayes}, rather than defined \textit{a priori} as in the ExpCM.
+Finally, we compared the branch lengths estimated between two different models with site-specific stationary states. 
+We compared the branch lengths estimated by ExpCM+$\Gamma\omega$ (avg) with the branch lengths estimated by the mutation-selection model implemented in \texttt{phylobayes}. 
+The \texttt{phylobayes} mutation-selection model estimates the amino-acid preferences which define the stationary state in a Bayesian framework rather than defining them from the DMS \textit{a priori}.
+Branch length estimates are very similar between the ExpCM+$\Gamma\omega$ (avg) and the \texttt{phylobayes} mutation-selection model. 
+The equivalence between the two models shows the general importance of site-specific stationary states. 
+The local accuracy of the DMS preferences and the ExpCM can still be seen. 
+The branch length estimates from either focal sequence (H1 and H3) are longer by the ExpCM than by the \texttt{phylobayes} mutation-selection model. 
+This indicates that the DMS preferences are still more accurate for the local branches around the focal sequence than the global preferences estimated by \texttt{phylobayes}.
 
-Both site-specific stationary state models estimate approximately equivalent branch lengths. 
-For a given branch, the two estimates are not exactly the same but each model estimate roughly the same number of branches as longer and to a very similar effect size. 
-However, the locality of the ExpCM can still be seen. 
-Branch lengths from the focal sequences are enriched in the branches which lengths are estimated to be longer by the ExpCM model. 
-That is, the preferences are still more accurate directly around the focal sequence than the amino-acid composition parameters estimated by \texttt{phylobayes} across the entire tree. 
-Altogether, this shows that site-specific models, even those which define the stationary state by different strategies, estimate longer branch lengths than uniform stationary state models but that epistasis plays a non-trivial role in how accurate a single stationary state can be.
+\skhcomment{What is the average difference in branch length between the two models?}
 

### 2018-02-14 manuscript/manuscript.tex
small tweaks to model feature section

@@
 \noindent  
-Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
+Molecular phylogenetics is used to estimate the time since the divergence of modern gene sequences.
 Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
@@
 This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
-Here we show that substitution models informed by experimental measurements of the purifying selection due to site-specific amino-acid preferences lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
+Here we show that substitution models informed by experimental measurements of the purifying selection \skhcomment{due to site-specific amino-acid preferences} lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
 This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
-The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
+The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites \skhcomment{repeat sentence start}.
 However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
@@
 
-Proteins evolve under both purifying selection to maintain their structure and function. 
+Proteins evolve under purifying selection to maintain their structure and function. 
 This purifying selection is not homogenous across sites in a protein.
-It is also not homogenous across the different amino acids at a given site.
+It is also not homogenous among the different amino acids at a given site.
 For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
@@
 The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
-For example, in the M5 of \cite{yang2000codon} the implied dN/dS parameter, $\omega$, follows a gamma distribution (GY94+$\Gamma\omega$, \ref{fig:model_feature}A).
-Under the GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
+For example, in GY94+$\Gamma\omega$ (M5 of \cite{yang2000codon}), the implied dN/dS parameter, $\omega$, follows a gamma distribution (\ref{fig:model_feature}A).
+Under GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
 However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.
@@
 These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
-The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
+However, the specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
 While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
@@
 Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
-Therefore, the number of remaining free parameters that are fit from the phylogenetic data for an ExpCM are similar to a non-site-specific substitution model.
+Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
 Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.
@@
 Some proteins may be better modeled as a combination of the two. 
-Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to examine the effects of each strategy on branch length estimation.
+Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to dissect the effect of each model of purifying selection on branch length estimation.
 

### 2018-02-14 manuscript/manuscript.tex
edits to decay results section

@@
 Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
-All phylogenetic substitution models \skhcomment{all?} are stochastic matrices and, as such, reach stationary state. 
-Stationary state describes the expected sequence composition after a very long evolutionary time and this stationary state sequence composition is independent to time. 
-In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.
+However, a complicating factor in this estimation procedure is the stationary state of the model. 
+Stationary state, as the principle eigenvector of a stochastic matrix, is a fundamental feature of any substitution model. 
+It describes the expected sequence composition after a very long evolutionary time. 
+Additionally, the stationary state does not change under application of the transition matrix, which means this sequence composition remains constant as time increases. 
+The invariance of the stationary state is represented by the long ``tails" in \ref{fig:decay}. 
+The exact sequence composition of stationary state is unique given the parametrization of the model.
 
-By definition, the expected sequence composition varies from site to site for the site-specific stationary state models (\ref{fig:decay} ExpCM, red) but is constant across sites for the uniform stationary state models (\ref{fig:decay} GY94, blue).
-The biggest difference in stationary state between the models is at constrained sites, such as site 305 in \ref{fig:decay}. 
-At such a site, the ExpCM would estimate a much longer branch than the GY94 given the observed divergence of a set of sequences.
+We visualized the effect of stationary state on branch length estimation by examining the expected amino-acid identity of a set of sequences given some branch length for site-specific and uniform stationary state models.
+By definition, the expected identity given a very long time varies from site to site for the site-specific stationary state models (\ref{fig:decay}, ExpCM, red) but is constant across sites for the uniform stationary state models (\ref{fig:decay}, GY94, blue).
+Due to the assumed equivalence of amino-acid preferences in the GY94 model, the biggest difference in stationary state between the ExpCM and the GY94 is seen at constrained sites, such as site 305 in \ref{fig:decay}.
+The ExpCM would estimate a much longer branch than the GY94 model given some observed divergence at such a site.
 
-Rate variation, as modeled by $\Gamma\omega$, affect the rate at which stationary state is approached by not the sequence composition at stationary state itself. 
-Stationary state is a fundamental feature of stochastic matrices and is invariant to multiplicative terms, such as $\omega$. 
-In \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical.
-
-This means even complex modeling of the $\omega$ to account for purifying selection will not affect the expectation at evolutionary equilibrium. 
-We inferred a site-specific $\omega$ value from the ExpCM \citep{spielman2015relationship} and applied these values to the GY94 model, GY94+$\Gamma\omega$ (\ref{fig:decay} yellow). 
-At constrained sites, this model takes even longer to reach stationary state than the GY94+$\Gamma\omega$ but the eventual stationary state is identical to both the GY94 and GY94+$\Gamma\omega$ stationary state.
-
-No matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.
+Modeling site-to-site rate variation by $\Gamma\omega$ affects the rate at which stationary state is approached but not the sequence composition at stationary state itself. 
+For example, in \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical. 
+Because the stationary state is invariant to multiplicative terms, increasingly complex modeling of $\omega$ will never fully recapitulate the difference in expected sequence composition between the ExpCM and the GY94 at constrained sites.
+We inferred an $\omega$ value from the ExpCM for each site \citep{spielman2015relationship} and applied this values to the GY94 model (\ref{fig:decay} yellow)
+These site-specific values drastically decreased the amount of time it takes to reach stationary state at the constrained sites but still does not affect the stationary state. 
+This means that no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.
 

### 2018-02-14 manuscript/manuscript.tex
small tweak to model feature figure

@@
 \caption{\label{fig:model_feature}
-\textbf{Different ways of modeling site-to-site variation in purifying selection.}
-(A) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
-In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
+\textbf{Different ways of modeling the site-to-site rate variation of purifying selection.}
+(A) The implied dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary following some statistical distribution (blue circles). 
+In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin. 
 A $\Gamma$ distribution with $K=4$ bins is shown here.
-(B) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
+(B) A substitution model stationary state defines the expected sequence composition after a very long evolutionary time. 
+Depending on how the model is parametrized, this expected sequence composition cab be uniform or variable across sites. 
+Each column is a site in the protein and the height of each letter is the expected proportion of that amino acid at stationary state. 
 (C) Substitution models can both, one, or neither of these features and we use models from the GY94 and ExpCM families to represent all possible combinations. 
+Green indicates presence of a feature and white indicates absence. 
 }

### 2018-02-14 manuscript/manuscript.tex
tweaks to decay and simulations figure legends

@@
 \caption{\label{fig:decay}
-\textbf{Effect of stationary state and rate variation on long branch estimation.}
-The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models calculated using \ref{eq:f}. 
-The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
-The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship} (\ref{eq:w_r}).
-\skhcomment{ Maybe change long branch length to expected asymptotic sequence divergence.}
+\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on expected asymptotic sequence divergence.}
+For each model, the expected amino-acid identity given some branch length is calculated using \ref{eq:f}. 
+The deep mutational scanning amino-acid preferences from \cite{doud2016accurate} defining the ExpCM (red) are shown as logoplots. 
+The implied dN/dS ($\omega_r$) for the GY94+$\omega_r$ (yellow) is calculated from the ExpCM using \ref{eq:w_r}. 
+\skhcomment{I use `implied dN/dS' to refer to two different concepts in the results.}
 }
@@
 \textbf{Model performance under simulated, site-specific data.} 
-Alignments were simulated under an ExpCM along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
-The averaged ExpCMs amino-acid frequencies defined by the average preference of that amino acid across all sites in the protein. 
-While these models extract information from the deep mutational scanning experiment, they are not site-specific. 
-Grey points represent the length of one branch and the black points are the mean branch lengths over ten simulations. 
-The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
+Alignments were simulated under an ExpCM and branch lengths were re-estimated using models from the ExpCM and GY94 families. 
+The averaged ExpCMs have a uniform stationary state defined by the average preference for each amino acid across all sites in the protein. 
+The length of one branch in one simulation is shown in grey and the average branch length of one branch across ten simulations is shown in black. 
+The grey, dashed line is the reference line $y=x$, representing the behavior of an unbiased estimator. 
 }

### 2018-02-14 manuscript/manuscript.tex
changes to simulations section

@@
 
-Next, we wanted to test the effect of substitution model on branch length estimation given sequences with site-specific amino-acid frequencies. 
-To this end, we simulated sequences under an ExpCM and re-inferred the branch lengths using the models  in \ref{fig:model_feature}C.
+Next, we tested the effect of stationary state and $\Gamma\omega$ rate variation on branch length estimation given sequences with site-specific amino-acid frequencies. 
+We simulated sequences under an ExpCM, re-estimated the branch lengths using the models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths.
 
-Models with uniform stationary state underestimated the branch lengths of the simulated sequences. 
-This underestimation is most pronounced on the long branches, with the GY94 estimating branches which are only $\sim \frac{2}{3}$ of the true branch length (\ref{fig:simulations}).
-Accounting for the site-to-site rate variation by $\Gamma\omega$ does not prevent this underestimation as the GY94+$\Gamma\omega$ estimates roughly the same branch length as the GY94. 
-We also estimated branch lengths using an average ExpCM. 
-These models have a uniform stationary state defined by the average preference of amino acid across all sites. 
-Even these models, which share some information with the model used to simulate the sequences, underestimate the long branches.
+The models with a uniform stationary state underestimated the branch lengths of the simulated sequences, especially the lengths of long branches. 
+The GY94 model estimated branches lengths which are only about 60\% of the true branch length for the longest branches. 
+Accounting for site-to-site rate variation by $\Gamma\omega$ did not prevent the underestimation of long branches, as the GY94+$\Gamma\omega$ inferred approximately the same branch lengths as the GY94 model.
+We also estimated branch lengths using averaged ExpCMs.
+The averaged ExpCMs have a uniform stationary state defined by the average DMS preference for a given amino acid across all sites from the deep mutational scan. 
+Even these models, which share information with the model used to simulate the sequences, underestimated the long branches. 
+The inability of these uniform stationary state models to specifically describe purifying selection on a site by site basis clearly hinder their ability to accurately estimate long branch lengths.
 
-As expected neither the ExpCM nor the ExpCM+$\Gamma\omega$ underestimate the long branches. 
-However, the variance between simulations increases as the branch length increases, though this error is not biased towards over- or underestimation. 
-This increase in error despite a perfect model match simply underscores the difficulty of estimating extremely long branches. 
-Despite this universal difficulty, it is clear that models with uniform stationary states will always drastically underestimate the long branch lengths.
+Neither the ExpCM nor the ExpCM+$\Gamma\omega$ underestimated the long branches. 
+This is an unsurprising result, as the sequences were simulated under an ExpCM. 
+However, while not biased towards over- or underestimation, the variance between simulations increased as the branch length increased. 
+The increase in error, even with a perfect model match, underscores the difficulty of estimating extremely long branches. 
+Despite this universal difficulty, it is clear from the GY94 and averaged ExpCM results that models with a uniform stationary state will always drastically underestimate long branches.
 

### 2018-02-14 manuscript/manuscript.tex
tweaks to empirical data table

@@
 \caption{\label{tab:empirical_data}
-Model comparison for Fig.~\ref{fig:empirical_trees}.\skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}} 
+Performance comparison for the models in~\ref{fig:empirical_trees}. 
+All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
+The mean $\omega$ value from each $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation, otherwise simply the fit, gene-wide $\omega$. 
+Each ExpCM has a stringency parameter $>1$.
+\skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}
+} 
      \begin{tabular}{cccccccccc}
         \hline
-         Model & {\shortstack{Stationary\\ State}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
-       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & yes & yes & 0 & -487510 & 0.19,  0.50,  0.90,  1.86 &  1.70, \\
-	ExpCM (H1+H3 avg) & yes & no &  950 & -492270 & 0.15 & 1.78\\
-	ExpCM + $\Gamma\omega$ (H1) & yes & yes & 13060 & -494040  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
-	ExpCM + $\Gamma\omega$ (H3) & yes & yes & 17370 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
-	ExpCM (H1) & yes & no & 2556 & -50030 &  0.13 & 1.22\\
-	ExpCM (H3) & yes & no &  3197 & -50350 & 0.12 & 1.45\\
-	GY94 + $\Gamma\omega$ & no & yes & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & N/A \\
-	GY94 & no & no & 7625 & -52560  & 0.07 & N/A\\
+         Model & {\shortstack{Site-specific\\ stationary state}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
+       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & + & + & 0 & -487510 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
+	ExpCM (H1+H3 avg) & + & - &  950 & -492270 & 0.15 & 1.78\\
+	ExpCM + $\Gamma\omega$ (H1) & + & + & 1306 & -494040  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
+	ExpCM + $\Gamma\omega$ (H3) & + & + & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
+	ExpCM (H1) & + & - & 2556 & -50030 &  0.13 & 1.22\\
+	ExpCM (H3) & + & - &  3197 & -50350 & 0.12 & 1.45\\
+	GY94 + $\Gamma\omega$ & - & + & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
+	GY94 & - & - & 7625 & -52560  & 0.07 & -\\
       \end{tabular}

### 2018-02-15 manuscript/manuscript.tex
rewrite of empirical data section

@@
 \caption{\label{fig:empirical_trees}
-\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to GY94 models.} ).
+\textbf{Effect of site-specific stationary state and $\Gamma\omega$ rate variation on HA branch length estimation.} 
+The HA tree is optimized by models from the ExpCM and GY94 families. 
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots. 
+The amino-acid preferences measured by the deep mutational scans defining the ExpCMs were measured in either an H1 background, an H3 background, or were the average measurements from the H1 and H3 scans.
+The circle denotes the H1 clade and the triangle denotes the H3 clade. 
 }
@@
 
-Next, we want to examine the effect of substitution model on branch length estimation for actual protein coding sequences. 
-We assume protein coding sequences have site-specific amino-acid preferences, like the simulations, but unlike the simulations, we do not know the underlying model.
+Next, we wanted to examine the effect of substitution model on branch length estimation for actual protein coding sequences. 
+We assume protein coding sequences have site-specific amino-acid preferences, but unlike the simulations, we do not know the underlying model.
+We used sequences from the influenza virus surface protein hemagglutinin (HA). 
+The HA tree structure makes it a useful model for studying long branch estimation; HA is conserved within any given subtype but the subtypes themselves are separated by long branches. 
+The most diverged subtypes on the HA tree are only about 40\% identical on the amino acid level. 
+We used an HA tree with 87 sequences from 14 of the 18 subtypes, excluding subtypes with very few sequences.
 
-To this end, we used sequences from the influenza virus surface protein hemagglutinin (HA). 
-HA is a good model for long branch estimation because the 18 HA subtypes are all separated by long branches. 
-Furthermore, HA homologs are incredibly diverse and the most diverged homologs on the tree are only 42\% identical on the amino-acid level. 
-We used a tree with 87 sequences from 14 of the 18 subtypes, excluding subtypes with very few sequences.
+We estimated the branch lengths on this HA tree using the models in \ref{fig:model_feature}C. 
+We used two different ExpCMs defined by different deep mutational scanning experiments from different HA homologs. 
+One scan measured the site-specific amino-acid preferences in an H1 background (\ref{fig:empirical_trees}, circle) and the other measured the preferences in an H3 background (\ref{fig:empirical_trees}, triangle).
+These two deep mutational scans sample the breadth of HA diversity as their respective focal sequences are only $\sim42\%$ identical on the amino acid level. 
+By comparing these two ExpCMs, we can examine the effect of the experiment's focal sequence on branch length estimation, along with the effect of site-specific stationary state and $\Gamma\omega$ rate variation.
 
-We optimized the branch lengths on the HA tree using the models in \ref{fig:model_feature}C. 
-We used two different ExpCMs defined by DMS preference sets measured independently in two different HA genetic backgrounds. 
-One DMS set was measured in an H1 background (\ref{fig:empirical_trees}, black circle) and the other DMS preference set was measured in an H3 background (\ref{fig:empirical_trees}, black triangle). 
-The two focal sequences of the DMS are only 42\% identical on the amino-acid level. 
-By comparing the two ExpCMs, we can examine the effect of the ExpCM focal sequence on branch length estimation. 
-Additionally, because the models we use represent all possible combinations of $\Gamma\omega$ rate variation and site-specific stationary state, we were able to examine the effect of each of these features while controlling for the presence of the other.
+Models with either site-specific stationary state or $\Gamma\omega$ rate variation estimated longer branches than models without (\ref{fig:empirical_trees}). 
+Furthermore, we can isolate the effect of each method to account for purifying selection from the other. 
+For example, the increase in branch length estimated by GY94+$\Gamma\omega$ compared to GY94 and by ExpCM+$\Gamma\omega$ compared to ExpCM shows that $\Gamma\omega$ rate variation increases branch length in the presence and absence of a site-specific stationary state.
+We also see an increase in branch length with the addition of a site-specific stationary state in the presence and absence of $\Gamma\omega$ rate variation (\ref{fig:empirical_trees}, GY94 vs.
+ExpCM and GY94+$\Gamma\omega$ vs.
+ExpCM+$\Gamma\omega$).
+The model with both methods, ExpCM+$\Gamma\omega$, estimated the longest branches of all the models. 
+However, the branch length extension from any of the ExpCM compared to any of the GY94 models was not uniform across the entire tree. 
+The branches with the largest difference in branch length are near the focal sequence of the deep mutational scans (\ref{fig:empirical_trees}, circle for H1 and triangle for H3)
+This local effect of the ExpCM indicates that the stationary state described by the deep mutational scanning preferences is most accurate near the focal sequences of the scan.
 
-Both site-specific stationary states and $\Gamma\omega$ rate variation increase branch length estimates (\ref{fig:empirical_trees}). 
-The addition of either site-specific stationary state or $\Gamma\omega$ rate variation leads to an increase but the longest branches were estimated with the ExpCM+$\Gamma\omega$, which includes both.
-
-However, the branch length extension for the ExpCM compared to the GY94 (or ExpCM+$\Gamma\omega$ compared to the GY94+$\Gamma\omega$) was not uniform across the entire tree. 
-The branches with the largest difference in branch length are near the focal sequence of the deep mutational scans (\ref{fig:empirical_trees}, circle for H1 and triangle for H3). 
-This local effect of the ExpCM indicates that the stationary state described by the DMS preferences is most accurate near the focal sequences of the scan.
-
-This result is not entirely surprising. 
-It is expected that protein homologs as diverged as HA would be affected by epistatic interactions, when the effect of a mutation at one site is dependent on the amino-acid identity at another site. 
-One of the strengths of deep mutational scanning is the ability to accurately measure the effect of a single amino-acid mutation. 
-However, these measurements are in the context of a single genetic background and are therefore completely blind to the effect of epistatic interactions. 
-Epistatic interactions would cause the DMS measurements to be different in different genetic backgrounds, resulting in ``shifting" preferences across the tree, and would explain the strong local effect we see with the ExpCM.
-When we average the H1 and H3 preferences, or balanced sampling the two different preference sets, we see estimation of longer branches from both the H1 and H3 clades compared to the GY94 model.  
-This is further evidence of shifts across the tree.
+The local effect of the ExpCM is not entirely surprising. 
+Proteins are affected by epistatic interactions, where the effect of a mutation at one site is dependent on the amino-acid identity at another site. 
+While one of the strengths of deep mutational scanning is the ability to accurately measure the effect of all single amino-acid mutations, these measurements are done in the context of a single genetic background and do not take into account epistatic interactions.
+Consequently, the stationary state defined by these preferences is most accurate for only a subset of the tree.
+An ExpCM with a more general stationary state, defined by the average of the H1 and H3 preferences, lengthen the preferences from both the H1 and H3 clades (\ref{fig:empirical_trees}).
 

### 2018-02-15 manuscript/manuscript.tex
small tweaks

@@
 The averaged ExpCMs have a uniform stationary state defined by the average preference for each amino acid across all sites in the protein. 
-The length of one branch in one simulation is shown in grey and the average branch length of one branch across ten simulations is shown in black. 
-The grey, dashed line is the reference line $y=x$, representing the behavior of an unbiased estimator. 
+The length of one branch in one simulation is shown in grey and the average length of one branch across ten simulations is shown in black. 
+The grey, dashed reference line $y=x$ represents the behavior of an unbiased estimator. 
 }
@@
 Models with either site-specific stationary state or $\Gamma\omega$ rate variation estimated longer branches than models without (\ref{fig:empirical_trees}). 
-Furthermore, we can isolate the effect of each method to account for purifying selection from the other. 
+Furthermore, we can isolate the effect of each method. 
 For example, the increase in branch length estimated by GY94+$\Gamma\omega$ compared to GY94 and by ExpCM+$\Gamma\omega$ compared to ExpCM shows that $\Gamma\omega$ rate variation increases branch length in the presence and absence of a site-specific stationary state.

### 2018-02-16 manuscript/manuscript.tex
revisions to abstract and Results section 1

@@
 \topmargin 0.0cm
-\oddsidemargin 0.2cm
-\textwidth 16cm 
+%\oddsidemargin 2cm
+%\textwidth 16cm 
 \textheight 21cm
@@
 This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
-Here we show that substitution models informed by experimental measurements of the purifying selection \skhcomment{due to site-specific amino-acid preferences} lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
+Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
 This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
-The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites \skhcomment{repeat sentence start}.
+The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
 However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
@@
 
-\subsection*{Different ways hat substitution models account for purifying selection}
+\subsection*{Different ways that substitution models account for purifying selection}
 
-\jdbcomment{Some other comments on this section, which I think in general is pretty good: $\omega$ shows up in the figure, but is never mentioned in the text.
-I feel like GY94 models need to be explained at least in terms of what they and $\Gamma$ distribution stand for.
-Just introducing nomenclature.
-I feel like the last paragraph should tie back to panel C of the figure.}
-
-\begin{figure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
@@
 \textbf{Different ways of modeling the site-to-site rate variation of purifying selection.}
-(A) The implied dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary following some statistical distribution (blue circles). 
+(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary following some statistical distribution (blue circles). 
 In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin. 
-A $\Gamma$ distribution with $K=4$ bins is shown here.
+A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
 (B) A substitution model stationary state defines the expected sequence composition after a very long evolutionary time. 
-Depending on how the model is parametrized, this expected sequence composition cab be uniform or variable across sites. 
-Each column is a site in the protein and the height of each letter is the expected proportion of that amino acid at stationary state. 
-(C) Substitution models can both, one, or neither of these features and we use models from the GY94 and ExpCM families to represent all possible combinations. 
-Green indicates presence of a feature and white indicates absence. 
+Most substitution models have stationary states that are uniform across sites.
+However, substitution models can have site-specific stationary states.
+In the logo plots, each column is a site in the protein and the height of each letter is the frequency of that amino acid at stationary state. 
+(C) Substitution models can incorporate neither, one, or both of these features.
+Here we will use substitution models from the \citet{goldman1994codon} (GY94) and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed rates to represent all possible combinations.
+%Green indicates presence of a feature and white indicates absence. 
 }
@@
 
-The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
-For example, in GY94+$\Gamma\omega$ (M5 of \cite{yang2000codon}), the implied dN/dS parameter, $\omega$, follows a gamma distribution (\ref{fig:model_feature}A).
-Under GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
-However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.
+Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use\jdbcomment{find a citation for this and replace word ``accurate'' with something else if appropriate.
+Perhaps this one or something cited within \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4620419/}}.
+Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
+The relative rate of these substitutions is referred to as dN/dS or $\omega$.
+In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
+Here we will use such substitution models in the form proposed by \citet{goldman1994codon}, which we denote GY94.
+Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called GY94 M0.
+This gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
 
-In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein.
-This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
-These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
-However, the specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
-While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
-This corresponds to $19\times L$ parameters for a protein of length $L$, or $\sim 10^4$ parameters for a typical size protein.
-As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting \skhcomment{cite rodrigue paper}.
-Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
+A single gene-wide $\omega$ ignores the fact that purifying selection is heterogeneous across sites.
+The most common strategy to ameliorate this defect is to allow the rate of nonsynonymous change ($\omega$) to vary among sites following some statistical distribution~\citep{yang1994maximum,yang2000codon}.
+For instance, in the M5 variant of the GY94 model~\citep{yang2000codon}, $\omega$ follows a gamma distribution as shown in \ref{fig:model_feature}A.
+We will denote this model as GY94+$\Gamma\omega$.
+A GY94+$\Gamma\omega$ captures the fact that the rate of nonsynonymous change can vary across sites. 
+However, this formulation treats all nonsynonymous substitutions equivalently, since the rate is agnostic to the identity of the specific amino-acid mutation.
+
+A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary}.
+These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
+This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
+These models better capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
+
+However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
+Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
+For instance, a GY94 M5 model with the typical F3X4 stationary state\jdbcomment{add citation, maybe just GY94 paper?} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
+However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
+This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
+It is challenging to obtain values for these parameters without overfitting the data\skhcomment{cite rodrigue paper}.
+Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
 Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
@@
 
-Finally, these two strategies to account for purifying selection, $\Gamma\omega$ rate variation and site-specific stationary states, are not mutually exclusive. 
-Some proteins may be better modeled as a combination of the two. 
-Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to dissect the effect of each model of purifying selection on branch length estimation.
+Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
+Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
+This $\omega$ parameter can be drawn from a distribution just like as for GY94-style models. 
+We will denote ExpCM's that also include a distribution over $\omega$ as ExpCM+$\Gamma\omega$.
+\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.
 

### 2018-02-16 manuscript/manuscript.tex
more progress on Results

@@
 
-\subsection*{Effect of stationary state and rate variation on branch length estimation}
+\subsection*{Effect of stationary state and rate variation on branch-length estimation}
 
-\begin{figure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
 \caption{\label{fig:decay}
-\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on expected asymptotic sequence divergence.}
-For each model, the expected amino-acid identity given some branch length is calculated using \ref{eq:f}. 
-The deep mutational scanning amino-acid preferences from \cite{doud2016accurate} defining the ExpCM (red) are shown as logoplots. 
-The implied dN/dS ($\omega_r$) for the GY94+$\omega_r$ (yellow) is calculated from the ExpCM using \ref{eq:w_r}. 
-\skhcomment{I use `implied dN/dS' to refer to two different concepts in the results.}
+\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
+The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by \citet{doud2016accurate}.
+The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
+For the GY94 model, the graphs are identical for all sites since this models does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
+The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
+However, all GY94 models including the one with site-specific $\omega_r$ values all approach the same asymptote since they all have the same stationary state.
+But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
 }
@@
 
-Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
-However, a complicating factor in this estimation procedure is the stationary state of the model. 
-Stationary state, as the principle eigenvector of a stochastic matrix, is a fundamental feature of any substitution model. 
-It describes the expected sequence composition after a very long evolutionary time. 
-Additionally, the stationary state does not change under application of the transition matrix, which means this sequence composition remains constant as time increases. 
-The invariance of the stationary state is represented by the long ``tails" in \ref{fig:decay}. 
-The exact sequence composition of stationary state is unique given the parametrization of the model.
+Given a single branch, a substitution model transforms sequence identity into branch length.
+Under a molecular-clock assumption, this branch length is proportional to time.
+The transformation from sequence identity to branch length is trivial and insensitive to the choice of substitution model when the sequence identity is high.
+For instance, in the case when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
+However, as more substitutions accumulate then it becomes progressively more likely for multiple changes to occur at the same site.
+In this regime, the accuracy of the substitution model becomes critical for transforming the sequence identity into branch length.
+Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic sequence identity.
+For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
+If the substitution model underestimates the asymptotic sequence divergence then it will also underestimate long branch lengths, since it will predict that long branches lead to more divergence lead than is actually the case.
+
+\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of an H1 influenza hemagglutinin (HA).
+The GY94 and GY94+$\Gamma\omega$ models each predict the same behavior for all sites, since they do not have any site-specific parameters.
+Both of these models predict an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored by the models due to more redundant codons and biases towards certain nucleotides.
+Intuitively, this asymptotic sequence identity seems low, since like many proteins HA has a highly conserved structure and function which imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs.
+
+\jdbcomment{Then talk about we have a sense of these constraints from the DMS, and if we use them to put an $\omega_r$ it changes rate of approach of asymptote but not asymptote.
+Then ExpCM with different asymptotes.
+Then point about stationary state.}
 
@@
 This means that no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.
+
+However, a complicating factor in this estimation procedure is the stationary state of the model. 
+Stationary state, as the principle eigenvector of a stochastic matrix, is a fundamental feature of any substitution model. 
+It describes the expected sequence composition after a very long evolutionary time. 
+Additionally, the stationary state does not change under application of the transition matrix, which means this sequence composition remains constant as time increases. 
+The invariance of the stationary state is represented by the long ``tails" in \ref{fig:decay}. 
+The exact sequence composition of stationary state is unique given the parametrization of the model.
 

### 2018-02-16 manuscript/manuscript.tex
updates to Results section 2

@@
 In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
-Here we will use such substitution models in the form proposed by \citet{goldman1994codon}, which we denote GY94.
-Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called GY94 M0.
-This gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
+Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
+Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
+We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
+The gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
 
@@
 This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
-These models better capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
+These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
 
@@
 This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
-It is challenging to obtain values for these parameters without overfitting the data\skhcomment{cite rodrigue paper}.
+It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
 Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
@@
 Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
-This $\omega$ parameter can be drawn from a distribution just like as for GY94-style models. 
-We will denote ExpCM's that also include a distribution over $\omega$ as ExpCM+$\Gamma\omega$.
+This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
+We will denote such models as ExpCM+$\Gamma\omega$.
 \ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.
@@
 Under a molecular-clock assumption, this branch length is proportional to time.
-The transformation from sequence identity to branch length is trivial and insensitive to the choice of substitution model when the sequence identity is high.
-For instance, in the case when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
+The transformation from sequence identity to branch length is trivial when the sequence identity is high.
+For instance, when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
 However, as more substitutions accumulate then it becomes progressively more likely for multiple changes to occur at the same site.
@@
 
-\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of an H1 influenza hemagglutinin (HA).
-The GY94 and GY94+$\Gamma\omega$ models each predict the same behavior for all sites, since they do not have any site-specific parameters.
-Both of these models predict an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored by the models due to more redundant codons and biases towards certain nucleotides.
-Intuitively, this asymptotic sequence identity seems low, since like many proteins HA has a highly conserved structure and function which imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs.
+\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
+The GY94 predicts the same behavior for all sites, since it does not have any site-specific parameters.
+This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
+Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function which imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs.
 
-\jdbcomment{Then talk about we have a sense of these constraints from the DMS, and if we use them to put an $\omega_r$ it changes rate of approach of asymptote but not asymptote.
-Then ExpCM with different asymptotes.
-Then point about stationary state.}
+Accounting for site-to-site rate variation in GY94 models affects the rate at which the asymptotic sequence identity is approached, but not the actual value of this asymptote. 
+For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but the asymptote for both models is identical. 
+This fact holds true even if we use experimental measurements of HA's site-specific amino-acid preferences~\citep{doud2016accurate} to calculate a different $\omega_r$ value for each site using the method of \citet[\ref{eq:w_r};][]{spielman2015relationship}.
+Specifically, this GY94+$\omega_r$ model predicts that different sites will approach the asymptote at different rates, but the asymptote is always the same (\ref{fig:decay}).
+The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of homogeneous substitution models.
+The asymptotic sequence divergence is a function of only the stationary state of the substitution matrix, and the stationary state of a stochastic matrix is invariant with respect to multiplication by any non-zero number\jdbcomment{But this isn't true, because $\omega$ only multiplies some elements of transition matrix.
+I think it's probably something like if you multiple pairs of elements on equivalent positions across diagonal?}.
+Therefore, different schemes for modeling $\omega$ can alter the time it takes to reach stationary state, but no matter how ``well" a model accounts for site-to-site variation in $\omega$ it will always have the same stationary state as a simple GY94 model.
+
+\jdbcomment{In contrast, ExpCM models have different stationary states at each site...}
 
@@
 The ExpCM would estimate a much longer branch than the GY94 model given some observed divergence at such a site.
-
-Modeling site-to-site rate variation by $\Gamma\omega$ affects the rate at which stationary state is approached but not the sequence composition at stationary state itself. 
-For example, in \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical. 
-Because the stationary state is invariant to multiplicative terms, increasingly complex modeling of $\omega$ will never fully recapitulate the difference in expected sequence composition between the ExpCM and the GY94 at constrained sites.
-We inferred an $\omega$ value from the ExpCM for each site \citep{spielman2015relationship} and applied this values to the GY94 model (\ref{fig:decay} yellow)
-These site-specific values drastically decreased the amount of time it takes to reach stationary state at the constrained sites but still does not affect the stationary state. 
-This means that no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.
 

### 2018-02-18 manuscript/manuscript.tex
added in Jesse's comments from slack (no text change)

@@
 \topmargin 0.0cm
-%\oddsidemargin 2cm
-%\textwidth 16cm 
+\oddsidemargin 1cm
+\textwidth 15cm 
 \textheight 21cm
@@
 
-\subsection*{Failure to account for site-specificity leads to branch length underestimation.}
+\subsection*{Simulations demonstrate how failure to account for site-specificity leads to branch-length underestimation.}
 
@@
 \end{figure}
+
+\skhcomment{Change "averaged" to "ExpCM averaged across sites". This will distinguish it from the other types of averages used later. Also clarify in text referring to figure}
 
@@
 
-\subsection*{empirical Data}
+\subsection*{Site-specific models estimate longer branches on real data.}
 
@@
 \skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}
+\skhcomment{check LL values}
 } 
@@
 \end{table}
+
+\skhcomment{In the beginning of the 4th Results section, you don't need to say we assume site-specific preferences.
+We aren't assuming anything, we're just analyzing the data.
+Give a better intro to why we chose HA: there are data sets for two diverged variants (H1 and H3), and this gene has been widely studied using phylogenetics.}
 
@@
 \subsection*{Competing effects of shifting preferences and long branches.}
+
+\skhcomment{I know I was pressing on this a lot, but I think that the three trees in Figure 5 turn out not to be useful as they aren't aligned in a helpful way.
+So I would suggest instead creating something that is a hybrid between the current Supplementary figure 4A and the Figure 5B.
+It shows the sequence subsets, and then just shows that the stringency parameter goes down for more diverged alignments.
+I like the Supplementary figure 4A as it shows the trees most clearly.
+I like the Figure 5B because it gives the stringency parameters but also shows their actual effects in logos, which is more intuitive than just plotting them on a graph.}
 
@@
 \subsection*{\texttt{phylobayes}}
+
+\skhcomment{For Figure 6, make a better x-legend. All the models are mutation-selection models. Instead write something like `phylobayes` for the x-legend}
 
@@
 \textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)}
-\skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.} 
  }
@@
 \textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.77$)}}
-\skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.}
 \end{suppfig}
-
-\begin{table}[t!]
-\caption{\label{tab:simulation_params}
-ExpCM parameters used to simulate sequences in Fig.~\ref{fig:simulation}.}
-      \begin{tabular}{ccccc}
-        \hline
-          Parameter & Value\\ \hline
-       	$\beta$ & $1.54$\\
-	$\kappa$ & $3.60$\\
-	$\omega$ & $0.20$\\
-	$\phi_A$, $\phi_C$, $\phi_G$& $0.38$, $0.17$, $0.23$\\
-      \end{tabular}
-\end{table}
-
-\begin{table}[t!]
-\caption{\label{tab:wsn_low_params}
-Model parameters used in  Fig.~\ref{fig:decay}.}
-      \begin{tabular}{ccccc}
-        \hline
-          Model & Parameters\\ \hline
-          ExpCM & $\beta=1.54196$\\
-           & $\kappa=3.47184$\\
-           & $\omega=0.219225$\\ 
-          YNGKP M0 & $\kappa=2.9984$\\
-          & $\omega=0.09076$\\
-          YNGKP M5 & $\kappa=2.9984$\\
-          & $\omega=0.09076$\\
-      \end{tabular}
-\end{table}
 

### 2018-02-19 manuscript/manuscript.tex
added comment about multiplicative term & matrices
changed axis label for phylobayes
changed averaged ExpCM name in simulations

@@
 \jdbcomment{In contrast, ExpCM models have different stationary states at each site...}
+\skhcomment{Google books: reversible matrix exchangeability stationary state}
 
@@
 Alignments were simulated under an ExpCM and branch lengths were re-estimated using models from the ExpCM and GY94 families. 
-The averaged ExpCMs have a uniform stationary state defined by the average preference for each amino acid across all sites in the protein. 
+The ExpCM's averaged across sites have a uniform stationary state defined by the average preference for each amino acid across all sites in the protein. 
 The length of one branch in one simulation is shown in grey and the average length of one branch across ten simulations is shown in black. 
@@
 \end{figure}
-
-\skhcomment{Change "averaged" to "ExpCM averaged across sites". This will distinguish it from the other types of averages used later. Also clarify in text referring to figure}
 
@@
 Accounting for site-to-site rate variation by $\Gamma\omega$ did not prevent the underestimation of long branches, as the GY94+$\Gamma\omega$ inferred approximately the same branch lengths as the GY94 model.
-We also estimated branch lengths using averaged ExpCMs.
-The averaged ExpCMs have a uniform stationary state defined by the average DMS preference for a given amino acid across all sites from the deep mutational scan. 
+We also estimated branch lengths using ExpCM's averaged across sites.
+The ExpCM's averaged across sites have a uniform stationary state defined by the average DMS preference for a given amino acid across all sites from the deep mutational scan. 
 Even these models, which share information with the model used to simulate the sequences, underestimated the long branches. 
@@
 \subsection*{\texttt{phylobayes}}
-
-\skhcomment{For Figure 6, make a better x-legend. All the models are mutation-selection models. Instead write something like `phylobayes` for the x-legend}
 

### 2018-02-19 manuscript/manuscript.tex
changed LL values in table

@@
          Model & {\shortstack{Site-specific\\ stationary state}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
-       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & + & + & 0 & -487510 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
-	ExpCM (H1+H3 avg) & + & - &  950 & -492270 & 0.15 & 1.78\\
-	ExpCM + $\Gamma\omega$ (H1) & + & + & 1306 & -494040  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
+       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & + & + & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
+	ExpCM (H1+H3 avg) & + & - &  950 & -49227 & 0.15 & 1.78\\
+	ExpCM + $\Gamma\omega$ (H1) & + & + & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
 	ExpCM + $\Gamma\omega$ (H3) & + & + & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\

### 2018-02-19 manuscript/manuscript.tex
removed old comment
added citations to model feature section

@@
 
-Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use\jdbcomment{find a citation for this and replace word ``accurate'' with something else if appropriate.
-Perhaps this one or something cited within \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4620419/}}.
+Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
 Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
@@
 Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
-For instance, a GY94 M5 model with the typical F3X4 stationary state\jdbcomment{add citation, maybe just GY94 paper?} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
+For instance, a GY94 M5 model with the typical F3X4 stationary state~\citep{yang1994maximum} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
 However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
@@
 \skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}
-\skhcomment{check LL values}
 } 

### 2018-02-19 manuscript/manuscript.tex
updates to references

@@
 The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of homogeneous substitution models.
-The asymptotic sequence divergence is a function of only the stationary state of the substitution matrix, and the stationary state of a stochastic matrix is invariant with respect to multiplication by any non-zero number\jdbcomment{But this isn't true, because $\omega$ only multiplies some elements of transition matrix.
-I think it's probably something like if you multiple pairs of elements on equivalent positions across diagonal?}.
+The asymptotic sequence divergence is a function of only the stationary state of the substitution matrix, and the stationary state of a stochastic matrix is invariant with respect to multiplication by any non-zero number~\citep{nielsen2006statistical}.
 Therefore, different schemes for modeling $\omega$ can alter the time it takes to reach stationary state, but no matter how ``well" a model accounts for site-to-site variation in $\omega$ it will always have the same stationary state as a simple GY94 model.
@@
 \jdbcomment{In contrast, ExpCM models have different stationary states at each site...}
-\skhcomment{Google books: reversible matrix exchangeability stationary state}
 

### 2018-02-20 manuscript/manuscript.tex
some revisions

@@
 \noindent  
-Molecular phylogenetics is used to estimate the time since the divergence of modern gene sequences.
-Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
-For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
-This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
+Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
+For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
+In the case of viruses, there is independent evidence that the inferences from molecular phylogenetics are underestimates of the true deep divergence times.  
+This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
 Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
-This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
+This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
 The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
@@
 \textbf{Different ways of modeling the site-to-site rate variation of purifying selection.}
-(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary following some statistical distribution (blue circles). 
+(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
 In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin. 
@@
 
-Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
-Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
+Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
+Codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
 The relative rate of these substitutions is referred to as dN/dS or $\omega$.
@@
 A single gene-wide $\omega$ ignores the fact that purifying selection is heterogeneous across sites.
-The most common strategy to ameliorate this defect is to allow the rate of nonsynonymous change ($\omega$) to vary among sites following some statistical distribution~\citep{yang1994maximum,yang2000codon}.
+The most common strategy to ameliorate this defect is to allow $\omega$ to vary among sites according to some statistical distribution~\citep{yang1994maximum,yang2000codon}.
 For instance, in the M5 variant of the GY94 model~\citep{yang2000codon}, $\omega$ follows a gamma distribution as shown in \ref{fig:model_feature}A.
@@
 A GY94+$\Gamma\omega$ captures the fact that the rate of nonsynonymous change can vary across sites. 
-However, this formulation treats all nonsynonymous substitutions equivalently, since the rate is agnostic to the identity of the specific amino-acid mutation.
+However, this formulation treats all nonsynonymous substitutions equivalently, since the rate is agnostic to the amino-acid identity of the mutation.
 
-A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary}.
+A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
 These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
@@
 Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
-For instance, a GY94 M5 model with the typical F3X4 stationary state~\citep{yang1994maximum} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
+For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
 However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
@@
 For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
-If the substitution model underestimates the asymptotic sequence divergence then it will also underestimate long branch lengths, since it will predict that long branches lead to more divergence lead than is actually the case.
+If the substitution model underestimates the asymptotic sequence divergence then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.
 
 \ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
-The GY94 predicts the same behavior for all sites, since it does not have any site-specific parameters.
+The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
 This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
-Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function which imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs.
+Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs\jdbcomment{citation, I think Juhye has something she cites that looks at really diverged HAs}.
 
@@
 For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but the asymptote for both models is identical. 
-This fact holds true even if we use experimental measurements of HA's site-specific amino-acid preferences~\citep{doud2016accurate} to calculate a different $\omega_r$ value for each site using the method of \citet[\ref{eq:w_r};][]{spielman2015relationship}.
+This fact holds true even if we use experimental measurements of HA's site-specific amino-acid preferences~\citep{doud2016accurate} to calculate a different $\omega_r$ value for each site using the method of \citet{spielman2015relationship} (see \ref{eq:w_r}).
 Specifically, this GY94+$\omega_r$ model predicts that different sites will approach the asymptote at different rates, but the asymptote is always the same (\ref{fig:decay}).
-The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of homogeneous substitution models.
-The asymptotic sequence divergence is a function of only the stationary state of the substitution matrix, and the stationary state of a stochastic matrix is invariant with respect to multiplication by any non-zero number~\citep{nielsen2006statistical}.
-Therefore, different schemes for modeling $\omega$ can alter the time it takes to reach stationary state, but no matter how ``well" a model accounts for site-to-site variation in $\omega$ it will always have the same stationary state as a simple GY94 model.
+The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of reversible substitution models.
+These models are mathematically represented by reversible stochastic matrices, which can be decomposed into stationary states and symmetric exchangeability matrices~\citep{nielsen2006statistical}.
+The stationary state is invariant with respect to multiplication of the symmetric exchangeability matrix by any non-zero number.
+Different schemes for modeling $\omega$ only affect elements of the symmetric exchangeability matrix.
+Therefore, no matter how ``well" a model accounts for site-to-site variation in $\omega$, it will always have the same stationary state as a simple GY94 model.
 
-\jdbcomment{In contrast, ExpCM models have different stationary states at each site...}
+However, mutation-selection models such as ExpCM's have site-specific stationary states.
+Therefore, they predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
+For instance, \ref{fig:decay} shows that at sites such as 348 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
+When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a very long branch length at modest sequence identities that GY94 model might attribute to a much shorter branch.
 
-We visualized the effect of stationary state on branch length estimation by examining the expected amino-acid identity of a set of sequences given some branch length for site-specific and uniform stationary state models.
-By definition, the expected identity given a very long time varies from site to site for the site-specific stationary state models (\ref{fig:decay}, ExpCM, red) but is constant across sites for the uniform stationary state models (\ref{fig:decay}, GY94, blue).
-Due to the assumed equivalence of amino-acid preferences in the GY94 model, the biggest difference in stationary state between the ExpCM and the GY94 is seen at constrained sites, such as site 305 in \ref{fig:decay}.
-The ExpCM would estimate a much longer branch than the GY94 model given some observed divergence at such a site.
-
-However, a complicating factor in this estimation procedure is the stationary state of the model. 
-Stationary state, as the principle eigenvector of a stochastic matrix, is a fundamental feature of any substitution model. 
-It describes the expected sequence composition after a very long evolutionary time. 
-Additionally, the stationary state does not change under application of the transition matrix, which means this sequence composition remains constant as time increases. 
-The invariance of the stationary state is represented by the long ``tails" in \ref{fig:decay}. 
-The exact sequence composition of stationary state is unique given the parametrization of the model.
-
-\subsection*{Simulations demonstrate how failure to account for site-specificity leads to branch-length underestimation.}
-
-\begin{figure}[H]
+\subsection*{Simulations demonstrate how failure to model site-specific stationary states leads to branch-length underestimation.}
+\begin{figure}
 \centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
 \caption{\label{fig:simulations}
-\textbf{Model performance under simulated, site-specific data.} 
-Alignments were simulated under an ExpCM and branch lengths were re-estimated using models from the ExpCM and GY94 families. 
-The ExpCM's averaged across sites have a uniform stationary state defined by the average preference for each amino acid across all sites in the protein. 
-The length of one branch in one simulation is shown in grey and the average length of one branch across ten simulations is shown in black. 
-The grey, dashed reference line $y=x$ represents the behavior of an unbiased estimator. 
+\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
+We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences measured by deep mutational scanning of an H1 HA~\citep{doud2016accurate}.
+We then inferred the branch lengths of this tree on the simulated alignments.
+The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
+We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
+The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
+\jdbcomment{Because of order we will discuss in text, switch order of ExpCM and ExpCM averaged across sites.}
 }
 \end{figure}
+To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
+Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees}) using ExpCM parameterized by the actual amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
+We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
 
-Next, we tested the effect of stationary state and $\Gamma\omega$ rate variation on branch length estimation given sequences with site-specific amino-acid frequencies. 
-We simulated sequences under an ExpCM, re-estimated the branch lengths using the models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths.
-
-The models with a uniform stationary state underestimated the branch lengths of the simulated sequences, especially the lengths of long branches. 
-The GY94 model estimated branches lengths which are only about 60\% of the true branch length for the longest branches. 
-Accounting for site-to-site rate variation by $\Gamma\omega$ did not prevent the underestimation of long branches, as the GY94+$\Gamma\omega$ inferred approximately the same branch lengths as the GY94 model.
-We also estimated branch lengths using ExpCM's averaged across sites.
-The ExpCM's averaged across sites have a uniform stationary state defined by the average DMS preference for a given amino acid across all sites from the deep mutational scan. 
-Even these models, which share information with the model used to simulate the sequences, underestimated the long branches. 
-The inability of these uniform stationary state models to specifically describe purifying selection on a site by site basis clearly hinder their ability to accurately estimate long branch lengths.
-
-Neither the ExpCM nor the ExpCM+$\Gamma\omega$ underestimated the long branches. 
-This is an unsurprising result, as the sequences were simulated under an ExpCM. 
-However, while not biased towards over- or underestimation, the variance between simulations increased as the branch length increased. 
-The increase in error, even with a perfect model match, underscores the difficulty of estimating extremely long branches. 
-Despite this universal difficulty, it is clear from the GY94 and averaged ExpCM results that models with a uniform stationary state will always drastically underestimate long branches.
+The models with a uniform stationary state underestimated the lengths of long branches separating the simulated sequences (\ref{fig:simulations}). 
+The GY94 model estimated branches lengths which are only about 60\% of the true values for the longest branches. 
+Accounting for site-to-site variation in $\omega$ did not fix the fundament problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the length of the longest branches.
+However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models, which accurately accounted for the site-specific amino-acid preferences used in the simulations.
+The improved performance of the ExpCM's is due to their site-specific modeling of the stationary state: if we parameterize these models by site-specific amino-acid preferences that have been averaged across HA sites, then they perform no better than GY94 models.
+Therefore, models with uniform stationary states are fundamentally incapable of accurately estimating the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.
 
@@
 
-\begin{figure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figures/empirical_trees.pdf}}
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
 \caption{\label{fig:empirical_trees}
@@
 The circle denotes the H1 clade and the triangle denotes the H3 clade. 
+\jdbcomment{You could perhaps do fewer preferences (just a bit fewer), or maybe better break the ExpCM labels onto two lines.
+Or maybe even better, but model labels above preferences so they can be bigger and maybe in bold, and it will also align the preferences with the y-axis on the plots.}
 }
@@
 Each ExpCM has a stringency parameter $>1$.
-\skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}
+\jdbcomment{Let's get rid of the 2nd and 3rd column}
 } 
@@
 
-\skhcomment{In the beginning of the 4th Results section, you don't need to say we assume site-specific preferences.
-We aren't assuming anything, we're just analyzing the data.
-Give a better intro to why we chose HA: there are data sets for two diverged variants (H1 and H3), and this gene has been widely studied using phylogenetics.}
+The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM models.
+But how do these models perform on real rather than simulated data?
+Real genes do evolve under site-specific constraint, but this constraint is almost certainly not modeled in all its complexity by an ExpCM.
+However, if ExpCM's do a substantially better job of modeling the constraint, we might still expect them to estimate more accurate branch lengths.
 
-Next, we wanted to examine the effect of substitution model on branch length estimation for actual protein coding sequences. 
-We assume protein coding sequences have site-specific amino-acid preferences, but unlike the simulations, we do not know the underlying model.
-We used sequences from the influenza virus surface protein hemagglutinin (HA). 
-The HA tree structure makes it a useful model for studying long branch estimation; HA is conserved within any given subtype but the subtypes themselves are separated by long branches. 
-The most diverged subtypes on the HA tree are only about 40\% identical on the amino acid level. 
-We used an HA tree with 87 sequences from 14 of the 18 subtypes, excluding subtypes with very few sequences.
+To test the models on real data, we used actual sequences of influenza HA. 
+The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
+HA consists of a number of different subtypes.
+Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
+However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\jdbcomment{?}) have a highly conserved structure that performs a highly conserved function\jdbcomment{citation, Juhye can probably help}.
+We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
+For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
+Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty probably lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
 
-We estimated the branch lengths on this HA tree using the models in \ref{fig:model_feature}C. 
-We used two different ExpCMs defined by different deep mutational scanning experiments from different HA homologs. 
+Given this fixed HA tree topology, we estimated the branch lengths on the tree using each of the substitution models in \ref{fig:model_feature}C. 
+There are two all amino-acid deep mutational scanning datasets for HA~\jdbcomment{citations}, and each parameterizes a different ExpCM.
 One scan measured the site-specific amino-acid preferences in an H1 background (\ref{fig:empirical_trees}, circle) and the other measured the preferences in an H3 background (\ref{fig:empirical_trees}, triangle).
@@
 
+\jdbcomment{The results in Figure ? show the distinct effects of modeling variation in $\omega$ and variation in the stationary state... (Basically, more clearly delineate the two effects)}
 Models with either site-specific stationary state or $\Gamma\omega$ rate variation estimated longer branches than models without (\ref{fig:empirical_trees}). 

### 2018-02-20 manuscript/manuscript.tex
added comment about competing effects figure
tweaks to empirical trees figure

@@
 The circle denotes the H1 clade and the triangle denotes the H3 clade. 
-\jdbcomment{You could perhaps do fewer preferences (just a bit fewer), or maybe better break the ExpCM labels onto two lines.
-Or maybe even better, but model labels above preferences so they can be bigger and maybe in bold, and it will also align the preferences with the y-axis on the plots.}
 }
@@
 I like the Figure 5B because it gives the stringency parameters but also shows their actual effects in logos, which is more intuitive than just plotting them on a graph.}
+\skhcomment{Two panels one for H1 and one for H3. Each panel has big tree with the subtrees labeled and the same prefs snippet as fig 4 with the different beta values for each subtree.}
 

### 2018-02-20 manuscript/manuscript.tex
removed columns from model comparison table

@@
 Each ExpCM has a stringency parameter $>1$.
-\jdbcomment{Let's get rid of the 2nd and 3rd column}
 } 
@@
         \hline
-         Model & {\shortstack{Site-specific\\ stationary state}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
-       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & + & + & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
-	ExpCM (H1+H3 avg) & + & - &  950 & -49227 & 0.15 & 1.78\\
-	ExpCM + $\Gamma\omega$ (H1) & + & + & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
-	ExpCM + $\Gamma\omega$ (H3) & + & + & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
-	ExpCM (H1) & + & - & 2556 & -50030 &  0.13 & 1.22\\
-	ExpCM (H3) & + & - &  3197 & -50350 & 0.12 & 1.45\\
-	GY94 + $\Gamma\omega$ & - & + & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
-	GY94 & - & - & 7625 & -52560  & 0.07 & -\\
+         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
+       	ExpCM + $\Gamma\omega$ (H1+H3 avg)  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
+	ExpCM (H1+H3 avg)  &  950 & -49227 & 0.15 & 1.78\\
+	ExpCM + $\Gamma\omega$ (H1)  & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
+	ExpCM + $\Gamma\omega$ (H3) & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
+	ExpCM (H1) & 2556 & -50030 &  0.13 & 1.22\\
+	ExpCM (H3) &  3197 & -50350 & 0.12 & 1.45\\
+	GY94 + $\Gamma\omega$  & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
+	GY94 & 7625 & -52560  & 0.07 & -\\
       \end{tabular}

### 2018-02-26 manuscript/manuscript.tex
small tweaks to simulations figure
added citation to model feature section

@@
 We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
-The gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
+The gene-wide $\omega$ is usually $<1$\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
 
@@
 The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
-\jdbcomment{Because of order we will discuss in text, switch order of ExpCM and ExpCM averaged across sites.}
 }

### 2018-02-26 manuscript/manuscript.tex
updates to competing effects text and figure

@@
 
-\jdbcomment{The results in Figure ? show the distinct effects of modeling variation in $\omega$ and variation in the stationary state... (Basically, more clearly delineate the two effects)}
-Models with either site-specific stationary state or $\Gamma\omega$ rate variation estimated longer branches than models without (\ref{fig:empirical_trees}). 
-Furthermore, we can isolate the effect of each method. 
-For example, the increase in branch length estimated by GY94+$\Gamma\omega$ compared to GY94 and by ExpCM+$\Gamma\omega$ compared to ExpCM shows that $\Gamma\omega$ rate variation increases branch length in the presence and absence of a site-specific stationary state.
-We also see an increase in branch length with the addition of a site-specific stationary state in the presence and absence of $\Gamma\omega$ rate variation (\ref{fig:empirical_trees}, GY94 vs.
-ExpCM and GY94+$\Gamma\omega$ vs.
-ExpCM+$\Gamma\omega$).
-The model with both methods, ExpCM+$\Gamma\omega$, estimated the longest branches of all the models. 
-However, the branch length extension from any of the ExpCM compared to any of the GY94 models was not uniform across the entire tree. 
-The branches with the largest difference in branch length are near the focal sequence of the deep mutational scans (\ref{fig:empirical_trees}, circle for H1 and triangle for H3)
-This local effect of the ExpCM indicates that the stationary state described by the deep mutational scanning preferences is most accurate near the focal sequences of the scan.
+The results in \ref{fig:empirical_trees} show the distinct effects of modeling variation in $\omega$ and variation in the stationary state. 
+The models with $\Gamma\omega$ rate variation estimated longer branches than the same base model without this variation. 
+For example, in \ref{fig:empirical_trees}, the GY94+$\Gamma\omega$ model estimated longer branches than the GY94 and the ExpCM+$\Gamma
+\omega$(H1 prefs) estimated longer branches than the ExpCM(H1 prefs).
 
-The local effect of the ExpCM is not entirely surprising. 
-Proteins are affected by epistatic interactions, where the effect of a mutation at one site is dependent on the amino-acid identity at another site. 
-While one of the strengths of deep mutational scanning is the ability to accurately measure the effect of all single amino-acid mutations, these measurements are done in the context of a single genetic background and do not take into account epistatic interactions.
-Consequently, the stationary state defined by these preferences is most accurate for only a subset of the tree.
-An ExpCM with a more general stationary state, defined by the average of the H1 and H3 preferences, lengthen the preferences from both the H1 and H3 clades (\ref{fig:empirical_trees}).
+The models with site-specific stationary states, the ExpCM's, estimated longer branches than the GY94 models. 
+Both the ExpCM(H1 prefs) and ExpCM(H3) prefs estimated longer branches than the GY94 model. 
+However, in contrast with the effect of adding $\Gamma\omega$ rate variation, the effect of site-specific stationary defined by deep mutational scanning preferences is not uniform across the entire tree.
+The effect is largest on the branches near the focal sequence of the deep mutational scan (\ref{fig:empirical_trees}). 
+While the increase in length estimated by the ExpCM's has a strong local effect, the increase is in seen even in the presence of $\Gamma\omega$ rate variation. 
+The ExpCM+$\Gamma\omega$ estimated longer branches than the GY94+$\Gamma\omega$. 
+The site-specificity of the ExpCM's stationary state is clearly important for estimating the long branches on the HA tree.
 
@@
 
-\skhcomment{I know I was pressing on this a lot, but I think that the three trees in Figure 5 turn out not to be useful as they aren't aligned in a helpful way.
-So I would suggest instead creating something that is a hybrid between the current Supplementary figure 4A and the Figure 5B.
-It shows the sequence subsets, and then just shows that the stringency parameter goes down for more diverged alignments.
-I like the Supplementary figure 4A as it shows the trees most clearly.
-I like the Figure 5B because it gives the stringency parameters but also shows their actual effects in logos, which is more intuitive than just plotting them on a graph.}
-\skhcomment{Two panels one for H1 and one for H3. Each panel has big tree with the subtrees labeled and the same prefs snippet as fig 4 with the different beta values for each subtree.}
-
 \begin{figure}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/doud_compete_2}}
-\caption{\label{fig:doud_compete}
-\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
-\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
-\textbf{(B)} The phylogenetic tree of the "high" alignment. 
-The colors denote the alignment and the black circle denotes the focal H3 sequence. 
-\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
-\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
-Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
-The branch lengths are in average number of codon substitutions per site. 
+\centerline{\includegraphics[width=0.85\textwidth]{figures/compete_3}}
+\caption{\label{fig:compete}
+\textbf{Effect of overall divergence on the congruency between natural sequences and the deep mutational scanning preferences.} 
+For each deep mutational scanning preference set, we fit an ExpCM to three trees representing high, intermediate, and low overall divergence from the experimental focal sequence. 
+We used the ExpCM stringency parameter ($\beta$) to determine the congruency between the natural sequences and the experimental measurements. 
+A $\beta >1$ indicates that the natural sequences favor the same amino-acid preferences as the experimental measurements but with a greater strength. 
+A $\beta < 1$ indicates the natural sequences do not favor the same amino-acid preferences as the experimental measurements or favor the same amino acids but less strongly. 
+As the overall divergence increased, the $\beta$ value for the ExpCM(H3 prefs) and ExpCM(H1 prefs) decreased, indicating that the preferences do ``shift" over the full HA tree. 
+\skhcomment{1. I need to clean up and do a better job aligning the different parts of this figure but what do you think of the overall layout? 2. \ref{supp:compete_4} shows an alternative layout.}
 }
@@
 
-We looked closer at the effect of shifting preferences on branch length estimation. 
-We examined the behavior of the ExpCM on trees with varying levels of overall sequence divergence from the ExpCM focal sequence. 
-For each tree, we asked ``does the site-specific stationary state model estimate longer branches than the uniform stationary state model?" and ``is the preference set defining the ExpCM relevant?"
+The local effect of the ExpCM is not entirely surprising. 
+Proteins are affected epistatic interactions, where the effect of a mutation at one site is dependent on the amino-acid identity at another site. 
+While one of the strengths of deep mutational scanning is the ability to accurately measure the effect of all single amino-acid mutations, these measurements are done in the context of a single genetic background and do not take into account epistatic interactions.
+The stationary state of the ExpCM's is only as accurate as deep mutational scanning preferences and the results in \ref{fig:empirical_trees} indicate that any given deep mutational scan is most accurate for only a subset of the tree with closely related sequences to the focal sequence.
 
-The effect of site-specific stationary state on branch length estimation is seen most strongly on long branches. 
-As the full HA tree represents ``high" divergence from the DMS focal sequence, we took subsets the tree to create trees with either ``intermediate" or ``low" divergence from the DMS focal sequence. 
-The majority of the branch length estimates are very similar between the GY94+$\Gamma\omega$ and the ExpCM+$\Gamma\omega$ for the ``low" and ``intermediate" alignments. 
-Only when the tree has a high overall divergence from the DMS focal sequence, and therefore the longest branches, is there a difference between the two models. 
-This result is not unsurprising. 
-One of the original motivations for the ``mutation-selection" models was the effect the model would have on long branches specifically.
+%An ExpCM with a more general stationary state, defined by the average of the H1 and H3 preferences, increases the estimates of branch length from both the H1 and H3 clade (\ref{fig:empirical_trees}).
 
-We wanted to determine the accuracy or relevance of the ExpCM on these subtrees. 
-Due to the local branch length extension in \ref{fig:empirical_trees}, we expect that the preferences are most accurate on the low divergence tree.
-We use the ExpCM stringency parameters as a proxy for accuracy of the preference set.
-The stringency parameter rescales the raw preferences from the deep mutational scan. 
-Large stringency parameter values ($>1$) indicate that selection in nature prefers the same amino acids as selection in lab but with a higher stringency.
-Or that the preferences measured in lab are relevant for describing the evolution of that protein. 
-The stringency parameter for ExpCM(H1 prefs) has the highest value for the low divergence alignment and declines as the overall divergence increases. 
-The ExpCM(H3 prefs) show the same pattern. 
-The inverse relationship between stringency parameter and divergence supports the hypothesis that epistatic interactions degrade the ExpCM stationary state as defined by DMS preferences. 
-\skhcomment{All epistatic? Could there just be a change?}
+We examined how the deep mutational scanning preferences shift across the full HA tree by comparing the ExpCM behavior on trees with varying levels of overall divergence from the ExpCM focal sequence.
+For each of the subtrees in \ref{fig:compete}, we asked "how well do the deep mutational scanning preferences agree with the natural sequences?" using the ExpCM stringency parameter $\beta$ (\ref{eq:Frxy}).
+The fit $\beta$ parameter rescales the raw preferences from the deep mutational scan and relates the selection in nature to selection in the lab. 
+We interpret $\beta >1$ as selection in nature favoring the same amino acids as selection in the experiments, or that the amino-acid preferences are a good model of purifying selection. 
+The larger the $\beta$ more strongly the selection in nature favors these same, experimental preferences. 
+Conversely, we interpret a $\beta<1$ as selection in nature favoring different amino acids than selection in lab and that the preferences are not a good description of natural evolution.  
+\skhcomment{mention pop-gen derivation and the relationship between lab vs. nature selection and pop size?}
 
-Current models which have one stationary state across the entire tree, site-specific or not, will not be able to address this tension.
+The results in \ref{fig:compete} show that as the overall divergence from the focal sequence of a deep mutational scan increase, the $\beta$ value decreases. 
+This inverse relationship between $\beta$ and overall divergence is seen for both the ExpCM(H1 prefs) and ExpCM(H3 prefs). 
+Along with the local effect in branch length extension seen \ref{fig:empirical_trees}, this indicates that the amino-acid preferences have shifted across the HA tree. 
+This leaves us in a regime with inherent tension. 
+Models which take into account the site-specific constraints of purifying selection in their stationary states are important for the accurate estimation of long branches but the site-specific constraints are not static throughout evolution.
+Models with a single stationary state regardless of lineage, site-specific or not, will not be able to capture the shifting functional constraints due to epistatic interactions. 
+\skhcomment{All betas are greater than 1 so the preferences are ``better" than the GY94 model even at high overall divergence.}
 
@@
 \centerline{\includegraphics[width=0.85\textwidth]{figures/lee_compete}}
-\caption{\label{fig:lee_compete}
+\caption{\label{supp:lee_compete}
 \textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
@@
 
+\begin{suppfig}[H]
+\centerline{\includegraphics[width=0.85\textwidth]{figures/compete_4}}
+\caption{\label{supp:compete_4}
+}
+\end{suppfig}
+
 \clearpage 

### 2018-02-26 manuscript/manuscript.tex
rewrite of phylobayes section

@@
 
-Finally, we compared the branch lengths estimated between two different models with site-specific stationary states. 
-We compared the branch lengths estimated by ExpCM+$\Gamma\omega$ (avg) with the branch lengths estimated by the mutation-selection model implemented in \texttt{phylobayes}. 
-The \texttt{phylobayes} mutation-selection model estimates the amino-acid preferences which define the stationary state in a Bayesian framework rather than defining them from the DMS \textit{a priori}.
-Branch length estimates are very similar between the ExpCM+$\Gamma\omega$ (avg) and the \texttt{phylobayes} mutation-selection model. 
-The equivalence between the two models shows the general importance of site-specific stationary states. 
-The local accuracy of the DMS preferences and the ExpCM can still be seen. 
-The branch length estimates from either focal sequence (H1 and H3) are longer by the ExpCM than by the \texttt{phylobayes} mutation-selection model. 
-This indicates that the DMS preferences are still more accurate for the local branches around the focal sequence than the global preferences estimated by \texttt{phylobayes}.
+The previous sections show that the ExpCM's, with their site-specific stationary states, estimate longer branches than the GY94. 
+However, the ExpCM stationary state has a strong local effect as the amino-acid preferences from deep mutational scanning experiments are most accurate near their focal sequence. 
+There are other models with site-specific stationary states which infer all of the parameters from the phylogenetic data rather than from independent experiments. 
+For example, the CAT model, an amino-acid rather than codon model, describes the stationary state at each site as a mixture model of distinct, finite amino-acid profiles and the mutation-selection model implemented in \texttt{phylobayes} is almost identical to the ExpCM except the amino-acid preferences are inferred in a Bayesian framework rather than defined by deep mutational scanning experiments.
+We compared the ExpCM with the \texttt{phylobayes} mutation-selection model to examine the effect of site-specific stationary states on branch length estimation with and without experimental preferences.
 
-\skhcomment{What is the average difference in branch length between the two models?}
+The results in \ref{fig:phylobayes} show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \texttt{phylobayes} mutation-selection model, which includes $\Gamma\omega$, estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}.
+While the two models estimated the same branch length on average, the tension between local and global accuracy is still apparent in the results. 
+All of the long branches from either the H1 or H3 focal sequence were estimated to be longer by the ExpCM+$\Gamma\omega$ while all other branches were estimated to be longer by the \texttt{phylobayes} mutation-selection model.
+This enrichment of branches from the focal sequence in the branches estimated to be longer by the ExpCM+$\Gamma\omega$(H1+H3 prefs) indicates that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the focal sequence.
+The accuracy of the deep mutational scanning preferences is sacrificed for more a universal stationary state. 
+While the parameters inferred by the \texttt{phylobayes} model are a good compromise, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
 
@@
 \caption{\label{fig:phylobayes}
-\textbf{Comparison of ExpCM and phylobayes}
-\skhcomment{y=x line too faint?}
+\textbf{Comparison of ExpCM and \texttt{phylobayes}}
+We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the mutation-selection model implemented in \texttt{phylobayes}. 
+In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
 }

### 2018-02-26 manuscript/manuscript.tex
tweaks to empirical trrees section

@@
 The results in \ref{fig:empirical_trees} show the distinct effects of modeling variation in $\omega$ and variation in the stationary state. 
-The models with $\Gamma\omega$ rate variation estimated longer branches than the same base model without this variation. 
-For example, in \ref{fig:empirical_trees}, the GY94+$\Gamma\omega$ model estimated longer branches than the GY94 and the ExpCM+$\Gamma
-\omega$(H1 prefs) estimated longer branches than the ExpCM(H1 prefs).
+For two models with the same stationary state, the model which allows $\omega$ to vary estimates longer branches. 
+For example, the GY94+$\Gamma\omega$ estimates longer branches on the HA tree than the GY94 and the ExpCM+$\Gamma\omega$(H1 prefs) estimates longer branches than the ExpCM(H1 prefs).
+In addition, the models with site-specific stationary states estimate longer branches than the models with a uniform stationary state given the same $\omega$ parametrization.  
+Both the ExpCM(H1 prefs) and ExpCM(H3 prefs) estimate longer branches than the GY94 model.
 
-The models with site-specific stationary states, the ExpCM's, estimated longer branches than the GY94 models. 
-Both the ExpCM(H1 prefs) and ExpCM(H3) prefs estimated longer branches than the GY94 model. 
-However, in contrast with the effect of adding $\Gamma\omega$ rate variation, the effect of site-specific stationary defined by deep mutational scanning preferences is not uniform across the entire tree.
-The effect is largest on the branches near the focal sequence of the deep mutational scan (\ref{fig:empirical_trees}). 
-While the increase in length estimated by the ExpCM's has a strong local effect, the increase is in seen even in the presence of $\Gamma\omega$ rate variation. 
-The ExpCM+$\Gamma\omega$ estimated longer branches than the GY94+$\Gamma\omega$. 
-The site-specificity of the ExpCM's stationary state is clearly important for estimating the long branches on the HA tree.
+However, in contrast to the effect of adding $\Gamma\omega$ rate variation, the effect of the site-specific stationary state defined by deep mutational scanning preferences in not uniform across the entire tree.
+The effect is largest on the branches near the focal sequence of the deep mutational scan for both the ExpCM(H1 prefs) and ExpCM (H3 prefs). 
+While the ExpCM's show an increase in branch length with a strong local effect, the increase is seen even in the presence of $\Gamma\omega$ rate variation. 
+The ExpCM+$\Gamma\omega$ estimates the longest branches for all of the models tested. 
+These results indicate that the site-specificity of the ExpCM stationary state is important for long branch underestimation but this stationary state is not constant across the entire HA tree.
 

### 2018-02-26 manuscript/manuscript.tex
small tweaks to model feature section

@@
 Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
-Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
-We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
+When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon} but we will refer to them as GY94 models. 
 The gene-wide $\omega$ is usually $<1$\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
@@
 However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
-This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
+This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
 It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
-Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
-Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
-Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.
+Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} to address this concern as ExpCM's define these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so the parameters do not need to be fit from phylogenetic data.
+The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
+Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation\citep{si2008empirical} are discussed in the last section of the Results.
 

### 2018-02-27 manuscript/manuscript.tex
started to outline materials and methods

@@
 
+\subsection*{Influenza hemagglutinin sequences.}
+We downloaded all full-length protein-coding sequences for each of the 18 influenza HA subtypes. 
+We used \texttt{phydms\_prepalignment} \citep{hilton2017phydms} to filter and align the sequences. 
+We subsampled to five per year plus the required sequences.
+
 \subsection*{Substitution models}

### 2018-02-27 manuscript/manuscript.tex
added Juhye refs to text

@@
 This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
-Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs\jdbcomment{citation, I think Juhye has something she cites that looks at really diverged HAs}.
+Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
 
@@
 Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
-However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\jdbcomment{?}) have a highly conserved structure that performs a highly conserved function\jdbcomment{citation, Juhye can probably help}.
+However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\jdbcomment{?}) have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
 We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).

### 2018-02-27 manuscript/manuscript.tex
added H17/H18 refs to paper

@@
 Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
-However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\jdbcomment{?}) have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
+However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\citep{zhu2013hemagglutinin,hoffmann2016hemagglutinin,sun2013bat}) have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
 We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).

### 2018-02-27 manuscript/manuscript.tex
some edits up to competing effects section

@@
 \renewcommand{\figurename}{{}}
-\renewcommand{\thefigure}{{Figure \arabic{figure}}}
+\renewcommand{\thefigure}{{Figure~\arabic{figure}}}
 
 \renewcommand{\tablename}{{}}
-\renewcommand{\thetable}{{Table \arabic{table}}}
+\renewcommand{\thetable}{{Table~\arabic{table}}}
 
 \newfloat{suppfile}{thp}{losuppfile}
-\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
+\renewcommand{\thesuppfile}{Supplementary~file~\arabic{suppfile}}
 \floatname{suppfile}{}
@@
 \newfloat{suppfig}{thp}{losuppfig}
-\renewcommand{\thesuppfig}{Supplementary figure \arabic{suppfig}}
+\renewcommand{\thesuppfig}{Supplementary~figure~\arabic{suppfig}}
 \floatname{suppfig}{}
@@
 \newfloat{supptable}{thp}{losupptable}
-\renewcommand{\thesupptable}{Supplementary table \arabic{supptable}}
+\renewcommand{\thesupptable}{Supplementary~table~\arabic{supptable}}
 \floatname{supptable}{}
@@
 
-\renewcommand{\theequation}{Equation \arabic{equation}}
-
-\newcommand{\mutDNA}{\textbf{mutDNA}}
-\newcommand{\mutvirus}{\textbf{mutvirus}}
-\newcommand{\DNA}{\textbf{DNA}}
-\newcommand{\virus}{\textbf{virus}}
+\renewcommand{\theequation}{Equation~\arabic{equation}}
 
@@
 For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
-In the case of viruses, there is independent evidence that the inferences from molecular phylogenetics are underestimates of the true deep divergence times.  
+In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
 This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
 Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
-This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
+This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
 The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
-However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
-Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times. 
+However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
+Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how site-specific functional constraints shift over time due to epistasis.
 \end{abstract}
@@
 \caption{\label{fig:model_feature}
-\textbf{Different ways of modeling the site-to-site rate variation of purifying selection.}
+\textbf{Different ways that codon models account for purifying selection.}
 (A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
-In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin. 
+For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin~\citep{yang1994maximum,yang2000codon}. 
 A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
@@
 (C) Substitution models can incorporate neither, one, or both of these features.
-Here we will use substitution models from the \citet{goldman1994codon} (GY94) and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed rates to represent all possible combinations.
-%Green indicates presence of a feature and white indicates absence. 
+Here we will use substitution models from the Goldman-Yang~\citep[GY94;][]{goldman1994codon,yang2000codon} and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed $\omega$ to represent all possible combinations.
 }
@@
 The relative rate of these substitutions is referred to as dN/dS or $\omega$.
-In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
+In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average rate of fixation of nonsynonymous mutations relative to synonymous ones.
 Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
-When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon} but we will refer to them as GY94 models. 
-The gene-wide $\omega$ is usually $<1$\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
+When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon}.
+Here we will refer to M0 Goldman-Yang models simply as GY94 models. 
+The gene-wide $\omega$ is usually $<1$~\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
 
@@
 Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
-For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
+For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing the nucleotide composition of the stationary state.
 However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
@@
 It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
-Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} to address this concern as ExpCM's define these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so the parameters do not need to be fit from phylogenetic data.
+Here we will primarily use experimentally informed codon models (ExpCM's)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
 The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
-Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation\citep{si2008empirical} are discussed in the last section of the Results.
+Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical}\jdbcomment{check that this should be Si Quang and not Quang} are discussed in the last section of the Results.
 
 Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
-Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
+Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter\jdbcomment{cite Haddox et al paper}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
 This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
@@
 The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
-However, all GY94 models including the one with site-specific $\omega_r$ values all approach the same asymptote since they all have the same stationary state.
+However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
 But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
+\jdbcomment{Can you choose something instead of site 348 where the preferences and asymptotic divergence distributions look similar?
+Here they look very different since there are 6 Leu codons and 1 Met codon}
 }
@@
 For instance, when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
-However, as more substitutions accumulate then it becomes progressively more likely for multiple changes to occur at the same site.
-In this regime, the accuracy of the substitution model becomes critical for transforming the sequence identity into branch length.
+However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
+In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
 Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic sequence identity.
 For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
-If the substitution model underestimates the asymptotic sequence divergence then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.
+If the substitution model underestimates the asymptotic sequence identity then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.
 
@@
 The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of reversible substitution models.
-These models are mathematically represented by reversible stochastic matrices, which can be decomposed into stationary states and symmetric exchangeability matrices~\citep{nielsen2006statistical}.
+These models are reversible stochastic matrices, which can be decomposed into stationary states and symmetric exchangeability matrices~\citep{nielsen2006statistical}.
 The stationary state is invariant with respect to multiplication of the symmetric exchangeability matrix by any non-zero number.
-Different schemes for modeling $\omega$ only affect elements of the symmetric exchangeability matrix.
+Different schemes for modeling $\omega$ only multiply elements of the symmetric exchangeability matrix.
 Therefore, no matter how ``well" a model accounts for site-to-site variation in $\omega$, it will always have the same stationary state as a simple GY94 model.
@@
 For instance, \ref{fig:decay} shows that at sites such as 348 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
-When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a very long branch length at modest sequence identities that GY94 model might attribute to a much shorter branch.
+When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a long branch length at modest sequence identities that a GY94 model might attribute to a shorter branch.
 
 \subsection*{Simulations demonstrate how failure to model site-specific stationary states leads to branch-length underestimation.}
+
 \begin{figure}
@@
 \end{figure}
+
 To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
-Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees}) using ExpCM parameterized by the actual amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
+Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees}) using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
 We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
@@
 The GY94 model estimated branches lengths which are only about 60\% of the true values for the longest branches. 
-Accounting for site-to-site variation in $\omega$ did not fix the fundament problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the length of the longest branches.
+Accounting for site-to-site variation in $\omega$ did not fix the fundamental problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the longest branches.
 However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models, which accurately accounted for the site-specific amino-acid preferences used in the simulations.
@@
 \textbf{Effect of site-specific stationary state and $\Gamma\omega$ rate variation on HA branch length estimation.} 
-The HA tree is optimized by models from the ExpCM and GY94 families. 
-The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots. 
-The amino-acid preferences measured by the deep mutational scans defining the ExpCMs were measured in either an H1 background, an H3 background, or were the average measurements from the H1 and H3 scans.
+The branch lengths of the HA tree are optimized using the indicated ExpCM and GY94 models. 
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for \jdbcomment{?
+example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in ?}.
+The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\cite{doud2016accurate}, an H3 HA~\jdbcomment{Lee citation}, or the average of the measurements for these two HAs.
 The circle denotes the H1 clade and the triangle denotes the H3 clade. 
@@
 \caption{\label{tab:empirical_data}
-Performance comparison for the models in~\ref{fig:empirical_trees}. 
+{\bf Fitting of substitution models to the HA phylogenetic tree.}
+The models fit here are the same ones in~\ref{fig:empirical_trees}. 
 All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
-The mean $\omega$ value from each $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation, otherwise simply the fit, gene-wide $\omega$. 
-Each ExpCM has a stringency parameter $>1$.
+The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
+All ExpCM's fit a stringency parameter $>1$.
 } 
@@
         \hline
-         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
+         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
        	ExpCM + $\Gamma\omega$ (H1+H3 avg)  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
@@
 The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM models.
-But how do these models perform on real rather than simulated data?
-Real genes do evolve under site-specific constraint, but this constraint is almost certainly not modeled in all its complexity by an ExpCM.
-However, if ExpCM's do a substantially better job of modeling the constraint, we might still expect them to estimate more accurate branch lengths.
+But how do these models perform on real data?
+Real genes do evolve under functional constraint, but these constraints are almost certainly more complex than what is modeled by an ExpCM.
+However, if ExpCM's do a substantially better job than GY94 models of capturing the true constraints, then we might still expect them to estimate more accurate branch lengths.
 
@@
 
-Given this fixed HA tree topology, we estimated the branch lengths on the tree using each of the substitution models in \ref{fig:model_feature}C. 
-There are two all amino-acid deep mutational scanning datasets for HA~\jdbcomment{citations}, and each parameterizes a different ExpCM.
-One scan measured the site-specific amino-acid preferences in an H1 background (\ref{fig:empirical_trees}, circle) and the other measured the preferences in an H3 background (\ref{fig:empirical_trees}, triangle).
-These two deep mutational scans sample the breadth of HA diversity as their respective focal sequences are only $\sim42\%$ identical on the amino acid level. 
-By comparing these two ExpCMs, we can examine the effect of the experiment's focal sequence on branch length estimation, along with the effect of site-specific stationary state and $\Gamma\omega$ rate variation.
+There are two deep mutational scanning datasets for HA that estimated amino-acid preferences for all sites.
+One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA\jdbcomment{Lee citation}.
+These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
+The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites\jdbcomment{Lee citation, maybe also refer to your own supplementary figures}.
+Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.
+These three sets of HA amino-acid preferences define three different ExpCM's.
 
-The results in \ref{fig:empirical_trees} show the distinct effects of modeling variation in $\omega$ and variation in the stationary state. 
-For two models with the same stationary state, the model which allows $\omega$ to vary estimates longer branches. 
-For example, the GY94+$\Gamma\omega$ estimates longer branches on the HA tree than the GY94 and the ExpCM+$\Gamma\omega$(H1 prefs) estimates longer branches than the ExpCM(H1 prefs).
-In addition, the models with site-specific stationary states estimate longer branches than the models with a uniform stationary state given the same $\omega$ parametrization.  
-Both the ExpCM(H1 prefs) and ExpCM(H3 prefs) estimate longer branches than the GY94 model.
+We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each of these models with $\Gamma\omega$ rate variation.
+\ref{tab:empirical_data} shows that all of the ExpCM's fit the actual data much better than the GY94 model.
+The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
+For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
+\jdbcomment{Probably comment a bit on $\omega$ here: they are typically less than one for all models, but are higher for ExpCM than GY94. Possibly also comment on the stringency parameter.}
 
-However, in contrast to the effect of adding $\Gamma\omega$ rate variation, the effect of the site-specific stationary state defined by deep mutational scanning preferences in not uniform across the entire tree.
-The effect is largest on the branches near the focal sequence of the deep mutational scan for both the ExpCM(H1 prefs) and ExpCM (H3 prefs). 
-While the ExpCM's show an increase in branch length with a strong local effect, the increase is seen even in the presence of $\Gamma\omega$ rate variation. 
-The ExpCM+$\Gamma\omega$ estimates the longest branches for all of the models tested. 
-These results indicate that the site-specificity of the ExpCM stationary state is important for long branch underestimation but this stationary state is not constant across the entire HA tree.
+\ref{fig:empirical_trees} shows that modeling purifying selection via either $\Gamma\omega$ rate variation or a site-specific stationary state increases the estimated branch lengths.
+But important, the effects from these two methods of modeling purifying selection are largely independent.
+For every model (GY94 or any of the three ExpCM's), the $\Gamma\omega$ version estimates longer deep branches.
+In addition, all ExpCM's estimate longer deep branches than the GY94 model, and all ExpCM+$\Gamma\omega$ models estimate longer deep branches than the GY94+$\Gamma\omega$ model.
+
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a fashion that appears roughly uniform across the tree (\ref{fig:empirical_trees}, the branch lengthening from the ExpCM's is not uniform across the tree.
+In particular, the branch lengthening is most pronounced near the sequence of the HA that was used in the deep mutational scanning experiment that parameterized the ExpCM.
+For instance, the ExpCM informed by the H1 data most dramatically lengthens branches near the H1 clade of the tree, while the ExpCM informed by the H3 data has the largest effect on branches near the H3 clade.
+The ExpCM informed by the average of the H1 and H3 data has a more uniform effect, but still most strongly affects branches near either the H1 or H3 clades.
+Therefore, \jdbcomment{something general summarizing the key finding from this section: ExpCM's help, but clearly any given set of preferences from an experiment has an effect strongest near that sequence.}
 
@@
 
-\begin{figure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.85\textwidth]{figures/compete_3}}
@@
 
-\begin{figure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.5\textwidth]{figures/phylobayes.pdf}}
@@
 In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
+\jdbcomment{The diagonal dotted line is too faint}
 }

### 2018-02-27 manuscript/manuscript.tex
more outline to materials and methods

@@
 We subsampled to five per year plus the required sequences. 
+Total of 87 sequences. 
+Inferred tree topology using \texttt{RAxML}.
+
+\subsection*{Simulations.}
+How was the model fit? 
+How many simulations? 
+How was the tree constructed?
+
+\subsection*{Expected Asymptotic Sequence Identity.}
+How were the models fit? 
+What is the equation? 
+What is the equation for the $\omega_r$ term? (does it include the ExpCM $\omega$?)
+How is the GY94$\omega_r$ model constructed? ie How does it interact with the $\omega$ value in the GY94 model?
+
+\subsection*{Deep mutational scanning amino acid preferences.}
+Where did we get the preferences? 
+How were the average preferences calculated? 
+How was the ``hybrid" sequence made? 
+Numbering scheme.
 

### 2018-02-27 manuscript/manuscript.tex
add Juhye reference to paper

@@
 example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in ?}.
-The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\cite{doud2016accurate}, an H3 HA~\jdbcomment{Lee citation}, or the average of the measurements for these two HAs.
+The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\cite{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
 The circle denotes the H1 clade and the triangle denotes the H3 clade. 
@@
 There are two deep mutational scanning datasets for HA that estimated amino-acid preferences for all sites.
-One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA\jdbcomment{Lee citation}.
+One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
 These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
-The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites\jdbcomment{Lee citation, maybe also refer to your own supplementary figures}.
+The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep{lee2018deep}\jdbcomment{maybe also refer to your own supplementary figures}.
 Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.

### 2018-02-27 manuscript/manuscript.tex
changed site choice for decay figure

@@
 The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
-Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical}\jdbcomment{check that this should be Si Quang and not Quang} are discussed in the last section of the Results.
+Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical} are discussed in the last section of the Results.
 
 Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
-Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter\jdbcomment{cite Haddox et al paper}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
+Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
 This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
@@
 But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
-\jdbcomment{Can you choose something instead of site 348 where the preferences and asymptotic divergence distributions look similar?
-Here they look very different since there are 6 Leu codons and 1 Met codon}
+\skhcomment{Currently, this plot is made using the hybridDoud preferences. I should change the numbering to reflect WSN sequential numbering}
 }

### 2018-02-27 manuscript/manuscript.tex
changed decay figure to WSN sequential

@@
 \textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
-The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by \citet{doud2016accurate}.
+The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by~\citet{doud2016accurate}.
 The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
@@
 But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
-\skhcomment{Currently, this plot is made using the hybridDoud preferences. I should change the numbering to reflect WSN sequential numbering}
 }

### 2018-02-27 manuscript/manuscript.tex
fixed reference line in phylobayes figure
add in line reference to supp figures

@@
 The branch lengths of the HA tree are optimized using the indicated ExpCM and GY94 models. 
-The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for \jdbcomment{?
-example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in ?}.
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
 The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\cite{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
@@
 In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
-\jdbcomment{The diagonal dotted line is too faint}
 }
@@
 \centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
-\caption{\label{fig:prefs_average}
+\caption{\label{suppfig:prefs_average}
 \textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.77$)}}

### 2018-02-27 manuscript/manuscript.tex
revisions to all parts of results

@@
 For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
-\jdbcomment{Probably comment a bit on $\omega$ here: they are typically less than one for all models, but are higher for ExpCM than GY94. Possibly also comment on the stringency parameter.}
+\jdbcomment{Probably comment a bit on $\omega$ here: they are typically less than one for all models, but are higher for ExpCM than GY94.
+Possibly also comment on the stringency parameter, aligning with description in the next section.}
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/compete_3}}
+\centerline{\includegraphics[width=0.75\textwidth]{figures/compete_3}}
 \caption{\label{fig:compete}
-\textbf{Effect of overall divergence on the congruency between natural sequences and the deep mutational scanning preferences.} 
-For each deep mutational scanning preference set, we fit an ExpCM to three trees representing high, intermediate, and low overall divergence from the experimental focal sequence. 
-We used the ExpCM stringency parameter ($\beta$) to determine the congruency between the natural sequences and the experimental measurements. 
-A $\beta >1$ indicates that the natural sequences favor the same amino-acid preferences as the experimental measurements but with a greater strength. 
-A $\beta < 1$ indicates the natural sequences do not favor the same amino-acid preferences as the experimental measurements or favor the same amino acids but less strongly. 
-As the overall divergence increased, the $\beta$ value for the ExpCM(H3 prefs) and ExpCM(H1 prefs) decreased, indicating that the preferences do ``shift" over the full HA tree. 
-\skhcomment{1. I need to clean up and do a better job aligning the different parts of this figure but what do you think of the overall layout? 2. \ref{supp:compete_4} shows an alternative layout.}
+\textbf{The congruency between natural selection and the deep mutational scanning measurements decreases with sequence divergence.} 
+We fit an ExpCM informed by the H1 or H3 deep mutational scanning experiments to trees spanning sequences with low, intermediate, and high divergence from the sequence used in the experiment. 
+The ExpCM stringency parameter ($\beta$) is a measure of the congruency between natural selection and the experimental measurements. 
+Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with greater stringency. 
+As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases.
+Therefore, the preferences measured in each experiment are progressively less congruent with those of the HA sequences on the tree as we include increasingly diverged sequences. 
+\jdbcomment{\emph{Maybe} if logo stacks a bit higher, the point is a bit clearer.}
 }
@@
 
-The local effect of the ExpCM is not entirely surprising. 
-Proteins are affected epistatic interactions, where the effect of a mutation at one site is dependent on the amino-acid identity at another site. 
-While one of the strengths of deep mutational scanning is the ability to accurately measure the effect of all single amino-acid mutations, these measurements are done in the context of a single genetic background and do not take into account epistatic interactions.
-The stationary state of the ExpCM's is only as accurate as deep mutational scanning preferences and the results in \ref{fig:empirical_trees} indicate that any given deep mutational scan is most accurate for only a subset of the tree with closely related sequences to the focal sequence.
+The fact that an ExpCM has the greatest effect on branches close to the sequence used in the experiment can be rationalized in terms of existing knowledge about how epistasis shifts protein amino-acid preferences during evolution~\jdbcomment{Stokes shift paper, Shah et al paper, Doud et al 2015, Haddox et al 2018, look at other papers cited about Stokes shift in Haddox et al 2018}.
+In reality, sites in proteins do not have completely conserved site-specific amino-acid preferences. 
+Rather, the effect of a mutation at one site in a protein can depend on the amino-acid identity of other sites in the protein~\jdbcomment{Maybe cite a couple of general papers on epistasis, some good ones in Haddox et al 2018}.
+The deep mutational scanning experiments that inform our ExpCM's are each done in the context of a single HA genetic background, and so do not account for how such epistatic interactions can cause amino-acid preferences to shift as substitutions accumulate.
+Therefore, it makes sense that an ExpCM would most accurately describe the site-specific amino-acid preferences of sequences closely related to the one used in the experiment.
 
-%An ExpCM with a more general stationary state, defined by the average of the H1 and H3 preferences, increases the estimates of branch length from both the H1 and H3 clade (\ref{fig:empirical_trees}).
-
-We examined how the deep mutational scanning preferences shift across the full HA tree by comparing the ExpCM behavior on trees with varying levels of overall divergence from the ExpCM focal sequence.
-For each of the subtrees in \ref{fig:compete}, we asked "how well do the deep mutational scanning preferences agree with the natural sequences?" using the ExpCM stringency parameter $\beta$ (\ref{eq:Frxy}).
+We can observe this phenomenon of shifting amino-acid preferences degrading the accuracy of the ExpCM by fitting trees containing increasingly diverged sequences.
+Specifically, for each HA for which we have deep mutational scanning data, we created three HA phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge ?$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge ?$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
+\ref{fig:compete} shows these sets of HA sequences.
+For each of the subtrees in \ref{fig:compete}, we examined the congruence between site-specific natural selection over the tree and the amino-acid preferences measured in the deep mutational scanning experiment.
+\jdbcomment{Some general description of $\beta$ here.
+I think we'd noted above possibly describing it then?
+If so, just put a short recap here.
+If you decide not to describe it above in any detail, have something longer here.}
 The fit $\beta$ parameter rescales the raw preferences from the deep mutational scan and relates the selection in nature to selection in the lab. 
@@
 Conversely, we interpret a $\beta<1$ as selection in nature favoring different amino acids than selection in lab and that the preferences are not a good description of natural evolution.
-\skhcomment{mention pop-gen derivation and the relationship between lab vs. nature selection and pop size?}
 
-The results in \ref{fig:compete} show that as the overall divergence from the focal sequence of a deep mutational scan increase, the $\beta$ value decreases. 
-This inverse relationship between $\beta$ and overall divergence is seen for both the ExpCM(H1 prefs) and ExpCM(H3 prefs). 
-Along with the local effect in branch length extension seen \ref{fig:empirical_trees}, this indicates that the amino-acid preferences have shifted across the HA tree. 
-This leaves us in a regime with inherent tension. 
+\ref{fig:compete} shows that as the divergence from the sequence used in the deep mutational scan increases, the value of $\beta$ decreases. 
+This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
+Specifically, \jdbcomment{something about how the decreasing $\beta$ flattens the preferences and so loses information from the experiment.}
+The flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
+As shown in \jdbcomment{simulation figure}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
+This fact also helps rationalize why the average of the H1 and H3 experiments does better across the whole tree \jdbcomment{refer to the figure showing that and give some explanation.}
+
+The fact that amino-acid preferences shift as proteins evolve leaves us with an inherent tension: models with site-specific stationary states only become important for accurate branch-length estimation as sequences become increasingly diverged, but this same divergence degrades the accuracy of extrapolating the amino-acid preferences from any given experiment across the phylogenetic tree.
+More importantly, the fact that amino-acid preferences shift over time means that there will not be any model with a single set of site-specific stationary states that accurately describes a phylogenetic tree that covers a wide span of sequences.
+\jdbcomment{Use a slightly more succinct version of what you have below to drive home the point:
+``This leaves us in a regime with inherent tension. 
 Models which take into account the site-specific constraints of purifying selection in their stationary states are important for the accurate estimation of long branches but the site-specific constraints are not static throughout evolution.
-Models with a single stationary state regardless of lineage, site-specific or not, will not be able to capture the shifting functional constraints due to epistatic interactions. 
-\skhcomment{All betas are greater than 1 so the preferences are ``better" than the GY94 model even at high overall divergence.}
+Models with a single stationary state regardless of lineage, site-specific or not, will not be able to capture the shifting functional constraints due to epistatic interactions.''}
 
-\subsection*{\texttt{phylobayes}}
+\subsection*{Models with stationary states estimated from natural sequences give similar results to ExpCM's}
+The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured experiments.
+However, there are other implementations of mutation-selection models that infer the amino-acid preferences from natural sequence data.
+These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of thousands of parameters giving the site-specific stationary states.
+For example, the CAT model\jdbcomment{citation} implements site-specific stationary states for amino-acid substitution models by drawing the stationary state at each site from a mixture model of distinct, finite amino-acid profiles \jdbcomment{wording in preceding sentence might need some tweaking}.
+However, the model most comparable to our ExpCM's is the mutation-selection model implemented in \texttt{phylobayes}, \jdbcomment{and referred to as MutSel.}
+This \jbcomment{MutSel} model is mathematically identical to the ExpCM's except the amino-acid preferences are sampled from a Bayesian mixture models rather than derived from experiments\jdbcomment{The preceding might not be accurate either in the wording about Bayesian mixture models, or because the nucleotide frequency parameters might be different?
+Also might have site-specific $\omega_r$.
+Check these and adjust to be accurate.}
+Like ExpCM's, these MutSel models still assume a single set of site-specific amino-acid preferences for the entire tree.
 
-The previous sections show that the ExpCM's, with their site-specific stationary states, estimate longer branches than the GY94. 
-However, the ExpCM stationary state has a strong local effect as the amino-acid preferences from deep mutational scanning experiments are most accurate near their focal sequence. 
-There are other models with site-specific stationary states which infer all of the parameters from the phylogenetic data rather than from independent experiments. 
-For example, the CAT model, an amino-acid rather than codon model, describes the stationary state at each site as a mixture model of distinct, finite amino-acid profiles and the mutation-selection model implemented in \texttt{phylobayes} is almost identical to the ExpCM except the amino-acid preferences are inferred in a Bayesian framework rather than defined by deep mutational scanning experiments.
-We compared the ExpCM with the \texttt{phylobayes} mutation-selection model to examine the effect of site-specific stationary states on branch length estimation with and without experimental preferences.
-
-The results in \ref{fig:phylobayes} show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \texttt{phylobayes} mutation-selection model, which includes $\Gamma\omega$, estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}.
-While the two models estimated the same branch length on average, the tension between local and global accuracy is still apparent in the results. 
-All of the long branches from either the H1 or H3 focal sequence were estimated to be longer by the ExpCM+$\Gamma\omega$ while all other branches were estimated to be longer by the \texttt{phylobayes} mutation-selection model.
-This enrichment of branches from the focal sequence in the branches estimated to be longer by the ExpCM+$\Gamma\omega$(H1+H3 prefs) indicates that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the focal sequence.
-The accuracy of the deep mutational scanning preferences is sacrificed for more a universal stationary state. 
-While the parameters inferred by the \texttt{phylobayes} model are a good compromise, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
+Comparing ExpCM and MutSel models can therefore help resolve whether the limitations of the ExpCM's are due to the experiments not accurately measuring the amino-acid preferences that best describe site-specific selection across the entire tree, or whether the limitation is simply that there is no such set of preferences\jdbcomment{probably too wordy}.
+The best-performing ExpCM is the one that uses the average of the H1 and H3 deep mutational scanning\jdbcomment{reference why this is true}.
+We therefore compared this ExpCM with the \texttt{phylobayes} mutation-selection model to examine the effect of site-specific stationary states on branch length estimation with and without experimental preferences.
+The results in \ref{fig:phylobayes} show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \jdbcomment{MutSel} model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}. 
+\jdbcomment{Should add note somewhere about how branch lengths were normalized. This might actually be best in the figure legend, not here.}
+While the two models estimated similar branch lengths overall, the tension between local and global accuracy of the amino-acid preferences is still apparent in the results. 
+All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the \jdbcomment{MutSel} model.
+The relatively longer branches from the experimental sequences when using the ExpCM+$\Gamma\omega$ suggests that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the experimental sequences.
+However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the MutSel model appear to be slightly better than the average of the experimental values.
+Therefore, while the site-specific MutSel models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
 
@@
 In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
+\jdbcomment{Is this the $\Gamma\omega$ model? If so, indicate in y-axis}
 }
 \end{figure}
+
+\jdbcomment{Somewhere that makes sense (possibly a few paragraphs above) add more material on correlation of prefs from phylobayes and DMS.
+We want to show how phylobayes prefs are more similar to H1+H3 avg than H1 and H3 are too each other, so maybe show several correlation plots.
+However, avoid correlating H1+H3 with H1 or H3, since this is confounded since H1 and H3 are both part of H1+H3.
+Also, add a supplementary figure showing the phylobayes prefs in the same format as the DMS prefs}
 

### 2018-02-28 manuscript/manuscript.tex
fixed typo in comment so doc compiles

@@
 However, the model most comparable to our ExpCM's is the mutation-selection model implemented in \texttt{phylobayes}, \jdbcomment{and referred to as MutSel.}
-This \jbcomment{MutSel} model is mathematically identical to the ExpCM's except the amino-acid preferences are sampled from a Bayesian mixture models rather than derived from experiments\jdbcomment{The preceding might not be accurate either in the wording about Bayesian mixture models, or because the nucleotide frequency parameters might be different?
+This \jdbcomment{MutSel} model is mathematically identical to the ExpCM's except the amino-acid preferences are sampled from a Bayesian mixture models rather than derived from experiments\jdbcomment{The preceding might not be accurate either in the wording about Bayesian mixture models, or because the nucleotide frequency parameters might be different?
 Also might have site-specific $\omega_r$.

### 2018-02-28 manuscript/manuscript.tex
cleaned up compete figure naming in manuscript

@@
 \begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figures/compete_3}}
+\centerline{\includegraphics[width=0.75\textwidth]{figures/compete}}
 \caption{\label{fig:compete}
@@
 
-\begin{suppfig}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/lee_compete}}
-\caption{\label{supp:lee_compete}
-\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
-\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
-\textbf{(B)} The phylogenetic tree of the "high" alignment. 
-The colors denote the alignment and the black circle denotes the focal H3 sequence. 
-\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
-\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
-Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
-The branch lengths are in average number of codon substitutions per site. 
-}
-\end{suppfig}
-
-\begin{suppfig}[H]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/compete_4}}
-\caption{\label{supp:compete_4}
-}
-\end{suppfig}
-
 \clearpage 

### 2018-02-28 analysis/phylobayes/pymodules/utils.py
first draft of prefs correlation figure

@@ -93,3 +93,30 @@ def calc_distances(treePath):
         df["distance"].append(t.get_distance(seqs[0], seqs[1]))
     df = pd.DataFrame(df)
     return df
+
+
+def melt_df(df, new_name):
+    """
+    The purpose of this function reformat a preferences file from wide form
+    to long form.
+
+    input:
+        df: `pandas dataframe`, preferences in wide form
+        new_name: `str`, column name for values
+    output:
+        `pandas dataframe`: preferences in long form
+    """
+    df = pd.melt(df, id_vars=['site'], value_vars=[x for x in df.columns.values
+                 if x != "site"], var_name='amino_acid', value_name=new_name)
+    return df
+
+
+def rescale_prefs(df, beta):
+    """
+    This function rescales the preferences by raising them to the beta value.
+    """
+    df = df.drop("site", axis=1)
+    df = df.apply(pd.to_numeric)
+    df = df ** beta
+    df["site"] = [x+1 for x in range(len(df))]
+    return df

### 2018-02-28 analysis/phylobayes/pymodules/utils.py
fixed bug in preferences rescaling

@@ -118,5 +118,6 @@ def rescale_prefs(df, beta):
     df = df.drop("site", axis=1)
     df = df.apply(pd.to_numeric)
     df = df ** beta
+    df = df.div(df.sum(axis=1), axis=0)
     df["site"] = [x+1 for x in range(len(df))]
     return df

### 2018-02-28 analysis/decay_to_stationary/stationary_state.py
cleaned up decay to stationary directory

@@ -69,13 +69,13 @@ def get_stationarystate(model,start,length):
 def main():
     # set up params for runs
     start_site = 117
-    number_of_sites = 5
+    number_of_sites = 15
     phydms_dir = "../HA/branch_lengths/phydms/"
     prefs_dir = "../HA/data/references/"
-    YNGKP_M0_modelparams_fname = "{0}WSN_low_0_YNGKP_M0_modelparams.txt".format(phydms_dir)
-    YNGKP_M5_modelaprams_fname = "{0}WSN_low_0_YNGKP_M5_modelparams.txt".format(phydms_dir)
-    ExpCM_modelparams_fname = "{0}WSN_low_0_ExpCM_HA_Doud_prefs_modelparams.txt".format(phydms_dir)
-    Doud_prefs_fname = "{0}HA_Doud_prefs.csv".format(prefs_dir)
+    YNGKP_M0_modelparams_fname = "{0}hybrid_lowH1_0_YNGKP_M0_modelparams.txt".format(phydms_dir)
+    YNGKP_M5_modelaprams_fname = "{0}hybrid_lowH1_0_YNGKP_M5_modelparams.txt".format(phydms_dir)
+    ExpCM_modelparams_fname = "{0}hybrid_lowH1_0_ExpCM_HA_hybridDoud_prefs_modelparams.txt".format(phydms_dir)
+    Doud_prefs_fname = "{0}HA_hybridDoud_prefs.csv".format(prefs_dir)
     if not os.path.isdir("outputs"):
         os.makedirs("outputs")
 

### 2018-02-28 analysis/decay_to_stationary/temp.py

@@ -1,0 +1,170 @@
+
+import pandas as pd
+from phydmslib.constants import *
+import phydmslib.models
+import phydmslib.file_io
+import os
+
+
+# Define the models
+def read_phydms_model_params(param):
+    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    return params
+
+def create_model_ExpCM(param, prefs):
+    """
+    This is the basic `ExpCM` from `phydms`
+
+    `omega` is set to 1.0
+    """
+    params = read_phydms_model_params(param)
+    prefs = phydmslib.file_io.readPrefs(prefs)
+    sites = sorted(prefs.keys())
+    prefs = [prefs[r] for r in sites]
+    params["phiT"] = 1 - sum([params[x] for x in params.keys() if
+                              x.startswith("phi")])
+    phi = scipy.array([params["phi{0}".format(INDEX_TO_NT[x])] for x in
+                       range(N_NT)])
+    return phydmslib.models.ExpCM(prefs, kappa=params["kappa"],
+                                  omega=params["omega"], beta=params["beta"],
+                                  mu=0.3, phi=phi, freeparams=['mu'])
+
+
+# Calculations
+def f_calculation(x, pr, Mt, model_name):
+    """
+    This function calcuates the value of `f` (the y-axis) for some amino-acid
+    `x`, site `r`, and time `t`.
+
+    The inputs for this function are the stationary state subsetted to site `r`
+    and the transition matrix (M) subsetted to site `r` and time `t`.
+
+    If the model is `YNGKP_M5`, then the input is a list of stationary states
+    and transition matrices the length of the number of categories. The `f`
+    value will be the average `f` value for the categories.
+
+    >>> print(f_calculation(8, scipy.array([0.1,0.2,0.3,0.4]), \
+    scipy.array([[1,-100,3,-100],[-100,-100,-100,-100],[9,-100,11,-100],\
+    [-100,-100,-100,-100]]), "ExpCM"))
+    6.4
+    """
+    target_codons = scipy.where(CODON_TO_AA == x)[0]  # amino-acid `x` codons
+    if model_name == "GY94 + Gr":  # need to avg `f` over categories `k`
+        prx = [pr[k][target_codons] for k in range(len(pr))]  # ss
+        # transition to syn. codons
+        Mtxy = [Mt[k][target_codons][:, target_codons] for k in range(len(Mt))]
+        # sum the transition to syn codons
[…]

### 2018-02-28 manuscript/manuscript.tex
updated phylobayes to indicate gammaomega
added supp figure of phylobayes prefs
made multipanel phylobayes figure
small tweaks to manuscript

@@
 These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
-The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep{lee2018deep}\jdbcomment{maybe also refer to your own supplementary figures}.
+The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
+\ref{suppfig:prefs_lee};][]{lee2018deep}.
 Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.
@@
 Therefore, the preferences measured in each experiment are progressively less congruent with those of the HA sequences on the tree as we include increasingly diverged sequences. 
-\jdbcomment{\emph{Maybe} if logo stacks a bit higher, the point is a bit clearer.}
 }
@@
 The flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
-As shown in \jdbcomment{simulation figure}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
+As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
 This fact also helps rationalize why the average of the H1 and H3 experiments does better across the whole tree \jdbcomment{refer to the figure showing that and give some explanation.}
@@
 \subsection*{Models with stationary states estimated from natural sequences give similar results to ExpCM's}
+
+\begin{figure}
+\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
+\caption{\label{fig:phylobayes}
+\textbf{Comparison of ExpCM and \texttt{phylobayes}}
+We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the mutation-selection model implemented in \texttt{phylobayes}. 
+In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
+\skhcomment{I am less sold on using colors on the x or y axis. Is it too confusing with the focal sequence color coding in panel B}
+}
+\end{figure}
+
 The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured experiments.
@@
 These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of thousands of parameters giving the site-specific stationary states.
-For example, the CAT model\jdbcomment{citation} implements site-specific stationary states for amino-acid substitution models by drawing the stationary state at each site from a mixture model of distinct, finite amino-acid profiles \jdbcomment{wording in preceding sentence might need some tweaking}.
+For example, the CAT model~\citep{si2008empirical} implements site-specific stationary states for amino-acid substitution models by drawing the stationary state at each site from a mixture model of distinct, finite amino-acid profiles \jdbcomment{wording in preceding sentence might need some tweaking}.
 However, the model most comparable to our ExpCM's is the mutation-selection model implemented in \texttt{phylobayes}, \jdbcomment{and referred to as MutSel.}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.5\textwidth]{figures/phylobayes.pdf}}
-\caption{\label{fig:phylobayes}
-\textbf{Comparison of ExpCM and \texttt{phylobayes}}
-We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the mutation-selection model implemented in \texttt{phylobayes}. 
-In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
-\jdbcomment{Is this the $\Gamma\omega$ model? If so, indicate in y-axis}
-}
-\end{figure}
-
 \jdbcomment{Somewhere that makes sense (possibly a few paragraphs above) add more material on correlation of prefs from phylobayes and DMS.
 We want to show how phylobayes prefs are more similar to H1+H3 avg than H1 and H3 are too each other, so maybe show several correlation plots.
-However, avoid correlating H1+H3 with H1 or H3, since this is confounded since H1 and H3 are both part of H1+H3.
-Also, add a supplementary figure showing the phylobayes prefs in the same format as the DMS prefs}
+However, avoid correlating H1+H3 with H1 or H3, since this is confounded since H1 and H3 are both part of H1+H3.}
 
@@
 
+\begin{suppfig}[H]
+\centerline{\includegraphics[width=\textwidth]{figures/prefs_mutSel}}
+\caption{\label{suppfig:prefs_mutSel}
+\textbf{Preferences inferred by the mutSel model on the full HA tree.}
+}
+\end{suppfig}
+
 \clearpage 

### 2018-03-01 manuscript/manuscript.tex
cleaned up the empirical trees section

@@
 For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
-\jdbcomment{Probably comment a bit on $\omega$ here: they are typically less than one for all models, but are higher for ExpCM than GY94.
-Possibly also comment on the stringency parameter, aligning with description in the next section.}
+As mentioned in the previous section, $\omega$ is generally fit to values $<1$ in phylogenetic studies~\citep{murrell2015gene} and our results follow this trend. 
+\ref{tab:empirical_data} shows that most models were fit to very small $\omega$ values. 
+However, the ExpCM's always fit $\omega$ values greater than the GY94 model and the ExpCM's with $\Gamma\omega$ all have one $\omega >1$. 
+This increase in $\omega$ values makes sense due to the different interpretation of $\omega$ in each family of models. 
+The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
+The ExpCM's more realistic null model gives it more power to detect diversifying selection, such as the \skhcomment{obvious} immune pressure on HA~\citep{bloom2017identification}.
 
@@
 
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a fashion that appears roughly uniform across the tree (\ref{fig:empirical_trees}, the branch lengthening from the ExpCM's is not uniform across the tree.
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a fashion that appears roughly uniform across the tree (\ref{fig:empirical_trees}), the branch lengthening from the ExpCM's is not uniform across the tree.
 In particular, the branch lengthening is most pronounced near the sequence of the HA that was used in the deep mutational scanning experiment that parameterized the ExpCM.
@@
 The ExpCM informed by the average of the H1 and H3 data has a more uniform effect, but still most strongly affects branches near either the H1 or H3 clades.
-Therefore, \jdbcomment{something general summarizing the key finding from this section: ExpCM's help, but clearly any given set of preferences from an experiment has an effect strongest near that sequence.}
+Therefore, \ref{fig:empirical_trees} shows that the ExpCM's estimate longer branches but that the effect is clearly biased by the given set of preferences used to define the model.
 

### 2018-03-02 manuscript/manuscript.tex
small tweaks to compete section

@@
 We can observe this phenomenon of shifting amino-acid preferences degrading the accuracy of the ExpCM by fitting trees containing increasingly diverged sequences.
-Specifically, for each HA for which we have deep mutational scanning data, we created three HA phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge ?$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge ?$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
+Specifically, for each HA for which we have deep mutational scanning data, we created three HA phylogenetic trees: a ``low'' divergence tree that contains only sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
 \ref{fig:compete} shows these sets of HA sequences.
 For each of the subtrees in \ref{fig:compete}, we examined the congruence between site-specific natural selection over the tree and the amino-acid preferences measured in the deep mutational scanning experiment.
-\jdbcomment{Some general description of $\beta$ here.
-I think we'd noted above possibly describing it then?
-If so, just put a short recap here.
-If you decide not to describe it above in any detail, have something longer here.}
-The fit $\beta$ parameter rescales the raw preferences from the deep mutational scan and relates the selection in nature to selection in the lab. 
-We interpret $\beta >1$ as selection in nature favoring the same amino acids as selection in the experiments, or that the amino-acid preferences are a good model of purifying selection. 
-The larger the $\beta$ more strongly the selection in nature favors these same, experimental preferences. 
-Conversely, we interpret a $\beta<1$ as selection in nature favoring different amino acids than selection in lab and that the preferences are not a good description of natural evolution.
+We used the ExpCM fit parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
+In essence, $\beta$ rescales the deep mutational scanning measurements~(\ref{fig:compete}). 
+Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, or a strong congruence between natural selection and the experimental preferences.
 
@@
 This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
-Specifically, \jdbcomment{something about how the decreasing $\beta$ flattens the preferences and so loses information from the experiment.}
+Specifically, as the $\beta$ value decreases, the preferences ``flatten" and draw less information from the experiment. 
+At the most extreme, $\beta = 0$, the preferences would be perfectly uniform and look identical to the GY94 preferences in \ref{fig:empirical_trees}. 
 The flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
 As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
-This fact also helps rationalize why the average of the H1 and H3 experiments does better across the whole tree \jdbcomment{refer to the figure showing that and give some explanation.}
+This fact also helps rationalize why the ExpCM defined by the average of the H1 and H3 experiments, representing the ``average" constraint on HA, does better across the whole tree (\ref{fig:empirical_trees}).
 
@@
 More importantly, the fact that amino-acid preferences shift over time means that there will not be any model with a single set of site-specific stationary states that accurately describes a phylogenetic tree that covers a wide span of sequences.
-\jdbcomment{Use a slightly more succinct version of what you have below to drive home the point:
-``This leaves us in a regime with inherent tension. 
-Models which take into account the site-specific constraints of purifying selection in their stationary states are important for the accurate estimation of long branches but the site-specific constraints are not static throughout evolution.
-Models with a single stationary state regardless of lineage, site-specific or not, will not be able to capture the shifting functional constraints due to epistatic interactions.''}
 

### 2018-03-03 manuscript/manuscript.tex
tweaks to phylobayes section

@@
 \newcommand\jdbcomment[1]{{\color{red}[#1]}}
+\newcommand\modelName[1]{{\color{magenta}[#1]}}
 
@@
 \caption{\label{fig:phylobayes}
-\textbf{Comparison of ExpCM and \texttt{phylobayes}}
-We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the mutation-selection model implemented in \texttt{phylobayes}. 
-In comparison to ExpCM's, the \texttt{phylobayes} model infers the site-specific amino-acid preferences from the phylogenetic data rather than defining them \textit{a priori} from deep mutational scanning experiments.
+\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
+We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the ExpCM(H1+H3 avg)+$\Gamma\omega$ and the \modelName{MutSel} model implemented in \texttt{phylobayes}. 
+(A) The amino-acid preferences inferred by \modelName{MutSel} are more similar to the average of the H1 and H3 preferences then the H1 and H3 individual preferences sets are to each other, indicating that the average preference set is able to capture the ``global" constraint across the tree.
+(B) The ExpCM and the \modelName{MutSel} model estimated similar branch lengths. 
+However, the branches near the sequences used in the deep mutationally scanning experiments were estimated to be slightly longer by the ExpCM.
+To aid direct comparison between the models, we normalized the branch lengths on each tree by the length between an early and a late seasonal human H1 sequence. 
+See Materials and Methods for more information. 
+\skhcomment{add in which $\beta$ value I used.}
 \skhcomment{I am less sold on using colors on the x or y axis. Is it too confusing with the focal sequence color coding in panel B}
@@
 
-The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured experiments.
-However, there are other implementations of mutation-selection models that infer the amino-acid preferences from natural sequence data.
-These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of thousands of parameters giving the site-specific stationary states.
-For example, the CAT model~\citep{si2008empirical} implements site-specific stationary states for amino-acid substitution models by drawing the stationary state at each site from a mixture model of distinct, finite amino-acid profiles \jdbcomment{wording in preceding sentence might need some tweaking}.
-However, the model most comparable to our ExpCM's is the mutation-selection model implemented in \texttt{phylobayes}, \jdbcomment{and referred to as MutSel.}
-This \jdbcomment{MutSel} model is mathematically identical to the ExpCM's except the amino-acid preferences are sampled from a Bayesian mixture models rather than derived from experiments\jdbcomment{The preceding might not be accurate either in the wording about Bayesian mixture models, or because the nucleotide frequency parameters might be different?
-Also might have site-specific $\omega_r$.
-Check these and adjust to be accurate.}
-Like ExpCM's, these MutSel models still assume a single set of site-specific amino-acid preferences for the entire tree.
+The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
+However, other implementations allow the stationary stationary state of a site-specific model to be inferred directly from the natural sequence data. 
+These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary sates.
+The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 
+The \modelName{MutSel} model is mathematically identical to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
+Like ExpCM's, \modelName{MutSel} models still assume a single set of site-specific amino-acid preferences for the entire tree.
 
-Comparing ExpCM and MutSel models can therefore help resolve whether the limitations of the ExpCM's are due to the experiments not accurately measuring the amino-acid preferences that best describe site-specific selection across the entire tree, or whether the limitation is simply that there is no such set of preferences\jdbcomment{probably too wordy}.
-The best-performing ExpCM is the one that uses the average of the H1 and H3 deep mutational scanning\jdbcomment{reference why this is true}.
-We therefore compared this ExpCM with the \texttt{phylobayes} mutation-selection model to examine the effect of site-specific stationary states on branch length estimation with and without experimental preferences.
-The results in \ref{fig:phylobayes} show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \jdbcomment{MutSel} model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}. 
-\jdbcomment{Should add note somewhere about how branch lengths were normalized. This might actually be best in the figure legend, not here.}
-While the two models estimated similar branch lengths overall, the tension between local and global accuracy of the amino-acid preferences is still apparent in the results. 
-All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the \jdbcomment{MutSel} model.
+Comparing ExpCM's and \modelName{MutSel} models can therefore help to explain the limitations of the ExpCM's. 
+Are the amino-acid preference sets defining the ExpCM's in \ref{fig:empirical_trees} poor descriptors of the site-specific natural selection across the whole tree, or is there simply not a single set that adequately describes this selection?
+If there is a single set, we would expect the \modelName{MutSel} model to outperform the ExpCM because its stationary state is not biased by the specific sequence used in an experiment. 
+The ExpCM that estimates the longest branches is the one that uses the average of the H1 and H3 deep mutational scanning measurements with $\Gamma\omega$ rate variation~(\ref{fig:empirical_trees}).  
+We therefore compared this ExpCM with a \modelName{MutSel} model given the tree in \ref{fig:empirical_trees}.
+
+First, the stationary state inferred by \modelName{MutSel} is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
+As \ref{fig:phylobayes}A shows, the site-specific amino-acid preferences estimated by \modelName{MutSel} are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
+This strong correlation between the \modelName{MutSel} and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.
+
+Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \modelName{MutSel} model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}.
+While overall the two models estimated similar lengths, the tension between local and global accuracy of the amino-acid preferences is still apparent in these results. 
+All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the \modelName{MutSel} model.
 The relatively longer branches from the experimental sequences when using the ExpCM+$\Gamma\omega$ suggests that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the experimental sequences.
-However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the MutSel model appear to be slightly better than the average of the experimental values.
-Therefore, while the site-specific MutSel models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
-
-\jdbcomment{Somewhere that makes sense (possibly a few paragraphs above) add more material on correlation of prefs from phylobayes and DMS.
-We want to show how phylobayes prefs are more similar to H1+H3 avg than H1 and H3 are too each other, so maybe show several correlation plots.
-However, avoid correlating H1+H3 with H1 or H3, since this is confounded since H1 and H3 are both part of H1+H3.}
+However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the \modelName{MutSel} model appear to be slightly better than the average of the experimental values.
+Therefore, while the site-specific \modelName{MutSel} models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
 
@@
   \item We don't allow any of the models to vary by lineage. 
+  \item discussion in \citet{rodrigue2010mutation}
 \end{enumerate}

### 2018-03-05 manuscript/manuscript.tex
ported Jesse's comments from pdf to latex document

@@
 
-\title{Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages}
+\title{Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.}
 
@@
 Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
+\skhcomment{I wonder if we should change this second sentence, because I'm not sure how many of the other cases for which there are actually other methods outside of viruses?Maybe it would be better to say: "For highly diverged sequences, such phylogenetic techniques often estimate surprisingly recent divergence times."Then the next sentence (as already written) talks about how for viruses there are other methods that show these are too recent.}
 For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
@@
 However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
-Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how site-specific functional constraints shift over time due to epistasis.
+Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time due to epistasis.
 \end{abstract}
@@
 A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
-(B) A substitution model stationary state defines the expected sequence composition after a very long evolutionary time. 
+(B) A substitution model's stationary state defines the expected sequence composition after a very long evolutionary time. 
 Most substitution models have stationary states that are uniform across sites.
@@
 It is also not homogenous among the different amino acids at a given site.
-For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
+For instance, some protein sites strongly prefer hydrophobic amino acids, others are constrained to just one or a few amino acids, and yet others tolerate many amino acids.
 In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.
@@
 Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
-Codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
+Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
 The relative rate of these substitutions is referred to as dN/dS or $\omega$.
@@
 This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
-These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
+These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and usually better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
 
@@
 The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
-Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical} are discussed in the last section of the Results.
+An alternative strategy of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} estimation is discussed in the last section of the Results.
 
 Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
-Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
+Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, \skhcomment{I realized this citation is wrong.
+For the fact that ExpCM and mutation-selection models can still incorporate omega, we should cite the relevant Rodrigue / Lartillot paper and Bloom (2017).The Haddox et al (2017) should be cited for the fact that omega can still be drawn from a gamma.
+At that point, we should also cite relevant Rodrigue / Lartillot paper for their approach to do this.} which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
 This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
@@
 \textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
-The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by~\citet{doud2016accurate}.
+The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as experimentally measured by~\citet{doud2016accurate}.
 The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
-For the GY94 model, the graphs are identical for all sites since this models does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
+For the GY94 model, the graphs are identical for all sites since this model does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
 The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
@@
 But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
+\skhcomment{I still feel like some of the fonts here are too small. Can you make the small ones up the same size as the legend titles on the axes.}
 }
@@
 The transformation from sequence identity to branch length is trivial when the sequence identity is high.
-For instance, when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
+For instance, when there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
 However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
 In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
-Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic sequence identity.
+Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some amino-acid asymptotic sequence identity.
 For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
@@
 This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
-Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
+Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
 
@@
 \textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
-We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences measured by deep mutational scanning of an H1 HA~\citep{doud2016accurate}.
+We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
 We then inferred the branch lengths of this tree on the simulated alignments.
@@
 
-The models with a uniform stationary state underestimated the lengths of long branches separating the simulated sequences (\ref{fig:simulations}). 
-The GY94 model estimated branches lengths which are only about 60\% of the true values for the longest branches. 
+The models with a uniform stationary state underestimated the lengths of long branches on the phylogenetic of the simulated sequences (\ref{fig:simulations}). 
+The GY94 model estimated branches lengths that are only about 60\% of the true values for the longest branches. 
 Accounting for site-to-site variation in $\omega$ did not fix the fundamental problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the longest branches.
 However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models, which accurately accounted for the site-specific amino-acid preferences used in the simulations.
-The improved performance of the ExpCM's is due to their site-specific modeling of the stationary state: if we parameterize these models by site-specific amino-acid preferences that have been averaged across HA sites, then they perform no better than GY94 models.
+The improved performance of the ExpCM's is due to their site-specific modeling of the stationary state: if we parameterize these models by site-specific amino-acid preferences that have been averaged across HA sites, then they perform no better than GY94 models~\ref{fig:simulations}.
 Therefore, models with uniform stationary states are fundamentally incapable of accurately estimating the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.
@@
 The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
-The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\cite{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
+The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
 The circle denotes the H1 clade and the triangle denotes the H3 clade. 
@@
         \hline
-         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
+         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & $\omega$ & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
        	ExpCM + $\Gamma\omega$ (H1+H3 avg)  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
@@
 Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
-However, HA proteins from all subtypes (with the exception of an unusual subtype of bat influenza virus that we exclude from this analysis~\citep{zhu2013hemagglutinin,hoffmann2016hemagglutinin,sun2013bat}) have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
-We used \texttt{RAxML} with nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
+However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
+We used \texttt{RAxML} with a nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
 For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
-Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty probably lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
+Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
 
@@
 These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
+\skhcomment{figure out how to put the lee paper first and then the supp figures}
 The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
@@
 
-We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each of these models with $\Gamma\omega$ rate variation.
-\ref{tab:empirical_data} shows that all of the ExpCM's fit the actual data much better than the GY94 model.
+We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
+\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
 The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
 For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
-As mentioned in the previous section, $\omega$ is generally fit to values $<1$ in phylogenetic studies~\citep{murrell2015gene} and our results follow this trend. 
-\ref{tab:empirical_data} shows that most models were fit to very small $\omega$ values. 
-However, the ExpCM's always fit $\omega$ values greater than the GY94 model and the ExpCM's with $\Gamma\omega$ all have one $\omega >1$. 
-This increase in $\omega$ values makes sense due to the different interpretation of $\omega$ in each family of models. 
+As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
+\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
+However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
+Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
+This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
 The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
-The ExpCM's more realistic null model gives it more power to detect diversifying selection, such as the \skhcomment{obvious} immune pressure on HA~\citep{bloom2017identification}.
+This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification}\skhcomment{rodrigue MBE}, for amino-acid change, which occurs at some sites in HA due to immune selection.
 
+\skhcomment{I don't like how this paragraph comes almost as an afterthought from the next one given that branch lengths are actually are main point.
+I wonder if we can add a stronger first sentence to draw more emphasis to this paragraph.
+I also feel like a sentence between the current first and second one would be good.
+It could comment a bit more on the specifics of the results in Figure 4 in a way that draws out the point better.}
 \ref{fig:empirical_trees} shows that modeling purifying selection via either $\Gamma\omega$ rate variation or a site-specific stationary state increases the estimated branch lengths.
-But important, the effects from these two methods of modeling purifying selection are largely independent.
+But importantly, the effects from these two methods of modeling purifying selection are largely independent.
 For every model (GY94 or any of the three ExpCM's), the $\Gamma\omega$ version estimates longer deep branches.
@@
 The ExpCM informed by the average of the H1 and H3 data has a more uniform effect, but still most strongly affects branches near either the H1 or H3 clades.
-Therefore, \ref{fig:empirical_trees} shows that the ExpCM's estimate longer branches but that the effect is clearly biased by the given set of preferences used to define the model.
+Therefore, \ref{fig:empirical_trees} shows that the ExpCM's estimate longer branches, but that the effect is biased by the set of preferences used to define the stationary state of the model.
 
@@
 Rather, the effect of a mutation at one site in a protein can depend on the amino-acid identity of other sites in the protein~\jdbcomment{Maybe cite a couple of general papers on epistasis, some good ones in Haddox et al 2018}.
-The deep mutational scanning experiments that inform our ExpCM's are each done in the context of a single HA genetic background, and so do not account for how such epistatic interactions can cause amino-acid preferences to shift as substitutions accumulate.
-Therefore, it makes sense that an ExpCM would most accurately describe the site-specific amino-acid preferences of sequences closely related to the one used in the experiment.
+The deep mutational scanning experiments that inform our ExpCM's were performed in the context of a single HA genetic background, and so do not account for how such epistatic interactions can cause amino-acid preferences to shift as substitutions accumulate.
+Therefore, it makes sense that an ExpCM would most accurately describe the evolution of sequences closely related to the one used in the experiment.
 
-We can observe this phenomenon of shifting amino-acid preferences degrading the accuracy of the ExpCM by fitting trees containing increasingly diverged sequences.
-Specifically, for each HA for which we have deep mutational scanning data, we created three HA phylogenetic trees: a ``low'' divergence tree that contains only sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
-\ref{fig:compete} shows these sets of HA sequences.
-For each of the subtrees in \ref{fig:compete}, we examined the congruence between site-specific natural selection over the tree and the amino-acid preferences measured in the deep mutational scanning experiment.
-We used the ExpCM fit parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
+We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to containing increasingly diverged sequences.
+Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
+\ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
+For each subtree, we examined the congruence between site-specific natural selection and the amino-acid preferences measured in the deep mutational scanning experiment. 
+We used the ExpCM stringency parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
 In essence, $\beta$ rescales the deep mutational scanning measurements~(\ref{fig:compete}). 
-Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, or a strong congruence between natural selection and the experimental preferences.
+Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, indicating a strong congruence between natural selection and the experimental preferences.
 
@@
 At the most extreme, $\beta = 0$, the preferences would be perfectly uniform and look identical to the GY94 preferences in \ref{fig:empirical_trees}.
-The flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
-As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
+In reality, $\beta$ never reaches a value this low, indicating the deep mutational scanning experiments remain somewhat informative about the real purifying selection across the entire swath of HAs. 
+However, the flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
+As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate.
 This fact also helps rationalize why the ExpCM defined by the average of the H1 and H3 experiments, representing the ``average" constraint on HA, does better across the whole tree (\ref{fig:empirical_trees}).
@@
 \skhcomment{I am less sold on using colors on the x or y axis. Is it too confusing with the focal sequence color coding in panel B}
+\skhcomment{Order in figure legend doesn't make sense.
+The first data in the figure is about prefs, not branch lengths, so describe that first.
+Also, either arrange figure to compare MutSel and avg first, or re-arrange legend to describe them in the order they are presented.
+I don't know if you need to add exact beta value?
+Just say re-scaled by stringency parameter in Table X.
+I don't feel strongly about coloring.}
 }
@@
 The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
-However, other implementations allow the stationary stationary state of a site-specific model to be inferred directly from the natural sequence data. 
-These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary sates.
+However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
+These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states\skhcomment{overfitting problems associated with trying to make}.
 The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 
-The \modelName{MutSel} model is mathematically identical to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
+The \modelName{MutSel} model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
 Like ExpCM's, \modelName{MutSel} models still assume a single set of site-specific amino-acid preferences for the entire tree.
 
-Comparing ExpCM's and \modelName{MutSel} models can therefore help to explain the limitations of the ExpCM's. 
-Are the amino-acid preference sets defining the ExpCM's in \ref{fig:empirical_trees} poor descriptors of the site-specific natural selection across the whole tree, or is there simply not a single set that adequately describes this selection?
-If there is a single set, we would expect the \modelName{MutSel} model to outperform the ExpCM because its stationary state is not biased by the specific sequence used in an experiment. 
+Comparing ExpCM's and \modelName{MutSel} models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree.
+If the limits of the ExpCM's reflected in Figure X are simply because the deep mutational scanning do not accurately measure the "true" amino-acid preferences of HA across the tree, then we would expect the \modelName{MutSel} models (which derive their stationary states from the entire tree) to perform better.
+On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect \modelName{MutSel} and ExpCM models to perform similarly.
+
 The ExpCM that estimates the longest branches is the one that uses the average of the H1 and H3 deep mutational scanning measurements with $\Gamma\omega$ rate variation~(\ref{fig:empirical_trees}).  
 We therefore compared this ExpCM with a \modelName{MutSel} model given the tree in \ref{fig:empirical_trees}. 
-
 First, the stationary state inferred by \modelName{MutSel} is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 

### 2018-03-05 manuscript/manuscript.tex
small tweaks to results based on Jesse's comments

@@
 Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
-\skhcomment{I wonder if we should change this second sentence, because I'm not sure how many of the other cases for which there are actually other methods outside of viruses?Maybe it would be better to say: "For highly diverged sequences, such phylogenetic techniques often estimate surprisingly recent divergence times."Then the next sentence (as already written) talks about how for viruses there are other methods that show these are too recent.}
-For highly diverged sequences, such phylogenetic techniques often estimate substantially more recent times than other methods.
+For highly diverged sequences, such phylogenetic techniques of- ten estimate surprisingly recent divergence times. 
 In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
@@
 Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
-Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, \skhcomment{I realized this citation is wrong.
-For the fact that ExpCM and mutation-selection models can still incorporate omega, we should cite the relevant Rodrigue / Lartillot paper and Bloom (2017).The Haddox et al (2017) should be cited for the fact that omega can still be drawn from a gamma.
-At that point, we should also cite relevant Rodrigue / Lartillot paper for their approach to do this.} which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
-This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
+Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{rodrigue2016detecting, bloom2017identification}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
+This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models~\citep{haddox2017mapping, rodrigue2014site}. 
 We will denote such models as ExpCM+$\Gamma\omega$.
@@
 The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
-This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification}\skhcomment{rodrigue MBE}, for amino-acid change, which occurs at some sites in HA due to immune selection.
+This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, odrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.
 
@@
 See Materials and Methods for more information. 
-\skhcomment{add in which $\beta$ value I used.}
-\skhcomment{I am less sold on using colors on the x or y axis. Is it too confusing with the focal sequence color coding in panel B}
 \skhcomment{Order in figure legend doesn't make sense.
@@
 However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
-These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states\skhcomment{overfitting problems associated with trying to make}.
+These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
 The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 

### 2018-03-05 manuscript/manuscript.tex
updated beta values in prefs logoplots

@@
 \caption{\label{suppfig:prefs_doud}
-\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.19$)} 
-\skhcomment{I need to change the $\beta$ value when the new \texttt{phydms} results finish running.}
+\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.12$)} 
 }
@@
 \caption{\label{suppfig:prefs_lee}
-\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)}
+\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.28$)}
  }
@@
 \caption{\label{suppfig:prefs_average}
-\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.77$)}}
+\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.70$)}}
 \end{suppfig}

### 2018-03-05 manuscript/manuscript.tex
tweaks to phylobayes figre

@@
 \textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
-We estimated the branch lengths of the HA tree in \ref{fig:empirical_trees} using the ExpCM(H1+H3 avg)+$\Gamma\omega$ and the \modelName{MutSel} model implemented in \texttt{phylobayes}. 
-(A) The amino-acid preferences inferred by \modelName{MutSel} are more similar to the average of the H1 and H3 preferences then the H1 and H3 individual preferences sets are to each other, indicating that the average preference set is able to capture the ``global" constraint across the tree.
-(B) The ExpCM and the \modelName{MutSel} model estimated similar branch lengths. 
-However, the branches near the sequences used in the deep mutationally scanning experiments were estimated to be slightly longer by the ExpCM.
-To aid direct comparison between the models, we normalized the branch lengths on each tree by the length between an early and a late seasonal human H1 sequence. 
-See Materials and Methods for more information. 
-\skhcomment{Order in figure legend doesn't make sense.
-The first data in the figure is about prefs, not branch lengths, so describe that first.
-Also, either arrange figure to compare MutSel and avg first, or re-arrange legend to describe them in the order they are presented.
-I don't know if you need to add exact beta value?
-Just say re-scaled by stringency parameter in Table X.
-I don't feel strongly about coloring.}
+(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a \modelName{MutSel} to the tree in \ref{fig:empirical_trees}. 
+The \modelName{MutSel} amino-acid preferences are inferred from the natural sequences while the ExpCM amino-acid preferences are experimentally measured and then rescaled using the stringency parameter in \ref{tab:empirical_data}.
+The \modelName{MutSel} preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual experiments are to each other (Pearson's r: 0.74 vs.
+0.52).
+(B) The \modelName{MutSel} and the ExpCM(H1+H3 avg)+$\Gamma\omega$ estimated similar branch lengths. 
+To aid direct comparison between the models, we normalized the branch lengths by the length between and early and late seasonal human H1 sequence. 
 }

### 2018-03-05 manuscript/manuscript.tex
outline simulations materials and methods

@@
 But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
-\skhcomment{I still feel like some of the fonts here are too small. Can you make the small ones up the same size as the legend titles on the axes.}
 }
@@
 
+\subsection*{Decay to stationary state}
+
+I built the models using the results for the hybrid\_lowH1 tree. 
+I used YNKGP M0, YNGKP M5, and the ExpCM with the hybrid doud preferences. 
+The $\omega_r$ values were calculated without the fit $\omega$ from the ExpCM. 
+When making the YNGKP M0 + $\omega_r$ the $\omega$ value is set to spielman\_wr * $\omega$ from the fit YNGKP M0. 
+The preferences were rescaled with the same $\beta$ as the ExpCM.
+
+\subsubsection*{Spielman $\omega_{r}$ values inferred from the ExpCM} 
+We inferred the average nonsynonymous fixation rate from the ExpCM following~\citet{spielman2015relationship} as 
+\begin{equation}
+\label{eq:w_r}
+\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times P_{r,xy}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
+\end{equation}
+where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.
+
+\subsubsection*{Expected pairwise amino-acid identity}
+\textit{Do I need to talk about the branchScale scaling I used?}
+The expected pairwise amino-acid identity at a site $r$ over time $t$ for a given model is 
+\begin{equation}
+\label{eq:f}
+\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
+\end{equation}
+where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.
+
+\subsection*{Simulations}
+
+Used \texttt{pyvolve}. 
+The tree was the large tree, specifically high hybrid with doud prefs 
+total of 10 replicates 
+Model was the hybrid lowH1 with hybrid doud prefs
+
 \subsection*{Influenza hemagglutinin sequences.}
@@
 
-\subsubsection*{Spielman $\omega_{r}$ values inferred from the ExpCM} 
-We inferred the average nonsynonymous fixation rate from the ExpCM following~\citet{spielman2015relationship} as 
-\begin{equation}
-\label{eq:w_r}
-\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times P_{r,xy}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
-\end{equation}
-where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.
-
-\subsubsection*{Expected pairwise amino-acid identity}
-\textit{Do I need to talk about the branchScale scaling I used?}
-The expected pairwise amino-acid identity at a site $r$ over time $t$ for a given model is 
-\begin{equation}
-\label{eq:f}
-\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
-\end{equation}
-where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.
-
 \newpage

### 2018-03-06 manuscript/manuscript.tex
first pass at materials and methods

--- 2018-03-05 manuscript/manuscript.tex
@@
 We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to containing increasingly diverged sequences.
-Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
+Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge 59\%$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge 46\%$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
 \ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
@@
 
-\newpage
 \section*{Materials and Methods}
 
-\subsection*{Influenza hemagglutinin sequences.}
-We downloaded all full-length protein-coding sequences for each of the 18 influenza HA subtypes. 
-We used \texttt{phydms\_prepalignment} \citep{hilton2017phydms} to filter and align the sequences. 
-We subsampled to five per year plus the required sequences. 
-Total of 87 sequences. 
-Inferred tree topology using \texttt{RAxML}.
+\subsection*{Models}
 
-\subsection*{Simulations.}
-How was the model fit? 
-How many simulations? 
-How was the tree constructed?
+\subsubsection*{Experimentally Informed Codon Model (ExpCM)}
 
-\subsection*{Expected Asymptotic Sequence Identity.}
-How were the models fit? 
-What is the equation? 
-What is the equation for the $\omega_r$ term? (does it include the ExpCM $\omega$?)
-How is the GY94$\omega_r$ model constructed? ie How does it interact with the $\omega$ value in the GY94 model?
-
-\subsection*{Deep mutational scanning amino acid preferences.}
-Where did we get the preferences? 
-How were the average preferences calculated? 
-How was the ``hybrid" sequence made? 
-Numbering scheme.
-
-\subsection*{Substitution models}
-\subsubsection*{GY94 models}
-\subsubsection*{ExpCMs}
-We recap the \textbf{Exp}erimentally Informed \textbf{C}odon \textbf{M}odel (ExpCM) \citep{bloom2014experimentally,bloom2014informed,bloom2017identification,hilton2017phydms} to introduce nomenclature.
+The ExpCM is described in detail in \citet{hilton2017phydms}.
 
@@
 where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
-The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form:
-\begin{equation}
-Q_{xy} = 
-\begin{cases}
-\phi_w & \mbox{if $x$ and $y$ differ by a transversion to nucleotide $w$} \\
-\kappa \phi_w & \mbox{if $x$ and $y$ differ by a transition to nucleotide $w$} \\
-0 & \mbox{if $x$ and $y$ differ by $>1$ nucleotide.}
-\end{cases}
-\end{equation}
-The $\kappa$ parameter represents the transition-transversion ratio, and the $\phi_w$ values give the expected frequency of nucleotide $w$ in the absence of selection on amino-acid substitutions, and are constrained by $1 = \sum_w \phi_w$.
+The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form.
 
@@
 
-The ExpCM stationary state frequency $p_{r,x}$ of codon $x$ at site $r$ is~\citep{bloom2017identification} 
+\subsubsection*{GY94}
+
+We consider the basic Goldman-Yang style (GY94) substitution model defined as M0 in \citet{yang2000codon}. 
+$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by
+
 \begin{equation}
-\label{eq:p_rx}
-p_{r,x} = \frac{\left(\pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta} \phi_{x_0} \phi_{x_1} \phi_{x_2}}{\sum_z \left(\pi_{r,\mathcal{A}\left(z\right)}\right)^{\beta} \phi_{z_0} \phi_{z_1} \phi_{z_2}},
+\label{eq:GY94}
+P_xy = 
+\begin{cases}
+  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
+  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
+  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
+  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
+  \end{cases}
 \end{equation}
-\subsection*{Theoretical effect of model choice on branch length}
-\subsection*{Effect of model choice on natural sequences}
+where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
+codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.
 
-\subsubsection*{ExpCM + $\Gamma\omega$ and YNGKP M5}
-
-\subsubsection*{Spielman $\omega_{r}$ values inferred from the ExpCM} 
-We inferred the average nonsynonymous fixation rate from the ExpCM following~\citet{spielman2015relationship} as 
+\subsubsection*{GY94 with $\omega_r$}
+Following~\citet{spielman2015relationship}, we inferred an $\omega$ for each site in the protein ($\omega_r$) from an ExpCM. 
 \begin{equation}
 \label{eq:w_r}
-\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times P_{r,xy}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
+\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times \frac{P_{r,xy}}{\omega}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
 \end{equation}
@@
 
-\subsubsection*{Expected pairwise amino-acid identity}
-\textit{Do I need to talk about the branchScale scaling I used?}
-The expected pairwise amino-acid identity at a site $r$ over time $t$ for a given model is 
+We replaced the $\omega$ value in \ref{eq:GY94} with $\omega \times \omega_r$, resulting in a site-specific model with the GY94 stationary state.
+
+\subsection*{Adding $\Gamma\omega$}
+
+For the GY94+$\Gamma\omega$ and ExpCM+$\Gamma\omega$, $\omega$ is drawn from $K=4$ discrete categories and equal weight is given to each category. 
+Set to the mean of the discretized $\Gamma$-distribution.
+
+\subsection*{Deep Mutational Scanning Amino-Acid preferences.}
+We used deep mutational scanning amino-acid preferences from an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the ExpCM's. 
+We only considered homologous sites between the H1 and H3 protein (described in \ref{suppfile:WSN_Perth_map}). 
+The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.
+
+\subsection*{Influenza virus hemagglutinin (HA) sequences and tree topology.}
+
+We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18. 
+We filtered and aligned the sequences using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}. 
+Specifically, we used \texttt{phydms\_comprehensive} default parameters except \texttt{--minidentity 0.3} and we removed sites which are not homologous between the H1 and H3 deep mutational references sequences using the sequence in \ref{suppfile:hybrid_sequence}.
+We subsampled the sequences so that there were 5 sequences per subtype with $\le 1$ sequence per year per subtype. 
+%Finally, we ensured the ``landmark" sequences from the H1, H2, and H3 subtypes (\ref{suppfile:required_seqs}) were included. 
+The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.
+
+We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence. 
+The ``low divergence" alignments had   $\ge 59\%$ amino-acid identity from the reference sequence and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence.
+The resulting alignments are found in \ref{suppfile:alignment_lowH1}, \ref{suppfile:alignment_lowH3}, \ref{suppfile:alignment_intermediateH1}, and \ref{suppfile:alignment_intermediateH3}.
+
+We inferred the tree topology of each alignment using \texttt{RAxML}~\citep{stamatakis2006raxml} and the GTRCAT model. 
+We estimated the branch lengths of given this fixed topology for the ExpCM and GY94 models using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
+
+\subsection*{Asymptotic sequence divergence.}
+
+We built ExpCM and GY94 models using parameters fit to the ``low divergence from H1" tree.  
+Specifically, we built a GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, a GY94+$\Gamma\omega$ model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94+Gr}, an ExpCM using the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}, and a GY94+$\omega_r$ using the GY94 parameters in \ref{suppfile:hybrid_lowH1_GY94} and the ExpCM defined by\ref{suppfile:hybrid_lowH1_ExpCMdoud} and \ref{suppfile:H1_prefs}.
+
+For each model we calculated the expected sequence divergence given some branch length $t$ by 
 \begin{equation}
@@
 
+\subsection*{Simulations}
+
+We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along tree for the full alignment (\ref{fig:empirical_trees};\ref{suppfile:alignment_high}) using an ExpCM defined by the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}.
+We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}.
+
 \newpage
 \section*{Supplemental Information}
-
-\subsection*{Model Parameters for the simulations}
 
@@
 \caption{\label{suppfig:prefs_doud}
-\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.12$)} 
+\textbf{H1 preferences measured by \cite{doud2016accurate} (\ref{suppfile:H1_prefs}) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.12$)}. 
 }
@@
 \caption{\label{suppfig:prefs_lee}
-\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.28$)}
+\textbf{H3 preferences measured by~\citet{lee2018deep} (\ref{suppfile:H1_prefs}) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.28$)}
  }
@@
 \caption{\label{suppfig:prefs_average}
-\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.70$)}}
+\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by~\citet{lee2018deep} ((\ref{suppfile:avg_prefs})) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.70$)}}
 \end{suppfig}
@@
 
+\begin{suppfig}[H]
+\centerline{\includegraphics[width=0.50\textwidth]{figures/divergence_distances.pdf}}
+\caption{\label{suppfig:subalignments}
+\textbf{Subalignments.}
+}
+\end{suppfig}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:hybrid_sequence}
+The ``hybrid" sequence used to align the HA sequences.
+Specifically, this sequence is the H1 deep mutational scanning reference sequence A/Wilson Smith/1933 except it only contains the codon positions shared between A/Wilson Smith/1933 and the H3 deep mutational scanning reference sequence A/Perth/2009.
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:required_seqs}
+These sequences from the H1, H2, and H3 subtypes were required to be in the final alignment. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:alignment_high}
+HA sequences used in \ref{fig:empirical_trees}.  
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:alignment_lowH1}
+HA sequences for the ``low divergence from H1" tree in \ref{fig:compete}.  
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:alignment_lowH3}
+HA sequences for the ``low divergence from H3" tree in \ref{fig:compete}.  
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:alignment_intermediateH1}
+HA sequences for the ``low divergence from H1" tree in \ref{fig:compete}.  
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:alignment_intermediateH3}
+HA sequences for the ``low divergence from H3" tree in \ref{fig:compete}.  
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:WSN_Perth_map}
+Homologous sites in WSN and Perth using sequential numbering starting at the second (first non-M) codon. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:H1_prefs}
+Deep mutational scanning preferences from \citet{doud2016accurate} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:H3_prefs}
+Deep mutational scanning preferences from \citet{lee2018deep} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:avg_prefs}
+The average deep mutational scanning preferences from \citet{doud2016accurate} ad \citet{lee2018deep}. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:hybrid_lowH1_ExpCMdoud}
+ExpCM model parameters fit using the amino-acid preferences from~\citet{doud2016accurate}(\ref{suppfile:H1_prefs}) and the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1})
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:hybrid_lowH1_GY94}
+GY94 model parameters fit using the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1}). 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:hybrid_lowH1_GY94+Gr}
+GY94+$\Gamma\omega$ model parameters fit using the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1}). 
+}
+\end{suppfile}
+
 \clearpage 

### 2018-03-07 manuscript/manuscript.tex
second pass at materials and methods

@@
 The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
-This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, odrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.
+This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, rodrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.
 
@@
 \textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
-(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a \modelName{MutSel} to the tree in \ref{fig:empirical_trees}. 
-The \modelName{MutSel} amino-acid preferences are inferred from the natural sequences while the ExpCM amino-acid preferences are experimentally measured and then rescaled using the stringency parameter in \ref{tab:empirical_data}.
-The \modelName{MutSel} preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual experiments are to each other (Pearson's r: 0.74 vs.
-0.52).
-(B) The \modelName{MutSel} and the ExpCM(H1+H3 avg)+$\Gamma\omega$ estimated similar branch lengths. 
+(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the tree in \ref{fig:empirical_trees}. 
+The pbMutSel amino-acid preferences are inferred from the natural sequences while the ExpCM amino-acid preferences are experimentally measured and then rescaled using the stringency parameter in \ref{tab:empirical_data}.
+The pbMutSel preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual experiments are to each other (Pearson's r: 0.74 vs. 0.52). 
+(B) The pbMutSel and the ExpCM(H1+H3 avg)+$\Gamma\omega$ estimated similar branch lengths. 
 To aid direct comparison between the models, we normalized the branch lengths by the length between and early and late seasonal human H1 sequence. 
@@
 These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
-The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as \modelName{MutSel}. 
-The \modelName{MutSel} model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
-Like ExpCM's, \modelName{MutSel} models still assume a single set of site-specific amino-acid preferences for the entire tree.
+The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as pbMutSel. 
+The pbMutSel model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
+Like ExpCM's, pbMutSel models still assume a single set of site-specific amino-acid preferences for the entire tree.
 
-Comparing ExpCM's and \modelName{MutSel} models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree.
-If the limits of the ExpCM's reflected in Figure X are simply because the deep mutational scanning do not accurately measure the "true" amino-acid preferences of HA across the tree, then we would expect the \modelName{MutSel} models (which derive their stationary states from the entire tree) to perform better.
-On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect \modelName{MutSel} and ExpCM models to perform similarly.
+Comparing ExpCM's and pbMutSel models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree.
+If the limits of the ExpCM's reflected in Figure X are simply because the deep mutational scanning do not accurately measure the "true" amino-acid preferences of HA across the tree, then we would expect the pbMutSel models (which derive their stationary states from the entire tree) to perform better.
+On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect pbMutSel and ExpCM models to perform similarly.
 
 The ExpCM that estimates the longest branches is the one that uses the average of the H1 and H3 deep mutational scanning measurements with $\Gamma\omega$ rate variation~(\ref{fig:empirical_trees}).  
-We therefore compared this ExpCM with a \modelName{MutSel} model given the tree in \ref{fig:empirical_trees}. 
-First, the stationary state inferred by \modelName{MutSel} is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
-As \ref{fig:phylobayes}A shows, the site-specific amino-acid preferences estimated by \modelName{MutSel} are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
-This strong correlation between the \modelName{MutSel} and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.
+We therefore compared this ExpCM with a pbMutSel model given the tree in \ref{fig:empirical_trees}. 
+First, the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
+As \ref{fig:phylobayes}A shows, the site-specific amino-acid preferences estimated by pbMutSel are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
+This strong correlation between the pbMutSel and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.
 
-Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the \modelName{MutSel} model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}.
+Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the pbMutSel model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}. 
 While overall the two models estimated similar lengths, the tension between local and global accuracy of the amino-acid preferences is still apparent in these results. 
-All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the \modelName{MutSel} model.
+All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the pbMutSel model.
 The relatively longer branches from the experimental sequences when using the ExpCM+$\Gamma\omega$ suggests that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the experimental sequences.
-However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the \modelName{MutSel} model appear to be slightly better than the average of the experimental values.
-Therefore, while the site-specific \modelName{MutSel} models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
+However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the pbMutSel model appear to be slightly better than the average of the experimental values.
+Therefore, while the site-specific pbMutSel models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
 
@@
 
-\subsubsection*{Experimentally Informed Codon Model (ExpCM)}
+\subsubsection*{Experimentally Informed Codon Model (ExpCM).}
 
-The ExpCM is described in detail in \citet{hilton2017phydms}.
-
-In an ExpCM, rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
+In an ExpCM, the rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
 \begin{equation}
+\label{eq:ExpCM}
 P_{r,xy} = Q_{xy} \times F_{r,xy}
@@
 
-The deep mutational scanning data are incorporated into the ExpCM via the $F_{r,xy}$ terms.
+The deep mutational scanning amino-acid preferences are incorporated into the ExpCM via the $F_{r,xy}$ terms.
 The experiments measure the preference $\pi_{r,a}$ of every site $r$ for every amino-acid $a$.
@@
 where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\beta$ is the stringency parameter, and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
-The ExpCM has six free parameters (three $\phi_w$ values, $\kappa$, $\beta$, and $\omega$).
 The preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined by an experiment independent of the sequence alignment being analyzed.
+The ExpCM is described in detail in~\citet{hilton2017phydms}.
 
@@
 
-We consider the basic Goldman-Yang style (GY94) substitution model defined as M0 in \citet{yang2000codon}. 
+The GY94 model is defined as the M0 in \cite{yang2000codon}. 
 $P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by
@@
 \label{eq:GY94}
-P_xy = 
+P_{xy} = 
 \begin{cases}
@@
 
-\subsubsection*{GY94 with $\omega_r$}
-Following~\citet{spielman2015relationship}, we inferred an $\omega$ for each site in the protein ($\omega_r$) from an ExpCM. 
+\subsubsection*{GY94 with $\omega_r$.}
+
+We defined a site-specific model with the GY94 stationary state by multiplying the $\omega$ in \ref{eq:GY94} with a site-specific $\omega_r$ value calculated from an ExpCM (\ref{eq:ExpCM}). 
+Specifically, following \citet{spielman2015relationship}, 
 \begin{equation}
@@
 \end{equation}
-where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.
+where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, $\omega$ is the relative nonsynonymous substitution rate from the ExpCM and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.
 
-We replaced the $\omega$ value in \ref{eq:GY94} with $\omega \times \omega_r$, resulting in a site-specific model with the GY94 stationary state.
+\subsection*{$\Gamma\omega$ rate variation.}
 
-\subsection*{Adding $\Gamma\omega$}
+The GY94+$\Gamma\omega$ is defined as M5 in \citet{yang2000codon} and the ExpCM+$\Gamma\omega$ follows the same logic. 
+Briefly, $\omega$ is drawn from a $\Gamma$ distribution discretized into $K=4$ bins. 
+Each bin is equally weighted and $\omega$ takes on the mean value of the bin.
 
-For the GY94+$\Gamma\omega$ and ExpCM+$\Gamma\omega$, $\omega$ is drawn from $K=4$ discrete categories and equal weight is given to each category. 
-Set to the mean of the discretized $\Gamma$-distribution.
-
-\subsection*{Deep Mutational Scanning Amino-Acid preferences.}
-We used deep mutational scanning amino-acid preferences from an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the ExpCM's. 
-We only considered homologous sites between the H1 and H3 protein (described in \ref{suppfile:WSN_Perth_map}). 
+\subsection*{Deep mutational scanning amino-acid preferences.}
+We used amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the $\pi_{r, A\left(x\right)}$ parameters~(\ref{eq:Frxy}) of the ExpCM's.
+We only considered homologous sites between the H1 and H3 protein, described in \ref{suppfile:WSN_Perth_map}. 
 The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.
@@
 We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18. 
-We filtered and aligned the sequences using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}. 
-Specifically, we used \texttt{phydms\_comprehensive} default parameters except \texttt{--minidentity 0.3} and we removed sites which are not homologous between the H1 and H3 deep mutational references sequences using the sequence in \ref{suppfile:hybrid_sequence}.
-We subsampled the sequences so that there were 5 sequences per subtype with $\le 1$ sequence per year per subtype. 
-%Finally, we ensured the ``landmark" sequences from the H1, H2, and H3 subtypes (\ref{suppfile:required_seqs}) were included. 
+We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
+Specifically, we used \texttt{phydms\_prepalignment} with the flag \texttt{--minidentity 0.3} to sequences with ambiguous nucleotides, premature stops, or frameshift mutations as well as redundant sequences.
+We also removed any codon sites which are not shared between the H1 and H3 deep mutational scanning reference sequences by aligning to sequence in \ref{suppfile:hybrid_sequence}. 
+We subsampled the remaining sequences so there are $\ge 5$ sequences per subtype with $\le 1$ sequence per year per subtype. 
 The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.
@@
 We inferred the tree topology of each alignment using \texttt{RAxML}~\citep{stamatakis2006raxml} and the GTRCAT model. 
-We estimated the branch lengths of given this fixed topology for the ExpCM and GY94 models using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
+We estimated the branch lengths of this fixed topology using the ExpCM and GY94 models with \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
 
@@
 
-We built ExpCM and GY94 models using parameters fit to the ``low divergence from H1" tree.  
-Specifically, we built a GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, a GY94+$\Gamma\omega$ model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94+Gr}, an ExpCM using the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}, and a GY94+$\omega_r$ using the GY94 parameters in \ref{suppfile:hybrid_lowH1_GY94} and the ExpCM defined by\ref{suppfile:hybrid_lowH1_ExpCMdoud} and \ref{suppfile:H1_prefs}.
+We defined ExpCM and GY94 models using parameter values fit to the ``low divergence from H1" tree~(\ref{fig:empirical_trees}, \ref{suppfile:alignment_lowH1}). 
+Specifically, we defined the GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, the GY94+$\Gamma\omega$ using the parameter values in\ref{suppfile:hybrid_lowH1_GY94+Gr} and the ExpCM using the parameter values in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from~\citet{doud2016accurate} in \ref{suppfile:H1_prefs}.
 
-For each model we calculated the expected sequence divergence given some branch length $t$ by 
+For each model, we calculated the expected sequence divergence given branch length $t$ by 
 \begin{equation}
@@
 
-We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along tree for the full alignment (\ref{fig:empirical_trees};\ref{suppfile:alignment_high}) using an ExpCM defined by the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}.
-We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}.
+We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along the tree for the full alignment (\ref{fig:empirical_trees}, \ref{suppfile:alignment_high}) using an ExpCM defined by parameters fit to the ``low divergence from H1 tree" (\ref{suppfile:alignment_lowH1}).
+Specifically, we defined the ExpCM using the parameter values in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from \citet{doud2016accurate} in \ref{suppfile:H1_prefs}. 
+We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
+
+\subsection*{pbMutSel inference with \texttt{phylobayes}.}
+
+We inferred a pbMutSel model for the full tree (\ref{fig:empirical_trees}, \ref{suppfile:alignment_high}). 
+We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
+We normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
+These two H1 sequences are early and late representatives of the longest continuous sampled \skhcomment{in evolution but not in time...} lineage of human influenza.
+
+\subsection*{Code and analysis.}
+All code used is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. 
+For the main analysis, we used \texttt{phydms}~\citep{hilton2017phydms} version 2.2.dev2 (available at \url{github.com/jbloomlab/phydms}) to filter sequences and estimate branch lengths, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences and \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to infer the pbMutSel model.
+
+We also used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization.
 

### 2018-03-07 manuscript/manuscript.tex
changed MutSel name in figures
small edits

@@
 \newcommand\jdbcomment[1]{{\color{red}[#1]}}
-\newcommand\modelName[1]{{\color{magenta}[#1]}}
 
@@
 Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
-For highly diverged sequences, such phylogenetic techniques of- ten estimate surprisingly recent divergence times. 
+For highly diverged sequences, such phylogenetic techniques often estimate surprisingly recent divergence times. 
 In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
 This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
-Here we show that models informed by experimental measurements of purifying selection due to site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
+Here we show that models which account for purifying selection through experimental measurements of site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
 This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
@@
 
-\subsection*{Different ways that substitution models account for purifying selection}
+\subsection*{Different ways substitution models account for purifying selection}
 
@@
 \caption{\label{fig:model_feature}
-\textbf{Different ways that codon models account for purifying selection.}
+\textbf{Different ways codon models account for purifying selection.}
 (A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
@@
 The relative rate of these substitutions is referred to as dN/dS or $\omega$.
-In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average rate of fixation of nonsynonymous mutations relative to synonymous ones.
+In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average fixation rate of nonsynonymous mutations relative to synonymous ones.
 Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
@@
 We will denote this model as GY94+$\Gamma\omega$.
-A GY94+$\Gamma\omega$ captures the fact that the rate of nonsynonymous change can vary across sites. 
+A GY94+$\Gamma\omega$ captures the fact that the rate of nonsynonymous substitution can vary across sites. 
 However, this formulation treats all nonsynonymous substitutions equivalently, since the rate is agnostic to the amino-acid identity of the mutation.
@@
 A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
+These models explicitly define a different set of amino-acid preferences at each site in the protein. 
 This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
@@
 Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
-For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing the nucleotide composition of the stationary state.
+For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and nine parameters describing the nucleotide composition of the stationary state.
 However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
@@
 It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
-Here we will primarily use experimentally informed codon models (ExpCM's)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
-The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
-An alternative strategy of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} estimation is discussed in the last section of the Results.
+Here we will primarily use experimentally informed codon models~\citep[ExpCM's;][]{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
+The number of remaining free parameters for an ExpCM is similar to a non-site-specific substitution model.
+An alternative strategy of obtaining parameters for site-specific stationary states via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.
 
@@
 
-\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
+\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
 The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
@@
 To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
-Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees}) using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
+Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees} using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
 We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
 
-The models with a uniform stationary state underestimated the lengths of long branches on the phylogenetic of the simulated sequences (\ref{fig:simulations}). 
+The models with a uniform stationary state underestimated the lengths of long branches on the phylogenetic tree of the simulated sequences (\ref{fig:simulations}). 
 The GY94 model estimated branches lengths that are only about 60\% of the true values for the longest branches. 
@@
 
-\subsection*{Site-specific models estimate longer branches on real data.}
+\subsection*{Experimentally informed site-specific models estimate longer branches on real data.}
 
@@
 However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
-We used \texttt{RAxML} with a nucleotide model of substitution (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
+We used \texttt{RAxML}  with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
 For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
@@
 \skhcomment{figure out how to put the lee paper first and then the supp figures}
-The experimentally measured amino-acid preferences clearly differ between the the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
+The experimentally measured amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
 \ref{suppfig:prefs_lee};][]{lee2018deep}.
@@
 
-We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to containing increasingly diverged sequences.
+We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to trees containing increasingly diverged sequences.
 Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge 59\%$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge 46\%$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
@@
 
+%%%
+
+\begin{suppfile}
+\caption{
+\label{suppfile:WSN_Perth_map}
+Homologous sites in WSN and Perth using sequential numbering starting at the second (first non-M) codon. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:H1_prefs}
+Deep mutational scanning preferences from \citet{doud2016accurate} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:H3_prefs}
+Deep mutational scanning preferences from \citet{lee2018deep} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+\label{suppfile:avg_prefs}
+The average deep mutational scanning preferences from \citet{doud2016accurate} ad \citet{lee2018deep}. 
+}
+\end{suppfile}
+
 \begin{suppfile}
@@
 Specifically, this sequence is the H1 deep mutational scanning reference sequence A/Wilson Smith/1933 except it only contains the codon positions shared between A/Wilson Smith/1933 and the H3 deep mutational scanning reference sequence A/Perth/2009.
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-\label{suppfile:required_seqs}
-These sequences from the H1, H2, and H3 subtypes were required to be in the final alignment. 
 }
@@
 \caption{
-\label{suppfile:WSN_Perth_map}
-Homologous sites in WSN and Perth using sequential numbering starting at the second (first non-M) codon. 
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-\label{suppfile:H1_prefs}
-Deep mutational scanning preferences from \citet{doud2016accurate} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-\label{suppfile:H3_prefs}
-Deep mutational scanning preferences from \citet{lee2018deep} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-\label{suppfile:avg_prefs}
-The average deep mutational scanning preferences from \citet{doud2016accurate} ad \citet{lee2018deep}. 
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-\label{suppfile:hybrid_lowH1_ExpCMdoud}
-ExpCM model parameters fit using the amino-acid preferences from~\citet{doud2016accurate}(\ref{suppfile:H1_prefs}) and the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1})
-}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
 \label{suppfile:hybrid_lowH1_GY94}
@@
 
+\begin{suppfile}
+\caption{
+\label{suppfile:hybrid_lowH1_ExpCMdoud}
+ExpCM model parameters fit using the amino-acid preferences from~\citet{doud2016accurate}(\ref{suppfile:H1_prefs}) and the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1})
+}
+\end{suppfile}
+
 \clearpage 

### 2018-03-07 manuscript/manuscript.tex
edits to materials and methods

@@
 However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
-Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time due to epistasis.
+Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.
 \end{abstract}
@@
          Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & $\omega$ & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
-       	ExpCM + $\Gamma\omega$ (H1+H3 avg)  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
+       	ExpCM  (H1+H3 avg) + $\Gamma\omega$  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
 	ExpCM (H1+H3 avg)  &  950 & -49227 & 0.15 & 1.78\\
-	ExpCM + $\Gamma\omega$ (H1)  & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
-	ExpCM + $\Gamma\omega$ (H3) & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
+	ExpCM  (H1) + $\Gamma\omega$  & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
+	ExpCM (H3) + $\Gamma\omega$ & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
 	ExpCM (H1) & 2556 & -50030 &  0.13 & 1.22\\
@@
 These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
-\skhcomment{figure out how to put the lee paper first and then the supp figures}
-The experimentally measured amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites~\citep[\ref{suppfig:prefs_doud}, 
-\ref{suppfig:prefs_lee};][]{lee2018deep}.
+The experimentally measured amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites~\citep[][\ref{suppfig:prefs_doud}, 
+\ref{suppfig:prefs_lee}]{lee2018deep}.
 Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.
@@
 Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with greater stringency. 
-As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases.
+As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases and the preference sets ``flatten."
 Therefore, the preferences measured in each experiment are progressively less congruent with those of the HA sequences on the tree as we include increasingly diverged sequences. 
@@
 These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
-The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as pbMutSel. 
+The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as pbMutSel~\citep{rodrigue2014site}. 
 The pbMutSel model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
@@
 Comparing ExpCM's and pbMutSel models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree.
-If the limits of the ExpCM's reflected in Figure X are simply because the deep mutational scanning do not accurately measure the "true" amino-acid preferences of HA across the tree, then we would expect the pbMutSel models (which derive their stationary states from the entire tree) to perform better.
+If the limits of the ExpCM's reflected in \ref{fig:empirical_trees} are simply because the deep mutational scanning experiments do not accurately measure the ``true" amino-acid preferences of HA across the tree, then we would expect the pbMutSel models (which derive their stationary states from the entire tree) to perform better.
 On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect pbMutSel and ExpCM models to perform similarly.
@@
 We therefore compared this ExpCM with a pbMutSel model given the tree in \ref{fig:empirical_trees}. 
-First, the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
-As \ref{fig:phylobayes}A shows, the site-specific amino-acid preferences estimated by pbMutSel are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
+First, As \ref{fig:phylobayes}A shows, the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
+The site-specific amino-acid preferences estimated by pbMutSel are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
 This strong correlation between the pbMutSel and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.
 
-Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the pbMutSel model estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}. 
+Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the pbMutSel model also estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}.
 While overall the two models estimated similar lengths, the tension between local and global accuracy of the amino-acid preferences is still apparent in these results. 
@@
 \subsection*{Deep mutational scanning amino-acid preferences.}
-We used amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the $\pi_{r, A\left(x\right)}$ parameters~(\ref{eq:Frxy}) of the ExpCM's.
-We only considered homologous sites between the H1 and H3 protein, described in \ref{suppfile:WSN_Perth_map}. 
+We used amino-acid preferences measured in the deep mutational scans of an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the ExpCM $\pi_{r, A\left(x\right)}$ parameters~(\ref{eq:Frxy}).
+We only considered sites which are homologous between H1 HA and H3 HA. 
+These homologous sites, and their mapping between the H1 reference strain A/Wilson Smith/1933 and the H3 reference strain a A/Perth/2009, are described in \ref{suppfile:WSN_Perth_map}. 
 The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.
@@
 We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
-Specifically, we used \texttt{phydms\_prepalignment} with the flag \texttt{--minidentity 0.3} to sequences with ambiguous nucleotides, premature stops, or frameshift mutations as well as redundant sequences.
-We also removed any codon sites which are not shared between the H1 and H3 deep mutational scanning reference sequences by aligning to sequence in \ref{suppfile:hybrid_sequence}. 
-We subsampled the remaining sequences so there are $\ge 5$ sequences per subtype with $\le 1$ sequence per year per subtype. 
+Specifically, we used \texttt{phydms\_prepalignment} with the flag \texttt{--minidentity 0.3} to remove sequences with ambiguous nucleotides, premature stops, or frameshift mutations as well as redundant sequences.
+We also removed any codon sites which are not shared between the H1 HA and H3 HA used in the deep mutational scanning experiments. 
+We subsampled the remaining sequences to five sequences per subtype with $\le 1$ sequence per year per subtype. 
+We also included a small number of representative sequences from the major human and equine influenza lineages, described in \ref{suppfile:required_seqs}. 
 The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.
@@
 We normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
-These two H1 sequences are early and late representatives of the longest continuous sampled \skhcomment{in evolution but not in time...} lineage of human influenza.
+These two H1 sequences are early and late representatives of the longest known human influenza lineage.
 
@@
 For the main analysis, we used \texttt{phydms}~\citep{hilton2017phydms} version 2.2.dev2 (available at \url{github.com/jbloomlab/phydms}) to filter sequences and estimate branch lengths, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences and \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to infer the pbMutSel model.
-
 We also used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization.
@@
 \caption{\label{suppfig:prefs_doud}
-\textbf{H1 preferences measured by \cite{doud2016accurate} (\ref{suppfile:H1_prefs}) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.12$)}. 
+\textbf{H1 HA amino-acid preferences measured by deep mutational scanning.}
+Each column represents a site in the protein and the height of each letter is proportional to the preference for the amino acid as measured by~\citet{doud2016accurate}. 
+Here the preferences are rescaled by the ExpCM stringency parameter in \ref{tab:empirical_data} ($\beta=1.12$). 
+We only considered homologous sites between H1 and H3. 
+The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 is found in \ref{suppfile:WSN_Perth_map}. 
 }
@@
 \caption{\label{suppfig:prefs_lee}
-\textbf{H3 preferences measured by~\citet{lee2018deep} (\ref{suppfile:H1_prefs}) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.28$)}
+\textbf{H3 HA amino-acid preferences measured by deep mutational scanning.}
+Each column represents a site in the protein and the height of each letter is proportional to the preference for the amino acid as measured by~\citet{lee2018deep}. 
+Here the preferences are rescaled by the ExpCM stringency parameter in \ref{tab:empirical_data} ($\beta=1.28$). 
+We only considered homologous sites between H1 and H3. 
+The conversion between the numbering scheme in this figure and sequential numbering of the H3 HA reference sequence A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}. 
  }
@@
 \caption{\label{suppfig:prefs_average}
-\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by~\citet{lee2018deep} ((\ref{suppfile:avg_prefs})) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.70$)}}
+\textbf{Average of H1 HA  and H3 HA amino-acid preferences measured by deep mutational scanning.}
+Each column represents a site in the protein and the height of each letter is proportional to the average preference for the amino acid measured by~\citet{doud2016accurate} and~\citet{lee2018deep}. 
+Here the preferences are rescaled by the ExpCM stringency parameter in \ref{tab:empirical_data} ($\beta=1.70$). 
+We only considered homologous sites between H1 and H3. 
+The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 or H3 HA reference sequence A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}.
+}
 \end{suppfig}
@@
 \caption{\label{suppfig:prefs_mutSel}
-\textbf{Preferences inferred by the mutSel model on the full HA tree.}
+\textbf{Amino-acid preferences inferred by the pbMutSel model.}
+Each column represents a site in the protein and the height of each letter is proportional to the preference for the amino acid inferred by pbMutSel~\citep{rodrigue2014site} using the tree in \ref{fig:empirical_trees}.
+We only considered homologous sites between H1 and H3. 
+The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 or H3 HA reference sequence A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}.
 }
@@
 \label{suppfile:WSN_Perth_map}
-Homologous sites in WSN and Perth using sequential numbering starting at the second (first non-M) codon. 
+Map for homologous sites between H1 HA and H3 HA. 
+This file provides a conversion between the numbering scheme we use in the paper, including all logoplots and the sites in \ref{fig:decay}, to sequential numbering in either the H1 HA reference sequence A/Wilson Smith/1933 or the H3 HA reference sequence A/Perth/2009.
+The sequential numbering for the HA strains starts with site1 as the second codon (first non-methionine codon). 
 }
@@
 \label{suppfile:H1_prefs}
-Deep mutational scanning preferences from \citet{doud2016accurate} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
+Amino acid preferences measured by the deep mutational scanning of the H1 HA strain A/Wilson Smith/1933~\citep{doud2016accurate}. 
+This file only contains measurements for the homologous sites between H1 HA and H3 HA. 
+Conversion from this numbering scheme to sequential numbering in A/Wilson Smith/1933 is found in \ref{suppfile:WSN_Perth_map}. 
 }
@@
 \label{suppfile:H3_prefs}
-Deep mutational scanning preferences from \citet{lee2018deep} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
+Amino acid preferences measured by the deep mutational scanning of the H3 HA strain A/Perth/2009~\citep{lee2018deep}. 
+This file only contains measurements for the homologous sites between H1 HA and H3 HA. 
+Conversion from this numbering scheme to sequential numbering in A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}. 
 }
@@
 \label{suppfile:avg_prefs}
-The average deep mutational scanning preferences from \citet{doud2016accurate} ad \citet{lee2018deep}. 
-}
+Average of the amino acid preferences measured by the deep mutational scanning of the H1 HA strain A/Wilson Smith/1933~\citep{doud2016accurate} and the H3 HA strain A/Perth/2009~\citep{lee2018deep}. 
+This file only contains measurements for the homologous sites between H1 HA and H3 HA. 
+Conversion from this numbering scheme to sequential numbering in A/Wilson Smith/1933 or A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}. }
 \end{suppfile}
@@
 \caption{
-\label{suppfile:hybrid_sequence}
-The ``hybrid" sequence used to align the HA sequences.
-Specifically, this sequence is the H1 deep mutational scanning reference sequence A/Wilson Smith/1933 except it only contains the codon positions shared between A/Wilson Smith/1933 and the H3 deep mutational scanning reference sequence A/Perth/2009.
+\label{suppfile:required_seqs}
+This file contains the representative sequences of major human equine influenza lineages added to the alignment in \ref{suppfile:alignment_high}. 
 }
@@
 \label{suppfile:alignment_high}
-HA sequences used in \ref{fig:empirical_trees}.  
+The HA sequences used for the full HA tree in \ref{fig:empirical_trees}. 
+We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
 }
@@
 \label{suppfile:alignment_lowH1}
-HA sequences for the ``low divergence from H1" tree in \ref{fig:compete}.  
+The HA sequences using for the ``low divergence from H1" tree in \ref{fig:compete}. 
+All sequences have $\ge 59\%$ amino-acid identity to the H1 reference sequence A/Wilson Smith/1933. 
+We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
 }
@@
 \label{suppfile:alignment_lowH3}
-HA sequences for the ``low divergence from H3" tree in \ref{fig:compete}.  
+The HA sequences using for the ``low divergence from H3" tree in \ref{fig:compete}. 
+All sequences have $\ge 59\%$ amino-acid identity to the H3 reference sequence A/Perth/2009. 
+We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
 }
@@
 \label{suppfile:alignment_intermediateH1}
-HA sequences for the ``low divergence from H1" tree in \ref{fig:compete}.  
+The HA sequences using for the ``intermediate divergence from H1" tree in \ref{fig:compete}. 
+All sequences have $\ge 46\%$ amino-acid identity to the H1 reference sequence A/Wilson Smith/1933. 
+We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
 }
@@
 \label{suppfile:alignment_intermediateH3}
-HA sequences for the ``low divergence from H3" tree in \ref{fig:compete}.  
+The HA sequences using for the ``intermediate divergence from H3" tree in \ref{fig:compete}. 
+All sequences have $\ge 46\%$ amino-acid identity to the H3 reference sequence A/Perth/2009. 
+We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
 }
@@
 \label{suppfile:hybrid_lowH1_GY94}
-GY94 model parameters fit using the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1}). 
+The GY94 model parameters fit to the ``low divergence from H1" tree (\ref{fig:compete}, \ref{suppfile:alignment_lowH1}). 
+The model was fit using \texttt{phydms\_comprehensive}. 
 }
@@
 \label{suppfile:hybrid_lowH1_GY94+Gr}
-GY94+$\Gamma\omega$ model parameters fit using the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1}). 
+The GY94+$\Gamma\omega$ model parameters fit to the ``low divergence from H1" tree (\ref{fig:compete}, \ref{suppfile:alignment_lowH1}). 
+The model was fit using \texttt{phydms\_comprehensive}. 
 }
@@
 \label{suppfile:hybrid_lowH1_ExpCMdoud}
-ExpCM model parameters fit using the amino-acid preferences from~\citet{doud2016accurate}(\ref{suppfile:H1_prefs}) and the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1})
+The ExpCM model parameters fit to the ``low divergence from H1" tree (\ref{fig:compete}, \ref{suppfile:alignment_lowH1}). 
+The model was fit using \texttt{phydms\_comprehensive} and the H1 HA amino-acid preference set in \ref{suppfile:H1_prefs}. 
 }

### 2018-03-08 manuscript/manuscript.tex
finished empirical trees paragraph

@@
 
-\skhcomment{I don't like how this paragraph comes almost as an afterthought from the next one given that branch lengths are actually are main point.
-I wonder if we can add a stronger first sentence to draw more emphasis to this paragraph.
-I also feel like a sentence between the current first and second one would be good.
-It could comment a bit more on the specifics of the results in Figure 4 in a way that draws out the point better.}
-\ref{fig:empirical_trees} shows that modeling purifying selection via either $\Gamma\omega$ rate variation or a site-specific stationary state increases the estimated branch lengths.
+Models which account for purifying selection via either $\Gamma\omega$ rate variation or a site-specific stationary state are not just better fit models, they also increase the estimation of branch length on the HA tree.
+As \ref{fig:empirical_trees} shows, the tree optimized by without $\Gamma\omega$ rate variation and without a site-specific stationary state, the GY94 model, estimated the shortest branches of all eight models.
+Specifically, the ExpCM's defined by H1 HA or H3 HA amino-acid preferences estimated a tree which overall length is $\sim 8\%$ longer than the GY94 tree and the GY94+$\Gamma\omega$ model estimated tree which is $\sim 36\%$ longer than the GY94 tree.
 But importantly, the effects from these two methods of modeling purifying selection are largely independent.

### 2018-03-08 manuscript/manuscript.tex
added within subtype diversity to manuscript

@@
 The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
-This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
-Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
+This model predicts an asymptotic sequence divergence of 0.059, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
+Intuitively, this asymptotic sequence identity of 0.059 seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
 
@@
 HA consists of a number of different subtypes.
-Sequences within a subtype have \jdbcomment{?} amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
+Sequences within a subtype have $>75\%$ amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
 However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.

### 2018-03-08 manuscript/manuscript.tex
small teweaks to the text

@@
 When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon}.
-Here we will refer to M0 Goldman-Yang models simply as GY94 models. 
+Here we will refer to M0 Goldman-Yang models simply as GY94 models (\ref{eq:GY94}). 
 The gene-wide $\omega$ is usually $<1$~\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
@@
 This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
-It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
-Here we will primarily use experimentally informed codon models~\citep[ExpCM's;][]{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
+It is challenging to obtain values for these parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
+Here we will primarily use experimentally informed codon models~\citep[ExpCM's, \ref{eq:ExpCM};][]{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
 The number of remaining free parameters for an ExpCM is similar to a non-site-specific substitution model.
@@
 The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
-This model predicts an asymptotic sequence divergence of 0.059, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
+This model predicts an asymptotic sequence divergence of 0.059, which is slightly higher than 0.050 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
 Intuitively, this asymptotic sequence identity of 0.059 seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
@@
 However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
-We used \texttt{RAxML}  with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
+We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
 For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
@@
 We therefore compared this ExpCM with a pbMutSel model given the tree in \ref{fig:empirical_trees}. 
-First, As \ref{fig:phylobayes}A shows, the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
+First, \ref{fig:phylobayes}A shows that the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
 The site-specific amino-acid preferences estimated by pbMutSel are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
@@
 
-\subsubsection*{Experimentally Informed Codon Model (ExpCM).}
+\subsubsection*{GY94}
+
+The GY94 model is defined as the M0 in \cite{yang2000codon}. 
+$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by
+
+\begin{equation}
+\label{eq:GY94}
+P_{xy} = 
+\begin{cases}
+  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
+  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
+  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
+  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
+  \end{cases}
+\end{equation}
+where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
+codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.
+The stationary state of a GY94 model is defined as 
+\begin{equation}
+\label{eq:px}
+p_{x} = \Phi_x
+\end{equation}
+
+\subsubsection*{Experimentally Informed Codon Model (ExpCM)}
 
@@
 where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\beta$ is the stringency parameter, and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
+
+The stationary state of an ExpCM is defined as 
+\begin{equation}
+\label{eq:prx}
+p_{r,x} = \frac{\phi_{x_1}\phi_{x_2}\phi_{x_3}\left(\pi_{r,A\left(x\right)}\right)^\beta}{\sum_z \phi_{z_1}\phi_{z_2}\phi_{z_3}\left(\pi_{r,A\left(z\right)}\right)^\beta}
+\end{equation}
+where $\phi_{x_1}$, $\phi_{x_2}$, and $\phi_{x_3}$ are the nucleotides at position 1, 2, and 3 of codon $x$.
+
 The preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined by an experiment independent of the sequence alignment being analyzed.
 The ExpCM is described in detail in~\citet{hilton2017phydms}.
-
-\subsubsection*{GY94}
-
-The GY94 model is defined as the M0 in \cite{yang2000codon}. 
-$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by
-
-\begin{equation}
-\label{eq:GY94}
-P_{xy} = 
-\begin{cases}
-  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
-  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
-  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
-  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
-  \end{cases}
-\end{equation}
-where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
-codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.
 
@@
 We defined ExpCM and GY94 models using parameter values fit to the ``low divergence from H1" tree~(\ref{fig:empirical_trees}, \ref{suppfile:alignment_lowH1}). 
-Specifically, we defined the GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, the GY94+$\Gamma\omega$ using the parameter values in\ref{suppfile:hybrid_lowH1_GY94+Gr} and the ExpCM using the parameter values in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from~\citet{doud2016accurate} in \ref{suppfile:H1_prefs}.
+Specifically, we defined the GY94 model using the parameter values in~\ref{suppfile:hybrid_lowH1_GY94}, the GY94+$\Gamma\omega$ using the parameter values in~\ref{suppfile:hybrid_lowH1_GY94+Gr} and the ExpCM using the parameter values in~\ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from~\citet{doud2016accurate} in~\ref{suppfile:H1_prefs}.
 
@@
 This file provides a conversion between the numbering scheme we use in the paper, including all logoplots and the sites in \ref{fig:decay}, to sequential numbering in either the H1 HA reference sequence A/Wilson Smith/1933 or the H3 HA reference sequence A/Perth/2009.
-The sequential numbering for the HA strains starts with site1 as the second codon (first non-methionine codon). 
+The sequential numbering for the HA strains starts with ``site 1" as the second codon (first non-methionine codon). 
 }

### 2018-03-08 manuscript/manuscript.tex
added reference to stoke shift paper

@@
 
-The fact that an ExpCM has the greatest effect on branches close to the sequence used in the experiment can be rationalized in terms of existing knowledge about how epistasis shifts protein amino-acid preferences during evolution~\jdbcomment{Stokes shift paper, Shah et al paper, Doud et al 2015, Haddox et al 2018, look at other papers cited about Stokes shift in Haddox et al 2018}.
+The fact that an ExpCM has the greatest effect on branches close to the sequence used in the experiment can be rationalized in terms of existing knowledge about how epistasis shifts protein amino-acid preferences during evolution~\citep{pollock2012amino, shah2015contingency, doud2015site, haddox2017mapping, risso2014mutational}.
 In reality, sites in proteins do not have completely conserved site-specific amino-acid preferences. 
-Rather, the effect of a mutation at one site in a protein can depend on the amino-acid identity of other sites in the protein~\jdbcomment{Maybe cite a couple of general papers on epistasis, some good ones in Haddox et al 2018}.
+Rather, the effect of a mutation at one site in a protein can depend on the amino-acid identity of other sites in the protein~\citep{starr2018pervasive, pollock2012amino, shah2015contingency, bazykin2015changing}.
 The deep mutational scanning experiments that inform our ExpCM's were performed in the context of a single HA genetic background, and so do not account for how such epistatic interactions can cause amino-acid preferences to shift as substitutions accumulate.

### 2018-03-09 manuscript/manuscript.tex
added table with total tree lengths

@@
 As \ref{fig:empirical_trees} shows, the tree optimized by without $\Gamma\omega$ rate variation and without a site-specific stationary state, the GY94 model, estimated the shortest branches of all eight models.
-Specifically, the ExpCM's defined by H1 HA or H3 HA amino-acid preferences estimated a tree which overall length is $\sim 8\%$ longer than the GY94 tree and the GY94+$\Gamma\omega$ model estimated tree which is $\sim 36\%$ longer than the GY94 tree.
+Specifically, the ExpCM's defined by H1 HA or H3 HA amino-acid preferences estimated a tree which overall length is $\sim 8\%$ longer than the GY94 tree and the GY94+$\Gamma\omega$ model estimated tree which is $\sim 36\%$ longer than the GY94 tree~\ref{supptab:tree_lengths}.
 But importantly, the effects from these two methods of modeling purifying selection are largely independent.
@@
 In addition, all ExpCM's estimate longer deep branches than the GY94 model, and all ExpCM+$\Gamma\omega$ models estimate longer deep branches than the GY94+$\Gamma\omega$ model.
+\skhcomment{I think we can make a stronger statement about the independence of these effects.
+I did the calculation and 96\% of the branch length extension of GY94 vs.
+ExpCM(H1+H3)+$\Gamma\omega$ can be explained by simply adding the ``effect of adding $\Gamma\omega$" (GY94 vs.
+GY94+$\Gamma\omega$) and the ``effect of adding preferences" (GY94 vs.
+ExpCM(H1+H3).
+Do you think that is a useful analysis to include?}
 
@@
 \caption{\label{suppfig:subalignments}
-\textbf{Subalignments.}
+\textbf{Overall divergence for the subtrees in \ref{fig:compete}.} 
+We created two sub alignments from the alignment (\ref{suppfile:alignment_high}) used in \ref{fig:empirical_trees} for each deep mutational scanning reference HA. 
+The ``low" alignments have $\ge59\%$ amino-acid identity to the reference sequence (\ref{suppfile:alignment_lowH1}, \ref{suppfile:alignment_lowH3}) and the ``intermediate" alignments have $\ge46\%$ amino-acid identity to the reference sequence (\ref{suppfile:alignment_intermediateH1}, \ref{suppfile:alignment_intermediateH3}).
 }
 \end{suppfig}
+
+\begin{table}[t!]
+\caption{\label{supptab:tree_lengths}
+{\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
+The branch lengths are in average codon substitutions per site. 
+} 
+     \begin{tabular}{cccccccccc}
+        \hline
+         Model & no rate variation & {\shortstack{$\Gamma\omega$ rate\\ variation}} \\ \hline
+         GY94 &53.05 & 72.28\\
+         ExpCM (H1) & 57.32 & 81.31\\
+         ExpCM (H3) & 57.77 & 78.98\\
+         ExpCM(H1 + H3 avg) & 62.98 & 83.96 \\
+      \end{tabular}
+\end{table}
 

### 2018-03-09 manuscript/manuscript.tex
rough draft of introduction

@@
 \section*{Introduction}
-\skhcomment{from JDB: what is the "age" of a virus? Maybe "divergence time of viral lineages"}
-skhcomment{from JDB: what is the less than a million actually? "Old" is not the right phrase.}
-Estimating the divergence time of viral lineages of a virus is essential to understanding its evolutionary history, including its emergence, spread, and past zoonoses. 
-This estimation is commonly done using the concept a ``molecular clock" to transform the branch lengths of the viral phylogenetic tree into age in years. 
-However, this molecular dating technique often underestimates the age of many viruses, including measles, foamy virus, and ebola \skhcomment{(citations)}, compared to other methods which are independent of the viral phylogeny.
-For example, SIV (the original source of HIV) is estimated to be less than a million years old based on the viral phylogeny \citep{sharp2000origins,wertheim2009dating,worobey2010island} but estimated to be several million years old based on the host tree or endogenous retroviral elements  \citep{compton2013host} \skhcomment{(other citations)}.
-Overall, there is a systematic and substantially large underestimation of of branch length on viral phylogenies. 
-\skhcomment{long branches}
 
-Branch length underestimation is due, in part, to strong purifying selection masking the evolutionary signal in the observed sequences. 
-Purifying selection can lead to mutational saturation, where multiple unobserved, substitutions occur at a single site along a long branch and erase the divergence signal \citep{holmes2003molecular}.
-Furthermore, proteins do not have equal preference for all amino acids at all sites, this evident by a simple visual inspection of a multiple sequence alignment. 
-How many and which amino acids tolerated at each site of the protein generate a site-specific expected rate of change. 
-Failing to account for these site-specific constraints will lead to branch length underestimation. 
-\skhcomment{you will have mutational saturation no matter what - this is a separate, addressable issue?}
-\skhcomment{talk about the high mutation rate in viruses?}
+When studying a protein's evolutionary history, it is important to understand the timescale at which the evolutionary events have occurred. 
+For example, understanding how long a particular viral lineage has been circulating informs hypotheses about population size, evolutionary rate, and possible previous hosts.
+Molecular phylogenetic is commonly used to make sure estimates of time since the divergence of two lineages.
+Under a molecular clock assumption, the branch lengths on a phylogenetic tree can be transformed into estimates of absolute time. 
+However, it has been observed that these phylogenetic estimates often appear to estimate times since deep divergences that appear to be too recent. 
+That is, when two nodes are separated by a very long branch, the phylogenetic estimate of this branch length often appears to be too short, which leads to an estimate of time that is too recent.
 
-Substitution models that incorporate site-to-site rate variation have been developed to decrease the bias in long branch estimation. 
-The most common strategy is to allow a single rate-controlling parameter to vary according to some statistical distribution, such as a $\Gamma$-distributed $\omega$ (~dN/dS) \citep{yang2000codon}. 
-This flexibility in the value of $\omega$ accounts for the site-to-site rate variation by allow some sites to have a higher dN/dS value than others. 
-While this modification is simple and only requires the addition of one extra parameter, it does not describe site-specificity in its stationary state. 
-That is, at evolutionary equilibrium, this model still assumes that each site in the protein evolves identically.
+In the case of viruses, there is independent evidence that the estimates of these deep divergences are indeed too recent. 
+The times since the divergence of lineages within lentiviruses, paramyxoviruses and filoviruses have been estimated by methods which rely on information outside of or in addition to the viral phylogenetic tree.
+These independent estimates are often orders of magnitude older than the estimates using the viral phylogenetic tree only. 
+For example, there are multiple examples of filovirus integrations into the genomes of rodents. 
+Using these viral integrations and the fossil dates of the rodent species tree, the divergence of two main filovirus groups has been estimated to be $>7$ millions years ago. 
+This estimate stands in stark contrast to the phylogenetic estimate of filovirus divergence which estimates this event happened $\sim 10,000$ years ago. 
+All together, this indicates that phylogenetic branch length estimation does truly have a systematic bias towards underestimation.
 
-An alternative approach is to model the site-specific amino-acid frequencies explicitly, such as those models in the mutation-selection family \citep{halpern1998evolutionary}. 
-In these models, each amino-acid at each site in the protein is described by its own parameter and these differences are reflected in the stationary state of the model. 
-The rate of change at a given site is controlled by these amino acid profiles and can now vary from site to site, as expected based on observations in nature. 
-Importantly, these rate variations are not constrained to an arbitrary statistical distribution but by parameters with a direct biological interpretation.
+Branch length underestimation is due in part to inadequate modeling of purifying selection. 
+During evolution, protein sites do not sample all 20 available amino-acids equally. 
+Instead, proteins have site-specific amino-acid preferences which are necessary to maintain the structure and function of the protein. 
+Failure to account for these constraints will lead to branch length underestimation because the model will assume that sequences which have been evolving for a very long period of time should be more diverged than is actually the case.
+Furthermore, the site-specific amino-acid preferences at one site are often dependent on the amino-acid identity at another site. 
+As a result, the preferences ``shift" over time time due to the accumulation of substitutions in other parts of the protein.
 
-Mutation-selection models are presumably more biologically relevant but pose more practical challenges than the $\Gamma\omega$ models. 
-These models are highly parametrized with 19 free parameters (the 20 amino acid preferences are constrained to sum to one) per site leading to thousands of parameters for the length of a normal protein.
-One way to avoid overfitting is to implement the model as a mixture model in either a bayesian \citep{lartillot2004bayesian} or maximum likelihood framework \citep{si2008empirical}.
+Most phylogenetic substitution models are time-homogenous and site-independent: they do not take into account how sites may interact with each other or how preferences may shift over time. 
+However, they can model purifying selection with varying degrees of complexity. 
+One method is to allow the parameter controlling the rate of nonsynonymous to synonymous substitution, the dN/dS parameter, to vary among the protein sites. 
+This variation accounts for the fact that constrained sites will have a lower rate of amino-acid substitution than mutationally constrained sites. 
+However, the stationary state of this model, which describes the expected sequence composition after a very long evolutionary time, will be exactly the same as a model without any variation in the dN/dS parameter.
+Alternatively, models can have site-specific stationary states which explicitly acknowledge the protein's site-specific amino-acid preferences. 
+Such models should estimate deep divergence times because model still expects relatively low sequence divergence even after a very long evolutionary time.
 
-Alternatively, you can reduce the parameter space by defining the amino-acid frequencies \textit{a priori}. 
-We have shown previously that we can define an Experimentally Informed Codon Model (ExpCM) \citep{bloom2014experimentally,bloom2014informed} from the mutation-selection family using measurements from deep mutational scanning \citep{fowler2014deep}, a high-throughput functional assay.
-ExpCM are therefore defined by amino-acid preferences measured in a \textit{single} genetic background and do not reflect any epistatic changes which may have occurred over the virus's evolutionary history.
-But they contain no more parameters than the traditional codon models while maintaining a site-specific stationary state. 
-We hypothesize that the ExpCM will estimate longer branches than the traditional models due to the protein-specific description of purifying selection. 
-\skhcomment{CAT model has been shown to work well (better) on saturated data.}
-
-In order to test this hypothesis, we compared the branch lengths of a influenza virus HA phylogenetic trees optimized by different substitution models. 
-We found that the ExpCM did extend the length of branches from the focal sequence on the tree \skhcomment{define focal} and that this extension was seen even in the context of $\Gamma$-distributed rate variation.
-Furthermore, we found this extension occurred even in the presence of $\Gamma$-distributed $\omega$, indicating that they are both important for modeling purifying selection. 
-This supports the conclusion that modeling purifying selection, especially in a model with a non-uniform stationary state, is important to estimating the branch lengths on phylogenetic trees.
+Here, we tested the effect of modeling site-specific amino-acid preferences though a model with a site-specific stationary state on an influenza hemagglutinin (HA) phylogenetic tree. 
+Specifically, we used Experimentally Informed Codon Models (ExpCM's) defined by site-specific amino-acid preferences measured by a high-throughput functional assay called deep mutational scanning. 
+We found that these ExpCM's estimated longer branches. 
+We found that the extension in branch length due modeling purifying selection via a site-specific stationary state is largely independent of the the effect of modeling purifying selection through rate variation.
+However, our results make it clear that the site-specific amino-preferences of HA are constant across the entire tree. 
+In particular, the branch lengthening is most pronounced near the sequence of the HA that was used in the deep mutational scanning experiment that parameterized the ExpCM.
+These results underscore the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.
 
@@
 It is challenging to obtain values for these parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
+\skhcomment{Do we need to comment on the tamuri paper?}
 Here we will primarily use experimentally informed codon models~\citep[ExpCM's, \ref{eq:ExpCM};][]{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
@@
 \begin{enumerate}
+	\item branch length underestimation
+	\item models of purifying selection - especially site-specific stationary states 
+	\item we looked at models which were defined we using site-specific amino-acid preferences measured by deep mutational scanning 
+	\item we found that the lengthen the branches but that the effect is not uniform across the entire tree - the preferences have shifted over time 
+	\item This points to two things 1) it is important to model purifying selection 2) just has there is an inherent limitation to models which are site-specific but also models which don't take into account epistasis or how preferences shift over time.
+\end{enumerate}
+
+\begin{enumerate}
   \item We don't allow any of the models to vary by lineage. 
   \item discussion in \citet{rodrigue2010mutation}
+  \item we could look into the epistasis because we used DMS sets from two, highly diverged viral homologs
 \end{enumerate}

### 2018-03-11 manuscript/manuscript.tex
first draft of discussion

@@
 
-\section*{Results and Discussion}
+\section*{Results}
 
@@
 
-\section*{Conclusion}
+\section*{Discussion}
 
-\begin{enumerate}
-	\item branch length underestimation
-	\item models of purifying selection - especially site-specific stationary states 
-	\item we looked at models which were defined we using site-specific amino-acid preferences measured by deep mutational scanning 
-	\item we found that the lengthen the branches but that the effect is not uniform across the entire tree - the preferences have shifted over time 
-	\item This points to two things 1) it is important to model purifying selection 2) just has there is an inherent limitation to models which are site-specific but also models which don't take into account epistasis or how preferences shift over time.
-\end{enumerate}
+Here, we tested the effect of models with site-specific stationary states on branch length estimation for the phylogenetic tree of highly diverged influenza HA sequences. 
+We used site-specific stationary state models called ExpCM's, which are defined by amino-acid preferences measured by deep mutational scanning. 
+We found that the ExpCM's estimated longer branches and did so independently from substitution rate variation via $\Gamma\omega$. 
+We also found that the ExpCM's estimated branch lengths of a similar length to models which infer the stationary state from the natural sequences in a Bayesian framework.
 
-\begin{enumerate}
-  \item We don't allow any of the models to vary by lineage. 
-  \item discussion in \citet{rodrigue2010mutation}
-  \item we could look into the epistasis because we used DMS sets from two, highly diverged viral homologs
-\end{enumerate}
+These results underscore the importance of modeling site-specific amino-acid preferences when estimating long branchs. 
+As the simulations \ref{fig:simulations} show, models with uniform stationary states will always underestimate the lengths of branches which have evolved under site-specific constraints. 
+The results with the influenza HA sequences shows that models with site-specific stationary states estimate longer branches when the stationary state is accurate. 
+However, it is also clear that the accuracy of the ExpCM's stationary state degrades as the tree becomes more diverged from the HA used in the deep mutational scan. 
+The desired effect of estimating long, accurate branches can only be achieved when the stationary state is accurate across the entire tree.
+
+However, it is clear from our results that a major limitation of site-independent, time-homogenous models is their inability to capture epistatic interactions. 
+The site-specific amino-acid preferences of a protein shift across time and models with a single stationary state across the entire tree cannot capture these dynamics. 
+The ExpCM's defined by preferences measured in one of two highly diverged HA's had the largest effect on branches near the HA from the experiment. 
+The local effect of the ExpCM's defined by a single preference set indicate that different parts of the tree have different stationary states. 
+This point is underscored by the comparison of the ExpCM defined by the average preference set and the pbMutSel model, which infers the stationary state from the natural sequences. 
+Even though the stationary state of the pbMutSel model is more representative of the global stationary state, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
+Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither one of these models is able to capture how these preferences shift over time.
+
+Clearly, the site-independence and time-homogeneity assumptions of the ExpCM and pbMutSel models inhibit their ability to accurately describe the effect of epistasis.
+In order to fully address the bias towards underestimation of divergence times by phylogenetic models, these models need to be able to accommodate how site-specific amino-acid preferences shift over time.
+One way to achieve this goal would be to void the site-independence assumption and create a model which explicitly describes the interactions between sites in a protein. 
+However, while even modeling simple pairwise interactions would be a daunting task, epistasis could have higher order interactions which be extremely difficult to model. 
+Another strategy would be to allow the stationary state to shift over lineages in the tree. 
+Such a model might be still be site-independent but, by breaking the time-homogeneity assumption, would capture some effect of the shifting preferences. 
+Models which account for epistasis would be inherently more complex but it clear from our results that it is important to capture how preferences shift over time to accurately estimate long branch lengths.
 

### 2018-03-12 manuscript/manuscript.tex
edits to Abstract

@@
 \usepackage{float}
+
+% allow bigger floats per here: https://tex.stackexchange.com/a/11382
+\renewcommand{\topfraction}{.85}
+\renewcommand{\bottomfraction}{.85}
+\renewcommand{\textfraction}{.15}
+\renewcommand{\floatpagefraction}{.85}
+\renewcommand{\dbltopfraction}{.66}
+\renewcommand{\dblfloatpagefraction}{.66}
+\setcounter{topnumber}{9}
+\setcounter{bottomnumber}{9}
+\setcounter{totalnumber}{20}
+\setcounter{dbltopnumber}{9}
+
 \renewcommand{\figurename}{{}}
@@
 \author
-{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2}$\\
+{Sarah K. Hilton$^{1,2}$  and Jesse D. Bloom$^{1,2*}$\\
 \\
-\normalsize{$^1$Division of Basic Sciences and Computational Biology Program,}\\
-\normalsize{Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}\\
-\normalsize{$^2$Department of Genome Sciences, University of Washington, Seattle, WA}\\
-\normalsize{E-mail:  jbloom@fredhutch.org.}\\
+\footnotesize{$^1$Basic Sciences and Computational Biology Program, Fred Hutchinson Cancer Research Center}\\
+\footnotesize{$^2$Department of Genome Sciences, University of Washington}\\
+\footnotesize{Seattle, WA, USA}\\
+\footnotesize{$^*$E-mail:  jbloom@fredhutch.org.}\\
 }
@@
 \noindent  
-Molecular phylogenetics can be used to estimate the time since the divergence of modern gene sequences.
-For highly diverged sequences, such phylogenetic techniques often estimate surprisingly recent divergence times. 
+Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
+For highly diverged sequences, such phylogenetic techniques can estimate surprisingly recent divergence times. 
 In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
 This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
-Here we show that models which account for purifying selection through experimental measurements of site-specific amino-acid preferences lengthen deep branches on phylogenies of influenza virus hemagglutinin.
+Here we examine the effect on branch-length estimation of using models that incorporate experimental measurements of purifying selection.
+We find that models informed by experimentally measured site-specific amino-acid preferences estimate longer deep branches on phylogenies of influenza virus hemagglutinin.
 This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
@@
 However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
-Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.
+Overall, our work underscores the importance of modeling site-specific amino-acid preferences when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to account for how these preferences shift over time.
 \end{abstract}
@@
 All code used is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. 
-For the main analysis, we used \texttt{phydms}~\citep{hilton2017phydms} version 2.2.dev2 (available at \url{github.com/jbloomlab/phydms}) to filter sequences and estimate branch lengths, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences and \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to infer the pbMutSel model.
+For the main analysis, we used \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2 (available at \url{github.com/jbloomlab/phydms}) to filter sequences and estimate branch lengths, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences and \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to infer the pbMutSel model.
 We also used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization.

### 2018-03-12 manuscript/manuscript.tex
edits to first section of Results

@@
 For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin~\citep{yang1994maximum,yang2000codon}. 
-A gamma distribution ($\Gamma$) with $K=4$ bins is shown here.
+A gamma distribution (denoted by $\Gamma$) with $K=4$ bins is shown here.
 (B) A substitution model's stationary state defines the expected sequence composition after a very long evolutionary time. 
@@
 A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-These models explicitly define a different set of amino-acid preferences at each site in the protein. 
+Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same protein are treated as selectively equivalent).
+Such models explicitly define a different set of amino-acid preferences at each site in the protein. 
 This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
-These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and usually better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
+These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and usually better describe actual evolution than models with only rate variation as assessed by Bayesian or maximum-likelihood criteria~\citep{lartillot2004bayesian, le2008phylogenetic, si2008empirical, wang2008class, rodrigue2010mutation,bloom2014experimentally,bloom2014informed,hilton2017phydms}.
 
@@
 For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and nine parameters describing the nucleotide composition of the stationary state.
-However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
+However, mutation-selection models must additionally specify 19 parameters defining the amino-acid preferences for \emph{each} site (there are 20 amino acids whose preferences are constrained to sum to one).
 This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
-It is challenging to obtain values for these parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
-\skhcomment{Do we need to comment on the tamuri paper?}
-Here we will primarily use experimentally informed codon models~\citep[ExpCM's, \ref{eq:ExpCM};][]{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
-The number of remaining free parameters for an ExpCM is similar to a non-site-specific substitution model.
-An alternative strategy of obtaining parameters for site-specific stationary states via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.
+It is challenging to obtain values for these amino-acid preference parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
+Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preferences parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
+Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
+An alternative strategy of obtaining the amino-acid preference parameters via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.
 
 Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
-Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{rodrigue2016detecting, bloom2017identification}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
-This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models~\citep{haddox2017mapping, rodrigue2014site}. 
+Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific amino-acid preferences~\citep{bloom2017identification,rodrigue2017detecting}.
+This $\omega$ parameter for an ExpCM can be drawn from a statistical distribution (e.g., a gamma distribution) just like for GY94-style models~\citep{rodrigue2014site,haddox2017mapping}. 
 We will denote such models as ExpCM+$\Gamma\omega$.
@@
 The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
-This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, rodrigue2016detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.
+This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, rodrigue2017detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.
 
@@
 
-\section*{Materials and Methods}
+\section*{Methods}
 

### 2018-03-12 manuscript/manuscript.tex
edits to second section of Results

@@
 \subsection*{Effect of stationary state and rate variation on branch-length estimation}
+Given a single branch, a substitution model transforms sequence identity into branch length.
+Under a molecular-clock assumption, this branch length is proportional to time.
+The transformation from sequence identity to branch length is trivial when the sequence identity is high.
+For instance, when there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
+However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
+In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
+Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic amino-acid sequence identity.
+For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be $\frac{1}{20} = 0.05$.
+If the substitution model underestimates the asymptotic sequence identity then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.
 
@@
 
-Given a single branch, a substitution model transforms sequence identity into branch length.
-Under a molecular-clock assumption, this branch length is proportional to time.
-The transformation from sequence identity to branch length is trivial when the sequence identity is high.
-For instance, when there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
-However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
-In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
-Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some amino-acid asymptotic sequence identity.
-For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
-If the substitution model underestimates the asymptotic sequence identity then it will also underestimate long branch lengths, since it will predict that sequences that have evolved for a very long time should be more diverged than is actually the case.
-
-\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
+\ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
 The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
-This model predicts an asymptotic sequence divergence of 0.059, which is slightly higher than 0.050 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
+This model predicts an asymptotic sequence divergence of 0.059, which is slightly higher than $\frac{1}{20} = 0.05$ since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
 Intuitively, this asymptotic sequence identity of 0.059 seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
@@
 Accounting for site-to-site rate variation in GY94 models affects the rate at which the asymptotic sequence identity is approached, but not the actual value of this asymptote. 
-For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but the asymptote for both models is identical. 
+For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but that the asymptote is identical for both models. 
 This fact holds true even if we use experimental measurements of HA's site-specific amino-acid preferences~\citep{doud2016accurate} to calculate a different $\omega_r$ value for each site using the method of \citet{spielman2015relationship} (see \ref{eq:w_r}).
@@
 Therefore, they predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
-For instance, \ref{fig:decay} shows that at sites such as 348 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
+For instance, \ref{fig:decay} shows that at sites such as 183 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
 When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a long branch length at modest sequence identities that a GY94 model might attribute to a shorter branch.

### 2018-03-12 manuscript/manuscript.tex
edits to 3rd section of Results

@@
 
-\subsection*{Simulations demonstrate how failure to model site-specific stationary states leads to branch-length underestimation.}
+\subsection*{Simulations demonstrate how failure to model site-specific amino-acid preferences leads to branch-length underestimation.}
+
+To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
+Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees} using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
+We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
 
@@
 
-To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
-Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees} using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
-We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
-
 The models with a uniform stationary state underestimated the lengths of long branches on the phylogenetic tree of the simulated sequences (\ref{fig:simulations}). 
-The GY94 model estimated branches lengths that are only about 60\% of the true values for the longest branches. 
+The GY94 model estimated branch lengths that are $\sim$60\% of the true values for the longest branches. 
 Accounting for site-to-site variation in $\omega$ did not fix the fundamental problem: the GY94+$\Gamma\omega$ did slightly better, but still substantially underestimated the longest branches.
-However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models, which accurately accounted for the site-specific amino-acid preferences used in the simulations.
-The improved performance of the ExpCM's is due to their site-specific modeling of the stationary state: if we parameterize these models by site-specific amino-acid preferences that have been averaged across HA sites, then they perform no better than GY94 models~\ref{fig:simulations}.
-Therefore, models with uniform stationary states are fundamentally incapable of accurately estimating the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.
+However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models.
+The improved performance of the ExpCM's is due to their modeling of the site-specific amino-acid preferences: if we parameterize ExpCM's by amino-acid preferences that have been averaged across HA sites (and so are no longer site-specific), then they perform no better than GY94 models (\ref{fig:simulations}).
+Therefore, models with uniform stationary states underestimate the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.
 

### 2018-03-13 manuscript/manuscript.tex
revisions to Results 4

@@
 % allow bigger floats per here: https://tex.stackexchange.com/a/11382
-\renewcommand{\topfraction}{.85}
-\renewcommand{\bottomfraction}{.85}
-\renewcommand{\textfraction}{.15}
-\renewcommand{\floatpagefraction}{.85}
+\renewcommand{\topfraction}{.95}
+\renewcommand{\bottomfraction}{.95}
+\renewcommand{\textfraction}{.05}
+\renewcommand{\floatpagefraction}{.95}
 \renewcommand{\dbltopfraction}{.66}
@@
 However, our results make it clear that the site-specific amino-preferences of HA are constant across the entire tree. 
-In particular, the branch lengthening is most pronounced near the sequence of the HA that was used in the deep mutational scanning experiment that parameterized the ExpCM.
+In particular, the branch lengthening is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that parameterized the ExpCM.
 These results underscore the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.
@@
 \caption{\label{fig:empirical_trees}
-\textbf{Effect of site-specific stationary state and $\Gamma\omega$ rate variation on HA branch length estimation.} 
-The branch lengths of the HA tree are optimized using the indicated ExpCM and GY94 models. 
-The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 example sites in HA; the full set of experimentally measured amino-acid preferences defining each ExpCM are shown in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
+\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
+The branch lengths of the HA tree are optimized using the indicated ExpCM or GY94 model. 
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
 The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
-The circle denotes the H1 clade and the triangle denotes the H3 clade. 
+Circle denotes the H1 clade and triangle denotes the H3 clade.
+\jdbcomment{Add a x-axis label (probably can just be for bottom-most plots) giving the units: codon substitutions per site.} 
+\jdbcomment{Add a line to this legend about how trees are rooted, such as: ``Each tree is separately midpoint rooted''---or if that is not what was done, then add something that indicates how it was done.
+If it isn't what was done, can you remind me of rationale for not doing that?}
 }
 \end{figure}
+
+The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM models.
+But how do these models perform on real data?
+Real genes do evolve under functional constraint, but these constraints are almost certainly more complex than what is modeled by an ExpCM.
+However, if ExpCM's do a substantially better job than GY94 models of capturing the true constraints, then we might still expect them to estimate more accurate branch lengths.
+
+To test the models on real data, we used actual sequences of influenza HA. 
+The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
+HA consists of a number of different subtypes.
+Sequences within a subtype have $>$75\% amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
+However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
+We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
+For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
+Although the nucleotide model used with \texttt{RAxML} to infer this tree topology is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
+
+Deep mutational scanning has been used to measure the amino-acid preferences of all sites in two different HAs.
+One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
+The amino-acid preferences measured for these two HAs are shown in \ref{suppfig:prefs_doud} and 
+\ref{suppfig:prefs_lee}.
+The H1 and H3 HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
+As described in \citet{lee2018deep}, the amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites (these differences are apparent in a simple visual comparison of \ref{suppfig:prefs_doud} and
+\ref{suppfig:prefs_lee}).
+Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree.
+These three sets of HA amino-acid preferences define three different ExpCM's.
 
@@
 
-The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM models.
-But how do these models perform on real data?
-Real genes do evolve under functional constraint, but these constraints are almost certainly more complex than what is modeled by an ExpCM.
-However, if ExpCM's do a substantially better job than GY94 models of capturing the true constraints, then we might still expect them to estimate more accurate branch lengths.
+We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
+\ref{tab:empirical_data} shows that all ExpCM's fit the actual data much better than the GY94 models.
+The best fit was for the ExpCM informed by the average of the H1 and H3 deep mutational scans.
+For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ model (\ref{tab:empirical_data}).
+As mentioned in the previous section, $\omega$ is generally $<1$ when a single value is fit to all sites in a gene~\citep{murrell2015gene}, and this is the case for all the models we tested (\ref{tab:empirical_data}).
+However, the ExpCM's always fit an $\omega$ greater than the GY94 model, suggesting that the site-specific amino-acid preferences capture some of the purifying selection that the GY94 models can represent only via a small $\omega$.
+Among the models with $\Gamma\omega$, the GY94+$\Gamma\omega$ model fits all four $\omega$ categories to values $\ll$1, but the ExpCM+$\Gamma\omega$ models fit one of the $\omega$ categories to a value $>1$.
+This increase in $\omega$ values makes sense given the different interpretation of $\omega$ for each family of models. 
+The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the amino-acid preferences.
+This more realistic null model gives ExpCM's enhanced power to detect diversifying selection for amino-acid change~\citep{bloom2017identification, rodrigue2017detecting}, which is known to occur at some sites in HA due to immune selection~\citep{bedford2014integrating}.
 
-To test the models on real data, we used actual sequences of influenza HA. 
-The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
-HA consists of a number of different subtypes.
-Sequences within a subtype have $>75\%$ amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
-However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
-We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
-For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
-Although the nucleotide model used with \texttt{RAxML} is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
+Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or a site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree.
+\ref{fig:empirical_trees} shows the branch lengths optimized by each model on a common scale.
+The tree's deepest branches are shortest when they are optimized by the GY94 model, which lacks both $\Gamma\omega$ and site-specific amino-acid preferences.
+Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
+Specifically, the tree's total height (distance from the two most divergent tips) is \jdbcomment{36\% needs to be updated for tree height} longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_lengths}).
+The total height is \jdbcomment{8\% needs to be updated for tree height} longer for ExpCM's informed by H1 or H3 amino-acid preferences than for GY94, and \jdbcomment{?} longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_lengths}).
+\jdbcomment{I am pretty sure that the standard quantity used in this context is not the average or total branch length, but rather the distance between the two most distant nodes on the tree, which is what I have called ``tree's total height'' here.
+So can you update \ref{supptab:tree_lengths} and the preceding numbers to use that measure instead?}
 
-There are two deep mutational scanning datasets for HA that estimated amino-acid preferences for all sites.
-One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
-These two HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
-The experimentally measured amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites~\citep[][\ref{suppfig:prefs_doud}, 
-\ref{suppfig:prefs_lee}]{lee2018deep}.
-Therefore, we also created a third set of amino-acid preferences by averaging the preferences measured in the deep mutational scanning of the H1 and H3 HAs, under the conjecture that these averaged preferences might do a better job of describing the ``average'' constraint on sites across the full HA tree.
-These three sets of HA amino-acid preferences define three different ExpCM's.
+The deepening of branch lengths that results from the $\Gamma\omega$ and site-specific amino-acid preference approaches to modeling purifying selection are largely independent.
+This can be seen by examining the ExpCM+$\Gamma\omega$ models, which combine $\Gamma\omega$ rate variation with site-specific amino-acid preferences.
+As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
+The near independence of these effects is quantified in \ref{supptab:tree_lengths}, which shows that \jdbcomment{96\%, needs to be fixed if using tree height} of the total tree height extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).
 
-We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
-\ref{tab:empirical_data} shows that all the ExpCM's fit the actual data much better than the GY94 models.
-The best fit was for the ExpCM that was informed by the average of the H1 and H3 deep mutational scans.
-For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ models.
-As mentioned in the previous section, $\omega$ generally takes on values $<1$ when fit to all sites in the gene~\citep{murrell2015gene}, and our results follow this trend. 
-\ref{tab:empirical_data} shows that all models that fit a single $\omega$ fit very small values, reflecting strong purifying selection against amino-acid change in HA. 
-However, the ExpCM's with a single $\omega$ always fit $\omega$ values greater than the GY94 model, suggesting that the site-specific stationary state of the ExpCM's captures some of the purifying selection that the GY94 models can represent only via a small omega.
-Among the models with $\Gamma\omega$, the GY94 fits all four $\omega$ categories to values $<< 1$, but the ExpCM's all fit one of the four $\omega$ categories to a value $>1$.
-This increase in $\omega$ values makes sense given the different interpretation of $\omega$ in each family of models. 
-The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the deep mutational scanning amino-acid preferences.
-This more realistic null model gives ExpCM's more power to detect diversifying selection~\citep{bloom2017identification, rodrigue2017detecting}, for amino-acid change, which occurs at some sites in HA due to immune selection.
-
-Models which account for purifying selection via either $\Gamma\omega$ rate variation or a site-specific stationary state are not just better fit models, they also increase the estimation of branch length on the HA tree.
-As \ref{fig:empirical_trees} shows, the tree optimized by without $\Gamma\omega$ rate variation and without a site-specific stationary state, the GY94 model, estimated the shortest branches of all eight models.
-Specifically, the ExpCM's defined by H1 HA or H3 HA amino-acid preferences estimated a tree which overall length is $\sim 8\%$ longer than the GY94 tree and the GY94+$\Gamma\omega$ model estimated tree which is $\sim 36\%$ longer than the GY94 tree~\ref{supptab:tree_lengths}.
-But importantly, the effects from these two methods of modeling purifying selection are largely independent.
-For every model (GY94 or any of the three ExpCM's), the $\Gamma\omega$ version estimates longer deep branches.
-In addition, all ExpCM's estimate longer deep branches than the GY94 model, and all ExpCM+$\Gamma\omega$ models estimate longer deep branches than the GY94+$\Gamma\omega$ model.
-\skhcomment{I think we can make a stronger statement about the independence of these effects.
-I did the calculation and 96\% of the branch length extension of GY94 vs.
-ExpCM(H1+H3)+$\Gamma\omega$ can be explained by simply adding the ``effect of adding $\Gamma\omega$" (GY94 vs.
-GY94+$\Gamma\omega$) and the ``effect of adding preferences" (GY94 vs.
-ExpCM(H1+H3).
-Do you think that is a useful analysis to include?}
-
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a fashion that appears roughly uniform across the tree (\ref{fig:empirical_trees}), the branch lengthening from the ExpCM's is not uniform across the tree.
-In particular, the branch lengthening is most pronounced near the sequence of the HA that was used in the deep mutational scanning experiment that parameterized the ExpCM.
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}).
+Instead, the branch lengthening is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
 For instance, the ExpCM informed by the H1 data most dramatically lengthens branches near the H1 clade of the tree, while the ExpCM informed by the H3 data has the largest effect on branches near the H3 clade.
-The ExpCM informed by the average of the H1 and H3 data has a more uniform effect, but still most strongly affects branches near either the H1 or H3 clades.
-Therefore, \ref{fig:empirical_trees} shows that the ExpCM's estimate longer branches, but that the effect is biased by the set of preferences used to define the stationary state of the model.
+The ExpCM informed by the average of the H1 and H3 data has a more uniform effect across the tree, but still most strongly extends branches leading to either the H1 or H3 clade.
+Therefore, \ref{fig:empirical_trees} shows that ExpCM's estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.
 
@@
 
-\begin{table}[t!]
+\begin{supptable}[t!]
 \caption{\label{supptab:tree_lengths}
@@
 The branch lengths are in average codon substitutions per site. 
+\jdbcomment{Change to be total tree height, which is distance between two most distant tip nodes.
+Also, to make it easier to relate back to the percentage increases in the text, make 8 rows, one for each model and get rid of the ``rate variation'' column. 
+Then add a new column that gives \% height relative to GY94 (the entry would just be 100\% for GY94).
+So this gives 3 columns: model, total tree height, \% of GY94 tree height}
 } 
@@
          GY94 &53.05 & 72.28\\
-         ExpCM (H1) & 57.32 & 81.31\\
-         ExpCM (H3) & 57.77 & 78.98\\
+         ExpCM(H1) & 57.32 & 81.31\\
+         ExpCM(H3) & 57.77 & 78.98\\
          ExpCM(H1 + H3 avg) & 62.98 & 83.96 \\
       \end{tabular}
-\end{table}
+\end{supptable}
 

### 2018-03-13 manuscript/manuscript.tex
changed "tree height" to "tree diameter" per Erick Matsen

@@
 Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
-Specifically, the tree's total height (distance from the two most divergent tips) is \jdbcomment{36\% needs to be updated for tree height} longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_lengths}).
-The total height is \jdbcomment{8\% needs to be updated for tree height} longer for ExpCM's informed by H1 or H3 amino-acid preferences than for GY94, and \jdbcomment{?} longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_lengths}).
-\jdbcomment{I am pretty sure that the standard quantity used in this context is not the average or total branch length, but rather the distance between the two most distant nodes on the tree, which is what I have called ``tree's total height'' here.
+Specifically, the tree's diameter (distance from the two most divergent tips) is \jdbcomment{36\% needs to be updated for tree diameter} longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_lengths}).
+The tree diameter is \jdbcomment{8\% needs to be updated for tree diameter} longer for ExpCM's informed by H1 or H3 amino-acid preferences than for GY94, and \jdbcomment{?} longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_lengths}).
+\jdbcomment{I am pretty sure that the standard quantity used in this context is not the average or total branch length, but rather the distance between the two most distant nodes on the tree, which is what I have called the ``tree diameter'' here.
+I checked with Erick Matsen, and he said that ``tree diameter'' is the correct term here.
 So can you update \ref{supptab:tree_lengths} and the preceding numbers to use that measure instead?}
@@
 As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
-The near independence of these effects is quantified in \ref{supptab:tree_lengths}, which shows that \jdbcomment{96\%, needs to be fixed if using tree height} of the total tree height extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).
+The near independence of these effects is quantified in \ref{supptab:tree_lengths}, which shows that \jdbcomment{96\%, needs to be fixed if using tree diameter} of the tree diameter extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).
 
@@
 The branch lengths are in average codon substitutions per site. 
-\jdbcomment{Change to be total tree height, which is distance between two most distant tip nodes.
+\jdbcomment{Change to be tree diameter, which is distance between two most distant tip nodes.
 Also, to make it easier to relate back to the percentage increases in the text, make 8 rows, one for each model and get rid of the ``rate variation'' column. 
-Then add a new column that gives \% height relative to GY94 (the entry would just be 100\% for GY94).
-So this gives 3 columns: model, total tree height, \% of GY94 tree height}
+Then add a new column that gives \% diameter relative to GY94 (the entry would just be 100\% for GY94).
+So this gives 3 columns: model, tree diameter, \% of GY94 tree diameter}
 } 

### 2018-03-13 manuscript/manuscript.tex
revisions to Results 5

@@
 
-\subsection*{Competing effects of shifting preferences and long branches.}
+\subsection*{Shifting amino-acid preferences limit the benefits of models with site-specific stationary states for estimating long branch lengths.}
 
@@
 \caption{\label{fig:compete}
-\textbf{The congruency between natural selection and the deep mutational scanning measurements decreases with sequence divergence.} 
+\textbf{The congruence between natural selection and the deep mutational scanning measurements decreases with sequence divergence.} 
 We fit an ExpCM informed by the H1 or H3 deep mutational scanning experiments to trees spanning sequences with low, intermediate, and high divergence from the sequence used in the experiment. 
-The ExpCM stringency parameter ($\beta$) is a measure of the congruency between natural selection and the experimental measurements. 
+The ExpCM stringency parameter ($\beta$) is a measure of the congruence between natural selection and the experimental measurements~\citep{bloom2014informed,hilton2017phydms}. 
 Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with greater stringency. 
-As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases and the preference sets ``flatten."
-Therefore, the preferences measured in each experiment are progressively less congruent with those of the HA sequences on the tree as we include increasingly diverged sequences. 
+As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases and the amino-acid preference ``flatten."
+Therefore, the preferences measured in each experiment are progressively less congruent with natural selection as we include increasingly diverged sequences. 
 }
@@
 
-The fact that an ExpCM has the greatest effect on branches close to the sequence used in the experiment can be rationalized in terms of existing knowledge about how epistasis shifts protein amino-acid preferences during evolution~\citep{pollock2012amino, shah2015contingency, doud2015site, haddox2017mapping, risso2014mutational}.
-In reality, sites in proteins do not have completely conserved site-specific amino-acid preferences. 
-Rather, the effect of a mutation at one site in a protein can depend on the amino-acid identity of other sites in the protein~\citep{starr2018pervasive, pollock2012amino, shah2015contingency, bazykin2015changing}.
-The deep mutational scanning experiments that inform our ExpCM's were performed in the context of a single HA genetic background, and so do not account for how such epistatic interactions can cause amino-acid preferences to shift as substitutions accumulate.
-Therefore, it makes sense that an ExpCM would most accurately describe the evolution of sequences closely related to the one used in the experiment.
+The fact that an ExpCM leads to the most profound lengthening of the branches leading to the sequence used in the experiment can be rationalized in terms of existing knowledge about epistasis during protein evolution.
+Each ExpCM is informed by a single set of experimentally measured amino-acid preferences.
+But in reality, the effect of a mutation at one site in a protein can depend on the amino-acid identities of other sites in the protein~\citep{ortlund2007crystal, gong2013stability, harms2014historical, tufts2014epistasis, starr2018pervasive}.
+This epistasis can lead to shifts in a protein's amino-acid preferences over evolutionary time~\citep{pollock2012amino, doud2015site, shah2015contingency, bazykin2015changing, haddox2017mapping}.
+Because the deep mutational scanning experiments that inform our ExpCM's were each performed in the context of a single HA genetic background, their measurements do not account for the epistatic shifts in amino-acid preferences that accumulate as HA evolves.
+Therefore, an ExpCM is expected to most accurately describe the evolution of sequences closely related to the one used in the experiment.
 
-We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to trees containing increasingly diverged sequences.
-Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge 59\%$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge 46\%$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
+We can observe how shifting amino-acid preferences degrade the accuracy of an ExpCM by fitting the model to trees containing increasingly diverged sequences.
+For both H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains sequences with $\ge$59\% amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge$46\% amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all HAs (which have as little as 38\% identity to the HA in the experiment).
 \ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
 For each subtree, we examined the congruence between site-specific natural selection and the amino-acid preferences measured in the deep mutational scanning experiment. 
-We used the ExpCM stringency parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
-In essence, $\beta$ rescales the deep mutational scanning measurements~(\ref{fig:compete}). 
-Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, indicating a strong congruence between natural selection and the experimental preferences.
+We used the ExpCM stringency parameter $\beta$ to measure this congruence~\citep{bloom2014informed,hilton2017phydms}. 
+Values of $\beta$ that are $>$1 indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, suggesting strong congruence between natural selection and the experimental preferences.
+In contrast, values of $\beta$ that are $<$1 flatten the preferences, suggesting that they provide a relatively poor description of natural selection on the protein.
 
@@
 This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
-Specifically, as the $\beta$ value decreases, the preferences ``flatten" and draw less information from the experiment. 
-At the most extreme, $\beta = 0$, the preferences would be perfectly uniform and look identical to the GY94 preferences in \ref{fig:empirical_trees}.
-In reality, $\beta$ never reaches a value this low, indicating the deep mutational scanning experiments remain somewhat informative about the real purifying selection across the entire swath of HAs. 
-However, the flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
-As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate.
-This fact also helps rationalize why the ExpCM defined by the average of the H1 and H3 experiments, representing the ``average" constraint on HA, does better across the whole tree (\ref{fig:empirical_trees}).
+Specifically, as the $\beta$ value decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
+At the most extreme value of $\beta = 0$, the preferences would be perfectly uniform and look similar to the GY94 preferences in \ref{fig:empirical_trees}.
+In reality, $\beta$ never reaches a value this low, indicating the deep mutational scanning experiments remain somewhat informative about real natural selection across the entire swath of HAs. 
+However, \ref{fig:compete} shows that the amino-acid preferences clearly become less informative about natural selection as we move away from the experimental sequence on the tree.
+This shifting of amino-acid preferences helps explain why the ExpCM informed by the average of the H1 and H3 experiments performs best (\ref{tab:empirical_data} and \ref{fig:empirical_trees}): averaging the measurements across these two HAs is a heuristic method of accounting for shifts in preferences during HA evolution.
 
-The fact that amino-acid preferences shift as proteins evolve leaves us with an inherent tension: models with site-specific stationary states only become important for accurate branch-length estimation as sequences become increasingly diverged, but this same divergence degrades the accuracy of extrapolating the amino-acid preferences from any given experiment across the phylogenetic tree.
-More importantly, the fact that amino-acid preferences shift over time means that there will not be any model with a single set of site-specific stationary states that accurately describes a phylogenetic tree that covers a wide span of sequences.
+The fact that amino-acid preferences shift as a protein evolves leaves us with an inherent tension: models with site-specific amino-acid preferences only become important for accurate branch-length estimation as sequences become increasingly diverged, but this same divergence degrades the accuracy of extrapolating the amino-acid preferences from any given experiment across the phylogenetic tree.
+Crucially, this problem is expected to be more fundamental than the inability of a single deep mutational scanning experiment to measure amino-acid preferences in more than one genetic background.
+If amino-acid preferences shift during evolution, there simply will not be any model with a single set of time-homogeneous site-specific stationary states that accurately describes evolution along the entirety of a phylogenetic tree that covers a wide span of sequences.
 

### 2018-03-13 manuscript/manuscript.tex
finished revisions to Results

@@
 
-\subsection*{Models with stationary states estimated from natural sequences give similar results to ExpCM's}
+\subsection*{A model with amino-acid preferences estimated from natural sequences give similar results to an ExpCM}
 
@@
 \textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
-(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the tree in \ref{fig:empirical_trees}. 
-The pbMutSel amino-acid preferences are inferred from the natural sequences while the ExpCM amino-acid preferences are experimentally measured and then rescaled using the stringency parameter in \ref{tab:empirical_data}.
-The pbMutSel preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual experiments are to each other (Pearson's r: 0.74 vs. 0.52). 
-(B) The pbMutSel and the ExpCM(H1+H3 avg)+$\Gamma\omega$ estimated similar branch lengths. 
-To aid direct comparison between the models, we normalized the branch lengths by the length between and early and late seasonal human H1 sequence. 
+(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
+The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}.
+The pbMutSel preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual H1 and H3 experiments are to each other (Pearson's $r$: 0.74 versus 0.52).
+(B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
+\jdbcomment{You need to mention and give a bit of explanation for the coloring.
+What does it mean for a branch to be ``from'' H1 or H3?
+Would it be better to say branch ``leading to''?
+And does that mean leading to it with respect to the root of the tree as shown in \ref{fig:empirical_trees}?}
+The \texttt{phydms} program implementing ExpCM's and the \texttt{phylobayes} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (\jdbcomment{say exact strain}) and latest (\jdbcomment{say exact strain}) seasonal human H1 sequences on the three.
 }
@@
 The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
-However, other implementations of mutation-selection models allow their stationary states to be inferred directly from the natural sequence data. 
-These models are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of parameters describing site-specific stationary states.
+However, there are other mathematically similar implementations of mutation-selection models that infer the amino-acid preferences directly from the natural sequence data. 
+When these models are designed for use in phylogenetic inference, they are generally implemented in a Bayesian framework, which avoids the overfitting problems associated with trying to make maximum-likelihood estimates of the thousands of amino-acid preference parameters~\citep{lartillot2014overcoming}.
+(Note that the maximum-likelihood implementations of \citet{tamuri2012estimating,tamuri2014penalized} are designed for estimating the amino-acid preferences, \emph{not} for phylogenetic inference.)
 The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as pbMutSel~\citep{rodrigue2014site}. 
-The pbMutSel model is mathematically very similar to the ExpCM's except the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
-Like ExpCM's, pbMutSel models still assume a single set of site-specific amino-acid preferences for the entire tree.
+\jdbcomment{Throughout this section, you refer to \texttt{phylobayes}.
+But are you sure that \texttt{phylobayes} is the same thing as the \texttt{PhyloBayes-MPI} program that implements the pbMutSel models?
+I didn't look into this in great detail, but I got different websites when I Googled the two.
+Check this, and if it's actually \texttt{PhyloBayes-MPI} then update all references to the program to indicate that.}
+In the pbMutSel model, the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
+However, like an ExpCM, a pbMutSel model still assumes a single set of time-homogeneous site-specific amino-acid preferences for the entire tree.
 
-Comparing ExpCM's and pbMutSel models can therefore help identify the ultimate limits of mutation-selection models where each site has a single time-homogeneous stationary state across the entire tree.
-If the limits of the ExpCM's reflected in \ref{fig:empirical_trees} are simply because the deep mutational scanning experiments do not accurately measure the ``true" amino-acid preferences of HA across the tree, then we would expect the pbMutSel models (which derive their stationary states from the entire tree) to perform better.
-On the other hand, if the major limitation is simply that there is not any single set of time-homogenous amino-acid preferences that describe HA evolution over the entire tree, then we would expect pbMutSel and ExpCM models to perform similarly.
+Comparing ExpCM and pbMutSel models can help determine the ultimate limits of mutation-selection models that assign each site a single set of amino-acid preferences. 
+If the limitations of ExpCM's described above arise simply because the deep mutational scanning experiments do not correctly measure the ``true" amino-acid preferences of HA across the entirety of a highly diverged phylogenetic tree, then we would expect the pbMutSel models (which infer these preferences from the entire tree) to perform better.
+On the other hand, if the major limitation is that no single set of time-homogenous amino-acid preferences can fully describe HA evolution over the entire tree, then we would expect ExpCM and pbMutSel models to perform similarly.
 
-The ExpCM that estimates the longest branches is the one that uses the average of the H1 and H3 deep mutational scanning measurements with $\Gamma\omega$ rate variation~(\ref{fig:empirical_trees}).  
-We therefore compared this ExpCM with a pbMutSel model given the tree in \ref{fig:empirical_trees}. 
-First, \ref{fig:phylobayes}A shows that the stationary state inferred by pbMutSel is similar to the average of the H1 and H3 preferences measured by deep mutational scanning. 
-The site-specific amino-acid preferences estimated by pbMutSel are more correlated with the average amino-acid preferences from the H1 and H3 experiments than the individual deep mutational scanning experiments are to each other.
-This strong correlation between the pbMutSel and the average deep mutational scanning preference set indicates that, when averaged, the experimental preferences are able to capture the global site-specific functional constraint fairly well.
+We fit a pbMutSel model to the entire HA phylogenetic tree, and compared the results to those from analyzing the same tree with the best ExpCM, which is the ExpCM(H1+H3 avg)+$\Gamma\omega$ variant.
+This is a direct apple-to-apples comparison, since the pbMutSel model also draws $\omega$ from a gamma-distribution.
+\jdbcomment{Is the previous sentence correct? If so, can you add citation for it?}
+First, we compared the amino-acid preferences inferred by the pbMutSel model to the preferences measured in the experiments.
+\ref{fig:phylobayes}A shows that the preferences inferred by pbMutSel are quite similar to the (H1+ H3 avg) obtained by averaging the deep mutational scanning measurements for the H1 and H3 HAs. 
+Notably, the amino-acid preferences from the pbMutSel model are more correlated with the (H1+ H3 avg) than the H1 and H3 measurements are with each other (\ref{fig:phylobayes}A).
+This strong correlation indicates that the ExpCM(H1+H3 avg)+$\Gamma\omega$ is unlikely to be much different than a pbMutSel model that is parameterized only using the natural sequence data.
 
-Second, the results in \ref{fig:phylobayes}B show that the ExpCM+$\Gamma\omega$(H1+H3 avg prefs) and the pbMutSel model also estimated similar branch lengths on the HA tree in \ref{fig:empirical_trees}.
-While overall the two models estimated similar lengths, the tension between local and global accuracy of the amino-acid preferences is still apparent in these results. 
-All the long branches from either the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM+$\Gamma\omega$, while many other branches were estimated to be slightly longer by the pbMutSel model.
-The relatively longer branches from the experimental sequences when using the ExpCM+$\Gamma\omega$ suggests that the ``global" stationary state inferred by \texttt{phylobayes} is not as accurate as the deep mutational scanning preferences for sequences close to the experimental sequences.
-However, at sequences distant from those used in the experiments, the ``global'' preferences estimated by the pbMutSel model appear to be slightly better than the average of the experimental values.
-Therefore, while the site-specific pbMutSel models inferred by \texttt{phylobayes} are probably more accurate for sequences distant from the H1 and H3 in the experiments, they do not avoid the tension between the importance of a site-specific stationary state for long branch estimation and shifting preferences.
+We next compared the branch lengths estimated by using the ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models.
+As shown in \ref{fig:phylobayes}B, these two models estimated similar branch lengths across the entire HA phylogenetic tree. 
+However, the estimates are not identical, and the tension between local and global accuracy of the amino-acid preferences is still apparent. 
+Specifically, the long branches leading to the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM, while many other branches were estimated to be slightly longer by the pbMutSel model.
+The relatively longer branches leading to the experimental sequences when using the ExpCM(H1+H3 avg)+$\Gamma\omega$ suggests that the ``tree average" amino-acid preferences inferred by the pbMutSel model are not as accurate as the preferences from the deep mutational scanning for sequences close to those used in the experiments.
+However, for sequences distant from those used in the experiments, the ``tree average'' preferences inferred by the pbMutSel model appear to be slightly better than the experimental values.
+Therefore, while the ExpCM and pbMutSel models differ slightly in the extent to which they lengthen different branches, neither model can avoid the tension between the local and global accuracy of amino-acid preferences.
+
+\subsection*{\jdbcomment{Possible new section?}}
+\jdbcomment{We should think about what happens if we try an ExpCM just on the set of sites that Juhye's paper found did \emph{not} substantially shift in amino-acid preferences between H1 and H3.}
 

### 2018-03-13 manuscript/manuscript.tex
edits to Methods and SI

@@
 The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}.
-The pbMutSel preferences are more correlated with the average of the H1 and H3 deep mutational scanning preference sets than the individual H1 and H3 experiments are to each other (Pearson's $r$: 0.74 versus 0.52).
+The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52).
 (B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
@@
 
-\subsection*{Models}
+\subsection*{Substitution models}
+All of the substitution models used in this paper have been described previously.
+However, here we briefly recap their exact mathematical implementations.
 
-\subsubsection*{GY94}
+\subsubsection*{GY94 model}
 
-The GY94 model is defined as the M0 in \cite{yang2000codon}. 
-$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by
-
+The GY94 model is M0 variant of the Goldman-Yang model described by \citet{yang2000codon}. 
+Specifically, the substitution rate $P_{xy}$ from codon $x$ to codon $y$ is 
 \begin{equation}
@@
   0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
-  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
-  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
-  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
+  \Phi_y & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transversion,} \\
+  \omega \Phi_{y} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transversion,} \\
+  \kappa \Phi_y & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transition,} \\
+  \omega \kappa \Phi_{y} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$ and $x$ is converted to $y$ by a single-nucleotide transition,} \\
+  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$,}
   \end{cases}
 \end{equation}
-where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
-codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.
-The stationary state of a GY94 model is defined as 
+where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\kappa$ is the transition-transversion rate, $\Phi_y$ is the equilibrium frequency of codon $y$, and $\omega$ is the relative rate of nonsynonymous and synonymous substitutions.
+\jdbcomment{I have dropped the $\mu$ from the equation above since it is confounded with branch lengths and so not estimated separately for any of our models.
+Also, the equation wasn't right as the $\omega$ term should only be there for nonsynonymous substitutions.
+Also, you need to add a line specifying how the $\Phi_y$ terms are defined. Probably the CF3X4 method? Add this, a proper citation, and one-sentence description.}
+The frequency $p_x$ of codon $x$ in the stationary state of a GY94 model is simply 
 \begin{equation}
 \label{eq:px}
-p_{x} = \Phi_x
+p_{x} = \Phi_x.
 \end{equation}
+Overall, a GY94 model has 11 free parameters: $\kappa$, $\omega$, and the 9 nucleotide frequency parameters used to define $\Phi_y$.
 
@@
 
-In an ExpCM, the rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
+The ExpCM models used in this paper are the ones described in \citet{bloom2017identification}. 
+Briefly, the rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is 
 \begin{equation}
@@
 where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
-The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form.
+The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form as described in \citet{bloom2017identification}.
+\jdbcomment{You need to add a more precise description of this because otherwise it doesn't make any sense when you refer to the $\phi_x$ parameters two paragraphs below if they have never been defined in this context.}
 
@@
 The experiments measure the preference $\pi_{r,a}$ of every site $r$ for every amino-acid $a$.
-The $F_{r,xy}$ terms are defined in terms of these experimentally measured amino-acid preferences as
+$F_{r,xy}$ is defined in terms of these experimentally measured amino-acid preferences as
 \begin{equation}
@@
 \begin{cases}
-   1 & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$} \\
-   \omega \times \frac{\ln\left[\left(\pi_{r,\mathcal{A}\left(y\right)} / \pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta}\right]}{1 - \left(\pi_{r,\mathcal{A}\left(x\right)} / \pi_{r,\mathcal{A}\left(y\right)}\right)^{\beta}} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$}
+   1 & \mbox{if $\mathcal{A}\left(x\right) = \mathcal{A}\left(y\right)$,} \\
+   \omega \times \frac{\ln\left[\left(\pi_{r,\mathcal{A}\left(y\right)} / \pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta}\right]}{1 - \left(\pi_{r,\mathcal{A}\left(x\right)} / \pi_{r,\mathcal{A}\left(y\right)}\right)^{\beta}} & \mbox{if $\mathcal{A}\left(x\right) \ne \mathcal{A}\left(y\right)$,}
    \end{cases}
 \end{equation}
-where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\beta$ is the stringency parameter, and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
+where $\beta$ is the stringency parameter~\citet{bloom2014informed,hilton2017phydms} and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
 
-The stationary state of an ExpCM is defined as 
+The stationary state of is 
 \begin{equation}
@@
 
-The preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined by an experiment independent of the sequence alignment being analyzed.
-The ExpCM is described in detail in~\citet{hilton2017phydms}.
+An ExpCM has five free parameters: $\kappa$, $\omega$, and the three independent $\phi_x$ values.
+The amino-acid preferences $\pi_{r,a}$ are \emph{not} free parameters since they are determined \textit{a priori} by an experiment independent of the sequence alignment being analyzed.
 
-\subsubsection*{GY94 with $\omega_r$.}
+\subsubsection*{$\Gamma\omega$ rate variation}
 
-We defined a site-specific model with the GY94 stationary state by multiplying the $\omega$ in \ref{eq:GY94} with a site-specific $\omega_r$ value calculated from an ExpCM (\ref{eq:ExpCM}). 
-Specifically, following \citet{spielman2015relationship}, 
+The GY94+$\Gamma\omega$ is equivalent to the M5 model in \citet{yang2000codon} with $\omega$ drawn from $K = 4$ categories.
+The ExpCM+$\Gamma\omega$ similarly draws $\omega$ from a $\Gamma$ distribution discretized into $K=4$ bins. 
+Each bin is equally weighted and $\omega$ takes on the mean value of the bin. 
+Because the $\Gamma$ distribution is defined by two parameters, adding $\Gamma\omega$ to a model with a single $\omega$ adds one free parameter.
+Therefore, the GY94+$\Gamma\omega$ model has 12 free parameters, and the ExpCM+$\Gamma\omega$ model has 6 free parameters.
+
+\subsubsection*{GY94 with $\omega_r$}
+In \ref{fig:decay}, we describe GY94 models where each site $r$ has its own $\omega_r$ value that is calculated from the amino-acid preferences using the relationship described by \citet{spielman2015relationship}.
+This relationship gives the expected rate of nonsynonymous to synonymous substitutions given the amino-acid preferences.
+Specifically, we first fit an ExpCM, which allows us to calculate $P_{r,xy}$ (\ref{eq:ExpCM}), $Q_{xy}$ (\jdbcomment{as noted above, you need to give this equation in the ExpCM section}), and $p_{r,x}$ (\ref{eq:prx}).
+We then calculated $\omega_r$ using the equation of \citet{spielman2015relationship}, normalizing by the gene-wide $\omega$ fit by the ExpCM:
 \begin{equation}
 \label{eq:w_r}
-\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times \frac{P_{r,xy}}{\omega}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
+\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times \frac{P_{r,xy}}{\omega}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}},
 \end{equation}
-where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, $\omega$ is the relative nonsynonymous substitution rate from the ExpCM and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.
+where $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.
 
-\subsection*{$\Gamma\omega$ rate variation.}
+\subsection*{HA amino-acid preferences from deep mutational scanning experiments}
+We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} to define the amino-acid preferences that inform the ExpCM's.
+We only used sites that can be unambiguously aligned in these H1 and H3 HAs. 
+These alignable sites and their mapping to sequential numbering of the HA sequences used in the deep mutational scanning experiments are in \ref{suppfile:WSN_Perth_map}. 
+The experimentally measured amino-acid preferences masked to just include these alignable sites are in \ref{suppfile:H1_prefs} and \ref{suppfile:H3_prefs}.
+\jdbcomment{Is this right, do the files just have the masked preferences rather than the full set of sites?
+Also, as described below, we are running up against a limit in supplementary files.
+So rather than also giving a file with the averaged values, just add a few sentences clearly explaining how the averaging was done.}
 
-The GY94+$\Gamma\omega$ is defined as M5 in \citet{yang2000codon} and the ExpCM+$\Gamma\omega$ follows the same logic. 
-Briefly, $\omega$ is drawn from a $\Gamma$ distribution discretized into $K=4$ bins. 
-Each bin is equally weighted and $\omega$ takes on the mean value of the bin.
-
-\subsection*{Deep mutational scanning amino-acid preferences.}
-We used amino-acid preferences measured in the deep mutational scans of an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the ExpCM $\pi_{r, A\left(x\right)}$ parameters~(\ref{eq:Frxy}).
-We only considered sites which are homologous between H1 HA and H3 HA. 
-These homologous sites, and their mapping between the H1 reference strain A/Wilson Smith/1933 and the H3 reference strain a A/Perth/2009, are described in \ref{suppfile:WSN_Perth_map}. 
-The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.
-
-\subsection*{Influenza virus hemagglutinin (HA) sequences and tree topology.}
+\subsection*{HA sequences and tree topology.}
 
 We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18.
+\jdbcomment{Add a sentence indicating \emph{why} you excluded these subtypes. Also you need to say where you downloaded the sequences from and give a date.} 
 We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
 Specifically, we used \texttt{phydms\_prepalignment} with the flag \texttt{--minidentity 0.3} to remove sequences with ambiguous nucleotides, premature stops, or frameshift mutations as well as redundant sequences.
-We also removed any codon sites which are not shared between the H1 HA and H3 HA used in the deep mutational scanning experiments. 
-We subsampled the remaining sequences to five sequences per subtype with $\le 1$ sequence per year per subtype. 
+We also removed all codon sites which that are not alignable between the H1 HA and H3 HA used in the deep mutational scanning experiments (these alignable sites are listed in  \ref{suppfile:WSN_Perth_map}).
+We subsampled the remaining sequences to five per subtype with $\le 1$ sequence per year per subtype. 
 We also included a small number of representative sequences from the major human and equine influenza lineages, described in \ref{suppfile:required_seqs}. 
-The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.
+\jdbcomment{We need to trim supplementary files, so get rid of \ref{suppfile:required_seqs} and put an approximate justification / explanation here.}
+The resulting alignment contains 87 sequences, and is provided in \ref{suppfile:alignment_high}.
 
-We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence. 
-The ``low divergence" alignments had   $\ge 59\%$ amino-acid identity from the reference sequence and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence.
+We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence \jdbcomment{for the analysis in Figure ?}. 
+The ``low divergence" alignments had $\ge 59\%$ amino-acid identity to the sequence used in the deep mutational scanning, and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence.
 The resulting alignments are found in \ref{suppfile:alignment_lowH1}, \ref{suppfile:alignment_lowH3}, \ref{suppfile:alignment_intermediateH1}, and \ref{suppfile:alignment_intermediateH3}.  
+\jdbcomment{Again, we should probably get rid of these four preceding files since we have too many. Presumably they can be re-generated using the description above of the identity cutoffs?}
 
 We inferred the tree topology of each alignment using \texttt{RAxML}~\citep{stamatakis2006raxml} and the GTRCAT model. 
-We estimated the branch lengths of this fixed topology using the ExpCM and GY94 models with \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
+We estimated the branch lengths of this fixed topology using each ExpCM and GY94 models with \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
 
-\subsection*{Asymptotic sequence divergence.}
+\subsection*{Asymptotic amino-acid sequence identity}
+For the analysis in \ref{fig:decay}, we fit models to the ``low divergence" H1 subtree.
+\jdbcomment{This gave the parameter values in Supplementary table... (You need to put these into one or more supplementary tables, or just list them here as we have too many Supplementary files.}
 
-We defined ExpCM and GY94 models using parameter values fit to the ``low divergence from H1" tree~(\ref{fig:empirical_trees}, \ref{suppfile:alignment_lowH1}). 
-Specifically, we defined the GY94 model using the parameter values in~\ref{suppfile:hybrid_lowH1_GY94}, the GY94+$\Gamma\omega$ using the parameter values in~\ref{suppfile:hybrid_lowH1_GY94+Gr} and the ExpCM using the parameter values in~\ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from~\citet{doud2016accurate} in~\ref{suppfile:H1_prefs}.
-
-For each model, we calculated the expected sequence divergence given branch length $t$ by 
+For each model, we calculated the expected amino-acid sequence identity for two sequences separated by a branch length of $t$ as 
 \begin{equation}
 \label{eq:f}
-\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
+\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} \left[e^{t\mathbf{P_r}}\right]_{xy}
 \end{equation}
-where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.
+where $a$ ranges over all 20 amino acids, $x \in a$ indicates that $x$ ranges over all codons that encode amino-acid $a$, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$ (given by \ref{eq:px} for GY94-family models, and \ref{eq:prx} for ExpCM-family models), and $\left[e^{t\mathbf{P_r}}\right]_{xy}$ is the value in row $x$ and column $y$ of the matrix obtained by exponentiating the product of $t$ and the substitution matrix $\mathbf{P_r}$ for site $r$ (defined by \ref{eq:GY94} for GY94-family models and \ref{eq:ExpCM} for ExpCM-family models).
 
 \subsection*{Simulations}
+For \ref{fig:simulations}, we simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along the full HA tree using an ExpCM defined by parameters fit to the ``low divergence'' H1 subtree.
 
-We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along the tree for the full alignment (\ref{fig:empirical_trees}, \ref{suppfile:alignment_high}) using an ExpCM defined by parameters fit to the ``low divergence from H1 tree" (\ref{suppfile:alignment_lowH1}).
-Specifically, we defined the ExpCM using the parameter values in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the H1 amino-acid preferences from \citet{doud2016accurate} in \ref{suppfile:H1_prefs}. 
 We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
@@
 \subsection*{pbMutSel inference with \texttt{phylobayes}.}
+For \ref{fig:phylobayes}, we fit a pbMutSel model to the full HA tree. 
+We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
+\jdbcomment{Presumably the program gives some measure of convergence? If so, describe what it was.}
+\jdbcomment{You need to better describe what the tree represents and what the pbMutSel preferences represent. Are they the posterior mean?}
 
-We inferred a pbMutSel model for the full tree (\ref{fig:empirical_trees}, \ref{suppfile:alignment_high}). 
-We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
-We normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
-These two H1 sequences are early and late representatives of the longest known human influenza lineage.
+In order to make the branch lengths in \ref{fig:phylobayes} comparable between the pbMutSel tree returned by \texttt{phylobayes} and the other trees returned by \texttt{phyddms}, we we normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
+These two H1 sequences are early and late representatives of the longest known human influenza lineage, and are of sufficiently high identity that different ExpCM and GY94 substitution models all estimate nearly identical branch lengths separating them.
 
-\subsection*{Code and analysis.}
-All code used is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. 
-For the main analysis, we used \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2 (available at \url{github.com/jbloomlab/phydms}) to filter sequences and estimate branch lengths, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences and \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to infer the pbMutSel model.
-We also used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization.
+\subsection*{Software versions and computer code}
+All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. 
+The external computer programs that we used were
+\begin{itemize} 
+\item \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2 (available at \url{github.com/jbloomlab/phydms}) to fit the ExpCM and GY94 models.
+\item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
+\item \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
+\item \jdbcomment{RAxML}
+\item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
+\end{itemize}
 
@@
 \section*{Supplemental Information}
+
+\jdbcomment{The MBE instructions appear to indicate that you can have at most five \emph{files} (see \url{https://academic.oup.com/mbe/pages/Supplementary_Information}). 
+I would take this to mean that we can have one merged PDF with figures / tables, plus four supplementary files.
+Looking at existing papers seems to confirm this.
+Note also that they have some limits on acceptable file formats, although you can turn FASTA and CSV into the allowable TXT just by adding that extension, as in \texttt{alignment.fasta.txt} and they won't complain.
+So you need to get rid of all but 4 supplementary files.
+
+I would say keep the map of site numbers, the H1 and H3 preferences, and the alignment. 
+I think all the other files can either be re-generated using the other ones if you provide adequate instructions in the methods (e.g., averaging the preferences or pruning the full alignment to sub-alignments), or simply have model parameters which could be put in a supplementary table.
+
+Also, put the supplementary information after the references.
+
+Also, put a page break after each supplementary figure / table using the \texttt{clearpage} command.
+}
 
@@
 \textbf{H1 HA amino-acid preferences measured by deep mutational scanning.}
-Each column represents a site in the protein and the height of each letter is proportional to the preference for the amino acid as measured by~\citet{doud2016accurate}. 
-Here the preferences are rescaled by the ExpCM stringency parameter in \ref{tab:empirical_data} ($\beta=1.12$). 
-We only considered homologous sites between H1 and H3. 
-The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 is found in \ref{suppfile:WSN_Perth_map}. 
+Each column represents a site in the HA protein, and the height of each letter is proportional to the preference for the amino acid measured by~\citet{doud2016accurate} and then re-scaled by the stringency parameter in \ref{tab:empirical_data}.
+The plot only shows sites that are alignable between the H1 and H3 HAs, and these alignable sites are numbered sequentially starting from 1.
+The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence is in \ref{suppfile:WSN_Perth_map}. 
+\jdbcomment{For this figure and the other logo plots, increase the number of rows by a few so things are a bit bigger.}
 }
@@
 \textbf{H3 HA amino-acid preferences measured by deep mutational scanning.}
-Each column represents a site in the protein and the height of each letter is proportional to the preference for the amino acid as measured by~\citet{lee2018deep}. 
-Here the preferences are rescaled by the ExpCM stringency parameter in \ref{tab:empirical_data} ($\beta=1.28$). 
-We only considered homologous sites between H1 and H3. 
-The conversion between the numbering scheme in this figure and sequential numbering of the H3 HA reference sequence A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}. 
+Similar to \ref{suppfig:prefs_doud} but shows the re-scaled preferences for the H3 HA as measured by \citet{lee2018deep}.
  }
@@
 \textbf{Average of H1 HA  and H3 HA amino-acid preferences measured by deep mutational scanning.}
-Each column represents a site in the protein and the height of each letter is proportional to the average preference for the amino acid measured by~\citet{doud2016accurate} and~\citet{lee2018deep}. 
-Here the preferences are rescaled by the ExpCM stringency parameter in \ref{tab:empirical_data} ($\beta=1.70$). 
-We only considered homologous sites between H1 and H3. 
-The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 or H3 HA reference sequence A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}.
+Similar to \ref{suppfig:prefs_doud} but shows the re-scaled average of the preferences for the H1 and H3 HAs.
 }
@@
 \textbf{Amino-acid preferences inferred by the pbMutSel model.}
-Each column represents a site in the protein and the height of each letter is proportional to the preference for the amino acid inferred by pbMutSel~\citep{rodrigue2014site} using the tree in \ref{fig:empirical_trees}.
-We only considered homologous sites between H1 and H3. 
-The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 or H3 HA reference sequence A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}.
+Similar to \ref{suppfig:prefs_doud}, but shows the preferences inferred by fitting the pbMutSel model to the full HA tree.
 }
@@
 \textbf{Overall divergence for the subtrees in \ref{fig:compete}.} 
+\jdbcomment{This legend is indecipherable because rather than saying what you are showing you just refer to a lot of other figures / files.
+Reduce the references to other figures / files to the minimum needed an explain what is actually being shown.}
 We created two sub alignments from the alignment (\ref{suppfile:alignment_high}) used in \ref{fig:empirical_trees} for each deep mutational scanning reference HA. 
@@
 \label{suppfile:WSN_Perth_map}
-Map for homologous sites between H1 HA and H3 HA. 
-This file provides a conversion between the numbering scheme we use in the paper, including all logoplots and the sites in \ref{fig:decay}, to sequential numbering in either the H1 HA reference sequence A/Wilson Smith/1933 or the H3 HA reference sequence A/Perth/2009.
+List of alignable sites between H1 HA and H3 HA. 
+This file provides a conversion between the numbering scheme we use in the paper (sequential numbering of just the alignable sites) to sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 and the H3 HA reference sequence A/Perth/2009.
 The sequential numbering for the HA strains starts with ``site 1" as the second codon (first non-methionine codon). 
+\jdbcomment{The last sentence is so unclear that even I'm not sure what it means.
+If this isn't apparent from the file, explain it much better!
+Wouldn't it be simpler if you just used numbering that started with 1 at the first codon?
+Also, add the actual file so we can see if it makes sense.}
 }
@@
 \label{suppfile:H1_prefs}
-Amino acid preferences measured by the deep mutational scanning of the H1 HA strain A/Wilson Smith/1933~\citep{doud2016accurate}. 
-This file only contains measurements for the homologous sites between H1 HA and H3 HA. 
-Conversion from this numbering scheme to sequential numbering in A/Wilson Smith/1933 is found in \ref{suppfile:WSN_Perth_map}. 
+Amino acid preferences measured by the deep mutational scanning of the H1 HA strain A/WSN/1933~\citep{doud2016accurate}. 
+This file only contains measurements for the alignable sites between H1 and H3 HAs. 
+Conversion from this numbering scheme to sequential numbering of A/WSN/1933 is in \ref{suppfile:WSN_Perth_map}. 
 }
@@
 Amino acid preferences measured by the deep mutational scanning of the H3 HA strain A/Perth/2009~\citep{lee2018deep}. 
-This file only contains measurements for the homologous sites between H1 HA and H3 HA. 
-Conversion from this numbering scheme to sequential numbering in A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}. 
+This file only contains measurements for the alignable sites between H1 and H3 HAs. 
+Conversion from this numbering scheme to sequential numbering of A/Perth/2009 is in in \ref{suppfile:WSN_Perth_map}. 
 }
@@
 We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
+\jdbcomment{This legend is not very good. If the file just has the alignment, then why are we talking about RAxML in its legend?
+Also, I think referring to \ref{fig:empirical_trees} to refer to the tree topology is not very clear, since that figure shows lots of stuff beyond just one tree.}
 }

### 2018-03-13 manuscript/manuscript.tex
small wording tweaks

@@
 And does that mean leading to it with respect to the root of the tree as shown in \ref{fig:empirical_trees}?}
-The \texttt{phydms} program implementing ExpCM's and the \texttt{phylobayes} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (\jdbcomment{say exact strain}) and latest (\jdbcomment{say exact strain}) seasonal human H1 sequences on the three.
+The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the three.
 }
@@
 (Note that the maximum-likelihood implementations of \citet{tamuri2012estimating,tamuri2014penalized} are designed for estimating the amino-acid preferences, \emph{not} for phylogenetic inference.)
-The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{phylobayes}, which we will refer to as pbMutSel~\citep{rodrigue2014site}. 
-\jdbcomment{Throughout this section, you refer to \texttt{phylobayes}.
-But are you sure that \texttt{phylobayes} is the same thing as the \texttt{PhyloBayes-MPI} program that implements the pbMutSel models?
-I didn't look into this in great detail, but I got different websites when I Googled the two.
-Check this, and if it's actually \texttt{PhyloBayes-MPI} then update all references to the program to indicate that.}
+The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{PhyloBayes-MPI}, which we will refer to as pbMutSel~\citep{rodrigue2014site}. 
 In the pbMutSel model, the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
@@
 We fit a pbMutSel model to the entire HA phylogenetic tree, and compared the results to those from analyzing the same tree with the best ExpCM, which is the ExpCM(H1+H3 avg)+$\Gamma\omega$ variant.
-This is a direct apple-to-apples comparison, since the pbMutSel model also draws $\omega$ from a gamma-distribution.
-\jdbcomment{Is the previous sentence correct? If so, can you add citation for it?}
+This is a direct apple-to-apples comparison, since the pbMutSel model also draws $\omega$ from a gamma-distribution~\citep{rodrigue2014site}.
 First, we compared the amino-acid preferences inferred by the pbMutSel model to the preferences measured in the experiments.
@@
 
-We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence \jdbcomment{for the analysis in Figure ?}. 
+We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence for the analysis in \ref{fig:compete}. 
 The ``low divergence" alignments had $\ge 59\%$ amino-acid identity to the sequence used in the deep mutational scanning, and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence.
@@
 
-\subsection*{pbMutSel inference with \texttt{phylobayes}.}
+\subsection*{pbMutSel inference with \texttt{PhyloBayes-MPI}.}
 For \ref{fig:phylobayes}, we fit a pbMutSel model to the full HA tree. 
@@
 
-In order to make the branch lengths in \ref{fig:phylobayes} comparable between the pbMutSel tree returned by \texttt{phylobayes} and the other trees returned by \texttt{phyddms}, we we normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
+In order to make the branch lengths in \ref{fig:phylobayes} comparable between the pbMutSel tree returned by \texttt{PhyloBayes-MPI} and the other trees returned by \texttt{phydms}, we we normalized the branch lengths on the pbMutSel consensus tree and the ExpCM(H1+H3 avg)+$\Gamma\omega$ by dividing each branch by the length from A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
 These two H1 sequences are early and late representatives of the longest known human influenza lineage, and are of sufficiently high identity that different ExpCM and GY94 substitution models all estimate nearly identical branch lengths separating them.
@@
 \subsection*{Software versions and computer code}
-All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. 
+All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. \skhcomment{make repo public}
 The external computer programs that we used were
@@
 \item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
-\item \texttt{phylobayes}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
-\item \jdbcomment{RAxML}
+\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
+\item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML} to infer tree topology.
 \item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
 \end{itemize}
+
+\clearpage 
+\bibliographystyle{mbe}
+\bibliography{references.bib}
 
@@
 I think all the other files can either be re-generated using the other ones if you provide adequate instructions in the methods (e.g., averaging the preferences or pruning the full alignment to sub-alignments), or simply have model parameters which could be put in a supplementary table.
-
-Also, put the supplementary information after the references.
-
-Also, put a page break after each supplementary figure / table using the \texttt{clearpage} command.
 }
@@
 \end{suppfig}
+\clearpage
 
@@
 \end{supptable}
+\clearpage
 
@@
 \end{suppfile}
+\clearpage
 
@@
 \end{suppfile}
-
 \clearpage
-\bibliographystyle{mbe}
-\bibliography{references.bib}
 

### 2018-03-13 manuscript/manuscript.tex
increased number of logo plot rows

@@
 The branch lengths of the HA tree are optimized using the indicated ExpCM or GY94 model. 
-The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logoplots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
 The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
@@
 The conversion between the numbering scheme in this figure and sequential numbering of the H1 HA reference sequence is in \ref{suppfile:WSN_Perth_map}. 
-\jdbcomment{For this figure and the other logo plots, increase the number of rows by a few so things are a bit bigger.}
 }

### 2018-03-15 manuscript/manuscript.tex
added x-axis label to empirical trees plot
began to reorganize supp files

@@
 Circle denotes the H1 clade and triangle denotes the H3 clade.
-\jdbcomment{Add a x-axis label (probably can just be for bottom-most plots) giving the units: codon substitutions per site.} 
 \jdbcomment{Add a line to this legend about how trees are rooted, such as: ``Each tree is separately midpoint rooted''---or if that is not what was done, then add something that indicates how it was done.
@@
 
+\begin{supptable}[t!]
+\caption{\label{supptab:fit_params}
+{\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
+The branch lengths are in average codon substitutions per site. 
+} 
+     \begin{tabular}{ccc}
+        \hline
+         Model & Parameters \\ \hline\hline
+         GY94 &  {\shortstack{$\kappa = 3.17$, $\omega = 0.10$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.13$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$,  $\phi_{2G} = 0.19$,\\ $\phi_{3, A} = 0.35$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
+         GY94 + $\Gamma\omega$ & {\shortstack{$\alpha_\omega = 0.51$, $\beta_\omega = 3.92$, $\kappa = 3.49$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.14$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$, $\phi_{2,G} = 0.20$, \\$\phi_{3,A} = 0.355$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
+         ExpCM(H1) &  {\shortstack{$\beta = 1.22$, $\kappa = 3.24$, $\omega = 0.13$,\\ $\phi_A = 0.37$, $\phi_C = 0.18$, $\phi_G = 0.23$}}\\ \hline
+      \end{tabular}
+\end{supptable}
+\clearpage
+
 %%%
@@
 \end{suppfile}
-\clearpage
 
@@
 \end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:avg_prefs}
-Average of the amino acid preferences measured by the deep mutational scanning of the H1 HA strain A/Wilson Smith/1933~\citep{doud2016accurate} and the H3 HA strain A/Perth/2009~\citep{lee2018deep}. 
-This file only contains measurements for the homologous sites between H1 HA and H3 HA. 
-Conversion from this numbering scheme to sequential numbering in A/Wilson Smith/1933 or A/Perth/2009 is found in \ref{suppfile:WSN_Perth_map}. }
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:required_seqs}
-This file contains the representative sequences of major human equine influenza lineages added to the alignment in \ref{suppfile:alignment_high}. 
-}
-\end{suppfile}
-\clearpage
 
@@
 \end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:alignment_lowH1}
-The HA sequences using for the ``low divergence from H1" tree in \ref{fig:compete}. 
-All sequences have $\ge 59\%$ amino-acid identity to the H1 reference sequence A/Wilson Smith/1933. 
-We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
-}
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:alignment_lowH3}
-The HA sequences using for the ``low divergence from H3" tree in \ref{fig:compete}. 
-All sequences have $\ge 59\%$ amino-acid identity to the H3 reference sequence A/Perth/2009. 
-We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
-}
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:alignment_intermediateH1}
-The HA sequences using for the ``intermediate divergence from H1" tree in \ref{fig:compete}. 
-All sequences have $\ge 46\%$ amino-acid identity to the H1 reference sequence A/Wilson Smith/1933. 
-We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
-}
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:alignment_intermediateH3}
-The HA sequences using for the ``intermediate divergence from H3" tree in \ref{fig:compete}. 
-All sequences have $\ge 46\%$ amino-acid identity to the H3 reference sequence A/Perth/2009. 
-We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
-}
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:hybrid_lowH1_GY94}
-The GY94 model parameters fit to the ``low divergence from H1" tree (\ref{fig:compete}, \ref{suppfile:alignment_lowH1}). 
-The model was fit using \texttt{phydms\_comprehensive}. 
-}
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:hybrid_lowH1_GY94+Gr}
-The GY94+$\Gamma\omega$ model parameters fit to the ``low divergence from H1" tree (\ref{fig:compete}, \ref{suppfile:alignment_lowH1}). 
-The model was fit using \texttt{phydms\_comprehensive}. 
-}
-\end{suppfile}
-\clearpage
-
-\begin{suppfile}
-\caption{
-\label{suppfile:hybrid_lowH1_ExpCMdoud}
-The ExpCM model parameters fit to the ``low divergence from H1" tree (\ref{fig:compete}, \ref{suppfile:alignment_lowH1}). 
-The model was fit using \texttt{phydms\_comprehensive} and the H1 HA amino-acid preference set in \ref{suppfile:H1_prefs}. 
-}
-\end{suppfile}
-\clearpage
 

### 2018-03-15 manuscript/manuscript.tex
changed from tree length to tree diameter

@@
 Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
-Specifically, the tree's diameter (distance from the two most divergent tips) is \jdbcomment{36\% needs to be updated for tree diameter} longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_lengths}).
-The tree diameter is \jdbcomment{8\% needs to be updated for tree diameter} longer for ExpCM's informed by H1 or H3 amino-acid preferences than for GY94, and \jdbcomment{?} longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_lengths}).
-\jdbcomment{I am pretty sure that the standard quantity used in this context is not the average or total branch length, but rather the distance between the two most distant nodes on the tree, which is what I have called the ``tree diameter'' here.
-I checked with Erick Matsen, and he said that ``tree diameter'' is the correct term here.
-So can you update \ref{supptab:tree_lengths} and the preceding numbers to use that measure instead?}
+Specifically, the tree's diameter (distance from the two most divergent tips) is 68\% longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_diameter}).
+The tree diameter is 24\% and 36\% longer for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, than for GY94, and 67\% longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
 
@@
 As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
-The near independence of these effects is quantified in \ref{supptab:tree_lengths}, which shows that \jdbcomment{96\%, needs to be fixed if using tree diameter} of the tree diameter extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).
+The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 83\% of the tree diameter extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).
 
@@
 \begin{supptable}[t!]
-\caption{\label{supptab:tree_lengths}
-{\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
+\caption{\label{supptab:tree_diameter}
+{\bf Tree diameter for trees in \ref{fig:empirical_trees}.}
 The branch lengths are in average codon substitutions per site. 
-\jdbcomment{Change to be tree diameter, which is distance between two most distant tip nodes.
-Also, to make it easier to relate back to the percentage increases in the text, make 8 rows, one for each model and get rid of the ``rate variation'' column. 
-Then add a new column that gives \% diameter relative to GY94 (the entry would just be 100\% for GY94).
-So this gives 3 columns: model, tree diameter, \% of GY94 tree diameter}
+\skhcomment{finish table legend}
 } 
@@
         \hline
-         Model & no rate variation & {\shortstack{$\Gamma\omega$ rate\\ variation}} \\ \hline
-         GY94 &53.05 & 72.28\\
-         ExpCM(H1) & 57.32 & 81.31\\
-         ExpCM(H3) & 57.77 & 78.98\\
-         ExpCM(H1 + H3 avg) & 62.98 & 83.96 \\
+         Model & Tree diameter & {\shortstack{Percentage increase compared\\ to GY94 tree diameter}} \\ \hline
+         GY94 & 11.42 & 0\%\\
+         ExpCM(H1) & 14.11& 23\%\\
+         ExpCM(H3) & 15.50 & 36\%\\
+         ExpCM(H1+H3 avg) & 19.08 & 67\%\\
+         GY94 + $\Gamma\omega$ & 19.13 & 68\%\\
+         ExpCM(H1) + $\Gamma\omega$ & 25.20 & 121\% \\
+         ExpCM(H3) + $\Gamma\omega$ & 26.56 & 133\% \\
+         ExpCM(H1+H3 avg) + $\Gamma\omega$ & 29.82 & 161\% \\
       \end{tabular}
@@
 {\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
-The branch lengths are in average codon substitutions per site. 
+\skhcomment{finish legend}
 } 

### 2018-03-15 manuscript/manuscript.tex
worked on sequence description

@@
 
-We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18.
-\jdbcomment{Add a sentence indicating \emph{why} you excluded these subtypes. Also you need to say where you downloaded the sequences from and give a date.} 
+We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes from the Influenza Virus Resource Database~\citep[June 2017;][]{bao2008influenza}. 
+We excluded rare subtypes 12, 15, 17, and 18, which have limited sequences in the database.  
 We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
@@
 We subsampled the remaining sequences to five per subtype with $\le 1$ sequence per year per subtype. 
-We also included a small number of representative sequences from the major human and equine influenza lineages, described in \ref{suppfile:required_seqs}. 
-\jdbcomment{We need to trim supplementary files, so get rid of \ref{suppfile:required_seqs} and put an approximate justification / explanation here.}
+We also included a small number of sequences from the major human and equine influenza lineages to ensure representation of these well-studied lineages. 
 The resulting alignment contains 87 sequences, and is provided in \ref{suppfile:alignment_high}.
@@
 We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence for the analysis in \ref{fig:compete}. 
-The ``low divergence" alignments had $\ge 59\%$ amino-acid identity to the sequence used in the deep mutational scanning, and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence.
-The resulting alignments are found in \ref{suppfile:alignment_lowH1}, \ref{suppfile:alignment_lowH3}, \ref{suppfile:alignment_intermediateH1}, and \ref{suppfile:alignment_intermediateH3}.  
-\jdbcomment{Again, we should probably get rid of these four preceding files since we have too many. Presumably they can be re-generated using the description above of the identity cutoffs?}
+The ``low divergence" alignments had $\ge 59\%$ amino-acid identity to the sequence used in the deep mutational scanning, and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence(~\ref{suppfig:subalignments}).
 
@@
 For the analysis in \ref{fig:decay}, we fit models to the ``low divergence" H1 subtree.
-\jdbcomment{This gave the parameter values in Supplementary table... (You need to put these into one or more supplementary tables, or just list them here as we have too many Supplementary files.}
+This gave the parameter values in \ref{supptab:fit_params}.
 
@@
 Looking at existing papers seems to confirm this.
-Note also that they have some limits on acceptable file formats, although you can turn FASTA and CSV into the allowable TXT just by adding that extension, as in \texttt{alignment.fasta.txt} and they won't complain.
-So you need to get rid of all but 4 supplementary files.
-
-I would say keep the map of site numbers, the H1 and H3 preferences, and the alignment. 
-I think all the other files can either be re-generated using the other ones if you provide adequate instructions in the methods (e.g., averaging the preferences or pruning the full alignment to sub-alignments), or simply have model parameters which could be put in a supplementary table.
 }
@@
 \caption{\label{suppfig:subalignments}
-\textbf{Overall divergence for the subtrees in \ref{fig:compete}.} 
-\jdbcomment{This legend is indecipherable because rather than saying what you are showing you just refer to a lot of other figures / files.
-Reduce the references to other figures / files to the minimum needed an explain what is actually being shown.}
-We created two sub alignments from the alignment (\ref{suppfile:alignment_high}) used in \ref{fig:empirical_trees} for each deep mutational scanning reference HA. 
-The ``low" alignments have $\ge59\%$ amino-acid identity to the reference sequence (\ref{suppfile:alignment_lowH1}, \ref{suppfile:alignment_lowH3}) and the ``intermediate" alignments have $\ge46\%$ amino-acid identity to the reference sequence (\ref{suppfile:alignment_intermediateH1}, \ref{suppfile:alignment_intermediateH3}).
+\textbf{Overall divergence for subtrees.}
+We created two subalignments for each HA used in our deep mutational scanning experiments. 
+The ``low divergence" alignments had $\ge$59\% amino-acid identity to either the H1 or H3 reference sequence. 
+The ``intermediate divergence" alignments had $\ge$46\% amino-acid identity to the reference sequences.
 }
@@
 \caption{\label{supptab:tree_diameter}
-{\bf Tree diameter for trees in \ref{fig:empirical_trees}.}
-The branch lengths are in average codon substitutions per site. 
-\skhcomment{finish table legend}
+{\bf Branch length extension as measured by tree diameter.}
+We calculated the tree diameter, or the distance between the two most divergent tips, for the trees in \ref{fig:empirical_trees}.
+For each tree, the diameter is reported as a raw value and a percentage of the GY94 model tree, the smallest of the eight trees. 
 } 
@@
         \hline
-         Model & Tree diameter & {\shortstack{Percentage increase compared\\ to GY94 tree diameter}} \\ \hline
-         GY94 & 11.42 & 0\%\\
-         ExpCM(H1) & 14.11& 23\%\\
-         ExpCM(H3) & 15.50 & 36\%\\
-         ExpCM(H1+H3 avg) & 19.08 & 67\%\\
-         GY94 + $\Gamma\omega$ & 19.13 & 68\%\\
-         ExpCM(H1) + $\Gamma\omega$ & 25.20 & 121\% \\
-         ExpCM(H3) + $\Gamma\omega$ & 26.56 & 133\% \\
-         ExpCM(H1+H3 avg) + $\Gamma\omega$ & 29.82 & 161\% \\
+         Model & {\shortstack{Tree diameter\\ (average codon substitutions per site)}} & {\shortstack{Percentage of\\ GY94 tree diameter}} \\ \hline
+         GY94 & 11.42 & 100\%\\
+         ExpCM(H1) & 14.11& 123\%\\
+         ExpCM(H3) & 15.50 & 136\%\\
+         ExpCM(H1+H3 avg) & 19.06 & 167\%\\
+         GY94 + $\Gamma\omega$ & 19.13 & 168\%\\
+         ExpCM(H1) + $\Gamma\omega$ & 25.21 & 221\% \\
+         ExpCM(H3) + $\Gamma\omega$ & 26.56 & 233\% \\
+         ExpCM(H1+H3 avg) + $\Gamma\omega$ & 29.82 & 261\% \\
       \end{tabular}
@@
 \caption{\label{supptab:fit_params}
-{\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
-\skhcomment{finish legend}
+{\bf Model parameters fit to a low divergence tree.}
+We fit GY94 models and an ExpCM defined by H1 deep mutational scanning preferences to the ``low divergence from H1" tree in \ref{fig:compete}. 
+We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay}. 
 } 
@@
         \hline
-         Model & Parameters \\ \hline\hline
+         Model & Parameters \\ \hline
          GY94 &  {\shortstack{$\kappa = 3.17$, $\omega = 0.10$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.13$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$,  $\phi_{2G} = 0.19$,\\ $\phi_{3, A} = 0.35$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline

### 2018-03-15 manuscript/manuscript.tex
finished touching up legends

@@
 We can observe how shifting amino-acid preferences degrade the accuracy of an ExpCM by fitting the model to trees containing increasingly diverged sequences.
-For both H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains sequences with $\ge$59\% amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge$46\% amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all HAs (which have as little as 38\% identity to the HA in the experiment).
+For both H1 and H3 HAs, we created three phylogenetic trees (\ref{suppfig:subalignments}): a ``low'' divergence tree that contains sequences with $\ge$59\% amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge$46\% amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all HAs (which have as little as 38\% identity to the HA in the experiment).
 \ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
@@
 List of alignable sites between H1 HA and H3 HA. 
-This file provides a conversion between the numbering scheme we use in the paper (sequential numbering of just the alignable sites) to sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 and the H3 HA reference sequence A/Perth/2009.
-The sequential numbering for the HA strains starts with ``site 1" as the second codon (first non-methionine codon). 
-\jdbcomment{The last sentence is so unclear that even I'm not sure what it means.
-If this isn't apparent from the file, explain it much better!
-Wouldn't it be simpler if you just used numbering that started with 1 at the first codon?
-Also, add the actual file so we can see if it makes sense.}
+This files provides a conversion between the numbering scheme we use in the paper (sequential numbering of just the alignable sites) to sequential numbering of the H1 HA reference sequence A/Wilson Smith/1933 and the H3 HA reference sequence A/Perth/2009.
 }
@@
 \label{suppfile:alignment_high}
-The HA sequences used for the full HA tree in \ref{fig:empirical_trees}. 
-We inferred the topology using \texttt{RAxML} and the GTRCAT model. 
-\jdbcomment{This legend is not very good. If the file just has the alignment, then why are we talking about RAxML in its legend?
-Also, I think referring to \ref{fig:empirical_trees} to refer to the tree topology is not very clear, since that figure shows lots of stuff beyond just one tree.}
+The HA sequences for the full HA tree. 
+The sequences in this alignment contain only the alignable sites between H1 and H3 HAs.
+Conversion from this numbering scheme to sequential numbering of A/Perth/2009 is in in \ref{suppfile:WSN_Perth_map}. 
 }

### 2018-03-17 manuscript/manuscript.tex

@@
 Circle denotes the H1 clade and triangle denotes the H3 clade.
-\jdbcomment{Add a line to this legend about how trees are rooted, such as: ``Each tree is separately midpoint rooted''---or if that is not what was done, then add something that indicates how it was done.
-If it isn't what was done, can you remind me of rationale for not doing that?}
+The \texttt{RAxML}-inferred base tree is midpoint rooted and each tree here maintains this rooting.
 }
@@
 where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\kappa$ is the transition-transversion rate, $\Phi_y$ is the equilibrium frequency of codon $y$, and $\omega$ is the relative rate of nonsynonymous and synonymous substitutions.
-\jdbcomment{I have dropped the $\mu$ from the equation above since it is confounded with branch lengths and so not estimated separately for any of our models.
-Also, the equation wasn't right as the $\omega$ term should only be there for nonsynonymous substitutions.
-Also, you need to add a line specifying how the $\Phi_y$ terms are defined. Probably the CF3X4 method? Add this, a proper citation, and one-sentence description.}
+We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citep{pond2010correcting}. 
+This method calculates the $\Phi_y$ values from the empirical alignment frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis.
+
 The frequency $p_x$ of codon $x$ in the stationary state of a GY94 model is simply 
@@
 where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
-The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form as described in \citet{bloom2017identification}.
-\jdbcomment{You need to add a more precise description of this because otherwise it doesn't make any sense when you refer to the $\phi_x$ parameters two paragraphs below if they have never been defined in this context.}
+
+The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form as 
+\begin{equation}
+\label{eq:Qxy}
+   Q_{xy} = \begin{cases}
+            0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,} \\
+            \phi_w & \mbox{if $x$ can be converted to $y$ by a transversion of a nucleotide to $w$,} \\
+            \kappa \times \phi_w & \mbox{if $x$ can be converted to $y$ by a transition of a nucleotide to $w$} \\
+            \end{cases}
+\end{equation}
+where $\phi_w$ is the nucleotide frequency of nucleotide $w$ and $\kappa$ is the transition-transversion rate.
 
@@
 \end{equation}
-where $\beta$ is the stringency parameter~\citet{bloom2014informed,hilton2017phydms} and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
+where $\beta$ is the stringency parameter~\citep{bloom2014informed,hilton2017phydms} and $\omega$ is the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences.
 
@@
 This relationship gives the expected rate of nonsynonymous to synonymous substitutions given the amino-acid preferences.
-Specifically, we first fit an ExpCM, which allows us to calculate $P_{r,xy}$ (\ref{eq:ExpCM}), $Q_{xy}$ (\jdbcomment{as noted above, you need to give this equation in the ExpCM section}), and $p_{r,x}$ (\ref{eq:prx}).
+Specifically, we first fit an ExpCM, which allows us to calculate $P_{r,xy}$ (\ref{eq:ExpCM}), $Q_{xy}$ (\ref{eq:Qxy}), and $p_{r,x}$ (\ref{eq:prx}).
 We then calculated $\omega_r$ using the equation of \citet{spielman2015relationship}, normalizing by the gene-wide $\omega$ fit by the ExpCM:

### 2018-03-19 manuscript/manuscript.tex
began to work on phylobayes terminology

@@
 (B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
-\jdbcomment{You need to mention and give a bit of explanation for the coloring.
-What does it mean for a branch to be ``from'' H1 or H3?
-Would it be better to say branch ``leading to''?
-And does that mean leading to it with respect to the root of the tree as shown in \ref{fig:empirical_trees}?}
-The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the three.
+Points denote branch lengths between all pairs of tips on the tree. 
+Blue and orange denote branches that terminate at the H1 and H3 deep mutational scanning reference sequences respectively. 
+\skhcomment{leading to? from? terminate at?}
+The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree.
 }
@@
 Therefore, while the ExpCM and pbMutSel models differ slightly in the extent to which they lengthen different branches, neither model can avoid the tension between the local and global accuracy of amino-acid preferences.
-
-\subsection*{\jdbcomment{Possible new section?}}
-\jdbcomment{We should think about what happens if we try an ExpCM just on the set of sites that Juhye's paper found did \emph{not} substantially shift in amino-acid preferences between H1 and H3.}
 
@@
 The experimentally measured amino-acid preferences masked to just include these alignable sites are in \ref{suppfile:H1_prefs} and \ref{suppfile:H3_prefs}.
-\jdbcomment{Is this right, do the files just have the masked preferences rather than the full set of sites?
-Also, as described below, we are running up against a limit in supplementary files.
-So rather than also giving a file with the averaged values, just add a few sentences clearly explaining how the averaging was done.}
+For the average preference set, we took the normalized pairwise average of the H1 and H3 preferences. 
+The preference for every amino acid, $a$, at every site, $r$, in the average preference set is
+\begin{equation}
+\label{eq:pi_avg}
+\pi_{r,a,\rm{(H1+H3\ avg)}} = \frac{\pi_{r,a,\rm{H1}} + \pi_{r,a,\rm{H3}}}{\sum_{a^{\prime}} \pi_{r,a^{\prime},\rm{(H1+H3\ avg)}}}
+\end{equation}
 
@@
 \section*{Supplemental Information}
-
-\jdbcomment{The MBE instructions appear to indicate that you can have at most five \emph{files} (see \url{https://academic.oup.com/mbe/pages/Supplementary_Information}). 
-I would take this to mean that we can have one merged PDF with figures / tables, plus four supplementary files.
-Looking at existing papers seems to confirm this.
-}
 

### 2018-03-20 manuscript/manuscript.tex
small tweaks to the first three sections of the results

@@
 We find that models informed by experimentally measured site-specific amino-acid preferences estimate longer deep branches on phylogenies of influenza virus hemagglutinin.
-This deepening of branch lengths is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
+This lengthening of branches is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
 The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
@@
 
-Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
+Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
 Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
@@
 It is challenging to obtain values for these amino-acid preference parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
-Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preferences parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
+Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preference parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
 Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
@@
 Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
-Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific amino-acid preferences~\citep{bloom2017identification,rodrigue2017detecting}.
+Mutation-selection models such as an ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific amino-acid preferences~\citep{bloom2017identification,rodrigue2017detecting}.
 This $\omega$ parameter for an ExpCM can be drawn from a statistical distribution (e.g., a gamma distribution) just like for GY94-style models~\citep{rodrigue2014site,haddox2017mapping}. 
@@
 For the GY94 model, the graphs are identical for all sites since this model does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
-The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
+The graphs do differ among sites if we calculate a different $\omega_r$ for each site $r$ in the GY94 model using the amino-acid preferences~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
 However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
-But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
+The ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
 }
@@
 \ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
-The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters.
-This model predicts an asymptotic sequence divergence of 0.059, which is slightly higher than $\frac{1}{20} = 0.05$ since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
-Intuitively, this asymptotic sequence identity of 0.059 seems low, since like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
+The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence divergence of 0.059. 
+While this predicted sequence divergence is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it intuitively feels to low. 
+Like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
 
-Accounting for site-to-site rate variation in GY94 models affects the rate at which the asymptotic sequence identity is approached, but not the actual value of this asymptote. 
+Accounting for site-to-site dN/dS rate variation in GY94 models affects the rate at which the asymptotic sequence identity is approached, but not the actual value of this asymptote. 
 For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but that the asymptote is identical for both models. 
@@
 However, mutation-selection models such as ExpCM's have site-specific stationary states.
-Therefore, they predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
+They predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
 For instance, \ref{fig:decay} shows that at sites such as 183 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
+\centerline{\includegraphics[width=0.9\textwidth]{figures/simulations}}
 \caption{\label{fig:simulations}
 \textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
-We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
-We then inferred the branch lengths of this tree on the simulated alignments.
+We simulated alignments along a HA phylogenetic tree (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
+We then inferred the branch lengths of this tree from the simulated alignments.
 The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.

### 2018-03-20 manuscript/manuscript.tex
small tweaks to the last sections

@@
 Circle denotes the H1 clade and triangle denotes the H3 clade.
-The \texttt{RAxML}-inferred base tree is midpoint rooted and each tree here maintains this rooting.
+The trees are midpoint rooted based on the tree inferred by \texttt{RAxML} using the GTRCAT model. 
 }
@@
 
-The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM models.
+The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM's.
 But how do these models perform on real data?
@@
 As described in \citet{lee2018deep}, the amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites (these differences are apparent in a simple visual comparison of \ref{suppfig:prefs_doud} and
-\ref{suppfig:prefs_lee}).
+\ref{suppfig:prefs_lee}, see site 33 as an example).
 Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree.
@@
 Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
-Specifically, the tree's diameter (distance from the two most divergent tips) is 68\% longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_diameter}).
-The tree diameter is 24\% and 36\% longer for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, than for GY94, and 67\% longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
+Specifically, the tree's diameter (the distance from the two most divergent tips) for the GY94+$\Gamma\omega$ model is 168\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
+The tree diameter is 123\% and 136\% of the GY94 model tree diameter for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, and 167\% of the GY94 model for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
 
@@
 As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
-The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 83\% of the tree diameter extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).
+The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 83\% of the tree diameter extension of ExpCM(H1+H3 avg)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3 avg) versus GY94).
 
 However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}).
-Instead, the branch lengthening is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
+Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
 For instance, the ExpCM informed by the H1 data most dramatically lengthens branches near the H1 clade of the tree, while the ExpCM informed by the H3 data has the largest effect on branches near the H3 clade.
@@
 
-The fact that an ExpCM leads to the most profound lengthening of the branches leading to the sequence used in the experiment can be rationalized in terms of existing knowledge about epistasis during protein evolution.
+The fact that an ExpCM leads to the most profound increase in branch length leading to the sequence used in the experiment can be rationalized in terms of existing knowledge about epistasis during protein evolution.
 Each ExpCM is informed by a single set of experimentally measured amino-acid preferences.
@@
 This epistasis can lead to shifts in a protein's amino-acid preferences over evolutionary time~\citep{pollock2012amino, doud2015site, shah2015contingency, bazykin2015changing, haddox2017mapping}.
-Because the deep mutational scanning experiments that inform our ExpCM's were each performed in the context of a single HA genetic background, their measurements do not account for the epistatic shifts in amino-acid preferences that accumulate as HA evolves.
+Because the deep mutational scanning experiments that inform our ExpCM's were each performed in the context of a single HA genetic background, their measurements do not account for the accumulation of epistatic shifts in amino-acid preferences as HA evolves.
 Therefore, an ExpCM is expected to most accurately describe the evolution of sequences closely related to the one used in the experiment.
@@
 \ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
-For each subtree, we examined the congruence between site-specific natural selection and the amino-acid preferences measured in the deep mutational scanning experiment. 
-We used the ExpCM stringency parameter $\beta$ to measure this congruence~\citep{bloom2014informed,hilton2017phydms}. 
+For each subtree, we examined the congruence between site-specific natural selection and the amino-acid preferences measured in the deep mutational scanning experiment using the ExpCM stringency parameter $\beta$~\citep{bloom2014informed,hilton2017phydms}.
 Values of $\beta$ that are $>$1 indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, suggesting strong congruence between natural selection and the experimental preferences.
@@
 This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
-Specifically, as the $\beta$ value decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
+As $\beta$ value decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
 At the most extreme value of $\beta = 0$, the preferences would be perfectly uniform and look similar to the GY94 preferences in \ref{fig:empirical_trees}.
@@
 Points denote branch lengths between all pairs of tips on the tree. 
-Blue and orange denote branches that terminate at the H1 and H3 deep mutational scanning reference sequences respectively. 
-\skhcomment{leading to? from? terminate at?}
+Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
 The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree.
@@
 This is a direct apple-to-apples comparison, since the pbMutSel model also draws $\omega$ from a gamma-distribution~\citep{rodrigue2014site}.
+\skhcomment{It is not clear how many bins \texttt{Phylobayes MPI} uses. I am going to go back and re-run the analysis specifying $k=4$ bins.}
 First, we compared the amino-acid preferences inferred by the pbMutSel model to the preferences measured in the experiments.
@@
 However, the estimates are not identical, and the tension between local and global accuracy of the amino-acid preferences is still apparent. 
-Specifically, the long branches leading to the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM, while many other branches were estimated to be slightly longer by the pbMutSel model.
+Specifically, the long branches between the H1 or H3 sequences used in the experiments and all other sequences were estimated to be slightly longer by the ExpCM, while many other branches were estimated to be slightly longer by the pbMutSel model.
 The relatively longer branches leading to the experimental sequences when using the ExpCM(H1+H3 avg)+$\Gamma\omega$ suggests that the ``tree average" amino-acid preferences inferred by the pbMutSel model are not as accurate as the preferences from the deep mutational scanning for sequences close to those used in the experiments.

### 2018-03-20 manuscript/manuscript.tex
tweaks to materials and methods

@@
 
-The stationary state of is 
+The stationary state of an ExpCM is 
 \begin{equation}
@@
 In \ref{fig:decay}, we describe GY94 models where each site $r$ has its own $\omega_r$ value that is calculated from the amino-acid preferences using the relationship described by \citet{spielman2015relationship}.
-This relationship gives the expected rate of nonsynonymous to synonymous substitutions given the amino-acid preferences.
-Specifically, we first fit an ExpCM, which allows us to calculate $P_{r,xy}$ (\ref{eq:ExpCM}), $Q_{xy}$ (\ref{eq:Qxy}), and $p_{r,x}$ (\ref{eq:prx}).
+This relationship defines the expected rate of nonsynonymous to synonymous substitutions given the amino-acid preferences.
+We first fit an ExpCM to the ``low divergence" H1 subtree (parameter values in \ref{supptab:fit_params}), which allows us to calculate $P_{r,xy}$ (\ref{eq:ExpCM}), $Q_{xy}$ (\ref{eq:Qxy}), and $p_{r,x}$ (\ref{eq:prx}).
 We then calculated $\omega_r$ using the equation of \citet{spielman2015relationship}, normalizing by the gene-wide $\omega$ fit by the ExpCM:
@@
 The experimentally measured amino-acid preferences masked to just include these alignable sites are in \ref{suppfile:H1_prefs} and \ref{suppfile:H3_prefs}.
-For the average preference set, we took the normalized pairwise average of the H1 and H3 preferences. 
-The preference for every amino acid, $a$, at every site, $r$, in the average preference set is
+For the average preference set, we took the pairwise average of the H1 and H3 preferences. 
+The preference for every amino acid $a$ at every site $r$ in the average preference set is
 \begin{equation}
 \label{eq:pi_avg}
-\pi_{r,a,\rm{(H1+H3\ avg)}} = \frac{\pi_{r,a,\rm{H1}} + \pi_{r,a,\rm{H3}}}{\sum_{a^{\prime}} \pi_{r,a^{\prime},\rm{(H1+H3\ avg)}}}
+\pi_{r,a,\rm{(H1+H3\ avg)}} = \frac{\pi_{r,a,\rm{H1}} + \pi_{r,a,\rm{H3}}}{2}
 \end{equation}
@@
 
-We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes from the Influenza Virus Resource Database~\citep[June 2017;][]{bao2008influenza}. 
+We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes from the Influenza Virus Resource Database~\citep{bao2008influenza} in June of 2017. 
 We excluded rare subtypes 12, 15, 17, and 18, which have limited sequences in the database.  
@@
 We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
-\jdbcomment{Presumably the program gives some measure of convergence? If so, describe what it was.}
-\jdbcomment{You need to better describe what the tree represents and what the pbMutSel preferences represent. Are they the posterior mean?}
+\skhcomment{re: measure of convergence.
+I need add some analysis to fully address this point.
+1.
+I am going to run 2 chains and assess the congruency between these two chains.
+2.
+I am going to look at the traces using Andrew Rambaut's program tracer.
+Both of these are in progress}
+We used \texttt{PhyloBayes-MPI} program \texttt{readpb_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences.
 
@@
 \textbf{Overall divergence for subtrees.}
-We created two subalignments for each HA used in our deep mutational scanning experiments. 
+We created two subalignments for each HA used in the deep mutational scanning experiments. 
 The ``low divergence" alignments had $\ge$59\% amino-acid identity to either the H1 or H3 reference sequence. 
@@
 {\bf Branch length extension as measured by tree diameter.}
-We calculated the tree diameter, or the distance between the two most divergent tips, for the trees in \ref{fig:empirical_trees}.
-For each tree, the diameter is reported as a raw value and a percentage of the GY94 model tree, the smallest of the eight trees. 
+We calculated the tree diameter, the distance between the two most divergent tips, for the trees in \ref{fig:empirical_trees}.
+For each tree, the diameter is reported as a raw value and as a percentage of the GY94 model tree, the smallest of the eight trees. 
 } 
@@
 We fit GY94 models and an ExpCM defined by H1 deep mutational scanning preferences to the ``low divergence from H1" tree in \ref{fig:compete}. 
-We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay}. 
+We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay} and simulate the sequences in \ref{fig:simulations}. 
 } 

### 2018-03-20 analysis/HA/data/_temp_min_id.py
calculated within subtype amino acid identity with sampling
I estimated the min within subtype identity by sampling
pairs of sequences from each subtype. I used 50000 samples
and found that the min identity was from HA 4 with ~68% id.

@@ -1,0 +1,59 @@
+"""
+The purpose of this script is to randomly sample within each subtype to get
+an id for the min identity within a given subtype.
+
+SKH 20180320
+"""
+
+import pandas as pd
+from Bio import SeqIO
+from collections import Counter
+import itertools
+import glob
+from pymodules.utils import *
+import numpy as np
+import scipy
+# for pair in itertools.combinations(group["seq_id"], 2)
+
+# Set up for translation and pairwise
+# codon to index, add codon `---`
+max_codon_index = (max(list(constants.CODON_TO_INDEX.values())))
+if "---" not in list(constants.CODON_TO_INDEX.keys()):
+    constants.CODON_TO_INDEX["---"] = max_codon_index + 1
+# index to AA, add AA `-`
+max_AA_index = max(list(constants.INDEX_TO_AA.keys()))
+if "-" not in list(constants.INDEX_TO_AA.values()):
+    constants.INDEX_TO_AA[max_AA_index+1] = "-"
+# condon to AA, add mapping of `---` to `-`
+constants.CODON_TO_AA = scipy.append(constants.CODON_TO_AA, [[max_AA_index+1]])
+
+def main():
+    np.random.seed(0)
+    n_samples = 50000
+    final = {"HA":[], "min_aa_id":[]}
+    NUMBERS = [x for x in range(1, 19) if x not in [15, 17, 18]]
+    for HA in NUMBERS:
+        m = 1.0
+        seqs = SeqIO.to_dict(SeqIO.parse("aligned/HA_{0}_hybrid_align.fasta".format(HA), "fasta"))
+        ids = list(seqs.keys())
+        ids.sort()
+        for i in range(n_samples):
+            sample = np.random.choice(ids, 2, replace=False)
+            p1 = translate_with_gaps(str(seqs[sample[0]].seq))
+            p2 = translate_with_gaps(str(seqs[sample[1]].seq))
+            aa_id = [1 if p1[x] == p2[x] else 0 for
+                          x in range(len(p1))]
+            aa_id = sum(aa_id)/len(aa_id)
+            if aa_id < m:
+                m = aa_id
+        final["HA"].append(HA)
+        final["min_aa_id"].append(m)
+        print(HA, m)
+    final = pd.DataFrame(final)
+    final.to_csv("_temp_subtype_aa_id.csv", index=False)
+
+
+
+
+if __name__ == '__main__':
+    main()

### 2018-03-20 analysis/HA/data/_temp_min_id.py
removed the min identity scripts

@@ -1,59 +1,0 @@
-"""
-The purpose of this script is to randomly sample within each subtype to get
-an id for the min identity within a given subtype.
-
-SKH 20180320
-"""
-
-import pandas as pd
-from Bio import SeqIO
-from collections import Counter
-import itertools
-import glob
-from pymodules.utils import *
-import numpy as np
-import scipy
-# for pair in itertools.combinations(group["seq_id"], 2)
-
-# Set up for translation and pairwise
-# codon to index, add codon `---`
-max_codon_index = (max(list(constants.CODON_TO_INDEX.values())))
-if "---" not in list(constants.CODON_TO_INDEX.keys()):
-    constants.CODON_TO_INDEX["---"] = max_codon_index + 1
-# index to AA, add AA `-`
-max_AA_index = max(list(constants.INDEX_TO_AA.keys()))
-if "-" not in list(constants.INDEX_TO_AA.values()):
-    constants.INDEX_TO_AA[max_AA_index+1] = "-"
-# condon to AA, add mapping of `---` to `-`
-constants.CODON_TO_AA = scipy.append(constants.CODON_TO_AA, [[max_AA_index+1]])
-
-def main():
-    np.random.seed(0)
-    n_samples = 50000
-    final = {"HA":[], "min_aa_id":[]}
-    NUMBERS = [x for x in range(1, 19) if x not in [15, 17, 18]]
-    for HA in NUMBERS:
-        m = 1.0
-        seqs = SeqIO.to_dict(SeqIO.parse("aligned/HA_{0}_hybrid_align.fasta".format(HA), "fasta"))
-        ids = list(seqs.keys())
-        ids.sort()
-        for i in range(n_samples):
-            sample = np.random.choice(ids, 2, replace=False)
-            p1 = translate_with_gaps(str(seqs[sample[0]].seq))
-            p2 = translate_with_gaps(str(seqs[sample[1]].seq))
-            aa_id = [1 if p1[x] == p2[x] else 0 for
-                          x in range(len(p1))]
-            aa_id = sum(aa_id)/len(aa_id)
-            if aa_id < m:
-                m = aa_id
-        final["HA"].append(HA)
-        final["min_aa_id"].append(m)
-        print(HA, m)
-    final = pd.DataFrame(final)
-    final.to_csv("_temp_subtype_aa_id.csv", index=False)
-
-
-
-
-if __name__ == '__main__':
-    main()

### 2018-03-23 analysis/decay_to_stationary/limit.py
small tweaks to model comparison figure

@@ -1,0 +1,279 @@
+"""
+The purpose of this script is to perform the calculations to make the
+"decay to stationary state" plot.
+
+This includes calculating the "spielman w_r" rates of non-syn change from the
+`ExpCM` applying these site-specific values to the `YNGKP_M0`.
+
+To run doctest, type:
+`python3 -m doctest -v test_doctest.py`
+
+SKH 20170910
+"""
+
+import pandas as pd
+from phydmslib.constants import *
+import phydmslib.models
+import phydmslib.file_io
+import os
+
+
+# Define the models
+def read_phydms_model_params(param):
+    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    return params
+
+
+def create_model_YNGKP_M0(param, nsites):
+    """
+    This is the basic `YNGKP_M0` from `phydms`
+    """
+    params = read_phydms_model_params(param)
+    e_pw = [[] for x in range(3)]
+    for i in range(3):
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
+                                               if x.startswith("phi{0}"
+                                               .format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
+                   range(N_NT)]
+    e_pw = scipy.array(e_pw)
+    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"],
+                                     omega=params["omega"], mu=0.3,
+                                     freeparams=['mu'])
+
+
+def create_model_YNGKP_M5(param, nsites):
+    """
+    This is the basic `YNGKP_M5` from `phydms` with 4 categories.
+    """
+    params = read_phydms_model_params(param)
+    e_pw = [[] for x in range(3)]
+    for i in range(3):
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
+                                               if x.startswith("phi{0}"
+                                               .format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
+                   range(N_NT)]
+    e_pw = scipy.array(e_pw)
+    m = phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], mu=0.3,
[…]

### 2018-03-23 analysis/decay_to_stationary/stationary_state.py

@@ -45,7 +45,7 @@ def create_model_ExpCM(param, prefs):
     return phydmslib.models.ExpCM(prefs, kappa=params["kappa"], omega=params["omega"],
         beta=params["beta"], mu=0.3, phi=phi, freeparams=['mu'])
 
-def get_stationarystate(model,start,length):
+def get_stationarystate(model ,target_sites):
     amino_acids = list(AA_TO_INDEX.keys())
     prefs = {}
     for amino_acid in amino_acids:
@@ -55,7 +55,7 @@ def get_stationarystate(model,start,length):
     final_cols = ["site"] + amino_acids
 
     # extract the stationary state
-    for r in range(start, start + length):
+    for r in target_sites:
         prefs["site"].append(r)
         for amino_acid in amino_acids:
             amino_acid_index = AA_TO_INDEX[amino_acid]
@@ -68,12 +68,11 @@ def get_stationarystate(model,start,length):
 
 def main():
     # set up params for runs
-    start_site = 117
-    number_of_sites = 15
+    target_sites = [111, 109, 113, 114, 115]
     phydms_dir = "../HA/branch_lengths/phydms/"
     prefs_dir = "../HA/data/references/"
     YNGKP_M0_modelparams_fname = "{0}hybrid_lowH1_0_YNGKP_M0_modelparams.txt".format(phydms_dir)
-    YNGKP_M5_modelaprams_fname = "{0}hybrid_lowH1_0_YNGKP_M5_modelparams.txt".format(phydms_dir)
+    YNGKP_M5_modelprams_fname = "{0}hybrid_lowH1_0_YNGKP_M5_modelparams.txt".format(phydms_dir)
     ExpCM_modelparams_fname = "{0}hybrid_lowH1_0_ExpCM_HA_hybridDoud_prefs_modelparams.txt".format(phydms_dir)
     Doud_prefs_fname = "{0}HA_hybridDoud_prefs.csv".format(prefs_dir)
     if not os.path.isdir("outputs"):
@@ -81,16 +80,16 @@ def main():
 
     # ExpCM
     model = create_model_ExpCM(ExpCM_modelparams_fname, Doud_prefs_fname)
-    prefs = get_stationarystate(model, start_site, number_of_sites)
+    prefs = get_stationarystate(model, target_sites)
     prefs.to_csv("outputs/ExpCM_stationarystate.csv", index=False)
 
     # YNGKP M0
[…]

### 2018-03-23 manuscript/manuscript.tex
began adding notes about limit to manuscript

@@
 HA consists of a number of different subtypes.
-Sequences within a subtype have $>$75\% amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
+Sequences within a subtype have $>$68\% amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
 However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
@@
 Both of these are in progress}
-We used \texttt{PhyloBayes-MPI} program \texttt{readpb_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences.
+We used \texttt{PhyloBayes-MPI} program \texttt{readpb\_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences.
 
@@
 We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay} and simulate the sequences in \ref{fig:simulations}. 
+\skhcomment{check this numbers}
 } 
@@
 
+\skhcomment{Notes from Sarah to Jesse}
+
+\skhcomment{(1) Within subtype amino acid identity.
+I didn't want to do a brute force approach so I took 1 million samples the amino-acid identity between pairs of sequences and took the minimum identity of these samples within a subtype.
+Do you think this is sufficient?}
+
+\skhcomment{(2) The asymptotic sequence divergence of the GY94 model.
+Currently, I have a value of 0.059 for the asymptotic divergence.
+I got this value by simply putting a very large value of t into the formula.
+I tried to do out the math and find the \textit{actual} limit but I got a value 0.0193.
+I will attach my notes and maybe there is an obvious place where the formula is wrong...}
+
+\begin{equation}
+\label{eq:limit}
+\end{equation}
+
 \end{document}

### 2018-03-23 manuscript/manuscript.tex
removed old, alternative figure
updated empirical trees figures and text with HA12 data

@@
 However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
-We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
+We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 92 HA sequences drawn from 15 of the 18 subtypes (we excluded bat influenza and one other rare subtypes).
 For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
@@
          Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & $\omega$ & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
-       	ExpCM  (H1+H3 avg) + $\Gamma\omega$  & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
-	ExpCM (H1+H3 avg)  &  950 & -49227 & 0.15 & 1.78\\
-	ExpCM  (H1) + $\Gamma\omega$  & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
-	ExpCM (H3) + $\Gamma\omega$ & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
-	ExpCM (H1) & 2556 & -50030 &  0.13 & 1.22\\
-	ExpCM (H3) &  3197 & -50350 & 0.12 & 1.45\\
-	GY94 + $\Gamma\omega$  & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
-	GY94 & 7625 & -52560  & 0.07 & -\\
+       	ExpCM  (H1+H3 avg) + $\Gamma\omega$  & 0 & -51083 & 0.19,  0.50,  0.91,  1.86 &  1.69\\
+	ExpCM (H1+H3 avg)  &  1063 & -51616 & 0.14 & 1.77\\
+	ExpCM  (H1) + $\Gamma\omega$  & 1321 & -51744  & 0.12 ,  0.42,  0.89,  2.13 & 1.11\\
+	ExpCM (H3) + $\Gamma\omega$ & 1777 & -51972 & 0.10,  0.36,  0.76,  1.84 & 1.28\\
+	ExpCM (H1) & 2670 & -52419 &  0.12 & 1.21\\
+	ExpCM (H3) &  3377 & -52773 & 0.12 & 1.43\\
+	GY94 + $\Gamma\omega$  & 4817 & -53487 & 0.00,  0.03,  0.08,  0.24 & - \\
+	GY94 & 7892 & -55025  & 0.07 & -\\
       \end{tabular}
@@
 Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
-Specifically, the tree's diameter (the distance from the two most divergent tips) for the GY94+$\Gamma\omega$ model is 168\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
-The tree diameter is 123\% and 136\% of the GY94 model tree diameter for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, and 167\% of the GY94 model for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
+Specifically, the tree's diameter (the distance from the two most divergent tips) for the GY94+$\Gamma\omega$ model is 159\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
+The tree diameter is 122\% and 135\% of the GY94 model tree diameter for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, and 160\% of the GY94 model for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
 
@@
 As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
-The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 83\% of the tree diameter extension of ExpCM(H1+H3 avg)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3 avg) versus GY94).
+The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 76\% of the tree diameter extension of ExpCM(H1+H3 avg)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3 avg) versus GY94).
 
@@
          Model & {\shortstack{Tree diameter\\ (average codon substitutions per site)}} & {\shortstack{Percentage of\\ GY94 tree diameter}} \\ \hline
-         GY94 & 11.42 & 100\%\\
-         ExpCM(H1) & 14.11& 123\%\\
-         ExpCM(H3) & 15.50 & 136\%\\
-         ExpCM(H1+H3 avg) & 19.06 & 167\%\\
-         GY94 + $\Gamma\omega$ & 19.13 & 168\%\\
-         ExpCM(H1) + $\Gamma\omega$ & 25.21 & 221\% \\
-         ExpCM(H3) + $\Gamma\omega$ & 26.56 & 233\% \\
-         ExpCM(H1+H3 avg) + $\Gamma\omega$ & 29.82 & 261\% \\
+         GY94 & 12.04 & 100\%\\
+         ExpCM(H1) & 14.70 & 122\%\\
+         ExpCM(H3) & 16.28 & 135\%\\
+         ExpCM(H1+H3 avg) & 19.21 & 160\%\\
+         GY94 + $\Gamma\omega$ & 19.15 & 159\%\\
+         ExpCM(H1) + $\Gamma\omega$ & 24.75 & 206\% \\
+         ExpCM(H3) + $\Gamma\omega$ & 25.03 & 208\% \\
+         ExpCM(H1+H3 avg) + $\Gamma\omega$ & 30.78 & 256\% \\
       \end{tabular}
@@
          Model & Parameters \\ \hline
-         GY94 &  {\shortstack{$\kappa = 3.17$, $\omega = 0.10$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.13$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$,  $\phi_{2G} = 0.19$,\\ $\phi_{3, A} = 0.35$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
-         GY94 + $\Gamma\omega$ & {\shortstack{$\alpha_\omega = 0.51$, $\beta_\omega = 3.92$, $\kappa = 3.49$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.14$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$, $\phi_{2,G} = 0.20$, \\$\phi_{3,A} = 0.355$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
-         ExpCM(H1) &  {\shortstack{$\beta = 1.22$, $\kappa = 3.24$, $\omega = 0.13$,\\ $\phi_A = 0.37$, $\phi_C = 0.18$, $\phi_G = 0.23$}}\\ \hline
+         GY94 &  {\shortstack{$\kappa = 3.17$, $\omega = 0.10$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.14$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$,  $\phi_{2G} = 0.20$,\\ $\phi_{3, A} = 0.36$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
+         GY94 + $\Gamma\omega$ & {\shortstack{$\alpha_\omega = 0.51$, $\beta_\omega = 3.92$, $\kappa = 3.49$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.14$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$, $\phi_{2,G} = 0.20$, \\$\phi_{3,A} = 0.36$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
+         ExpCM(H1) &  {\shortstack{$\beta = 1.56$, $\kappa = 3.64$, $\omega = 0.24$,\\ $\phi_A = 0.378$, $\phi_C = 0.17$, $\phi_G = 0.23$}}\\ \hline
       \end{tabular}

### 2018-03-23 manuscript/manuscript.tex
updated compete figure with HA12 tree

@@
 
-We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes from the Influenza Virus Resource Database~\citep{bao2008influenza} in June of 2017. 
-We excluded rare subtypes 12, 15, 17, and 18, which have limited sequences in the database.  
+We downloaded all full-length, coding sequences for 15 of the 18 Influenza A virus HA subtypes from the Influenza Virus Resource Database~\citep{bao2008influenza} in June of 2017. 
+We excluded rare subtypes 15, 17, and 18, which have limited sequences in the database.  
 We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
@@
 We also included a small number of sequences from the major human and equine influenza lineages to ensure representation of these well-studied lineages. 
-The resulting alignment contains 87 sequences, and is provided in \ref{suppfile:alignment_high}.
+The resulting alignment contains 92 sequences, and is provided in \ref{suppfile:alignment_high}.
 

### 2018-03-23 manuscript/manuscript.tex
small tweaks to manuscript

@@
 However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
-We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 92 HA sequences drawn from 15 of the 18 subtypes (we excluded bat influenza and one other rare subtypes).
+We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 92 HA sequences drawn from 15 of the 18 subtypes (we excluded bat influenza and one other rare subtype).
 For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
@@
 We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence for the analysis in \ref{fig:compete}. 
-The ``low divergence" alignments had $\ge 59\%$ amino-acid identity to the sequence used in the deep mutational scanning, and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence(~\ref{suppfig:subalignments}).
+The ``low divergence" alignments had $\ge 59\%$ amino-acid identity to the sequence used in the deep mutational scanning, and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence~(\ref{suppfig:subalignments}).
 
@@
 \subsection*{Simulations}
-For \ref{fig:simulations}, we simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along the full HA tree using an ExpCM defined by parameters fit to the ``low divergence'' H1 subtree.
-
+For \ref{fig:simulations}, we simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along the full HA tree using an ExpCM defined by parameters fit to the ``low divergence'' H1 subtree~(\ref{supptab:fit_params}).
 We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.
@@
 We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay} and simulate the sequences in \ref{fig:simulations}. 
-\skhcomment{check this numbers}
 } 
@@
 
-\skhcomment{Notes from Sarah to Jesse}
+\noindent\skhcomment{Notes from Sarah to Jesse}
 
-\skhcomment{(1) Within subtype amino acid identity.
-I didn't want to do a brute force approach so I took 1 million samples the amino-acid identity between pairs of sequences and took the minimum identity of these samples within a subtype.
-Do you think this is sufficient?}
+\noindent\skhcomment{\textbf{(1) Within subtype amino acid identity.} When describing the structure of the HA tree we comment that within a subtype HA sequences are $\ge 68\%$ identical on an amino acid level.
+To calculate this number I didn't want to look just at the tree of 92 sequences we used for the analysis because we might have missed the most diverged pair of sequences within a subtype.
+I also didn't want to take a brute force approach and calculate the pairwise identity of \textit{every} sequence I downloaded fro IVR or build a tree with all of these sequences.
+To get the number 68\% I took 1 million samples of pairs of sequence for each subtype and calculated the mean divergence between these 18 million samples.
+Do you think this is sufficient?
+Is there an alternative way which makes more sense?}
 
-\skhcomment{(2) The asymptotic sequence divergence of the GY94 model.
-Currently, I have a value of 0.059 for the asymptotic divergence.
-I got this value by simply putting a very large value of t into the formula.
-I tried to do out the math and find the \textit{actual} limit but I got a value 0.0193.
-I will attach my notes and maybe there is an obvious place where the formula is wrong...}
+\noindent\skhcomment{\textbf{(2) The asymptotic sequence divergence of the GY94 model.} In the discussion of the decay to stationary state plot, we comment that the asymptotic sequence divergence of the GY94 model is 0.059.
+I got this value by simply putting a very large value of $t$ (time) into the equation.
+I tried to find the true asymptotic divergence by solving the limit of the equation as t approaches infinity.
+When I did this I got a value of 0.0193.
+This is too low.
+I've attached my notes, is there a place where my math is obviously wrong?}
 
-\begin{equation}
-\label{eq:limit}
-\end{equation}
+\noindent\skhcomment{\textbf{(3) pbMutSel re-run.} I am going to re-run the pbMutSel analysis making two major changes.
+The first is I am going to specify the number of bins for the gamma distribution so there are exactly the same number of bins as in the ExpCM+$\Gamma\omega$ ($K=4$).
+The second change I am going to make is to run two parallel chains.
+I will use the amount of  congruency between the parallel chains along with the trace files to assess the convergence of the model.
+The server is going to be undergoing some maintenance this weekend so I might not have the results until mid next week.
+However, when I was first getting the pbMutSel model up and running I ran two chains in parallel on a similar (but not identical) tree and the results were very congruent.
+That is, I am not too worried about the results changing in the pbMutSel section.
+}
 

### 2018-03-25 manuscript/manuscript.tex
Jesse edits / comments

@@
 
+\jdbcomment{I have commented on the general structure of the Introduction, which I think is mostly OK. 
+Add in references next, as this is crucial.
+The Intro should frame the general problem and describe specifically what is already known.
+Right now it does a bit too much of repeating your first two sections of Results.
+In particular, there is some balance between how much goes in Intro and Results. 
+Right now we have a lot of good background in Results, so don't get too repetitive in Intro.}
+
 When studying a protein's evolutionary history, it is important to understand the timescale at which the evolutionary events have occurred. 
@@
 That is, when two nodes are separated by a very long branch, the phylogenetic estimate of this branch length often appears to be too short, which leads to an estimate of time that is too recent. 
+\jdbcomment{I would make the first sentence more general than just proteins.
+Second sentence: good idea to use viruses as examples, but try to make the sentence a bit more specific and less general---also, population size may not be relevant to estimates of deep divergence times.
+For the last part, can we in fact cite general non-viral references about times being too recent? 
+If you can't find several good ones, then just drop that and make first paragraph solely about divergence estimates and have next paragraph go into underestimation.}
 
@@
 All together, this indicates that phylogenetic branch length estimation does truly have a systematic bias towards underestimation. 
+\jdbcomment{Good content, although obviously needs references.
+Also, I think it is best to have at most one or two general sentences and then get right into specific examples.
+So I think paragraph would be stronger if you just have one (or two) general sentences and then bring up at least two specific examples, and finally return to general with ``similar things have been seen in viruses X, Y, and Z (citing each).}
 
@@
 As a result, the preferences ``shift" over time time due to the accumulation of substitutions in other parts of the protein.  
+\jdbcomment{Start with: "The under-estimation of deep branches is known to occur in part due to...
+Also, we need to make this paragraph about what has \emph{already} been shown. 
+It is not clear how much of what you argue here is things you can clearly cite to previous papers versus your new results.
+So overall content is good, but make this paragraph talk about prior literature with specific citations to specific points.}
 
@@
 Such models should estimate deep divergence times because model still expects relatively low sequence divergence even after a very long evolutionary time. 
+\jdbcomment{Generally good (except for lack of citations), but again is last point the one you are making or something that you can clearly cite from literature? See comment after next paragraph.}
 
@@
 These results underscore the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.
+\jdbcomment{Looking at this final paragraph, I would suggest structuring it and the two above it (last three) like this:
+\begin{itemize}
+\item Under-estimation can result from failure to model purifying selection.
+\item Most models are homogenous / independent, and model by rate variation.
+\item Some models use site-specific stationary states.
+\item Here we use ExpCMs and find...
+\end{itemize}
+So overall, pretty good structure that you already have but don't put too much of your results into the Introduction.
+
+The one thing that you need to add is a better description of deep mutational scanning. 
+So I would say spend less time repeating your results and more time providing that background in the last paragraph (or an additional added paragraph).
+}
 
@@
 Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preference parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
-Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
+Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters is similar to a non-site-specific substitution model.
 An alternative strategy of obtaining the amino-acid preference parameters via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.
@@
 \ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
-The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence divergence of 0.059. 
-While this predicted sequence divergence is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it intuitively feels to low. 
-Like many proteins HA has a highly conserved structure and function that imposes constraints that cause most sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
+The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence identity of 0.059. 
+While this predicted sequence identity is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it is much lower than the pairwise identity of even the most diverged HAs in nature.
+While it is of course possible that the identity of HAs in nature would become even lower given more time, it seems biochemically improbable that it would ever become as low as 0.059. 
+The reason is that like many proteins HA has a highly conserved structure and function that imposes constraints that cause many sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.
 
@@
 They predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
-For instance, \ref{fig:decay} shows that at sites such as 183 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the divergence will always be low. 
+For instance, \ref{fig:decay} shows that at sites such as 183 and 305 in the H1 HA, an ExpCM but not a GY94-style model predicts that the identity will always be relatively high. 
 When sites with highly constrained amino-acid preferences such as these are common, an ExpCM can estimate a long branch length at modest sequence identities that a GY94 model might attribute to a shorter branch.
@@
 \textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
-We simulated alignments along a HA phylogenetic tree (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
+We simulated alignments along a HA phylogenetic tree (see \ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
 We then inferred the branch lengths of this tree from the simulated alignments.
@@
 \textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
-The branch lengths of the HA tree are optimized using the indicated ExpCM or GY94 model. 
+The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
 The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
@@
 Deep mutational scanning has been used to measure the amino-acid preferences of all sites in two different HAs.
-One scan measured the site-specific amino-acid preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
+One scan measured the preferences of an H1 HA~\citep{doud2016accurate} and the other measured the preferences of an H3 HA~\citep{lee2018deep}.
 The amino-acid preferences measured for these two HAs are shown in \ref{suppfig:prefs_doud} and 
@@
 As described in \citet{lee2018deep}, the amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites (these differences are apparent in a simple visual comparison of \ref{suppfig:prefs_doud} and
-\ref{suppfig:prefs_lee}, see site 33 as an example).
-Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree.
+\ref{suppfig:prefs_lee}; see site 33 as an example).
+Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree \jdbcomment{reference suppfig with the average logo plot here}.
 These three sets of HA amino-acid preferences define three different ExpCM's.
@@
 \textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
+\jdbcomment{Move the general text that applies to the all the figure (e.g., the first two sentences below) to \emph{before} the (A) label.}
 (A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
@@
 \section*{Discussion}
-
+\jdbcomment{Add references to this section.
+Also, I think we are mostly using ``preferences'' rather than ``stationary state'', so use the former except in cases where the latter clearly makes more sense (e.g., talking about math rather than model.
+}
 Here, we tested the effect of models with site-specific stationary states on branch length estimation for the phylogenetic tree of highly diverged influenza HA sequences. 
@@
 This point is underscored by the comparison of the ExpCM defined by the average preference set and the pbMutSel model, which infers the stationary state from the natural sequences. 
-Even though the stationary state of the pbMutSel model is more representative of the global stationary state, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
+Even though the stationary state of the pbMutSel model is more representative of the global stationary state \jdbcomment{maybe ``tree average'' rather than ``global''}, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
 Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither one of these models is able to capture how these preferences shift over time.
+
+\jdbcomment{In particular, you need to add literature about how preferences shift.
+These can mostly be citations worked into above. 
+In particular, there are two types of citations: papers about how preferences shift over time. 
+This would include Hugh's paper and ones cited therein. 
+You don't need to cite all general papers on epistasis, but just the ones that focus on this in terms of preferences.
+
+Second, papers about the idea of modeling non-independence in phylogenetics.
+I believe that Richard Goldstein and David Pollock have a recent \textit{Nature Ecology \& Evolution} paper on this---look it up and also pull in relevant citations that they make.
+Also, there may be some paper in this area by Rodrigue and Lartillot.
+There are also a few cited in the third-to-last paragraph of discussion here (\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4626756/}).
+}
 
@@
 
+\clearpage
 \section*{Methods}
@@
 where $\mathcal{A}\left(x\right)$ is the amino-acid encoded by codon $x$, $\kappa$ is the transition-transversion rate, $\Phi_y$ is the equilibrium frequency of codon $y$, and $\omega$ is the relative rate of nonsynonymous and synonymous substitutions.
-We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citep{pond2010correcting}. 
+We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citet{pond2010correcting}. 
 This method calculates the $\Phi_y$ values from the empirical alignment frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis. 
+\jdbcomment{Also add a sentence saying how under these F3X4 there are 9 parameters defining $\Phi_x$.}
 
@@
 \end{equation}
-where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
+where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, $F_{r,xy}$ is proportional to the probability that this mutation fixes, and the diagonal elements $P_{xx}$ are set by $P_{xx} = -\sum_{z \ne x} P_{xz}$.
 
@@
 \section*{Supplemental Information}
+
+\jdbcomment{Further increase the number of rows in the logo plots to make them larger.
+Since we aren't under length constraints for supplement, might as well make them readable.
+Maybe 8 or so rows instead?}
 
@@
 
+\jdbcomment{This seems fine, except you should take the \emph{minimum} rather than the \emph{mean} divergence between these samples.
+Did you actually take minimum and just have typo in your comment above?
+}
+
 \noindent\skhcomment{\textbf{(2) The asymptotic sequence divergence of the GY94 model.} In the discussion of the decay to stationary state plot, we comment that the asymptotic sequence divergence of the GY94 model is 0.059.
@@
 I've attached my notes, is there a place where my math is obviously wrong?}
+
+\jdbcomment{
+It is not true that all elements of $\mathbf{D}$ are $< 0$.
+In fact, the principle eigenvector of a irreducible and acyclcic stochastic matrix is 1, and since $\mathbf{D}$ is a stochastic matrix minus the identity matrix, that means that it has one eigenvalue that is 0 and the rest are $< 0$.
+I'm not sure how all the math works out, but I know that it should end up being the case that $\left[\lim_{t \rightarrow \infty} e^{t \mathbf{P_r}}\right]_{xy} = p_{r,x}$ (see \url{https://en.wikipedia.org/wiki/Stochastic_matrix}).
+
+So the correct formula for the asymptotic sequence divergence at site $r$ must be
+\begin{equation}
+\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} p_{r, y}.
+\end{equation}
+You can see that this is true without even doing all the complicated matrix exponential limits: at stationary state, the probability of being $x$ at site $r$ is $p_{r,x}$.
+}
 
@@
 
+\jdbcomment{OK, sounds fine.}
+
 \end{document}

### 2018-03-26 manuscript/manuscript.tex
addressed Jesse's minor comments

@@
 \ref{suppfig:prefs_lee}; see site 33 as an example).
-Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree \jdbcomment{reference suppfig with the average logo plot here}.
+Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree~(\ref{suppfig:prefs_average}).
 These three sets of HA amino-acid preferences define three different ExpCM's.
@@
 \textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
-\jdbcomment{Move the general text that applies to the all the figure (e.g., the first two sentences below) to \emph{before} the (A) label.}
-(A) We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
+We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
 The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}.
-The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52).
+(A) The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52).
 (B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
@@
 We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citet{pond2010correcting}. 
-This method calculates the $\Phi_y$ values from the empirical alignment frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis. 
-\jdbcomment{Also add a sentence saying how under these F3X4 there are 9 parameters defining $\Phi_x$.}
+There are nine parameters describing the nucleotide frequencies at each codon site (the nucleotides are constrained to sum to one at each codon position) \skhcomment{which} are calculated from the empirical alignment frequencies.
+The ``corrected F3X4" method calculates the $\Phi_y$ values from these nucleotide frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis. 
+\skhcomment{check}
 
@@
 
-\jdbcomment{Further increase the number of rows in the logo plots to make them larger.
-Since we aren't under length constraints for supplement, might as well make them readable.
-Maybe 8 or so rows instead?}
-
 \begin{suppfig}[H]
@@
 
-\noindent\skhcomment{Notes from Sarah to Jesse}
-
-\noindent\skhcomment{\textbf{(1) Within subtype amino acid identity.} When describing the structure of the HA tree we comment that within a subtype HA sequences are $\ge 68\%$ identical on an amino acid level.
-To calculate this number I didn't want to look just at the tree of 92 sequences we used for the analysis because we might have missed the most diverged pair of sequences within a subtype.
-I also didn't want to take a brute force approach and calculate the pairwise identity of \textit{every} sequence I downloaded fro IVR or build a tree with all of these sequences.
-To get the number 68\% I took 1 million samples of pairs of sequence for each subtype and calculated the mean divergence between these 18 million samples.
-Do you think this is sufficient?
-Is there an alternative way which makes more sense?}
-
-\jdbcomment{This seems fine, except you should take the \emph{minimum} rather than the \emph{mean} divergence between these samples.
-Did you actually take minimum and just have typo in your comment above?
-}
-
-\noindent\skhcomment{\textbf{(2) The asymptotic sequence divergence of the GY94 model.} In the discussion of the decay to stationary state plot, we comment that the asymptotic sequence divergence of the GY94 model is 0.059.
-I got this value by simply putting a very large value of $t$ (time) into the equation.
-I tried to find the true asymptotic divergence by solving the limit of the equation as t approaches infinity.
-When I did this I got a value of 0.0193.
-This is too low.
-I've attached my notes, is there a place where my math is obviously wrong?}
-
-\jdbcomment{
-It is not true that all elements of $\mathbf{D}$ are $< 0$.
-In fact, the principle eigenvector of a irreducible and acyclcic stochastic matrix is 1, and since $\mathbf{D}$ is a stochastic matrix minus the identity matrix, that means that it has one eigenvalue that is 0 and the rest are $< 0$.
-I'm not sure how all the math works out, but I know that it should end up being the case that $\left[\lim_{t \rightarrow \infty} e^{t \mathbf{P_r}}\right]_{xy} = p_{r,x}$ (see \url{https://en.wikipedia.org/wiki/Stochastic_matrix}).
-
-So the correct formula for the asymptotic sequence divergence at site $r$ must be
-\begin{equation}
-\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} p_{r, y}.
-\end{equation}
-You can see that this is true without even doing all the complicated matrix exponential limits: at stationary state, the probability of being $x$ at site $r$ is $p_{r,x}$.
-}
-
-\noindent\skhcomment{\textbf{(3) pbMutSel re-run.} I am going to re-run the pbMutSel analysis making two major changes.
-The first is I am going to specify the number of bins for the gamma distribution so there are exactly the same number of bins as in the ExpCM+$\Gamma\omega$ ($K=4$).
-The second change I am going to make is to run two parallel chains.
-I will use the amount of  congruency between the parallel chains along with the trace files to assess the convergence of the model.
-The server is going to be undergoing some maintenance this weekend so I might not have the results until mid next week.
-However, when I was first getting the pbMutSel model up and running I ran two chains in parallel on a similar (but not identical) tree and the results were very congruent.
-That is, I am not too worried about the results changing in the pbMutSel section.
-}
-
-\jdbcomment{OK, sounds fine.}
-
 \end{document}

### 2018-03-26 analysis/decay_to_stationary/limit.py
added script to calculate asymp. divergence

@@ -230,10 +230,10 @@ def main():
 
         for x in target_codons:
             prx = pr[x]
+            answer = 0
             for y in target_codons:
-                A_Ainv_xy = A_Ainv[x,y]
-                answer = prx * A_Ainv_xy
-                final += answer
+                answer += pr[y]
+            final += (answer * prx)
     print(final)
 
 

### 2018-03-26 manuscript/manuscript.tex
updated asymp. seq identity in manuscript

@@
 \ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
-The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence identity of 0.059. 
+The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence identity of 0.062. 
 While this predicted sequence identity is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it is much lower than the pairwise identity of even the most diverged HAs in nature.
-While it is of course possible that the identity of HAs in nature would become even lower given more time, it seems biochemically improbable that it would ever become as low as 0.059. 
+While it is of course possible that the identity of HAs in nature would become even lower given more time, it seems biochemically improbable that it would ever become as low as 0.062. 
 The reason is that like many proteins HA has a highly conserved structure and function that imposes constraints that cause many sites to sample only a small subset of the 20 amino acids among all known HA homologs~\citep{nobusawa1991comparison}.

### 2018-04-01 manuscript/manuscript.tex
new draft of introduction

@@
 
-\jdbcomment{I have commented on the general structure of the Introduction, which I think is mostly OK. 
-Add in references next, as this is crucial.
-The Intro should frame the general problem and describe specifically what is already known.
-Right now it does a bit too much of repeating your first two sections of Results.
-In particular, there is some balance between how much goes in Intro and Results. 
-Right now we have a lot of good background in Results, so don't get too repetitive in Intro.}
+Phylogenetics and the concept of a molecular clock are commonly used to estimate the time since the divergence of two lineages, which places evolutionary events on an absolute timescale.
+However, in the case of viruses, there is independent evidence that the phylogenetic estimates of deep divergences are too recent. 
+For example, integration of filoviruses into their host genomes indicate that the filoviridiae family diverged 7-12 million years ago while the phylogenetic estimate of this event is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
+Similarly, the phylogenetic estimate of when major simian immunodeficiency virus (SIV) groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of SIV host species~\citep{wertheim2009dating, worobey2010island}.
+These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.
 
-When studying a protein's evolutionary history, it is important to understand the timescale at which the evolutionary events have occurred. 
-For example, understanding how long a particular viral lineage has been circulating informs hypotheses about population size, evolutionary rate, and possible previous hosts.
-Molecular phylogenetic is commonly used to make sure estimates of time since the divergence of two lineages.
-Under a molecular clock assumption, the branch lengths on a phylogenetic tree can be transformed into estimates of absolute time. 
-However, it has been observed that these phylogenetic estimates often appear to estimate times since deep divergences that appear to be too recent. 
-That is, when two nodes are separated by a very long branch, the phylogenetic estimate of this branch length often appears to be too short, which leads to an estimate of time that is too recent. 
-\jdbcomment{I would make the first sentence more general than just proteins.
-Second sentence: good idea to use viruses as examples, but try to make the sentence a bit more specific and less general---also, population size may not be relevant to estimates of deep divergence times.
-For the last part, can we in fact cite general non-viral references about times being too recent? 
-If you can't find several good ones, then just drop that and make first paragraph solely about divergence estimates and have next paragraph go into underestimation.}
+This underestimation is known to occur in part due to inadequate models of purifying selection. 
+During evolution protein sites do not sample all amino acids evenly.  
+Instead, functional constraint leads to different rates of amino-acid substitution at different sites in the protein and repeated substitutions at a given site~\citep{brown1982mitochondrial}. 
+If phylogenetic substitution models fail to account for this constraint, they will underestimate deep branches~\citep{wertheim2011purifying, halpern1998evolutionary}.
 
-In the case of viruses, there is independent evidence that the estimates of these deep divergences are indeed too recent. 
-The times since the divergence of lineages within lentiviruses, paramyxoviruses and filoviruses have been estimated by methods which rely on information outside of or in addition to the viral phylogenetic tree.
-These independent estimates are often orders of magnitude older than the estimates using the viral phylogenetic tree only. 
-For example, there are multiple examples of filovirus integrations into the genomes of rodents. 
-Using these viral integrations and the fossil dates of the rodent species tree, the divergence of two main filovirus groups has been estimated to be $>7$ millions years ago. 
-This estimate stands in stark contrast to the phylogenetic estimate of filovirus divergence which estimates this event happened $\sim 10,000$ years ago. 
-All together, this indicates that phylogenetic branch length estimation does truly have a systematic bias towards underestimation. 
-\jdbcomment{Good content, although obviously needs references.
-Also, I think it is best to have at most one or two general sentences and then get right into specific examples.
-So I think paragraph would be stronger if you just have one (or two) general sentences and then bring up at least two specific examples, and finally return to general with ``similar things have been seen in viruses X, Y, and Z (citing each).}
+Predicting the functional constraints of protein in nature is difficult but laboratory experiments allow for the direct test of a mutation on protein function~\citep{dean2007mechanistic, harms2013evolutionary}.
+Generally in such experiments,  protein variants are generated and then assayed for their functionality relative to wildtype. 
+Now, with the the advent of deep mutational scanning~\citep{fowler2014deep}, these experiments can be performed in a high-throughput, comprehensive, and quantitative fashion. 
+A deep mutational scan of a protein measured the effect of every amino-acid change on a protein's functionality in lab, which creates a map of the protein's site-specific amino-acid preferences.
+These experimental measurements can be incorporated into site-specific phylogenetic models called experimentally informed codon models (ExpCM's) which tend to describe the evolution of a protein better than standard, site-uniform models~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification, haddox2018mapping, lee2018deep, doud2015site}.
 
-Branch length underestimation is due in part to inadequate modeling of purifying selection. 
-During evolution, protein sites do not sample all 20 available amino-acids equally. 
-Instead, proteins have site-specific amino-acid preferences which are necessary to maintain the structure and function of the protein. 
-Failure to account for these constraints will lead to branch length underestimation because the model will assume that sequences which have been evolving for a very long period of time should be more diverged than is actually the case.
-Furthermore, the site-specific amino-acid preferences at one site are often dependent on the amino-acid identity at another site. 
-As a result, the preferences ``shift" over time time due to the accumulation of substitutions in other parts of the protein.  
-\jdbcomment{Start with: "The under-estimation of deep branches is known to occur in part due to...
-Also, we need to make this paragraph about what has \emph{already} been shown. 
-It is not clear how much of what you argue here is things you can clearly cite to previous papers versus your new results.
-So overall content is good, but make this paragraph talk about prior literature with specific citations to specific points.}
-
-Most phylogenetic substitution models are time-homogenous and site-independent: they do not take into account how sites may interact with each other or how preferences may shift over time. 
-However, they can model purifying selection with varying degrees of complexity. 
-One method is to allow the parameter controlling the rate of nonsynonymous to synonymous substitution, the dN/dS parameter, to vary among the protein sites. 
-This variation accounts for the fact that constrained sites will have a lower rate of amino-acid substitution than mutationally constrained sites. 
-However, the stationary state of this model, which describes the expected sequence composition after a very long evolutionary time, will be exactly the same as a model without any variation in the dN/dS parameter.
-Alternatively, models can have site-specific stationary states which explicitly acknowledge the protein's site-specific amino-acid preferences. 
-Such models should estimate deep divergence times because model still expects relatively low sequence divergence even after a very long evolutionary time. 
-\jdbcomment{Generally good (except for lack of citations), but again is last point the one you are making or something that you can clearly cite from literature? See comment after next paragraph.}
-
-Here, we tested the effect of modeling site-specific amino-acid preferences though a model with a site-specific stationary state on an influenza hemagglutinin (HA) phylogenetic tree. 
-Specifically, we used Experimentally Informed Codon Models (ExpCM's) defined by site-specific amino-acid preferences measured by a high-throughput functional assay called deep mutational scanning. 
-We found that these ExpCM's estimated longer branches. 
-We found that the extension in branch length due modeling purifying selection via a site-specific stationary state is largely independent of the the effect of modeling purifying selection through rate variation.
-However, our results make it clear that the site-specific amino-preferences of HA are constant across the entire tree. 
-In particular, the branch lengthening is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that parameterized the ExpCM.
-These results underscore the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to model how this selection shifts over time.
-\jdbcomment{Looking at this final paragraph, I would suggest structuring it and the two above it (last three) like this:
-\begin{itemize}
-\item Under-estimation can result from failure to model purifying selection.
-\item Most models are homogenous / independent, and model by rate variation.
-\item Some models use site-specific stationary states.
-\item Here we use ExpCMs and find...
-\end{itemize}
-So overall, pretty good structure that you already have but don't put too much of your results into the Introduction.
-
-The one thing that you need to add is a better description of deep mutational scanning. 
-So I would say spend less time repeating your results and more time providing that background in the last paragraph (or an additional added paragraph).
-}
+Here we use ExpCM's to estimate branch lengths on an influenza hemagglutinin phylogenetic tree and find that these site-specific models estimate longer deep branches. 
+We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
+We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.
+However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models - how amino-acid preferences shift over time due to epistasis.
 
@@
 Mutation-selection models such as an ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific amino-acid preferences~\citep{bloom2017identification,rodrigue2017detecting}.
-This $\omega$ parameter for an ExpCM can be drawn from a statistical distribution (e.g., a gamma distribution) just like for GY94-style models~\citep{rodrigue2014site,haddox2017mapping}. 
+This $\omega$ parameter for an ExpCM can be drawn from a statistical distribution (e.g., a gamma distribution) just like for GY94-style models~\citep{rodrigue2014site,haddox2018mapping}. 
 We will denote such models as ExpCM+$\Gamma\omega$.
@@
 But in reality, the effect of a mutation at one site in a protein can depend on the amino-acid identities of other sites in the protein~\citep{ortlund2007crystal, gong2013stability, harms2014historical, tufts2014epistasis, starr2018pervasive}.
-This epistasis can lead to shifts in a protein's amino-acid preferences over evolutionary time~\citep{pollock2012amino, doud2015site, shah2015contingency, bazykin2015changing, haddox2017mapping}.
+This epistasis can lead to shifts in a protein's amino-acid preferences over evolutionary time~\citep{pollock2012amino, doud2015site, shah2015contingency, bazykin2015changing, haddox2018mapping}.
 Because the deep mutational scanning experiments that inform our ExpCM's were each performed in the context of a single HA genetic background, their measurements do not account for the accumulation of epistatic shifts in amino-acid preferences as HA evolves.
@@
 \section*{Discussion}
+
+\noindent\skhcomment{\textbf{reference summary}}
+
+\noindent\skhcomment{\textit{how preferences shift over time:}
+\begin{itemize} 
+\item \citet{haddox2018mapping}: The amino-acid preferences from two diverged HIV env homologs are significantly shifted at several sites. 
+\item \citet{ortlund2007crystal}: Small, permissive mutations are needed for a large functional/conformational change from ancestral to extant protein. 
+\item \citet{gong2013stability}: Stabilizing mutations allow for otherwise destabilizing mutations to occur at T-cell epitopes in NP. 
+\item \citet{harms2014historical}: Contingency is strong because only a few permissive genotypes exist. 
+\item \citet{tufts2014epistasis}: Mutations in Hb from ancestral to extant sequences are highly dependent on the order. 
+\item \citet{starr2018pervasive}: The evolutionary history of Hsp90 shows strong contingency/entrenchment and that the epistasis was primarily caused by specific interactions between sites. 
+\item \citet{pollock2012amino}: Introduces the idea of evolutionary stokes shift - proteins tend to adjust to substitutions such that the propensity for that mutation is higher after the substitution than before it (making reversion less likely).
+\item \citet{doud2015site}: The amino-acid preferences from influenza NP homologs are shifted at a small number of sites.
+\item \citet{shah2015contingency}: Most mutations which are selected for via purifying selection are epistatic, contingent on the previous substitutions, and become entrenched. Computational model. 
+\item \citet{bazykin2015changing}: Review article. The set of amino-acids at a given position which confer the greatest fitness advantage change over time. 
+\end{itemize}
+}
+
+\noindent\skhcomment{\textit{Phylogenetic models which account for shifting amino-acid preferences:}
+\begin{itemize} 
+\item \citet{rodrigue2005site}: Amino-acid model which accounts for the structural fitness of a sequence along with empirical amino-acid frequencies. 
+\item \citet{choi2007quantifying}: Codon model which accounts accounts for solvent accessibility and pairwise interactions (tertiary structure) along with empirical amino-acid frequencies.
+\item \citet{bordner2013new}: Modeling site interdependence using factor graphs.
+Similar mathematically structure to \citet{rodrigue2005site} but with a slightly different interpretation of the parameters.
+Amino acid
+\item \citet{goldstein2017sequence}: Model based on statistical mechanics which describes the stability of a sequence as a function of the interaction of a given site an all other sites as well as a thermodynamic bath.
+\end{itemize}
+}
 \jdbcomment{Add references to this section.
@@
 \item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
-\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
+\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to \skhcomment{fit} the pbMutSel model. 
 \item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML} to infer tree topology.

### 2018-04-02 manuscript/manuscript.tex
new draft of the conclusion

@@
 
-\noindent\skhcomment{\textbf{reference summary}}
-
-\noindent\skhcomment{\textit{how preferences shift over time:}
-\begin{itemize} 
-\item \citet{haddox2018mapping}: The amino-acid preferences from two diverged HIV env homologs are significantly shifted at several sites. 
-\item \citet{ortlund2007crystal}: Small, permissive mutations are needed for a large functional/conformational change from ancestral to extant protein. 
-\item \citet{gong2013stability}: Stabilizing mutations allow for otherwise destabilizing mutations to occur at T-cell epitopes in NP. 
-\item \citet{harms2014historical}: Contingency is strong because only a few permissive genotypes exist. 
-\item \citet{tufts2014epistasis}: Mutations in Hb from ancestral to extant sequences are highly dependent on the order. 
-\item \citet{starr2018pervasive}: The evolutionary history of Hsp90 shows strong contingency/entrenchment and that the epistasis was primarily caused by specific interactions between sites. 
-\item \citet{pollock2012amino}: Introduces the idea of evolutionary stokes shift - proteins tend to adjust to substitutions such that the propensity for that mutation is higher after the substitution than before it (making reversion less likely).
-\item \citet{doud2015site}: The amino-acid preferences from influenza NP homologs are shifted at a small number of sites.
-\item \citet{shah2015contingency}: Most mutations which are selected for via purifying selection are epistatic, contingent on the previous substitutions, and become entrenched. Computational model. 
-\item \citet{bazykin2015changing}: Review article. The set of amino-acids at a given position which confer the greatest fitness advantage change over time. 
-\end{itemize}
-}
-
-\noindent\skhcomment{\textit{Phylogenetic models which account for shifting amino-acid preferences:}
-\begin{itemize} 
-\item \citet{rodrigue2005site}: Amino-acid model which accounts for the structural fitness of a sequence along with empirical amino-acid frequencies. 
-\item \citet{choi2007quantifying}: Codon model which accounts accounts for solvent accessibility and pairwise interactions (tertiary structure) along with empirical amino-acid frequencies.
-\item \citet{bordner2013new}: Modeling site interdependence using factor graphs.
-Similar mathematically structure to \citet{rodrigue2005site} but with a slightly different interpretation of the parameters.
-Amino acid
-\item \citet{goldstein2017sequence}: Model based on statistical mechanics which describes the stability of a sequence as a function of the interaction of a given site an all other sites as well as a thermodynamic bath.
-\end{itemize}
-}
-\jdbcomment{Add references to this section.
-Also, I think we are mostly using ``preferences'' rather than ``stationary state'', so use the former except in cases where the latter clearly makes more sense (e.g., talking about math rather than model.
-}
 Here, we tested the effect of models with site-specific stationary states on branch length estimation for the phylogenetic tree of highly diverged influenza HA sequences. 
@@
 
-These results underscore the importance of modeling site-specific amino-acid preferences when estimating long branchs. 
+These results underscore the importance of modeling site-specific amino-acid preferences when estimating long branches. 
 As the simulations \ref{fig:simulations} show, models with uniform stationary states will always underestimate the lengths of branches which have evolved under site-specific constraints. 
-The results with the influenza HA sequences shows that models with site-specific stationary states estimate longer branches when the stationary state is accurate. 
-However, it is also clear that the accuracy of the ExpCM's stationary state degrades as the tree becomes more diverged from the HA used in the deep mutational scan. 
-The desired effect of estimating long, accurate branches can only be achieved when the stationary state is accurate across the entire tree.
+The results with the influenza HA sequences show that models with site-specific amino-acid preferences estimate longer branches when the preferences are accurate. 
+However, it is also clear that the accuracy amino-acid preferences defining the ExpCM's degrade as the tree becomes more diverged from the HA used in the deep mutational scan. 
+The desired effect of estimating long, accurate branches will only be achieve when the stationary state is accurate across the entire tree.
 
-However, it is clear from our results that a major limitation of site-independent, time-homogenous models is their inability to capture epistatic interactions. 
+Therefore, it is clear from our results that a major limitation of site-independent, time-homogenous models is their inability to capture epistatic interactions. 
+It is well established that a protein's site-specific amino-acid preferences are contextually dependent~\citep{gong2013stability, harms2014historical, tufts2014epistasis, ortlund2007crystal} and shift as the protein accumulates substitutions overtime~\citep{haddox2018mapping, starr2018pervasive, pollock2012amino, doud2015site, shah2015contingency,bazykin2015changing}.
+The site-specific amino-acid preferences we used to define the ExpCM's were measured in only one genetic background. 
+Therefore, these preferences are only one snapshot of the functional constraints on a protein and do not capture any of the changes over time. 
+The model's ignorance of shifting preferences is reflected in our results: the ExpCM's have the largest effect on branch length extension near the HA from the experiment. 
+Furthermore, this is a limitation of all time-homogenous models. 
+Even though the stationary state of the pbMutSel model is more representative of the ``tree average" stationary state, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
 The site-specific amino-acid preferences of a protein shift across time and models with a single stationary state across the entire tree cannot capture these dynamics. 
-The ExpCM's defined by preferences measured in one of two highly diverged HA's had the largest effect on branches near the HA from the experiment. 
-The local effect of the ExpCM's defined by a single preference set indicate that different parts of the tree have different stationary states. 
-This point is underscored by the comparison of the ExpCM defined by the average preference set and the pbMutSel model, which infers the stationary state from the natural sequences. 
-Even though the stationary state of the pbMutSel model is more representative of the global stationary state \jdbcomment{maybe ``tree average'' rather than ``global''}, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
-Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither one of these models is able to capture how these preferences shift over time.
-
-\jdbcomment{In particular, you need to add literature about how preferences shift.
-These can mostly be citations worked into above. 
-In particular, there are two types of citations: papers about how preferences shift over time. 
-This would include Hugh's paper and ones cited therein. 
-You don't need to cite all general papers on epistasis, but just the ones that focus on this in terms of preferences.
-
-Second, papers about the idea of modeling non-independence in phylogenetics.
-I believe that Richard Goldstein and David Pollock have a recent \textit{Nature Ecology \& Evolution} paper on this---look it up and also pull in relevant citations that they make.
-Also, there may be some paper in this area by Rodrigue and Lartillot.
-There are also a few cited in the third-to-last paragraph of discussion here (\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4626756/}).
-}
+Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither the ExpCM nor the pbMutSel is able to capture how these preferences shift over time.
 
 Clearly, the site-independence and time-homogeneity assumptions of the ExpCM and pbMutSel models inhibit their ability to accurately describe the effect of epistasis.
-In order to fully address the bias towards underestimation of divergence times by phylogenetic models, these models need to be able to accommodate how site-specific amino-acid preferences shift over time.
-One way to achieve this goal would be to void the site-independence assumption and create a model which explicitly describes the interactions between sites in a protein. 
-However, while even modeling simple pairwise interactions would be a daunting task, epistasis could have higher order interactions which be extremely difficult to model. 
-Another strategy would be to allow the stationary state to shift over lineages in the tree. 
-Such a model might be still be site-independent but, by breaking the time-homogeneity assumption, would capture some effect of the shifting preferences. 
-Models which account for epistasis would be inherently more complex but it clear from our results that it is important to capture how preferences shift over time to accurately estimate long branch lengths.
+Models which account for the context-dependence of amino-acid preferences have been developed but thus far have been restricted to a few sequences~\citep{rodrigue2005site, choi2007quantifying}, short evolutionary distances~\citep{bordner2013new}, or simulated sequences~\citep{goldstein2017sequence}.
+As these models develop further, it may be able to estimate the branches on highly diverged trees, such as influenza HA, while accounting for both the site-specific amino-acid preferences and how these preferences shift over time.
+Models which account for epistasis will always be inherently more complex but it is clear from our results that it is important to capture how preferences shift in order to accurately model the evolution of highly diverged proteins in nature and accurately estimate long branch lengths.
 
@@
 We define the codon frequency parameters, $\Phi_y$, using the ``corrected F3X4" method from~\citet{pond2010correcting}. 
-There are nine parameters describing the nucleotide frequencies at each codon site (the nucleotides are constrained to sum to one at each codon position) \skhcomment{which} are calculated from the empirical alignment frequencies.
+There are nine parameters describing the nucleotide frequencies at each codon site (the nucleotides are constrained to sum to one at each codon position),  and these parameter values are calculated from the empirical alignment frequencies.
 The ``corrected F3X4" method calculates the $\Phi_y$ values from these nucleotide frequencies but corrects for the exclusion of sequences with premature stop codons from the analysis.
-\skhcomment{check}
 
@@
 \item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
-\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to \skhcomment{fit} the pbMutSel model. 
+\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
 \item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML} to infer tree topology.

### 2018-04-03 manuscript/manuscript.tex
comments on Intro

@@
 Phylogenetics and the concept of a molecular clock are commonly used to estimate the time since the divergence of two lineages, which places evolutionary events on an absolute timescale.
+\jdbcomment{First sentence not good.
+The first part basically makes sense, but the second part doesn't follow.
+Under a molecular clock you can estimate relative divergence times using phylogenetics.
+Absolute timescales require a molecular clock \emph{plus} some absolute calibration points.
+I would elaborate by breaking this into 2 or 3 sentences that independently discuss phylogenetics, molecular clock, and absolute timescale in a general fashion. 
+Include some citations to general reviews or papers in this area.
+Then have a paragraph break and go into next paragraph.
+This will yield two shorter paragraphs that cleanly divide the general concept from the specifics about viruses.}
 However, in the case of viruses, there is independent evidence that the phylogenetic estimates of deep divergences are too recent. 
-For example, integration of filoviruses into their host genomes indicate that the filoviridiae family diverged 7-12 million years ago while the phylogenetic estimate of this event is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
+For example, integration of filoviruses into their host genomes indicate that the filoviridiae family diverged 7-12 million years ago\jdbcomment{rather than going ``...
+million years ago while the...'' instead break with a dash: ``...
+million years ago---but the phylogenetic estimate...''' This sentence structure more clearly draws the contrast.
+Also, the wording is a bit unclear about what is being dated: ``the filiviridae family diverged'''.
+Diverged from what?
+Then ``event'' in the next part.
+What ``event''?
+Choose more precise wording.} while the phylogenetic estimate of this event is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
 Similarly, the phylogenetic estimate of when major simian immunodeficiency virus (SIV) groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of SIV host species~\citep{wertheim2009dating, worobey2010island}.
 These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.
+\jdbcomment{This ending is good, except that you have not introduced the nomenclature of branches or deep branches yet.
+Perhaps that could be remedied in the re-structured first paragraph above, which would take more time to talk about phylogenetics, what branch lengths mean, and absolute time in 2-3 sentences rather than all mashed together in one.}
 
@@
 If phylogenetic substitution models fail to account for this constraint, they will underestimate deep branches~\citep{wertheim2011purifying, halpern1998evolutionary}.
+
+\jdbcomment{I have some more subtle wording suggestions the previous paragraph, but it is pretty good structure wise.
+However, I don't think it transitions well to the next paragraph.
+There is a bit of a conceptual gap.
+Add something that bridges about how current approaches to better model this purifying selection are fundamentally statistical in that they try to \emph{infer} the purifying selection from the evolutionary process, such as by statistical distributions of rate variation or site-specific stationary states.
+Cite some papers for that.
+You don't need to go into a lot of detail as that is covered in first Results section, but do mention that most approaches do this by trying to statistically infer selection.
+Then transition into next paragraph about how these constrains can be measured experimentally.
+However, for this paper, I think the first two general sentences about experiments aren't needed.
+Just jump directly to deep mutational, talking about how this new experimental approach can measure the site-specific purifying selection quantitatively for all sites.}
 
@@
 Here we use ExpCM's to estimate branch lengths on an influenza hemagglutinin phylogenetic tree and find that these site-specific models estimate longer deep branches.
+\jdbcomment{Make this first sentence a bit broader, and say that we examine how using ExpCM's and other forms of modeling purifying selection affects estimates of deep divergence times of HA or something like that.
+Idea is correct, but the sentence seems a bit too narrow.
+Also, it is probably better to break into two sentences: one is what you investigate, and the second is what you find.
+Right now, both are in the same sentence.}
 We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
 We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.
-However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models - how amino-acid preferences shift over time due to epistasis.
+However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models - \jdbcomment{Look up the difference between a hyphen and em-dash} how amino-acid preferences shift over time due to epistasis.
 

### 2018-04-03 manuscript/manuscript.tex
some comments on Discussion

@@
 
-\subsection*{A model with amino-acid preferences estimated from natural sequences give similar results to an ExpCM}
+\subsection*{A model with amino-acid preferences estimated from natural sequences gives similar results to an ExpCM}
 
@@
 We also found that the ExpCM's estimated branch lengths of a similar length to models which infer the stationary state from the natural sequences in a Bayesian framework.  
+\jdbcomment{Make this paragraph a bit more general and less focused on ExpCM's.
+The content is mostly good, but phrase it as we tested how rate variation and site-specific amino-acid preferences affect these things, and we used ExpCMs' to model preferences.
+In general, the language should be a bit more general in Discussion and broader.}
 
@@
 The desired effect of estimating long, accurate branches will only be achieve when the stationary state is accurate across the entire tree. 
+\jdbcomment{I don't love the structure here.
+It is better to have one major point per paragraph.
+But the preceding paragraph has a topic sentence that is about the importance of preferences, but then the last part is about how they don't extrapolate well.
+That second point belongs in the next paragraph.
+The one above would be a good one to talk about why modeling prefs are important.}
 
@@
 Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither the ExpCM nor the pbMutSel is able to capture how these preferences shift over time.
+\jdbcomment{Again, break preceding into two paragraphs for two points.
+The first point is that ExpCM's don't extrapolate well when you get too far away from the focal sequence.
+The next point (which should be a new paragraph) is that this is a \emph{general} feature of this type of model as shown by pbMutSeq.}
 
+\jdbcomment{In the next paragraph, try to use the term shifting preferences rather than epistasis, as that is what we are talking about.
+The preferences do shift because of epistasis, so you can use that word.
+But we are not trying to use our models to accurately describe the effect of epistasis.
+We're trying to use them to model long-term evolution, and run into problems because preferences shift, which is due to epistasis.}
 Clearly, the site-independence and time-homogeneity assumptions of the ExpCM and pbMutSel models inhibit their ability to accurately describe the effect of epistasis.
@@
 Models which account for epistasis will always be inherently more complex but it is clear from our results that it is important to capture how preferences shift in order to accurately model the evolution of highly diverged proteins in nature and accurately estimate long branch lengths.
+\jdbcomment{Make clear that there are two challenges associated with context-dependent models: evaluating them mathematically, and knowing the right function to model the shifting preferences.
+Right now these two challenges are not clearly separated in the logical flow.}
 

### 2018-04-06 manuscript/manuscript.tex
first pass on introduction comments

--- 2018-04-02 manuscript/manuscript.tex
@@
 
-Phylogenetics and the concept of a molecular clock are commonly used to estimate the time since the divergence of two lineages, which places evolutionary events on an absolute timescale.
+Phylogenetics is commonly used to estimate the time since the divergence of two lineages. 
+Under the assumption of a molecular clock, branch lengths on the phylogenetic tree can be transformed into estimates of time. 
+When a calibration points exists, these timing estimates of evolutionary events can be placed on an absolute timescale.
+
 However, in the case of viruses, there is independent evidence that the phylogenetic estimates of deep divergences are too recent. 
-For example, integration of filoviruses into their host genomes indicate that the filoviridiae family diverged 7-12 million years ago while the phylogenetic estimate of this event is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
+For example, the integration of filoviruses into their host genomes indicate that the two main clades of filoviridae, ebolovirus and marburg virus, diverged from their common ancestor 7-12 million years ago---but the phylogenetic estimate of this divergence of ebolavirus and margburgvirus is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
 Similarly, the phylogenetic estimate of when major simian immunodeficiency virus (SIV) groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of SIV host species~\citep{wertheim2009dating, worobey2010island}.
@@
 If phylogenetic substitution models fail to account for this constraint, they will underestimate deep branches~\citep{wertheim2011purifying, halpern1998evolutionary}. 
+Most substitution models account for constraint by trying to statistically infer selection. 
+Some models allow the rate of amino-acid substitution to vary among sites according to some statistical distribution~\citep{yang1994maximum,yang2000codon}\skhcomment{models from wertheim2001purifying?}.
+Other models account for this site-specific constraints via site-specific stationary state describing the frequency of every amino acid at every site in the protein~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling, le2008phylogenetic}
 
-Predicting the functional constraints of protein in nature is difficult but laboratory experiments allow for the direct test of a mutation on protein function~\citep{dean2007mechanistic, harms2013evolutionary}.
-Generally in such experiments,  protein variants are generated and then assayed for their functionality relative to wildtype. 
-Now, with the the advent of deep mutational scanning~\citep{fowler2014deep}, these experiments can be performed in a high-throughput, comprehensive, and quantitative fashion. 
-A deep mutational scan of a protein measured the effect of every amino-acid change on a protein's functionality in lab, which creates a map of the protein's site-specific amino-acid preferences.
+A protein's functional constraints can be measured experimentally rather than inferred statistically~\citep{dean2007mechanistic, harms2013evolutionary}. 
+Recently, a new experimental approach called deep mutational scanning~\citep{fowler2014deep} has been developed which can measure the site-specific purifying selection quantitatively for all sites in a protein.
 These experimental measurements can be incorporated into site-specific phylogenetic models called experimentally informed codon models (ExpCM's) which tend to describe the evolution of a protein better than standard, site-uniform models~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification, haddox2018mapping, lee2018deep, doud2015site}.
 
-Here we use ExpCM's to estimate branch lengths on an influenza hemagglutinin phylogenetic tree and find that these site-specific models estimate longer deep branches. 
+Here we examined how ExpCM's and other forms of modeling purifying selection affects the estimates of deep divergence times of influenza virus hemagglutinin (HA). 
+We found that models with site-specific measurements of purifying selection estimates longer deep branches. 
 We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
 We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.
-However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models - how amino-acid preferences shift over time due to epistasis.
+However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models---how amino-acid preferences shift over time due to epistasis.
 
@@
 \item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
+\item \skhcomment{snakemake}
 \end{itemize}

### 2018-04-06 manuscript/manuscript.tex
more work on the introduction

@@
 Phylogenetics is commonly used to estimate the time since the divergence of two lineages. 
-Under the assumption of a molecular clock, branch lengths on the phylogenetic tree can be transformed into estimates of time. 
-When a calibration points exists, these timing estimates of evolutionary events can be placed on an absolute timescale.
+Under the assumption of a molecular clock~\citep{zuckerkandl1965}, the branch lengths on a phylogenetic tree can be transformed into estimates of time~\citep{yang2012molecular}. 
+When a calibration point exists, these timing estimates of evolutionary events can be placed on an absolute timescale.
 
@@
 We found that models with site-specific measurements of purifying selection estimates longer deep branches. 
-We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
+We found this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
 We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.

### 2018-04-06 manuscript/manuscript.tex
revision to intro

--- 2018-04-06 manuscript/manuscript.tex
@@
 
-Phylogenetics is commonly used to estimate the time since the divergence of two lineages. 
-Under the assumption of a molecular clock, branch lengths on the phylogenetic tree can be transformed into estimates of time. 
-When a calibration points exists, these timing estimates of evolutionary events can be placed on an absolute timescale.
+Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular}. 
+However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, aiewsakun2016time}.
+For example, the integration of filoviruses into their host genomes indicate that Ebola and Marburg virus diverged from their common ancestor 7 to 12 million years ago---but the estimate of this divergence time based on phylogenetic analyses of the viral sequences is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
+Similarly, the phylogenetic estimate of when major simian immunodeficiency virus groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of their host species~\citep{wertheim2009dating, worobey2010island}.
+These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}\jdbcomment{add paper specifically emphasizing discrepancy}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.
 
-However, in the case of viruses, there is independent evidence that the phylogenetic estimates of deep divergences are too recent. 
-For example, the integration of filoviruses into their host genomes indicate that the two main clades of filoviridae, ebolovirus and marburg virus, diverged from their common ancestor 7-12 million years ago---but the phylogenetic estimate of this divergence of ebolavirus and margburgvirus is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.
-Similarly, the phylogenetic estimate of when major simian immunodeficiency virus (SIV) groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of SIV host species~\citep{wertheim2009dating, worobey2010island}.
-These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.
+This underestimation occurs in part because phylogenetic models do a poor job of describing the real natural selection on protein-coding genes. 
+These genes evolve under purifying selection to maintain the structure and function of the proteins they encode. 
+In general, these constraints are highly idiosyncratic among sites~\citep{echave2016causes}.
+However, most phylogenetic models try to account for these constraints using relatively simple approaches such as allowing the rate of substitution to vary across sites according to some statistical distribution~\citep{yang1994maximum,yang2000codon}.
+These models of purifying selection are usually inadequate~\citep{duchene2015substitution, duchene2015evaluating}, potentially causing branch lengths to be severely underestimated~\citep{wertheim2011purifying, halpern1998evolutionary}.
 
-This underestimation is known to occur in part due to inadequate models of purifying selection. 
-During evolution protein sites do not sample all amino acids evenly.  
-Instead, functional constraint leads to different rates of amino-acid substitution at different sites in the protein and repeated substitutions at a given site~\citep{brown1982mitochondrial}. 
-If phylogenetic substitution models fail to account for this constraint, they will underestimate deep branches~\citep{wertheim2011purifying, halpern1998evolutionary}. 
-Most substitution models account for constraint by trying to statistically infer selection. 
-Some models allow the rate of amino-acid substitution to vary among sites according to some statistical distribution~\citep{yang1994maximum,yang2000codon}\skhcomment{models from wertheim2001purifying?}.
-Other models account for this site-specific constraints via site-specific stationary state describing the frequency of every amino acid at every site in the protein~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling, le2008phylogenetic}
+More recent work has used mutation-selection models to better account for purifying selection~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
+These models explicitly account for the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\jdbcomment{cite any papers that show this is true.
+Maybe the CAT model and the one on long-branch attraction?}.
+However, these approaches still ultimately involving inferring the site-specific purifying selection from natural sequence data.
 
-A protein's functional constraints can be measured experimentally rather than inferred statistically~\citep{dean2007mechanistic, harms2013evolutionary}. 
-Recently, a new experimental approach called deep mutational scanning~\citep{fowler2014deep} has been developed which can measure the site-specific purifying selection quantitatively for all sites in a protein.
-These experimental measurements can be incorporated into site-specific phylogenetic models called experimentally informed codon models (ExpCM's) which tend to describe the evolution of a protein better than standard, site-uniform models~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification, haddox2018mapping, lee2018deep, doud2015site}.
+Even more recently, it has become possible to directly measure purifying selection on proteins using deep mutational scanning.
+This high-throughput approach involves experimentally measuring how each amino-acid mutation affects protein function in the lab~\citep{fowler2014deep}.
+The resulting experimental measurements of which amino acids are preferred at each protein site can be used to inform phylogenetic substitution models~\citep{bloom2014experimentally}.
+These experimentally informed codon models (ExpCM's) generally exhibit much better phylogenetic fit than most standard substitution models~\citep{hilton2017phydms, haddox2018mapping, lee2018deep, doud2015site}.
 
-Here we examined how ExpCM's and other forms of modeling purifying selection affects the estimates of deep divergence times of influenza virus hemagglutinin (HA). 
-We found that models with site-specific measurements of purifying selection estimates longer deep branches. 
-We find this extension of branch length is mostly independent from common method of account for purifying selection by rate variation. 
-We also found that the ExpCM's estimated similar branch lengths to a site-specific model which infers the amino-acid preferences from the phylogenetic data~\citep{lartillot2004bayesian, rodrigue2014site}.
-However, the results from these site-specific models are limited by an effect of purifying selection which is not accounted for in current models---how amino-acid preferences shift over time due to epistasis.
+Here we examined how using ExpCM's and other models of purifying selection affects the estimates of deep divergence times on a phylogenetic tree of influenza virus hemagglutinin (HA). 
+We found that ExpCMs that model site-specific amino-acid preferences estimate longer deep branches. 
+This extension of branch length is mostly independent and additive with that achieved by the more conventional approach of modeling rate variation. 
+ExpCM's that use experimental data estimate similar branch lengths to a mutation-selection model that infers the amino-acid preferences from the natural sequence data~\citep{lartillot2004bayesian, rodrigue2014site}.
+However, we also show that all of these mutation-selection models are limited by their failure to account for another feature of purifying selection: the fact that a site's amino-acid preferences shift over time due to epistasis.
+Therefore, truly accurate analyses of deep phylogenies need to account for the fact that amino-acid preferences vary across time as well as across sites.
 
@@
 \end{figure}
-
-Proteins evolve under purifying selection to maintain their structure and function. 
-This purifying selection is not homogenous across sites in a protein.
-It is also not homogenous among the different amino acids at a given site.
-For instance, some protein sites strongly prefer hydrophobic amino acids, others are constrained to just one or a few amino acids, and yet others tolerate many amino acids.
-In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.
 

### 2018-04-10 manuscript/manuscript.tex
edits to conclusion

@@
 
-Here, we tested the effect of models with site-specific stationary states on branch length estimation for the phylogenetic tree of highly diverged influenza HA sequences. 
-We used site-specific stationary state models called ExpCM's, which are defined by amino-acid preferences measured by deep mutational scanning. 
-We found that the ExpCM's estimated longer branches and did so independently from substitution rate variation via $\Gamma\omega$. 
-We also found that the ExpCM's estimated branch lengths of a similar length to models which infer the stationary state from the natural sequences in a Bayesian framework.  
-\jdbcomment{Make this paragraph a bit more general and less focused on ExpCM's.
-The content is mostly good, but phrase it as we tested how rate variation and site-specific amino-acid preferences affect these things, and we used ExpCMs' to model preferences.
-In general, the language should be a bit more general in Discussion and broader.}
+Here we tested how rate variation and site-specific amino-acid preferences affect the estimation of deep branches on phylogenetic trees. 
+We used ExpCM's defined by deep mutational scanning experiments to model site-specific amino-acid preferences and found that these models estimate longer branches than the GY94 model with a uniform stationary state.
+We found that ExpCM's estimate similar branch lengths to models which infer their stationary state from the natural sequences. 
+Finally, we found that the extension in branch length from site-specific amino-acid preferences is mostly independent from rate variation.
 
 These results underscore the importance of modeling site-specific amino-acid preferences when estimating long branches. 
-As the simulations \ref{fig:simulations} show, models with uniform stationary states will always underestimate the lengths of branches which have evolved under site-specific constraints. 
-The results with the influenza HA sequences show that models with site-specific amino-acid preferences estimate longer branches when the preferences are accurate. 
-However, it is also clear that the accuracy amino-acid preferences defining the ExpCM's degrade as the tree becomes more diverged from the HA used in the deep mutational scan. 
-The desired effect of estimating long, accurate branches will only be achieve when the stationary state is accurate across the entire tree. 
-\jdbcomment{I don't love the structure here.
-It is better to have one major point per paragraph.
-But the preceding paragraph has a topic sentence that is about the importance of preferences, but then the last part is about how they don't extrapolate well.
-That second point belongs in the next paragraph.
-The one above would be a good one to talk about why modeling prefs are important.}
+As the simulations in \ref{fig:simulations} show, models with uniform stationary states will always underestimate the lengths of branches which have evolved under strong site-specific constraints. 
+We also found that the ExpCM's with the site-specific stationary states estimated longer branches than the GY94 model on the phylogenetic tree of influenza HA.
 
-Therefore, it is clear from our results that a major limitation of site-independent, time-homogenous models is their inability to capture epistatic interactions. 
+However, it is also clear that the amino-acid preferences defining the ExpCM's do not extrapolate well over a very diverged tree. 
 It is well established that a protein's site-specific amino-acid preferences are contextually dependent~\citep{gong2013stability, harms2014historical, tufts2014epistasis, ortlund2007crystal} and shift as the protein accumulates substitutions overtime~\citep{haddox2018mapping, starr2018pervasive, pollock2012amino, doud2015site, shah2015contingency,bazykin2015changing}.
-The site-specific amino-acid preferences we used to define the ExpCM's were measured in only one genetic background. 
-Therefore, these preferences are only one snapshot of the functional constraints on a protein and do not capture any of the changes over time. 
-The model's ignorance of shifting preferences is reflected in our results: the ExpCM's have the largest effect on branch length extension near the HA from the experiment. 
-Furthermore, this is a limitation of all time-homogenous models. 
+The site-specific amino-acid preferences we used to define the ExpCM's were only measured in one genetic background. 
+These preferences are only one snapshot of the functional constraint and do not capture any of the changes over time. 
+The model's ignorance of shifting preferences is reflected in our results: The ExpCM's have the largest effect on branch length extension near the HA from the experiment.
+
+The effect of shifting preferences on branch length estimation is not specific to ExpCM's but is a general limitation of site-independent, time-homogenous models. 
 Even though the stationary state of the pbMutSel model is more representative of the ``tree average" stationary state, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
@@
 Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither the ExpCM nor the pbMutSel is able to capture how these preferences shift over time.
-\jdbcomment{Again, break preceding into two paragraphs for two points.
-The first point is that ExpCM's don't extrapolate well when you get too far away from the focal sequence.
-The next point (which should be a new paragraph) is that this is a \emph{general} feature of this type of model as shown by pbMutSeq.}
 
-\jdbcomment{In the next paragraph, try to use the term shifting preferences rather than epistasis, as that is what we are talking about.
-The preferences do shift because of epistasis, so you can use that word.
-But we are not trying to use our models to accurately describe the effect of epistasis.
-We're trying to use them to model long-term evolution, and run into problems because preferences shift, which is due to epistasis.}
-Clearly, the site-independence and time-homogeneity assumptions of the ExpCM and pbMutSel models inhibit their ability to accurately describe the effect of epistasis.
-Models which account for the context-dependence of amino-acid preferences have been developed but thus far have been restricted to a few sequences~\citep{rodrigue2005site, choi2007quantifying}, short evolutionary distances~\citep{bordner2013new}, or simulated sequences~\citep{goldstein2017sequence}.
-As these models develop further, it may be able to estimate the branches on highly diverged trees, such as influenza HA, while accounting for both the site-specific amino-acid preferences and how these preferences shift over time.
-Models which account for epistasis will always be inherently more complex but it is clear from our results that it is important to capture how preferences shift in order to accurately model the evolution of highly diverged proteins in nature and accurately estimate long branch lengths.
-\jdbcomment{Make clear that there are two challenges associated with context-dependent models: evaluating them mathematically, and knowing the right function to model the shifting preferences.
-Right now these two challenges are not clearly separated in the logical flow.}
+Clearly, the site-independence and time-homogeneity assumptions of the ExpCM's and pbMutSel models inhibit their ability to accurately describe the effect of shifting preferences. 
+Models which account for context-dependent of amino-acid preferences have been developed but are still in their infancy. 
+Currently, there are two main challenges associated with context-dependence models. 
+The first challenge is how to efficiently evaluate these inherently more complex models mathematically so the models can be applied to large phylogenetic trees. 
+The second challenge is developing the correct function to model shifting preferences. 
+One strategy to to account for context-dependence is to evaluate amino-acid changes as how they might affect protein structure.
+However, in nature a protein may undergo conformational changes in order to perform some function. 
+Models which account for shifting preferences should also be flexible enough to be able to describe multiple, necessary protein conformations. 
+Such models will be critical for describing the long-term evolution of a protein and accurately estimating the length of deep branches on phylogenetic trees.
 

### 2018-04-10 analysis/decay_to_stationary/limit.py
added limit.py for stationary state calculate

@@ -1,0 +1,279 @@
+"""
+The purpose of this script is to perform the calculations to make the
+"decay to stationary state" plot.
+
+This includes calculating the "spielman w_r" rates of non-syn change from the
+`ExpCM` applying these site-specific values to the `YNGKP_M0`.
+
+To run doctest, type:
+`python3 -m doctest -v test_doctest.py`
+
+SKH 20170910
+"""
+
+import pandas as pd
+from phydmslib.constants import *
+import phydmslib.models
+import phydmslib.file_io
+import os
+
+
+# Define the models
+def read_phydms_model_params(param):
+    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
+    params = dict(zip(params[0], params[1]))
+    return params
+
+
+def create_model_YNGKP_M0(param, nsites):
+    """
+    This is the basic `YNGKP_M0` from `phydms`
+    """
+    params = read_phydms_model_params(param)
+    e_pw = [[] for x in range(3)]
+    for i in range(3):
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
+                                               if x.startswith("phi{0}"
+                                               .format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
+                   range(N_NT)]
+    e_pw = scipy.array(e_pw)
+    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"],
+                                     omega=params["omega"], mu=0.3,
+                                     freeparams=['mu'])
+
+
+def create_model_YNGKP_M5(param, nsites):
+    """
+    This is the basic `YNGKP_M5` from `phydms` with 4 categories.
+    """
+    params = read_phydms_model_params(param)
+    e_pw = [[] for x in range(3)]
+    for i in range(3):
+        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
+                                               if x.startswith("phi{0}"
+                                               .format(i))])
+        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
+                   range(N_NT)]
+    e_pw = scipy.array(e_pw)
+    m = phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], mu=0.3,
[…]

### 2018-04-10 analysis/decay_to_stationary/limit.py
cleaned up decay directory

@@ -1,279 +1,0 @@
-"""
-The purpose of this script is to perform the calculations to make the
-"decay to stationary state" plot.
-
-This includes calculating the "spielman w_r" rates of non-syn change from the
-`ExpCM` applying these site-specific values to the `YNGKP_M0`.
-
-To run doctest, type:
-`python3 -m doctest -v test_doctest.py`
-
-SKH 20170910
-"""
-
-import pandas as pd
-from phydmslib.constants import *
-import phydmslib.models
-import phydmslib.file_io
-import os
-
-
-# Define the models
-def read_phydms_model_params(param):
-    params = pd.read_csv(param, engine="python", sep=" = ", header=None)
-    params = dict(zip(params[0], params[1]))
-    return params
-
-
-def create_model_YNGKP_M0(param, nsites):
-    """
-    This is the basic `YNGKP_M0` from `phydms`
-    """
-    params = read_phydms_model_params(param)
-    e_pw = [[] for x in range(3)]
-    for i in range(3):
-        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
-                                               if x.startswith("phi{0}"
-                                               .format(i))])
-        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
-                   range(N_NT)]
-    e_pw = scipy.array(e_pw)
-    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"],
-                                     omega=params["omega"], mu=0.3,
-                                     freeparams=['mu'])
-
-
-def create_model_YNGKP_M5(param, nsites):
-    """
-    This is the basic `YNGKP_M5` from `phydms` with 4 categories.
-    """
-    params = read_phydms_model_params(param)
-    e_pw = [[] for x in range(3)]
-    for i in range(3):
-        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys()
-                                               if x.startswith("phi{0}"
-                                               .format(i))])
-        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in
-                   range(N_NT)]
-    e_pw = scipy.array(e_pw)
-    m = phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], mu=0.3,
[…]

### 2018-04-10 analysis/decay_to_stationary/scratch/how_long.py

@@ -1,164 +1,0 @@
-"""
-The purpose of this script is to calculate how long it takes a model to reach
-an expected pairwise identity of 50%.
-
-SKH 20171215
-"""
-
-import pandas as pd
-from phydmslib.constants import *
-import phydmslib.models
-import phydmslib.file_io
-
-
-## Define the models
-def read_phydms_model_params(param):
-    params = pd.read_csv(param, engine="python", sep= " = ", header=None)
-    params = dict(zip(params[0], params[1]))
-    return params
-
-def create_model_YNGKP_M0(param, nsites):
-    """
-    This is the basic `YNGKP_M0` from `phydms`
-    """
-    params = read_phydms_model_params(param)
-    e_pw = [[] for x in range(3)]
-    for i in range(3):
-        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys() if x.startswith("phi{0}".format(i))])
-        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in range(N_NT)]
-    e_pw = scipy.array(e_pw)
-    return phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], omega=params["omega"], mu=0.3,
-            freeparams=['mu'])
-
-def create_model_YNGKP_M5(param, nsites):
-    """
-    This is the basic `YNGKP_M5` from `phydms` with 4 categories.
-    """
-    params = read_phydms_model_params(param)
-    e_pw = [[] for x in range(3)]
-    for i in range(3):
-        params["phi{0}T".format(i)] = 1 - sum([params[x] for x in params.keys() if x.startswith("phi{0}".format(i))])
-        e_pw[i] = [params["phi{0}{1}".format(i, INDEX_TO_NT[x])] for x in range(N_NT)]
-    e_pw = scipy.array(e_pw)
-    m =  phydmslib.models.YNGKP_M0(e_pw, nsites, kappa=params["kappa"], mu=0.3,
-            freeparams=['mu', 'omega'])
-    return phydmslib.models.GammaDistributedOmegaModel(m, ncats=4, alpha_lambda=params["alpha_omega"], beta_lambda=params["beta_omega"])
-
-def create_model_ExpCM(param, prefs):
-    """
-    This is the basic `ExpCM` from `phydms`
-
-    `omega` is set to 1.0
-    """
-    params = read_phydms_model_params(param)
[…]

### 2018-04-10 manuscript/manuscript.tex
more work on conclusion

@@
 Similarly, the phylogenetic estimate of when major simian immunodeficiency virus groups diverged is almost 100 times more recent than the estimate based on the geographic isolation of their host species~\citep{wertheim2009dating, worobey2010island}.
-These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin, rhazes1848treatise}\jdbcomment{add paper specifically emphasizing discrepancy}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.
+These examples, along with other similar discrepancies with measles virus~\citep{furuse2010origin}, coronavirus~\citep{wertheim2013case}, and hepatitis B virus~\citep{fares2002revised, holmes2003molecular}, indicate that phylogenetic methods have a systematic bias toward underestimation of deep branches.
 
@@
 More recent work has used mutation-selection models to better account for purifying selection~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-These models explicitly account for the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\jdbcomment{cite any papers that show this is true.
-Maybe the CAT model and the one on long-branch attraction?}.
+These models explicitly account for the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression}.
 However, these approaches still ultimately involving inferring the site-specific purifying selection from natural sequence data.
@@
 The site-specific amino-acid preferences we used to define the ExpCM's were only measured in one genetic background. 
-These preferences are only one snapshot of the functional constraint and do not capture any of the changes over time. 
+Therefore, these preferences are only one snapshot of the functional constraint and do not capture any of the changes over time. 
 The model's ignorance of shifting preferences is reflected in our results: The ExpCM's have the largest effect on branch length extension near the HA from the experiment.
@@
 Models which account for context-dependent of amino-acid preferences have been developed but are still in their infancy.
+Thus far their application has been restricted to a few sequences~\citep{rodrigue2005site, choi2007quantifying}, short evolutionary distances~\citep{bordner2013new}, or simulated sequences~\citep{goldstein2017sequence}.
+
 Currently, there are two main challenges associated with context-dependence models. 
-The first challenge is how to efficiently evaluate these inherently more complex models mathematically so the models can be applied to large phylogenetic trees. 
+The first challenge is how to efficiently evaluate these inherently more complex models mathematically so the models can be applied to large phylogenetic trees~\citep{rodrigue2009computational}. 
 The second challenge is developing the correct function to model shifting preferences. 
-One strategy to to account for context-dependence is to evaluate amino-acid changes as how they might affect protein structure.
+For example, \citet{rodrigue2009computational} showed that while site-dependent amino-acid models outperformed site-independent amino-acid models, they were outperformed by sophisticated site-independent codon models.
+Most context-dependent models evaluate amino-acid changes as how they might affect a single protein conformation~\citep{rodrigue2005site, choi2007quantifying,bordner2013new, goldstein2017sequence}.
 However, in nature a protein may undergo conformational changes in order to perform some function. 
@@
 \subsection*{HA amino-acid preferences from deep mutational scanning experiments}
-We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} to define the amino-acid preferences that inform the ExpCM's.
+We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} \skhcomment{change citation} to define the amino-acid preferences that inform the ExpCM's.
 We only used sites that can be unambiguously aligned in these H1 and H3 HAs. 
@@
 These two H1 sequences are early and late representatives of the longest known human influenza lineage, and are of sufficiently high identity that different ExpCM and GY94 substitution models all estimate nearly identical branch lengths separating them.
+\skhcomment{add new run parameters}
 

### 2018-04-10 manuscript/manuscript.tex
updated phylobayes materials and methods

@@
 This is a direct apple-to-apples comparison, since the pbMutSel model also draws $\omega$ from a gamma-distribution~\citep{rodrigue2014site}.
-\skhcomment{It is not clear how many bins \texttt{Phylobayes MPI} uses. I am going to go back and re-run the analysis specifying $k=4$ bins.}
 First, we compared the amino-acid preferences inferred by the pbMutSel model to the preferences measured in the experiments.
@@
 For \ref{fig:phylobayes}, we fit a pbMutSel model to the full HA tree. 
-We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
-\skhcomment{re: measure of convergence.
-I need add some analysis to fully address this point.
-1.
-I am going to run 2 chains and assess the congruency between these two chains.
-2.
-I am going to look at the traces using Andrew Rambaut's program tracer.
-Both of these are in progress}
+We ran one chain for 5500 steps, saved every sample, and discarded the first 550 samples as a burnin. 
 We used \texttt{PhyloBayes-MPI} program \texttt{readpb\_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences. 
+Convergence was assessed visually using Tracer~\skhcomment{Rambaut et al, 2018} and by the correlation of amino-acid preferences inferred by two independent chains (r=0.996).
 
@@
 These two H1 sequences are early and late representatives of the longest known human influenza lineage, and are of sufficiently high identity that different ExpCM and GY94 substitution models all estimate nearly identical branch lengths separating them.
-\skhcomment{add new run parameters}
 

### 2018-04-10 manuscript/manuscript.tex
clean up some references

@@
 \subsection*{HA amino-acid preferences from deep mutational scanning experiments}
-We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} \skhcomment{change citation} to define the amino-acid preferences that inform the ExpCM's.
+We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} to define the amino-acid preferences that inform the ExpCM's.
 We only used sites that can be unambiguously aligned in these H1 and H3 HAs. 

### 2018-04-11 manuscript/manuscript.tex
added Ho (2015) reference to Intro

@@
 Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular}. 
-However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, aiewsakun2016time}.
+However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, ho2015time, aiewsakun2016time}.
 For example, the integration of filoviruses into their host genomes indicate that Ebola and Marburg virus diverged from their common ancestor 7 to 12 million years ago---but the estimate of this divergence time based on phylogenetic analyses of the viral sequences is only $\sim$10,000 years ago~\citep{carroll2013molecular, taylor2014evidence}.

### 2018-04-11 manuscript/manuscript.tex
minor edits to all but Discussion

@@
 For highly diverged sequences, such phylogenetic techniques can estimate surprisingly recent divergence times. 
-In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are too recent.
+In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are sometimes too recent.
 This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
@@
 We find that models informed by experimentally measured site-specific amino-acid preferences estimate longer deep branches on phylogenies of influenza virus hemagglutinin.
-This lengthening of branches is due to more realistic stationary states of the models, and is independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
+This lengthening of branches is due to more realistic stationary states of the models, and is mostly independent of the branch-length-extension from modeling site-to-site variation in amino-acid substitution rate.
 The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
@@
 More recent work has used mutation-selection models to better account for purifying selection~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-These models explicitly account for the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression}.
+These models explicitly incorporate the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression, le2008phylogenetic, si2008empirical, wang2008class}.
 However, these approaches still ultimately involving inferring the site-specific purifying selection from natural sequence data.
@@
 The resulting experimental measurements of which amino acids are preferred at each protein site can be used to inform phylogenetic substitution models~\citep{bloom2014experimentally}.
-These experimentally informed codon models (ExpCM's) generally exhibit much better phylogenetic fit than most standard substitution models~\citep{hilton2017phydms, haddox2018mapping, lee2018deep, doud2015site}.
+These experimentally informed codon models (ExpCM's) generally exhibit much better phylogenetic fit than standard substitution models~\citep{hilton2017phydms, haddox2018mapping, lee2018deep, doud2015site}.
 
-Here we examined how using ExpCM's and other models of purifying selection affects the estimates of deep divergence times on a phylogenetic tree of influenza virus hemagglutinin (HA). 
-We found that ExpCMs that model site-specific amino-acid preferences estimate longer deep branches. 
-This extension of branch length is mostly independent and additive with that achieved by the more conventional approach of modeling rate variation. 
-ExpCM's that use experimental data estimate similar branch lengths to a mutation-selection model that infers the amino-acid preferences from the natural sequence data~\citep{lartillot2004bayesian, rodrigue2014site}.
-However, we also show that all of these mutation-selection models are limited by their failure to account for another feature of purifying selection: the fact that a site's amino-acid preferences shift over time due to epistasis.
+Here we examine how ExpCM's and other models of purifying selection estimate deep divergence times on a phylogenetic tree of influenza virus hemagglutinin (HA). 
+We find that ExpCMs estimate longer deep branches, and show that this extension of branch length is mostly independent and additive with that achieved by the more conventional approach of modeling rate variation.
+We also show that ExpCM's estimate similar branch lengths to a mutation-selection model that infers the amino-acid preferences from the natural sequence data rather than using values obtained in experiments.
+However, all of these mutation-selection models are limited by their failure to account for another feature of purifying selection: the fact that a site's amino-acid preferences shift over time due to epistasis.
 Therefore, truly accurate analyses of deep phylogenies need to account for the fact that amino-acid preferences vary across time as well as across sites.
@@
 \subsection*{Different ways substitution models account for purifying selection}
+Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
+Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
+The relative rate of these substitutions is referred to as dN/dS or $\omega$.
+In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average fixation rate of nonsynonymous mutations relative to synonymous ones.
+Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
+When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon}.
+We will refer to M0 Goldman-Yang models simply as GY94 models (\ref{eq:GY94}). 
+The gene-wide $\omega$ is usually $<1$~\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
 
@@
 
-Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
-Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
-The relative rate of these substitutions is referred to as dN/dS or $\omega$.
-In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average fixation rate of nonsynonymous mutations relative to synonymous ones.
-Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
-When these models have a single gene-wide $\omega$ they are classified as M0 by \citet{yang2000codon}.
-Here we will refer to M0 Goldman-Yang models simply as GY94 models (\ref{eq:GY94}). 
-The gene-wide $\omega$ is usually $<1$~\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.
-
 A single gene-wide $\omega$ ignores the fact that purifying selection is heterogeneous across sites.
@@
 A GY94+$\Gamma\omega$ captures the fact that the rate of nonsynonymous substitution can vary across sites. 
-However, this formulation treats all nonsynonymous substitutions equivalently, since the rate is agnostic to the amino-acid identity of the mutation.
+However, these models do not capture the fact that the same amino-acid mutation can have very different effects at different sites.
 
-A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
+A model formulation that accounts for the fact that purifying selection depends idiosyncratically on the specific amino-acid mutation at each site is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
 Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same protein are treated as selectively equivalent).
@@
 The GY94 model predicts the same behavior for all sites, since it does not have any site-specific parameters, with an asymptotic sequence identity of 0.062. 
-While this predicted sequence identity is higher than $\frac{1}{20} = 0.05$, due to redundant codon and nucleotide biases favoring certain amino acids, it is much lower than the pairwise identity of even the most diverged HAs in nature.
+While this predicted sequence identity is higher than $\frac{1}{20} = 0.05$ due to redundant codon and nucleotide biases favoring certain amino acids, it is much lower than the pairwise identity of even the most diverged HAs in nature.
 While it is of course possible that the identity of HAs in nature would become even lower given more time, it seems biochemically improbable that it would ever become as low as 0.062. 
@@
 We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
+Note that these simulations closely parallel those performed by \citet{halpern1998evolutionary} and \citet{wertheim2011purifying}.
 
@@
 We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
-The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
+The grey dashed line at $y=x$ represents what would be seen if the inferred branch lengths exactly matched those used in the simulations. 
 }
@@
 Circle denotes the H1 clade and triangle denotes the H3 clade.
-The trees are midpoint rooted based on the tree inferred by \texttt{RAxML} using the GTRCAT model. 
+The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
 }
@@
 {\bf Fitting of substitution models to the HA phylogenetic tree.}
+All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
 The models fit here are the same ones in~\ref{fig:empirical_trees}. 
-All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
 The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
@@
 
-Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or a site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree.
+Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree.
 \ref{fig:empirical_trees} shows the branch lengths optimized by each model on a common scale.
@@
 As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
-The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 76\% of the tree diameter extension of ExpCM(H1+H3 avg)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3 avg) versus GY94).
+The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 76\% of the tree diameter extension of ExpCM(H1+H3 avg)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (compare ExpCM(H1+H3 avg) to GY94).
 
@@
 This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
-As $\beta$ value decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
+As $\beta$ decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
 At the most extreme value of $\beta = 0$, the preferences would be perfectly uniform and look similar to the GY94 preferences in \ref{fig:empirical_trees}.
@@
 The fact that amino-acid preferences shift as a protein evolves leaves us with an inherent tension: models with site-specific amino-acid preferences only become important for accurate branch-length estimation as sequences become increasingly diverged, but this same divergence degrades the accuracy of extrapolating the amino-acid preferences from any given experiment across the phylogenetic tree.
-Crucially, this problem is expected to be more fundamental than the inability of a single deep mutational scanning experiment to measure amino-acid preferences in more than one genetic background.
-If amino-acid preferences shift during evolution, there simply will not be any model with a single set of time-homogeneous site-specific stationary states that accurately describes evolution along the entirety of a phylogenetic tree that covers a wide span of sequences.
+Crucially, this problem is more fundamental than the inability of a single deep mutational scanning experiment to measure amino-acid preferences in more than one genetic background.
+If amino-acid preferences shift during evolution, there simply will not be any single set of time-homogeneous site-specific preferences that accurately describes evolution along the entirety of a phylogenetic tree that covers a wide span of sequences.
 
@@
 Comparing ExpCM and pbMutSel models can help determine the ultimate limits of mutation-selection models that assign each site a single set of amino-acid preferences. 
-If the limitations of ExpCM's described above arise simply because the deep mutational scanning experiments do not correctly measure the ``true" amino-acid preferences of HA across the entirety of a highly diverged phylogenetic tree, then we would expect the pbMutSel models (which infer these preferences from the entire tree) to perform better.
-On the other hand, if the major limitation is that no single set of time-homogenous amino-acid preferences can fully describe HA evolution over the entire tree, then we would expect ExpCM and pbMutSel models to perform similarly.
+If the limitations of ExpCM's described above arise simply because the deep mutational scanning experiments do not correctly measure the ``true" amino-acid preferences across the entirety of a highly diverged phylogenetic tree, then we would expect the pbMutSel models (which infer these preferences from the entire tree) to perform better.
+On the other hand, if the major limitation is that no single set of time-homogenous amino-acid preferences can fully describe evolution over the entire tree, then we would expect ExpCM and pbMutSel models to perform similarly.
 
@@
 However, the estimates are not identical, and the tension between local and global accuracy of the amino-acid preferences is still apparent. 
-Specifically, the long branches between the H1 or H3 sequences used in the experiments and all other sequences were estimated to be slightly longer by the ExpCM, while many other branches were estimated to be slightly longer by the pbMutSel model.
+Specifically, the branches leading to the H1 or H3 sequences used in the experiments were estimated to be slightly longer by the ExpCM, while some other branches were estimated to be slightly longer by the pbMutSel model.
 The relatively longer branches leading to the experimental sequences when using the ExpCM(H1+H3 avg)+$\Gamma\omega$ suggests that the ``tree average" amino-acid preferences inferred by the pbMutSel model are not as accurate as the preferences from the deep mutational scanning for sequences close to those used in the experiments.
@@
 
-We downloaded all full-length, coding sequences for 15 of the 18 Influenza A virus HA subtypes from the Influenza Virus Resource Database~\citep{bao2008influenza} in June of 2017. 
-We excluded rare subtypes 15, 17, and 18, which have limited sequences in the database.  
+We downloaded all full-length, coding sequences for 15 of the 18 influenza A virus HA subtypes from the Influenza Virus Resource Database~\citep{bao2008influenza} in June of 2017. 
+We excluded rare subtypes 15, 17, and 18, which have few sequences in the database.  
 We filtered and aligned the sequences using \texttt{phydms\_prepalignment}~\citep{hilton2017phydms}. 
@@
 We used \texttt{PhyloBayes-MPI} program \texttt{readpb\_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences. 
-Convergence was assessed visually using Tracer~\skhcomment{Rambaut et al, 2018} and by the correlation of amino-acid preferences inferred by two independent chains (r=0.996).
+Convergence was assessed visually using Tracer (\url{http://tree.bio.ed.ac.uk/software/tracer/}) and by the correlation of amino-acid preferences inferred by two independent chains (r=0.996).
 
@@
 \subsection*{Software versions and computer code}
-All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}. \skhcomment{make repo public}
+All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}.
 The external computer programs that we used were
@@
 \item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
-\item \skhcomment{snakemake}
+\item \texttt{snakemake}~\citep{koster2012snakemake} version 3.11.2 (available at \url{https://snakemake.readthedocs.io/en/stable/}) to run the pipelines.
 \end{itemize}

### 2018-04-11 manuscript/manuscript.tex
revisions to Discussion

@@
 
-Here we tested how rate variation and site-specific amino-acid preferences affect the estimation of deep branches on phylogenetic trees. 
-We used ExpCM's defined by deep mutational scanning experiments to model site-specific amino-acid preferences and found that these models estimate longer branches than the GY94 model with a uniform stationary state.
-We found that ExpCM's estimate similar branch lengths to models which infer their stationary state from the natural sequences. 
-Finally, we found that the extension in branch length from site-specific amino-acid preferences is mostly independent from rate variation.
+We examined how estimates of deep branch lengths on phylogenetic trees are affected by accounting for the fact that proteins prefer specific amino acids at specific sites. 
+We did this by comparing inferences from models informed by experimental measurements of site-specific amino-acid preferences with more conventional codon substitution models, as well as with models that infer the amino-acid preferences from the natural sequences.
+We found that models that account for site-specific amino-acid preferences estimated deeper long branches, regardless of whether these preferences are measured experimentally or inferred from the sequence alignment.
+Additionally, we showed that the extension in branch length from site-specific amino-acid preferences is mostly independent of the extension that results from simply modeling rate variation.
 
-These results underscore the importance of modeling site-specific amino-acid preferences when estimating long branches. 
-As the simulations in \ref{fig:simulations} show, models with uniform stationary states will always underestimate the lengths of branches which have evolved under strong site-specific constraints. 
-We also found that the ExpCM's with the site-specific stationary states estimated longer branches than the GY94 model on the phylogenetic tree of influenza HA.
+Overall, our results underscore the importance of modeling purifying selection in a way that is more nuanced than simply allowing the substitution rate to vary across sites.
+Protein sites do not simply differ in their rates of substitution---different sites also prefer different amino acids.
+There are now two ways to account for this fact: using models informed by deep mutational scanning experiments, or using models that infer site-specific amino-acid preferences from the natural sequence alignment.
+Combining either type of model with rate variation increases the inferred length of deep branches relative to models that only incorporate rate variation.
 
-However, it is also clear that the amino-acid preferences defining the ExpCM's do not extrapolate well over a very diverged tree. 
-It is well established that a protein's site-specific amino-acid preferences are contextually dependent~\citep{gong2013stability, harms2014historical, tufts2014epistasis, ortlund2007crystal} and shift as the protein accumulates substitutions overtime~\citep{haddox2018mapping, starr2018pervasive, pollock2012amino, doud2015site, shah2015contingency,bazykin2015changing}.
-The site-specific amino-acid preferences we used to define the ExpCM's were only measured in one genetic background. 
-Therefore, these preferences are only one snapshot of the functional constraint and do not capture any of the changes over time. 
-The model's ignorance of shifting preferences is reflected in our results: The ExpCM's have the largest effect on branch length extension near the HA from the experiment.
+However, assuming a single set of site-specific amino-acid preferences is still an imperfect way to model evolution over a highly diverged phylogenetic tree. 
+In the case of the experimentally informed models, it is fairly obvious why this is true: the amino-acid preferences are measured in just one genetic background, and therefore provide only a single snapshot of preferences that shift over evolutionary time due to epistasis~\citep{pollock2012amino, shah2015contingency, bazykin2015changing, haddox2018mapping, starr2018pervasive, doud2015site}.
+As a result, experimentally measured amino-acid preferences are most accurate for sequences similar to the one used in the experiment, and so cause the largest increases in branch length in that region of the phylogenetic tree.
+However, this limitation is not unique to experimentally informed models, but is a general limitation of describing purifying selection using a single set of site-independent and time-homogenous amino-acid preferences.
+For instance, we showed that averaging experimental measurements on two protein homologs does a somewhat better job of capturing the ``average'' constraint across the tree, and performs similarly to approaches that infer the ``average'' preferences from natural sequence data~\citep{rodrigue2010mutation,rodrigue2014site}.
+But even these ``average'' preferences exhibit a tradeoff between local and global accuracy for the inference of deep branch lengths.
 
-The effect of shifting preferences on branch length estimation is not specific to ExpCM's but is a general limitation of site-independent, time-homogenous models. 
-Even though the stationary state of the pbMutSel model is more representative of the ``tree average" stationary state, the ExpCM with average preferences has a larger effect on the branches from the H1 and H3 sequences.
-The site-specific amino-acid preferences of a protein shift across time and models with a single stationary state across the entire tree cannot capture these dynamics. 
-Therefore, while it is clear that modeling site-specific amino-acid preferences is important, neither the ExpCM nor the pbMutSel is able to capture how these preferences shift over time.
-
-Clearly, the site-independence and time-homogeneity assumptions of the ExpCM's and pbMutSel models inhibit their ability to accurately describe the effect of shifting preferences. 
-Models which account for context-dependent of amino-acid preferences have been developed but are still in their infancy.
-Thus far their application has been restricted to a few sequences~\citep{rodrigue2005site, choi2007quantifying}, short evolutionary distances~\citep{bordner2013new}, or simulated sequences~\citep{goldstein2017sequence}.
-
-Currently, there are two main challenges associated with context-dependence models. 
-The first challenge is how to efficiently evaluate these inherently more complex models mathematically so the models can be applied to large phylogenetic trees~\citep{rodrigue2009computational}. 
-The second challenge is developing the correct function to model shifting preferences. 
-For example, \citet{rodrigue2009computational} showed that while site-dependent amino-acid models outperformed site-independent amino-acid models, they were outperformed by sophisticated site-independent codon models.
-Most context-dependent models evaluate amino-acid changes as how they might affect a single protein conformation~\citep{rodrigue2005site, choi2007quantifying,bordner2013new, goldstein2017sequence}.
-However, in nature a protein may undergo conformational changes in order to perform some function. 
-Models which account for shifting preferences should also be flexible enough to be able to describe multiple, necessary protein conformations. 
-Such models will be critical for describing the long-term evolution of a protein and accurately estimating the length of deep branches on phylogenetic trees.
+So while modeling site-specific amino-acid preferences is a clear improvement over most conventional models, the next step towards greater accuracy will require relaxing the assumption that these preferences are time homogeneous and site independent.
+Of course, many authors have pointed out the shortcomings of models that fail to account for the full site-interdependent complexity of purifying selection~\citep{rodrigue2005site, choi2007quantifying, pollock2012amino, goldstein2017sequence}.
+However, the challenge is to overcome these shortcomings with models that are tractable for real phylogenetic questions.
+There are two main issues: first, the Felsenstein pruning algorithm~\citep{felsenstein1981evolutionary} that is typically used to evaluate phylogenetic likelihoods breaks down when sites are no longer treated independently.
+Some alternative algorithms have been proposed~\citep{bordner2013new,rodrigue2009computational,rodrigue2005site, choi2007quantifying}, but they are still in their infancy.
+Second, site-interdependent models require a realistic ``fitness function'' that describes the interactions among sites.
+It appears that typical structural modeling programs are insufficient for this purpose~\citep{rodrigue2009computational}.
+But hope comes from experimental progress in measuring actual site-interdependent constraints on proteins~\citep{olson2014comprehensive,wu2016adaptation,steinberg2016shifting,li2016fitness}, combined with new methods for using these measurements to parameterize fitness functions~\citep{sailer2017detecting,otwinowski2018inferring,otwinowski2018protein}.
+Perhaps some day such truly realistic models might be useful for phylogenetic inference.
+Until that day, our work shows that modeling a single set of time homogenous amino-acid preferences provides at least some improvement.
 

### 2018-04-12 manuscript/manuscript.tex
fixed reference formatting

@@
 \item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
-\item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML} to infer tree topology.
+\item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML}) to infer tree topology.
 \item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.

### 2018-04-13 manuscript/manuscript.tex
fixed problematic sentence in Intro
added EM comments to manuscript text

@@
 % added by SKH
-%\usepackage{lineno}
-%\linenumbers
+\usepackage{lineno}
+\linenumbers
 
@@
 \noindent  
+\skhcomment{from EM: title change to ``Modeling site-specific amino-acid preferences deepens phylogenetic divergence time estimates of viral proteins"}
+\skhcomment{from EM: I know you know better, but the abstract reads like you are conflating branch length estimates and calendar time. I suggest making the connection more explicit}
 Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
-For highly diverged sequences, such phylogenetic techniques can estimate surprisingly recent divergence times. 
-In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are sometimes too recent.
+For highly diverged sequences, such phylogenetic techniques \skhcomment{EM removed ``can"} estimate surprisingly recent divergence times. 
+In the case of viruses, independent evidence indicates that the estimates of deep divergence times from molecular phylogenetics are sometimes too recent.
 This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
@@
 The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
-However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
-Overall, our work underscores the importance of modeling site-specific amino-acid preferences when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to account for how these preferences shift over time.
+However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves\skhcomment{EM: I don't see how this is a counter-point for your argument yet-- your work is time-homogeneous too.}.
+Overall, our work underscores the importance of modeling site-specific amino-acid preferences when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to account for how these preferences shift over time\skhcomment{More about deep below, but here I just want to make sure that this deep and the "deepens" of the title are the same deep}.
 \end{abstract}
@@
 
-Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular}. 
+Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular} \skhcomment{EM: Again, we need to hear about how BL estimates are converted to calendar time}.
+\skhcomment{I think EM made this comment ``Can you make a more explicit argument as to why this might be a bad idea?" about the sentence above but it was unclear from the pdf.
+I narrowed it down to somewhere in the introduction before the DMS paragraph.}
 However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, ho2015time, aiewsakun2016time}.
@@
 These models explicitly incorporate the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression, le2008phylogenetic, si2008empirical, wang2008class}.
-However, these approaches still ultimately involving inferring the site-specific purifying selection from natural sequence data.
+However, these approaches require inferring the site-specific purifying selection from natural sequence data.
 
@@
 The resulting experimental measurements of which amino acids are preferred at each protein site can be used to inform phylogenetic substitution models~\citep{bloom2014experimentally}.
-These experimentally informed codon models (ExpCM's) generally exhibit much better phylogenetic fit than standard substitution models~\citep{hilton2017phydms, haddox2018mapping, lee2018deep, doud2015site}.
+These experimentally informed codon models (ExpCM's) \skhcomment{ExpCM's vs.
+ExpCMs}generally exhibit much better phylogenetic fit than standard substitution models~\citep{doud2015site, hilton2017phydms, haddox2018mapping, lee2018deep}.
 
-Here we examine how ExpCM's and other models of purifying selection estimate deep divergence times on a phylogenetic tree of influenza virus hemagglutinin (HA). 
+Here we examine how ExpCM's and other models of purifying selection estimate deep divergence times on a phylogenetic tree of influenza virus hemagglutinin (HA)\skhcomment{At this point I don't have a clear idea of what "deep" means in this context.
+At this point I just think deep = far back in time}.
 We find that ExpCMs estimate longer deep branches, and show that this extension of branch length is mostly independent and additive with that achieved by the more conventional approach of modeling rate variation.
@@
 \subsection*{Different ways substitution models account for purifying selection}
-Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
+Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}\skhcomment{accurate in terms of what.
+Also, needs to be speci ed that this is for protein-coding genes}.
 Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
@@
 \textbf{Different ways codon models account for purifying selection.}
-(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
+(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution \skhcomment{(blue circles)}. 
 For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin~\citep{yang1994maximum,yang2000codon}. 
@@
 
-A model formulation that accounts for the fact that purifying selection depends idiosyncratically on the specific amino-acid mutation at each site is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
+Mutation-selection models account for the fact that purifying selection depends idiosyncratically on the specific amino-acid mutation at each site~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
 Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same protein are treated as selectively equivalent).
@@
 Given a single branch, a substitution model transforms sequence identity into branch length.
-Under a molecular-clock assumption, this branch length is proportional to time.
+Under a molecular-clock assumption, this branch length is proportional to time\skhcomment{finally!!}.
 The transformation from sequence identity to branch length is trivial when the sequence identity is high.
@@
 
+\skhcomment{You've run into a huge personal pet peeve here.
+Topic sentences should make statements about content.
+Tell me what I need to know from this paragraph about how the models make different predictions, then the rest of the paragraph can back this up using further statements and parenthetical references to the figure.}
 \ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
@@
 Accounting for site-to-site dN/dS rate variation in GY94 models affects the rate at which the asymptotic sequence identity is approached, but not the actual value of this asymptote.
+\skhcomment{This paragraph is much better, with a real topic sentence :+1} 
 For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but that the asymptote is identical for both models. 
@@
 
-Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree.
+Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree\skhcomment{The diameter analysis in this paragraph hasn't convinced me that the effect is especially pronounced for deep branches.}.
 \ref{fig:empirical_trees} shows the branch lengths optimized by each model on a common scale.
@@
 Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
-Specifically, the tree's diameter (the distance from the two most divergent tips) for the GY94+$\Gamma\omega$ model is 159\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
+Specifically, the tree's diameter (the distance between the two most divergent tips) for the GY94+$\Gamma\omega$ model is 159\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
 The tree diameter is 122\% and 135\% of the GY94 model tree diameter for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, and 160\% of the GY94 model for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
@@
 
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}).
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}) \skhcomment{If you really want to make this point, I think you'll need to make plots or tests of some kind.
+My eyeball test is somewhat inconclusive.}.
+\skhcomment{Second EM comment ``really, I'm just supposed to stare at this?"} 
 Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
@@
 
-\ref{fig:compete} shows that as the divergence from the sequence used in the deep mutational scan increases, the value of $\beta$ decreases. 
+The value of $\beta$ decreases as the divergence from the sequence used in the deep mutational scan increases~\ref{fig:compete}. 
 This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
 As $\beta$ decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
-At the most extreme value of $\beta = 0$, the preferences would be perfectly uniform and look similar to the GY94 preferences in \ref{fig:empirical_trees}.
+At the most extreme value of $\beta = 0$, the preferences would be perfectly uniform and look similar to the GY94 preferences in \ref{fig:empirical_trees}\skhcomment{If you had some random preferences, would they really hit beta = 0?}.
 In reality, $\beta$ never reaches a value this low, indicating the deep mutational scanning experiments remain somewhat informative about real natural selection across the entire swath of HAs. 
@@
 Until that day, our work shows that modeling a single set of time homogenous amino-acid preferences provides at least some improvement.
+
+\skhcomment{Overall, the effect seems kind of small compared to the size of problems you describe in the introduction. Do you really think that we can fix those time estimates with perfect modeling?
+
+I really think you need to quantify the effects you see in figure 4 with plots rather than saying "look at the trees". If I missed something let me know.
+
+Discussion: would it be possible to use a pbMutSel model, but fix its values to those corresponding to an ExpCM model at the point in the tree where the measurement was made?
+I don't know if you can make the two models talk to each other though.}
 
@@
 
-\subsection*{HA sequences and tree topology.}
+\subsection*{HA sequences and tree topology}
 

### 2018-04-13 manuscript/manuscript.tex
added revisions after Erick Matsen comments

--- 2018-04-12 manuscript/manuscript.tex
@@
 
-\title{Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.}
+\title{Modeling site-specific amino-acid preferences deepens phylogenetic divergence estimates for viral proteins.}
 
@@
 Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
-For highly diverged sequences, such phylogenetic techniques can estimate surprisingly recent divergence times. 
-In the case of viruses, there is independent evidence that the estimates of deep divergence times from molecular phylogenetics are sometimes too recent.
+For highly diverged sequences, such phylogenetic techniques sometimes estimate surprisingly recent divergence times. 
+In the case of viruses, independent evidence indicates that the estimates of deep divergence times from molecular phylogenetics are sometimes too recent.
 This discrepancy is caused in part by inadequate models of purifying selection leading to branch-length underestimation.
@@
 The branch-length extension from experimentally informed site-specific models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
-However, the improvements from these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
+However, the improvements from all of these site-specific but time-homogeneous and site-independent models are limited by the fact that a protein's amino-acid preferences gradually shift as it evolves.
 Overall, our work underscores the importance of modeling site-specific amino-acid preferences when estimating deep divergence times---but also shows the inherent limitations of approaches that fail to account for how these preferences shift over time.
@@
 \section*{Introduction} 
-
 Molecular phylogenetics is commonly used to estimate the historical timing of evolutionary events \citep{yang2012molecular}.
+This is done by estimating branch lengths based on the inferred number of substitutions, and then converting these branch lengths into units of time under the assumption of a molecular clock~\citep{zuckerkandl1965,drummond2006relaxed}.
 However, phylogenetic estimates of the divergence times of many viral lineages are clearly too recent~\citep{duchene2014analyses, ho2015time, aiewsakun2016time}.
@@
 The resulting experimental measurements of which amino acids are preferred at each protein site can be used to inform phylogenetic substitution models~\citep{bloom2014experimentally}.
-These experimentally informed codon models (ExpCM's) generally exhibit much better phylogenetic fit than standard substitution models~\citep{hilton2017phydms, haddox2018mapping, lee2018deep, doud2015site}.
+These experimentally informed codon models (ExpCMs) generally exhibit much better phylogenetic fit than standard substitution models~\citep{doud2015site, hilton2017phydms, haddox2018mapping, lee2018deep}.
 
-Here we examine how ExpCM's and other models of purifying selection estimate deep divergence times on a phylogenetic tree of influenza virus hemagglutinin (HA). 
+Here we examine how ExpCMs and other models of purifying selection estimate branch lengths on a phylogenetic tree of influenza virus hemagglutinin (HA). 
 We find that ExpCMs estimate longer deep branches, and show that this extension of branch length is mostly independent and additive with that achieved by the more conventional approach of modeling rate variation.
-We also show that ExpCM's estimate similar branch lengths to a mutation-selection model that infers the amino-acid preferences from the natural sequence data rather than using values obtained in experiments.
+We also show that ExpCMs estimate similar branch lengths to a mutation-selection model that infers the amino-acid preferences from the natural sequence data rather than using values obtained in experiments.
 However, all of these mutation-selection models are limited by their failure to account for another feature of purifying selection: the fact that a site's amino-acid preferences shift over time due to epistasis.
@@
 \subsection*{Different ways substitution models account for purifying selection}
-Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
+Here we consider how purifying selection is handled by codon models, which are the most sophisticated of the three classes (nucleotide, codon, and amino acid) of phylogenetic substitution models in widespread use for protein-coding genes~\citep{arenas2015trends}.
 Standard codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
@@
 \textbf{Different ways codon models account for purifying selection.}
-(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue circles). 
-For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin~\citep{yang1994maximum,yang2000codon}. 
+(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue line). 
+For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin (blue circles)~\citep{yang1994maximum,yang2000codon}. 
 A gamma distribution (denoted by $\Gamma$) with $K=4$ bins is shown here.
@@
 It is challenging to obtain values for these amino-acid preference parameters in a maximum-likelihood framework without overfitting the data~\citep{rodrigue2013statistical}.
-Here we will primarily use experimentally informed codon models (ExpCM's), which define the site-specific amino-acid preference parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
+Here we will primarily use experimentally informed codon models (ExpCMs), which define the site-specific amino-acid preference parameters \textit{a priori} from deep mutational scanning experiments so that they do not need to be fit from phylogenetic data~\citep[see Methods and][]{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
 Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters is similar to a non-site-specific substitution model.
@@
 
-However, mutation-selection models such as ExpCM's have site-specific stationary states.
+However, mutation-selection models such as ExpCMs have site-specific stationary states.
 They predict that different sites will have different asymptotic sequence identities (\ref{fig:decay})---a prediction that accords with the empirical observation that some sites are much more variable than others in alignments of highly diverged sequences.
@@
 However, there was no systematic underestimation of long branches by the ExpCM and ExpCM+$\Gamma\omega$ models.
-The improved performance of the ExpCM's is due to their modeling of the site-specific amino-acid preferences: if we parameterize ExpCM's by amino-acid preferences that have been averaged across HA sites (and so are no longer site-specific), then they perform no better than GY94 models (\ref{fig:simulations}).
+The improved performance of the ExpCMs is due to their modeling of the site-specific amino-acid preferences: if we parameterize ExpCMs by amino-acid preferences that have been averaged across HA sites (and so are no longer site-specific), then they perform no better than GY94 models (\ref{fig:simulations}).
 Therefore, models with uniform stationary states underestimate the length of long branches in phylogenies of sequences that have evolved under strong site-specific amino-acid preferences.
@@
 The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
-The ExpCM's use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
+The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
 Circle denotes the H1 clade and triangle denotes the H3 clade.
@@
 
-The foregoing section shows the superiority of ExpCM's to GY94 models for estimating long branches on phylogenies simulated with ExpCM's.
+The foregoing section shows the superiority of ExpCMs to GY94 models for estimating long branches on phylogenies simulated with ExpCMs.
 But how do these models perform on real data?
 Real genes do evolve under functional constraint, but these constraints are almost certainly more complex than what is modeled by an ExpCM.
-However, if ExpCM's do a substantially better job than GY94 models of capturing the true constraints, then we might still expect them to estimate more accurate branch lengths.
+However, if ExpCMs do a substantially better job than GY94 models of capturing the true constraints, then we might still expect them to estimate more accurate branch lengths.
 
 To test the models on real data, we used actual sequences of influenza HA. 
-The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
+The topology of HA phylogenetic trees makes these sequences an interesting test case for branch-length estimation.
 HA consists of a number of different subtypes.
@@
 Therefore, we also created a third set of amino-acid preferences by averaging the measurements for the H1 and H3 HAs, under the conjecture that these averaged preferences might better describe the ``average'' constraint on sites across the full HA tree~(\ref{suppfig:prefs_average}).
-These three sets of HA amino-acid preferences define three different ExpCM's.
+These three sets of HA amino-acid preferences define three different ExpCMs.
 
@@
 The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
-All ExpCM's fit a stringency parameter $>1$.
+All ExpCMs fit a stringency parameter $>1$.
 } 
@@
 
-We fit the GY94 model and each of the three ExpCM's to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
-\ref{tab:empirical_data} shows that all ExpCM's fit the actual data much better than the GY94 models.
+We fit the GY94 model and each of the three ExpCMs to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
+\ref{tab:empirical_data} shows that all ExpCMs fit the actual data much better than the GY94 models.
 The best fit was for the ExpCM informed by the average of the H1 and H3 deep mutational scans.
-For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCM's without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ model (\ref{tab:empirical_data}).
+For all models, incorporating $\Gamma\omega$ rate variation improved the fit, although even ExpCMs without $\Gamma\omega$ greatly outperformed the GY94+$\Gamma\omega$ model (\ref{tab:empirical_data}).
 As mentioned in the previous section, $\omega$ is generally $<1$ when a single value is fit to all sites in a gene~\citep{murrell2015gene}, and this is the case for all the models we tested (\ref{tab:empirical_data}).
-However, the ExpCM's always fit an $\omega$ greater than the GY94 model, suggesting that the site-specific amino-acid preferences capture some of the purifying selection that the GY94 models can represent only via a small $\omega$.
+However, the ExpCMs always fit an $\omega$ greater than the GY94 model, suggesting that the site-specific amino-acid preferences capture some of the purifying selection that the GY94 models can represent only via a small $\omega$.
 Among the models with $\Gamma\omega$, the GY94+$\Gamma\omega$ model fits all four $\omega$ categories to values $\ll$1, but the ExpCM+$\Gamma\omega$ models fit one of the $\omega$ categories to a value $>1$.
@@
 The ExpCM $\omega$ is the relative rate of fixation of nonsynonymous to synonymous mutations \textit{after} accounting for the functional constraints described by the amino-acid preferences.
-This more realistic null model gives ExpCM's enhanced power to detect diversifying selection for amino-acid change~\citep{bloom2017identification, rodrigue2017detecting}, which is known to occur at some sites in HA due to immune selection~\citep{bedford2014integrating}.
+This more realistic null model gives ExpCMs enhanced power to detect diversifying selection for amino-acid change~\citep{bloom2017identification, rodrigue2017detecting}, which is known to occur at some sites in HA due to immune selection~\citep{bedford2014integrating}.
 
-Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree.
+Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer branches on the HA tree.
 \ref{fig:empirical_trees} shows the branch lengths optimized by each model on a common scale.
@@
 Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
-Specifically, the tree's diameter (the distance from the two most divergent tips) for the GY94+$\Gamma\omega$ model is 159\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
-The tree diameter is 122\% and 135\% of the GY94 model tree diameter for ExpCM's informed by H1 or H3 amino-acid preferences, respectively, and 160\% of the GY94 model for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
+Specifically, the tree's diameter (the distance between the two most divergent tips) for the GY94+$\Gamma\omega$ model is 159\% of the GY94 model tree diameter (\ref{supptab:tree_diameter}).
+The tree diameter is 122\% and 135\% of the GY94 model tree diameter for ExpCMs informed by H1 or H3 amino-acid preferences, respectively, and 160\% of the GY94 model for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_diameter}).
 
@@
 This can be seen by examining the ExpCM+$\Gamma\omega$ models, which combine $\Gamma\omega$ rate variation with site-specific amino-acid preferences.
-As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
+As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCMs).
 The near independence of these effects is quantified in \ref{supptab:tree_diameter}, which shows that 76\% of the tree diameter extension of ExpCM(H1+H3 avg)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (compare ExpCM(H1+H3 avg) to GY94).
 
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}).
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}) \jdbcomment{I think we should make a Figure 5 that looks similar to the current Figure 6B, that would show GY94 versus each ExpCM, GY94+gamma versus each ExpCM+gamma, and each model versus it's gamma.
+These could be panels (one or two rows) and panel B.}
 Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
@@
 The ExpCM informed by the average of the H1 and H3 data has a more uniform effect across the tree, but still most strongly extends branches leading to either the H1 or H3 clade.
-Therefore, \ref{fig:empirical_trees} shows that ExpCM's estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.
+Therefore, \ref{fig:empirical_trees} shows that ExpCMs estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.
 
@@
 This epistasis can lead to shifts in a protein's amino-acid preferences over evolutionary time~\citep{pollock2012amino, doud2015site, shah2015contingency, bazykin2015changing, haddox2018mapping}.
-Because the deep mutational scanning experiments that inform our ExpCM's were each performed in the context of a single HA genetic background, their measurements do not account for the accumulation of epistatic shifts in amino-acid preferences as HA evolves.
+Because the deep mutational scanning experiments that inform our ExpCMs were each performed in the context of a single HA genetic background, their measurements do not account for the accumulation of epistatic shifts in amino-acid preferences as HA evolves.
 Therefore, an ExpCM is expected to most accurately describe the evolution of sequences closely related to the one used in the experiment.
@@
 
-\ref{fig:compete} shows that as the divergence from the sequence used in the deep mutational scan increases, the value of $\beta$ decreases. 
-This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
+The value of $\beta$ decreases as the divergence from the sequence used in the deep mutational scan increases~\ref{fig:compete}. 
+This inverse relationship between $\beta$ and overall divergence is seen for the ExpCMs informed by both the H1 and H3 experiments.
 As $\beta$ decreases, the preferences ``flatten" and so the ExpCM draws less information from the experiment. 
@@
 Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
-The \texttt{phydms} program implementing ExpCM's and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree.
+The \texttt{phydms} program implementing ExpCMs and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree.
 }
@@
 
-The previous sections used ExpCM's, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
+The previous sections used ExpCMs, which are mutation-selection models that use site-specific amino-acid preferences that have been measured by experiments. 
 However, there are other mathematically similar implementations of mutation-selection models that infer the amino-acid preferences directly from the natural sequence data. 
@@
 (Note that the maximum-likelihood implementations of \citet{tamuri2012estimating,tamuri2014penalized} are designed for estimating the amino-acid preferences, \emph{not} for phylogenetic inference.)
-The model most comparable to our ExpCM's is the codon mutation-selection model implemented in \texttt{PhyloBayes-MPI}, which we will refer to as pbMutSel~\citep{rodrigue2014site}. 
+The model most comparable to our ExpCMs is the codon mutation-selection model implemented in \texttt{PhyloBayes-MPI}, which we will refer to as pbMutSel~\citep{rodrigue2014site}. 
 In the pbMutSel model, the amino-acid preferences are modeled using Dirichlet processes rather than derived from experiments. 
@@
 Comparing ExpCM and pbMutSel models can help determine the ultimate limits of mutation-selection models that assign each site a single set of amino-acid preferences. 
-If the limitations of ExpCM's described above arise simply because the deep mutational scanning experiments do not correctly measure the ``true" amino-acid preferences across the entirety of a highly diverged phylogenetic tree, then we would expect the pbMutSel models (which infer these preferences from the entire tree) to perform better.
+If the limitations of ExpCMs described above arise simply because the deep mutational scanning experiments do not correctly measure the ``true" amino-acid preferences across the entirety of a highly diverged phylogenetic tree, then we would expect the pbMutSel models (which infer these preferences from the entire tree) to perform better.
 On the other hand, if the major limitation is that no single set of time-homogenous amino-acid preferences can fully describe evolution over the entire tree, then we would expect ExpCM and pbMutSel models to perform similarly.
@@
 \subsection*{HA amino-acid preferences from deep mutational scanning experiments}
-We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} to define the amino-acid preferences that inform the ExpCM's.
+We used amino-acid preferences measured in deep mutational scans of the A/WSN/1933 H1 HA~\citep{doud2016accurate} and the A/Perth/2009 H3 HA~\citep{lee2018deep} to define the amino-acid preferences that inform the ExpCMs.
 We only used sites that can be unambiguously aligned in these H1 and H3 HAs. 
@@
 \end{itemize}
+
+\subsection*{Acknowledgments}
+We thank Erick Matsen and Trevor Bedford for helpful comments about the project and manuscript.
+SKH is supported in part by training grant T32AI083203 from the NIAID of the National Institutes of Health.
+This work was supported by the NIAID and NIGMS of the NIH under grant numbers R01AI127893 and R01GM102198. 
+JDB is supported in part by a Faculty Scholars grant from the Howard Hughes Medical Institute and the Simons Foundation.
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 

### 2018-04-16 manuscript/manuscript.tex
removed period from title
added draft of empirical scatterplots

@@
 
-\title{Modeling site-specific amino-acid preferences deepens phylogenetic divergence estimates for viral proteins.}
+\title{Modeling site-specific amino-acid preferences deepens phylogenetic divergence estimates for viral proteins}
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=0.7\textwidth]{figures/empirical_scatter.pdf}}
+\caption{\label{fig:empirical_scatter}
+\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
+Points denote the branch lengths between all pairs of tips on the tree. 
+Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
+Each column has ExpCMs defined by a different deep mutational scanning amino-acid preference set. 
+(A) The ExpCM defined by the H1, H3, or (H1+H3) average deep mutational scanning amino-acid preferences estimated longer branches from reference sequence used in the deep mutational scanning experiment.
+(B) The addition of $\Gamma\omega$ rate variation leads to more uniform extension of branch length than the addition of site-specific amino-acid preferences.
+}
+\end{figure}
+
 Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer branches on the HA tree.
@@
 
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees}) \jdbcomment{I think we should make a Figure 5 that looks similar to the current Figure 6B, that would show GY94 versus each ExpCM, GY94+gamma versus each ExpCM+gamma, and each model versus it's gamma.
-These could be panels (one or two rows) and panel B.}
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_scatter})
 Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
@@
 The ExpCM informed by the average of the H1 and H3 data has a more uniform effect across the tree, but still most strongly extends branches leading to either the H1 or H3 clade.
-Therefore, \ref{fig:empirical_trees} shows that ExpCMs estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.
+Therefore, \ref{fig:empirical_trees} and \ref{fig:empirical_scatter} show that ExpCMs estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.
 

### 2018-04-16 manuscript/manuscript.tex
updates to the empirical scatter plots

@@
 \caption{\label{fig:empirical_scatter}
-\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
+\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments does not extend branch lengths uniformly across the HA tree.}
 Points denote the branch lengths between all pairs of tips on the tree. 
 Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
-Each column has ExpCMs defined by a different deep mutational scanning amino-acid preference set. 
-(A) The ExpCM defined by the H1, H3, or (H1+H3) average deep mutational scanning amino-acid preferences estimated longer branches from reference sequence used in the deep mutational scanning experiment.
+Each column is a different site-specific amino-acid preference set. 
+(A) The ExpCMs defined by the H1, H3, or H1+H3 average deep mutational scanning amino-acid preferences estimate longer branches from the reference sequence used in the respective deep mutational scanning experiment.
 (B) The addition of $\Gamma\omega$ rate variation leads to more uniform extension of branch length than the addition of site-specific amino-acid preferences.

### 2018-04-16 manuscript/manuscript.tex
tweaks to empirical scatter figure
made empirical scatter figure 2 not 3 lines

@@
 The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
+The branch lengths are plotted for quantitative comparison in \ref{fig:empirical_scatter}. 
 }
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.7\textwidth]{figures/empirical_scatter.pdf}}
+\centerline{\includegraphics[width=\textwidth]{figures/empirical_scatter.pdf}}
 \caption{\label{fig:empirical_scatter}
@@
 Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
-Each column is a different site-specific amino-acid preference set. 
+The site-specific amino-acid preferences defining the ExpCM's are labeled above each plot.
+These data are the branch lengths in \ref{fig:empirical_trees} plotted quantitatively for comparison.
 (A) The ExpCMs defined by the H1, H3, or H1+H3 average deep mutational scanning amino-acid preferences estimate longer branches from the reference sequence used in the respective deep mutational scanning experiment.
@@
 
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_scatter})
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_scatter}).
 Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.

### 2018-04-16 manuscript/manuscript.tex
small tweaks to texts re: empirical scatter

@@
 \textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments does not extend branch lengths uniformly across the HA tree.}
+We quantitatively plotted the branch lengths in \ref{fig:empirical_trees} for comparison.
 Points denote the branch lengths between all pairs of tips on the tree. 
 Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
-The site-specific amino-acid preferences defining the ExpCM's are labeled above each plot.
-These data are the branch lengths in \ref{fig:empirical_trees} plotted quantitatively for comparison.
+The site-specific amino-acid preference set defining the ExpCM is labeled above each each plot.
 (A) The ExpCMs defined by the H1, H3, or H1+H3 average deep mutational scanning amino-acid preferences estimate longer branches from the reference sequence used in the respective deep mutational scanning experiment.
@@
 However, \ref{fig:compete} shows that the amino-acid preferences clearly become less informative about natural selection as we move away from the experimental sequence on the tree.
-This shifting of amino-acid preferences helps explain why the ExpCM informed by the average of the H1 and H3 experiments performs best (\ref{tab:empirical_data} and \ref{fig:empirical_trees}): averaging the measurements across these two HAs is a heuristic method of accounting for shifts in preferences during HA evolution.
+This shifting of amino-acid preferences helps explain why the ExpCM informed by the average of the H1 and H3 experiments performs best (\ref{tab:empirical_data}, \ref{fig:empirical_trees}, and \ref{fig:empirical_scatter}): averaging the measurements across these two HAs is a heuristic method of accounting for shifts in preferences during HA evolution.
 

### 2018-04-16 manuscript/manuscript.tex
a few revisions

@@
 To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
-Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees} using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
+Specifically, we simulated alignments of sequences along the HA phylogenetic tree using an ExpCM parameterized by the amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}.
 We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
@@
 \textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
-We simulated alignments along a HA phylogenetic tree (see \ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
+We simulated alignments along a HA phylogenetic tree using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
 We then inferred the branch lengths of this tree from the simulated alignments.
@@
 \subsection*{Experimentally informed site-specific models estimate longer branches on real data.}
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
-\caption{\label{fig:empirical_trees}
-\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
-The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
-The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
-The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
-Circle denotes the H1 clade and triangle denotes the H3 clade.
-The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
-The branch lengths are plotted for quantitative comparison in \ref{fig:empirical_scatter}. 
-}
-\end{figure}
 
@@
 \ref{suppfig:prefs_lee}.
-The H1 and H3 HAs have only $\sim$42\% amino-acid identity, and so are separated by a large distance on the phylogenetic tree (see triangle and circle on \ref{fig:empirical_trees}).
+The H1 and H3 HAs have only $\sim$42\% amino-acid identity.
 As described in \citet{lee2018deep}, the amino-acid preferences clearly differ between the H1 and H3 HA at a substantial number of sites (these differences are apparent in a simple visual comparison of \ref{suppfig:prefs_doud} and
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
+\caption{\label{fig:empirical_trees}
+\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
+The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
+The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
+Circle denotes the H1 clade and triangle denotes the H3 clade.
+The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
+This figure enables qualitative visualization of the trees; for a quantitative comparison of branch lengths optimized by different models, see \ref{fig:empirical_scatter}. 
+}
+\end{figure}
+
 We fit the GY94 model and each of the three ExpCMs to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
@@
 \caption{\label{fig:empirical_scatter}
-\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments does not extend branch lengths uniformly across the HA tree.}
-We quantitatively plotted the branch lengths in \ref{fig:empirical_trees} for comparison.
-Points denote the branch lengths between all pairs of tips on the tree. 
-Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
-The site-specific amino-acid preference set defining the ExpCM is labeled above each each plot.
-(A) The ExpCMs defined by the H1, H3, or H1+H3 average deep mutational scanning amino-acid preferences estimate longer branches from the reference sequence used in the respective deep mutational scanning experiment.
-(B) The addition of $\Gamma\omega$ rate variation leads to more uniform extension of branch length than the addition of site-specific amino-acid preferences.
+\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments extends branch lengths, especially for branches leading to the HA used in the experiment.}
+The points indicate the total length of branches separating all pairs of tips on the HA phylogenetic tree when the tree is optimized under the indicated model. 
+Blue and orange denote branches that lead to the H1 and H3 HAs used in the deep mutational scanning. 
+The amino-acid preference set defining the ExpCM is labeled above each each plot.
+(A) ExpCMs defined by amino-acid preferences from any of the deep mutational scanning experiments estimate generally longer branches than the GY94 model, with the increase particularly pronounced for branches leading to the HA used in the experiment.
+(B) The addition of $\Gamma\omega$ rate variation further extends branch lengths, without any apparent bias towards the HAs used in the experiment.
+Note that this figure shows the same data as \ref{fig:empirical_trees} in a different form.
 }
@@
 
-However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_scatter}).
+However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_trees} and \ref{fig:empirical_scatter}).
 Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.

### 2018-04-16 manuscript/manuscript.tex
_MBE_ submitted version

@@
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 
@@
 
-\title{Modeling site-specific amino-acid preferences deepens phylogenetic divergence estimates for viral proteins}
+\title{Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral divergence}
 
@@
 All code used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}.
-The external computer programs that we used were
-\begin{itemize} 
-\item \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2 (available at \url{github.com/jbloomlab/phydms}) to fit the ExpCM and GY94 models.
-\item \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7 (available at \url{https://github.com/sjspielman/pyvolve}) to simulate the sequences.
-\item \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8 (available at \url{https://github.com/bayesiancook/pbmpi}) to fit the pbMutSel model. 
-\item \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML}) to infer tree topology.
-\item We used \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, and \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo} for visualization of the results.
-\item \texttt{snakemake}~\citep{koster2012snakemake} version 3.11.2 (available at \url{https://snakemake.readthedocs.io/en/stable/}) to run the pipelines.
-\end{itemize}
+External computer programs used were \texttt{phydms}~\citep{hilton2017phydms} version 2.2.2, \texttt{pyvolve}~\citep{spielman2015pyvolve} version 0.8.7, \texttt{PhyloBayes-MPI}~\citep{rodrigue2014site} version 1.8, \texttt{RAxML}~\citep{stamatakis2006raxml} version 8.2.11 (available at \url{https://github.com/stamatak/standard-RAxML}), \texttt{ggplot2}~\citep{wickham2016ggplot2}, \texttt{ggtree}~\citep{yu2017ggtree}, \texttt{ggseqlogo}~\citep{wagih2017ggseqlogo}, and \texttt{snakemake}~\citep{koster2012snakemake} version 3.11.2.
 
@@
 \bibliographystyle{mbe}
+{\small
 \bibliography{references.bib}
+}
 
 \newpage
+\setcounter{page}{1}
+\renewcommand{\thepage}{S\arabic{page}}
 \section*{Supplemental Information}

### 2018-05-09 manuscript/manuscript.tex
added more information to computer programs and added github to acknowledgements

--- 2018-04-16 manuscript/manuscript.tex
@@
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+All code and data used for the analyses in this paper is available at https://github.com/jbloomlab/ divergence_timing_manuscript.
 

### 2018-05-09 manuscript/manuscript.tex
updated references

@@
 More recent work has used mutation-selection models to better account for purifying selection~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-These models explicitly incorporate the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression, le2008phylogenetic, si2008empirical, wang2008class}.
+These models explicitly incorporate the fact that different protein sites prefer different amino acids, and so can improve phylogenetic estimates when there are deep branches~\citep{philippe1998good, lartillot2007suppression, le2008phylogenetic, si2008empirical, wang2008class, susko2018accelerated}.
 However, these approaches require inferring the site-specific purifying selection from natural sequence data.
@@
 We used \texttt{PhyloBayes-MPI} program \texttt{readpb\_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences. 
-Convergence was assessed visually using Tracer (\url{http://tree.bio.ed.ac.uk/software/tracer/}) and by the correlation of amino-acid preferences inferred by two independent chains (r=0.996).
+Convergence was assessed visually using Tracer \citep{rambaut2018posterior} and by the correlation of amino-acid preferences inferred by two independent chains (r=0.996).
 
@@
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-All code and data used for the analyses in this paper is available at https://github.com/jbloomlab/ divergence_timing_manuscript.
+All code and data used for the analyses in this paper is available at \url{https://github.com/jbloomlab/ divergence_timing_manuscript}.
 

### 2018-08-13 manuscript/reviewer_response.tex
cleaned up manuscript directory and started reviewer response

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb}
\usepackage{hyperref}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
\newcommand{\response}[1]{{\color{black}#1}}
\newcommand\skhcomment[1]{{\color{magenta}[#1]}}

\title{Response to reviews of ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral divergence'' for \textit{Virus Evolution}}
\author{Sarah K. Hilton and Jesse D. Bloom}

\begin{document}
\maketitle

\emph{Below, the reviewer comments {\color{blue} are in blue}, and our responses are in black.}
\skhcomment{Change title?}
\skhcomment{Here is what we have to do for submission/revision: We revise the manuscript by clicking ``Create a Revision" under ``Action" in on the author page of the submission site. We cannot make revisions to the original submitted version and instead must upload a new file. We should highlight changes using colored or boldface text. We can respond to author comments in a submission box. Try to be specific in the response to the reviewers. Our review is due on October 9th.}
\skhcomment{I don't know if we can upload this latex document but we can at least write out the responses here.}

\color{blue}

\subsection*{Reviewer \#1 Comments}

\subsubsection*{Comments:} 
This is a very interesting manuscript about the consequences of not modeling site-specific amino-acid preferences when estimating sequence divergence (branch lengths in a phylogeny). 
What the Authors are pointing out in this article is clearly a major conceptual and methodological problem in practical phylogenetic analyses of viral sequences, and I think it is really important to raise the awareness of people interested in viral evolution about this problem. 
The explanations given by the Authors about the issue and about the methods that can be used to address it are cristal clear, with nice simulations and convincing analyses of empirical sequence data. 
For those reasons, I strongly recommend this manuscript. 
I would have only very minor comments and suggestions:

\response{}

\subsubsection*{Minor points}

(1) The title is slightly misleading: it seems to suggest that the article introduces a molecular dating method accounting for site-specific amino-acid preferences, which it does not do (no problem here: developing such a method would require substantial software development, way beyond the scope of this article, so this is not the question) ? but then, perhaps the title should be a bit more explicit about the exact content. 
What about:  ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral sequence divergence" 
\response{}

(2) different codons for the same protein are treated as selectively equivalent: different codons for the same amino-acid? 
\response{}

(3) An alternative strategy of obtaining the amino-acid preference parameters via Bayesian inference (Lartillot and Philippe, 2004; Rodrigue and Lartillot, 2014) is discussed in the last section of the Results. 
Perhaps could be rephrased: an alternative strategy for dealing with amino-acid preferences across sites is to consider them formally, not as parameters, but as random-effects across sites, and to infer them using Bayesian inference.
\response{}

(4) when explaining the saturation problem, it could be useful to discriminate between sequence identity and sequence divergence, e.g.: 
a substitution model transforms sequence identity into branch length. I would say: ``a substitution model transforms sequence *divergence* into branch length." 
similarly: 
``the transformation of sequence *divergence* to branch length is trivial when sequence identity is high". 
\response{}

(5) The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of *this type of* reversible substitution models. 
\response{}

\subsection*{Reviewer \#2 Comments}

\subsubsection*{Comments:} 
In this paper, Hilton and Bloom examine the performace of substitution models that explicitly incorporate experimental information about purifying selection. 
They find that they can estimate much longer branches in deep sections of phylogenetic trees, compared to traditional substitution models, and those that account for purifying selection (such as those implemented in PAML and HyPhy). 
I find that this is an interesting and important piece of work that is well written. 
Indeed, I find that the introduction is illuminating in explaining the difference between substitution models for protein coding data. 
I am happy to recommend this paper for publication, but I found some key aspects were neglected that can be improved with simple analyses and some clarifications in the text: 
\response{}

\subsubsection*{Minor points:}

In page 9 it is explained that a fixed tree topology was used for the subsequent analyses. 
I certainly agree with the authors that the model used in RAxML to estimate this tree probably underestimates branch lengths and that topological errors are negligible, but can they comment on how this could be tested using the ExpCM or codon models, given that they require substantially more computation? 
In other words, how suitable are these models to infer the tree topology, in addition to branch lengths? 
It is likely that most virus phylogenetic analyses can estimate a reasonable topology, but what about cases where there are very short branches next to very long branches (Felsenstein zone)? 
I do not recommend additional analyses for this part, but I think that discussing this would be very useful. 
\response{}

One key point in this study, which the authors acknowledge, is that estimating branch lengths accurately is key to infer evolutionary timescales. 
In this respect, is it possible to show for the simulations and empirical data how a root-to-tip regression (i.e. TempEst/pathogen) differs between the different substitution models? 
I would imagine that using more realistic models leads to better clocklike behaviour. 
However, it may be that a time-dependent rate phenomenon will be clearer, such that deeper branches will exhibit a predictably slower rate than those that are more recent. 
\response{}

Finally, the authors refer to the adequacy of substitution models. 
There has been some work in this field, also known as absolute model fit. 
This raises the question as to whether it is possible and valuable to assess the adequacy of these models, for example, using posterior predictive methods. 
Moreover, can the authors comment on the adequacy of the models described here for the flu empirical data? 
\response{}

\end{document}

### 2018-08-15 manuscript/manuscript.tex
fixed GitHub typo
first round of reviewer responses

@@
 
-\title{Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral divergence}
+\title{Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral sequence divergence}
 
@@
 Mutation-selection models account for the fact that purifying selection depends idiosyncratically on the specific amino-acid mutation at each site~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same protein are treated as selectively equivalent).
+\textcolor{red}{Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same amino acid are treated as selectively equivalent)}.
 Such models explicitly define a different set of amino-acid preferences at each site in the protein. 
@@
 Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters is similar to a non-site-specific substitution model.
-An alternative strategy of obtaining the amino-acid preference parameters via Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site} is discussed in the last section of the Results.
+\textcolor{red}{An alternative strategy to deal with site-specific amino-acid preferences is to consider them formally random-effects across sites rather than parameters, and to infer them using Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site}.
+This strategy is discussed in the last section of the Results.}
 
@@
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-All code and data used for the analyses in this paper is available at \url{https://github.com/jbloomlab/ divergence_timing_manuscript}.
+All code and data used for the analyses in this paper is available at \url{https://github.com/jbloomlab/divergence_timing_manuscript}.
 

### 2018-08-17 manuscript/reviewer_response.tex
next round of reviewer responses

@@
 \usepackage{hyperref} 
+\usepackage[round,semicolon]{natbib}
 
@@
 
-\response{}
+\response{Thank you for the excellent summary of our work. 
+We appreciate the favorable evaluation of the clarity of the work, which was furthered improved by the suggestions below.}
 
@@
 What about:  ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral sequence divergence" 
-\response{}
+\response{This is a good suggestion. 
+We have updated the title to ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral sequence divergence"}
 
 (2) different codons for the same protein are treated as selectively equivalent: different codons for the same amino-acid? 
-\response{}
+\response{Thank you for pointing out this typo. 
+We have updated the text to read ``Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same amino acid are treated as selectively equivalent)"}
 
@@
 Perhaps could be rephrased: an alternative strategy for dealing with amino-acid preferences across sites is to consider them formally, not as parameters, but as random-effects across sites, and to infer them using Bayesian inference.
-\response{}
+\response{This is a good suggestion. We have updated the text accordingly.}
 
@@
 ``the transformation of sequence *divergence* to branch length is trivial when sequence identity is high". 
-\response{}
+\response{We have updated the text to reflect the difference between sequence identity and sequence divergence.}
 
 (5) The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of *this type of* reversible substitution models. 
-\response{}
+\response{This is true. We have clarified the text as suggested.}
 
@@
 \subsubsection*{Comments:} 
-In this paper, Hilton and Bloom examine the performace of substitution models that explicitly incorporate experimental information about purifying selection. 
+In this paper, Hilton and Bloom examine the performance of substitution models that explicitly incorporate experimental information about purifying selection. 
 They find that they can estimate much longer branches in deep sections of phylogenetic trees, compared to traditional substitution models, and those that account for purifying selection (such as those implemented in PAML and HyPhy).
@@
 I am happy to recommend this paper for publication, but I found some key aspects were neglected that can be improved with simple analyses and some clarifications in the text: 
-\response{}
+\response{Thank you for the accurate and fair summary of our work.}
 
@@
 I do not recommend additional analyses for this part, but I think that discussing this would be very useful. 
-\response{}
+\response{This is an interesting and pertinent point. 
+As the reviewer noted, we do not currently have a way of directly testing the effect of ExpCMs on tree topology inference. 
+However, the modular structure of the HA tree, along with sequence annotation, gives us confidence that the nucleotide model in \texttt{RAxML} produces a reasonable tree. 
+But our confidence in the topology inferred by the nucleotide model is certainly not universal. 
+We have updated the text both to underscore why we believe this specific scenario gives us the correct tree. 
+We have also commented on work others have done to show that site-specific amino-acid preferences may infer more accurate tree topologies, specifically in the Felsenstein Zone~\citep{lartillot2007suppression}.
+Given this work and some leeway for speculation, we would hypothesize that the ExpCMs would be less sensitive to the Felsenstein Zone than uniform stationary state codon models.
+But testing this hypothesis is outside the scope of this paper. 
+}
 
@@
 However, it may be that a time-dependent rate phenomenon will be clearer, such that deeper branches will exhibit a predictably slower rate than those that are more recent. 
-\response{}
+\response{This is an interesting suggestion. 
+It is certainly true that a clock-like behavior is necessary in order to estimate evolutionary timescales. 
+However, we do not believe that analyses which quantify clock-like behavior are appropriate for our datasets. 
+The HA sequences we used our from several different hosts. 
+It has been shown that these hosts have their own, specific evolutionary rate~\citep{worobey2014synchronized} but these differences are \textit{not} accounted for in our model. 
+Therefore, while the HA tree is a useful test case to examine branch length estimation, we do not believe that our model will improve the clock-like behavior of the sequences. 
+}
 
@@
 Moreover, can the authors comment on the adequacy of the models described here for the flu empirical data? 
-\response{}
+\response{
+Model adequacy is indeed a very interesting area of future research and we are agree this area of phylogenetics deserves more attention. 
+Up until recently, the majority of model adequacy tests, including posterior-predictive methods, used the multinomial test~\citep{goldman1993statistical, brown2018evaluating}. 
+We do not believe that this test is the most appropriate test for our model because it does not test site-specificity directly. 
+Therefore, while we believe it is possible to assess the adequacy of these models, development of the most appropriate test statistic is outside the scope of this paper. 
+While we agree that evaluating the adequacy of the model would be important when estimating evolutionary timescales, it is not necessary to test the specific question of this paper --- do models with site-specific stationary states estimate longer branches than models with uniform stationary states?
+}
+
+\clearpage 
+\color{black}
+\bibliographystyle{mbe}
+{\small
+\bibliography{references.bib}
+}
 

### 2018-08-17 manuscript/manuscript.tex
next round of reviewer responses

@@
 Mutation-selection models account for the fact that purifying selection depends idiosyncratically on the specific amino-acid mutation at each site~\citep{halpern1998evolutionary,yang2008mutation,rodrigue2010mutation,tamuri2012estimating,mccandlish2014modeling}.
-\textcolor{red}{Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same amino acid are treated as selectively equivalent)}.
+Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same amino acid are treated as selectively equivalent).
 Such models explicitly define a different set of amino-acid preferences at each site in the protein. 
@@
 Because the amino-acid preference parameters in an ExpCM are obtained from experiments, the number of ExpCM free parameters is similar to a non-site-specific substitution model.
-\textcolor{red}{An alternative strategy to deal with site-specific amino-acid preferences is to consider them formally random-effects across sites rather than parameters, and to infer them using Bayesian inference~\citep{lartillot2004bayesian, rodrigue2014site}.
-This strategy is discussed in the last section of the Results.}
+An alternative strategy to account for site-specific amino-acid preferences is to formally consider them as random effects across sites, rather than parameters, and infer them using Bayesian methods~\citep{lartillot2004bayesian, rodrigue2014site}.
+This strategy is discussed in the last section of the Results.
 
@@
 \subsection*{Effect of stationary state and rate variation on branch-length estimation}
-Given a single branch, a substitution model transforms sequence identity into branch length.
+Given a single branch, a substitution model transforms sequence divergence into branch length.
 Under a molecular-clock assumption, this branch length is proportional to time.
-The transformation from sequence identity to branch length is trivial when the sequence identity is high.
+The transformation from sequence divergence to branch length is trivial when the sequence identity is high.
 For instance, when there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
 However, as substitutions accumulate it becomes progressively more likely for multiple changes to occur at the same site.
-In this regime, the accuracy of the substitution model becomes critical for transforming sequence identity into branch length.
+In this regime, the accuracy of the substitution model becomes critical for transforming sequence divergence into branch length.
 Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic amino-acid sequence identity.
@@
 Specifically, this GY94+$\omega_r$ model predicts that different sites will approach the asymptote at different rates, but the asymptote is always the same (\ref{fig:decay}).
-The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of reversible substitution models.
+The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of this type of reversible substitution models.
 These models are reversible stochastic matrices, which can be decomposed into stationary states and symmetric exchangeability matrices~\citep{nielsen2006statistical}.
@@
 Although the nucleotide model used with \texttt{RAxML} to infer this tree topology is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
+This confidence in the \texttt{RAxML}(GTRCAT)-inferred tree is certainly not universal and in other cases more complex, biologically-realistic models may be necessary for accurate topology inference. 
+For instance, others have shown that models with site-specific amino-acid preferences can avoid topological errors such as long-branch attraction~\citep{lartillot2007suppression}.
 

### 2018-08-17 manuscript/reviewer_response.tex
Jesse updates to reviewer response / edits

--- 2018-08-13 manuscript/reviewer_response.tex
@@
 
-\title{Response to reviews of ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral divergence'' for \textit{Virus Evolution}}
+\title{Response to reviews for \textit{Virus Evolution}}
 \author{Sarah K. Hilton and Jesse D. Bloom}
@@
 \emph{Below, the reviewer comments {\color{blue} are in blue}, and our responses are in black.}
-\skhcomment{Change title?}
-\skhcomment{Here is what we have to do for submission/revision: We revise the manuscript by clicking ``Create a Revision" under ``Action" in on the author page of the submission site.
-We cannot make revisions to the original submitted version and instead must upload a new file.
-We should highlight changes using colored or boldface text.
-We can respond to author comments in a submission box.
-Try to be specific in the response to the reviewers.
-Our review is due on October 9th.}
-\skhcomment{I don't know if we can upload this latex document but we can at least write out the responses here.}
 
 \color{blue}
+
+\subsection*{Editor Comments}
+Many thanks for submitting your manuscript "Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral divergence" to \textit{Virus Evolution}.
+My apologies that the review process took a little longer than usual.
+
+Your work was considered by Section Editor Aris Katzourakis, with the support of two expert peer reviewers.
+Their comments are included below.
+All conclude that your manuscript represents valuable progress on an interesting and important question for virus evolution.
+I agree.
+
+We would therefore like to invite you to submit a minor revision that addresses the comments raised in the reviews.
+
+Thanks again for submitting to \textit{Virus Evolution} and we look forward to receiving your revision.
+
+\response{Thank you for the reviews.
+We were pleased that both reviewers liked the manuscript, and have submitted a revised version that includes the minor changes that they suggested.}
+
+\subsection*{Section Editor (Aris Katzourakis) Comments}
+
+Thank you for your submission to \textit{Virus Evolution}.
+Your manuscript has now been evaluated by two independent reviewers both of whom found it interesting and engaging, and recommend publication.
+Both reviewers have some minor and easily addressable comments for clarity and presentation; please consider these and let me know how you have incorporated them into the manuscript.
+None of the comments require additional analysis nor should require extensive re-reviewing of the manuscript.
+
+\response{The minor comments by the reviewers were both helpful and straightforward to address.
+In this re-submission, we have made revisions that address these comments.
+This response details those changes, and we have also uploaded a version of the manuscript with all changes tracked.}
 
@@
 
-\response{}
+\response{Thank you for the excellent summary of our work. 
+We appreciate the kind comments about the clarity of the work, which has been furthered improved by incorporating the suggestions that you made below.}
 
@@
 What about:  ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral sequence divergence"
-\response{}
+
+\response{This is a good suggestion. 
+We have updated the title to ``Modeling site-specific amino-acid preferences deepens phylogenetic estimates of viral \emph{sequence} divergence" as suggested.}
 
 (2) different codons for the same protein are treated as selectively equivalent: different codons for the same amino-acid?
-\response{}
+
+\response{Thank you for pointing out this typo. 
+We have updated the text to read ``Here we will consider mutation-selection models where the site-specific selection is assumed to act solely at the protein level (different codons for the same \emph{amino acid} are treated as selectively equivalent)."}
 
@@
 Perhaps could be rephrased: an alternative strategy for dealing with amino-acid preferences across sites is to consider them formally, not as parameters, but as random-effects across sites, and to infer them using Bayesian inference.
-\response{}
+
+\response{This is a good suggestion. We have updated the text accordingly.}
 
 (4) when explaining the saturation problem, it could be useful to discriminate between sequence identity and sequence divergence, e.g.: 
-a substitution model transforms sequence identity into branch length. I would say: ``a substitution model transforms sequence *divergence* into branch length." 
+a substitution model transforms sequence identity into branch length. I would say: ``a substitution model transforms sequence \emph{divergence} into branch length." 
 similarly: 
-``the transformation of sequence *divergence* to branch length is trivial when sequence identity is high". 
-\response{}
+``the transformation of sequence \emph{divergence} to branch length is trivial when sequence identity is high".
 
-(5) The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of *this type of* reversible substitution models. 
-\response{}
+\response{We have updated the text to refer to sequence \emph{divergence} rather than identity in this context, since we agree that this is a clearer way to explain the fundamental problem.}
+
+(5) The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of \emph{this type of} reversible substitution models.
+
+\response{This is a good point. 
+We have revised the text as suggested to indicate that the statement applies to the types of substitution models discussed in the sentences preceding the statement in question.}
 
@@
 I am happy to recommend this paper for publication, but I found some key aspects were neglected that can be improved with simple analyses and some clarifications in the text:
-\response{}
+
+\response{Thank you for the accurate and fair summary of our work.}
 
@@
 I do not recommend additional analyses for this part, but I think that discussing this would be very useful.
-\response{}
+
+\response{This is a highly relevant point when thinking about the larger context of our work.
+
+As the reviewer noted, our current work only assesses how the site-specific models affect branch length, but not the topology of the tree itself. 
+For the specific HA phylogenetic tree analyzed in our paper, the different models seem very unlikely to substantially affect tree topology since the tree is clearly divided into widely separated clades (a ``modular'' tree structure), meaning that event the nucleotide model in \texttt{RAxML} produces a reasonable tree.
+We have updated the text to more clearly explain why we think that the substitution model is unlikely to have much affect on topology \emph{for the specific HA tree that we analyze}.
+
+However, as the reviewer notes, for other trees the model may markedly affect the topology as well as the branch lengths.
+We now describe this point in detail, and refer to other work showing that models can affect the accuracy of the tree topology, particularly in the Felsenstein Zone~\citep{lartillot2007suppression}. 
+Given this work and some leeway for speculation, we would hypothesize that the ExpCMs would be less sensitive to the Felsenstein Zone than uniform stationary state codon models.
+However, directly testing this is beyond the scope of the current paper, since it requires the non-trivial computational implementation of software that can use the ExpCMs and estimate tree topology.
+We have therefore taken the reviewer's suggestion and added careful discussion of this issue.
+}
 
@@
 However, it may be that a time-dependent rate phenomenon will be clearer, such that deeper branches will exhibit a predictably slower rate than those that are more recent.
-\response{}
+
+\response{This is an interesting suggestion. 
+It is certainly true that a clock-like behavior is necessary in order to estimate evolutionary timescales. 
+However, we do not believe that root-to-tip regression is informative when applied to our datasets. 
+The reason is that the HA sequences we use are from influenza viruses from several different hosts, such as humans, pigs, and birds. 
+It has been shown that these hosts each have their own specific evolutionary rate~\citep{worobey2014synchronized} (e.g., HA evolves \emph{much} faster in humans than birds, probably due to immune selection).
+However, our current implementation does \emph{not} incorporate host-specific rates, which therefore will confound clocklike behavior.
+We agree with the reviewer that in principle accounting for the substitution model issues discussed in our manuscript in conjunction with the host-specific rate issues discussed by others should improve clocklike behavior.
+Since there is not currently software that does both of these, are are unable to actually do this.
+However, we have added mention of this point in the Discussion.
+}
 
@@
 Moreover, can the authors comment on the adequacy of the models described here for the flu empirical data?
-\response{}
+
+\response{
+Model adequacy is indeed a very interesting area of future research and we fully agree this area of phylogenetics deserves more attention. 
+However, most model adequacy tests, including posterior-predictive methods, typically use the multinomial test~\citep{goldman1993statistical, brown2018evaluating}. 
+However, the multinomial test itself is not applicable to the models we use.
+The reason is that this test is explicitly designed for substitution models that are i.i.d.
+(independent and identically distributed) as stated by \citet{goldman1993statistical} in his original introduction of these tests (second paragraph of page 184 of his paper).
+But the models we use are not identically distributed, since there is a different stationary state for each site.
+Therefore, the multinomial test cannot be used to evaluate model adequacy.
+
+The general framework of adequacy tests should still be applicable to the models we use.
+However, new specific implementations other than the established multinomial test will be necessary to assess the adequacy of these models.
+We are indeed interested in developing computationally feasible implementations of model adequacy tests for codon models with site-specific stationary states, but have not yet succeeded in doing so.
+Therefore, we are not yet in a position to address this important question.
+So the best we can do for now is briefly mention the topic of model adequacy in the text. 
+}
+
+\clearpage 
+\color{black}
+\bibliographystyle{mbe}
+{\small
+\bibliography{references.bib}
+}
 

### 2018-08-17 manuscript/manuscript.tex
Jesse updates to reviewer response / edits

@@
 Specifically, this GY94+$\omega_r$ model predicts that different sites will approach the asymptote at different rates, but the asymptote is always the same (\ref{fig:decay}).
-The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of this type of reversible substitution models.
+The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of this type of reversible substitution model.
 These models are reversible stochastic matrices, which can be decomposed into stationary states and symmetric exchangeability matrices~\citep{nielsen2006statistical}.
@@
 For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
-Although the nucleotide model used with \texttt{RAxML} to infer this tree topology is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
-This confidence in the \texttt{RAxML}(GTRCAT)-inferred tree is certainly not universal and in other cases more complex, biologically-realistic models may be necessary for accurate topology inference. 
-For instance, others have shown that models with site-specific amino-acid preferences can avoid topological errors such as long-branch attraction~\citep{lartillot2007suppression}.
+Although the nucleotide model used with \texttt{RAxML} to infer this tree topology is probably less accurate than codon models, the modular subtype structure of the HA phylogeny (the tree is clearly divided into widely separated clades) means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.
+We note that other genes may have phylogenetic structures that are more prone to topological uncertainty. 
+In such cases, the accuracy of the substitution model may be important for avoiding topological errors such as long-branch attraction~\citep{felsenstein1978cases,lartillot2007suppression}.
 
@@
 Combining either type of model with rate variation increases the inferred length of deep branches relative to models that only incorporate rate variation.
+We expect that further improvements could be achieved by also incorporating other factors such as host-specific substitution rates~\citep{worobey2014synchronized} that are known to be important for modeling the evolution of viral genes such as HA.
 
@@
 It appears that typical structural modeling programs are insufficient for this purpose~\citep{rodrigue2009computational}.
-But hope comes from experimental progress in measuring actual site-interdependent constraints on proteins~\citep{olson2014comprehensive,wu2016adaptation,steinberg2016shifting,li2016fitness}, combined with new methods for using these measurements to parameterize fitness functions~\citep{sailer2017detecting,otwinowski2018inferring,otwinowski2018protein}.
+But hope comes from experimental progress in measuring actual site-interdependent constraints on proteins~\citep{olson2014comprehensive,wu2016adaptation,steinberg2016shifting,li2016fitness}, combined with new methods for using these measurements to parameterize fitness functions~\citep{sailer2017detecting,otwinowski2018inferring}.
 Perhaps some day such truly realistic models might be useful for phylogenetic inference.

### 2018-08-17 manuscript/reviewer_response.tex
fixed small typos in reviewer response

@@
 As the reviewer noted, our current work only assesses how the site-specific models affect branch length, but not the topology of the tree itself. 
-For the specific HA phylogenetic tree analyzed in our paper, the different models seem very unlikely to substantially affect tree topology since the tree is clearly divided into widely separated clades (a ``modular'' tree structure), meaning that event the nucleotide model in \texttt{RAxML} produces a reasonable tree.
+For the specific HA phylogenetic tree analyzed in our paper, the different models seem very unlikely to substantially affect tree topology since the tree is clearly divided into widely separated clades (a ``modular'' tree structure), meaning that even the nucleotide model in \texttt{RAxML} produces a reasonable tree.
 We have updated the text to more clearly explain why we think that the substitution model is unlikely to have much affect on topology \emph{for the specific HA tree that we analyze}.
@@
 We agree with the reviewer that in principle accounting for the substitution model issues discussed in our manuscript in conjunction with the host-specific rate issues discussed by others should improve clocklike behavior.
-Since there is not currently software that does both of these, are are unable to actually do this.
+Since there is not currently software that does both of these, we are unable to actually do this.
 However, we have added mention of this point in the Discussion.

### 2018-09-22 manuscript/manuscript.tex
updated grants
renamed figure files for easy submission
created version for final submission

@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
-\caption{\label{fig:model_feature}
-\textbf{Different ways codon models account for purifying selection.}
-(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue line). 
-For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin (blue circles)~\citep{yang1994maximum,yang2000codon}. 
-A gamma distribution (denoted by $\Gamma$) with $K=4$ bins is shown here.
-(B) A substitution model's stationary state defines the expected sequence composition after a very long evolutionary time. 
-Most substitution models have stationary states that are uniform across sites.
-However, substitution models can have site-specific stationary states.
-In the logo plots, each column is a site in the protein and the height of each letter is the frequency of that amino acid at stationary state. 
-(C) Substitution models can incorporate neither, one, or both of these features.
-Here we will use substitution models from the Goldman-Yang~\citep[GY94;][]{goldman1994codon,yang2000codon} and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed $\omega$ to represent all possible combinations.
-}
-\end{figure}
-
 A single gene-wide $\omega$ ignores the fact that purifying selection is heterogeneous across sites.
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
-\caption{\label{fig:decay}
-\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
-The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as experimentally measured by~\citet{doud2016accurate}.
-The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
-For the GY94 model, the graphs are identical for all sites since this model does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
-The graphs do differ among sites if we calculate a different $\omega_r$ for each site $r$ in the GY94 model using the amino-acid preferences~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
-However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
-The ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
-}
-\end{figure}
-
 \ref{fig:decay} shows how different substitution models predict amino-acid sequence identity to decrease as a function of branch length using model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
@@
 Note that these simulations closely parallel those performed by \citet{halpern1998evolutionary} and \citet{wertheim2011purifying}.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.9\textwidth]{figures/simulations}}
-\caption{\label{fig:simulations}
-\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
-We simulated alignments along a HA phylogenetic tree using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
-We then inferred the branch lengths of this tree from the simulated alignments.
-The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
-We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
-The grey dashed line at $y=x$ represents what would be seen if the inferred branch lengths exactly matched those used in the simulations. 
-}
-\end{figure}
 
@@
 
-\begin{table}[t!]
-\caption{\label{tab:empirical_data}
-{\bf Fitting of substitution models to the HA phylogenetic tree.}
-All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
-The models fit here are the same ones in~\ref{fig:empirical_trees}. 
-The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
-All ExpCMs fit a stringency parameter $>1$.
-} 
-     \begin{tabular}{cccccccccc}
-        \hline
-         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & $\omega$ & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
-       	ExpCM  (H1+H3 avg) + $\Gamma\omega$  & 0 & -51083 & 0.19,  0.50,  0.91,  1.86 &  1.69\\
-	ExpCM (H1+H3 avg)  &  1063 & -51616 & 0.14 & 1.77\\
-	ExpCM  (H1) + $\Gamma\omega$  & 1321 & -51744  & 0.12 ,  0.42,  0.89,  2.13 & 1.11\\
-	ExpCM (H3) + $\Gamma\omega$ & 1777 & -51972 & 0.10,  0.36,  0.76,  1.84 & 1.28\\
-	ExpCM (H1) & 2670 & -52419 &  0.12 & 1.21\\
-	ExpCM (H3) &  3377 & -52773 & 0.12 & 1.43\\
-	GY94 + $\Gamma\omega$  & 4817 & -53487 & 0.00,  0.03,  0.08,  0.24 & - \\
-	GY94 & 7892 & -55025  & 0.07 & -\\
-      \end{tabular}
-\end{table}
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
-\caption{\label{fig:empirical_trees}
-\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
-The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
-The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
-The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
-Circle denotes the H1 clade and triangle denotes the H3 clade.
-The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
-This figure enables qualitative visualization of the trees; for a quantitative comparison of branch lengths optimized by different models, see \ref{fig:empirical_scatter}. 
-}
-\end{figure}
-
 We fit the GY94 model and each of the three ExpCMs to the fixed HA tree topology estimated using \texttt{RAxML}, and also tested a version of each model with $\Gamma\omega$ rate variation.
@@
 This more realistic null model gives ExpCMs enhanced power to detect diversifying selection for amino-acid change~\citep{bloom2017identification, rodrigue2017detecting}, which is known to occur at some sites in HA due to immune selection~\citep{bedford2014integrating}.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figures/empirical_scatter.pdf}}
-\caption{\label{fig:empirical_scatter}
-\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments extends branch lengths, especially for branches leading to the HA used in the experiment.}
-The points indicate the total length of branches separating all pairs of tips on the HA phylogenetic tree when the tree is optimized under the indicated model. 
-Blue and orange denote branches that lead to the H1 and H3 HAs used in the deep mutational scanning. 
-The amino-acid preference set defining the ExpCM is labeled above each each plot.
-(A) ExpCMs defined by amino-acid preferences from any of the deep mutational scanning experiments estimate generally longer branches than the GY94 model, with the increase particularly pronounced for branches leading to the HA used in the experiment.
-(B) The addition of $\Gamma\omega$ rate variation further extends branch lengths, without any apparent bias towards the HAs used in the experiment.
-Note that this figure shows the same data as \ref{fig:empirical_trees} in a different form.
-}
-\end{figure}
 
@@
 \subsection*{Shifting amino-acid preferences limit the benefits of models with site-specific stationary states for estimating long branch lengths.}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figures/compete}}
-\caption{\label{fig:compete}
-\textbf{The congruence between natural selection and the deep mutational scanning measurements decreases with sequence divergence.} 
-We fit an ExpCM informed by the H1 or H3 deep mutational scanning experiments to trees spanning sequences with low, intermediate, and high divergence from the sequence used in the experiment. 
-The ExpCM stringency parameter ($\beta$) is a measure of the congruence between natural selection and the experimental measurements~\citep{bloom2014informed,hilton2017phydms}. 
-Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with greater stringency. 
-As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases and the amino-acid preference ``flatten."
-Therefore, the preferences measured in each experiment are progressively less congruent with natural selection as we include increasingly diverged sequences. 
-}
-\end{figure}
 
@@
 \subsection*{A model with amino-acid preferences estimated from natural sequences gives similar results to an ExpCM}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.99\textwidth]{figures/phylobayes.pdf}}
-\caption{\label{fig:phylobayes}
-\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
-We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
-The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}.
-(A) The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52).
-(B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
-Points denote branch lengths between all pairs of tips on the tree. 
-Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
-The \texttt{phydms} program implementing ExpCMs and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree.
-}
-\end{figure}
 
@@
 SKH is supported in part by training grant T32AI083203 from the NIAID of the National Institutes of Health.
-This work was supported by the NIAID and NIGMS of the NIH under grant numbers R01AI127893 and R01GM102198. 
+This work was supported by the NIAID and NIGMS of the NIH under grant numbers R01AI127893 and R35GM126911. 
 JDB is supported in part by a Faculty Scholars grant from the Howard Hughes Medical Institute and the Simons Foundation.
@@
 \newpage 
+%% figures 
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.90\textwidth]{fig1.pdf}}
+\caption{\label{fig:model_feature}
+\textbf{Different ways codon models account for purifying selection.}
+(A) The dN/dS parameter, $\omega$, can be defined as one gene-wide average (orange triangle) or allowed to vary according to some statistical distribution (blue line). 
+For computational tractability, the distribution is discretized into $K$ bins and $\omega$ takes on the mean of each bin (blue circles)~\citep{yang1994maximum,yang2000codon}. 
+A gamma distribution (denoted by $\Gamma$) with $K=4$ bins is shown here.
+(B) A substitution model's stationary state defines the expected sequence composition after a very long evolutionary time. 
+Most substitution models have stationary states that are uniform across sites.
+However, substitution models can have site-specific stationary states.
+In the logo plots, each column is a site in the protein and the height of each letter is the frequency of that amino acid at stationary state. 
+(C) Substitution models can incorporate neither, one, or both of these features.
+Here we will use substitution models from the Goldman-Yang~\citep[GY94;][]{goldman1994codon,yang2000codon} and experimentally informed codon model~\citep[ExpCM;][]{hilton2017phydms} families with and without gamma-distributed $\omega$ to represent all possible combinations.
+}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.90\textwidth]{fig2.pdf}}
+\caption{\label{fig:decay}
+\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
+The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as experimentally measured by~\citet{doud2016accurate}.
+The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
+For the GY94 model, the graphs are identical for all sites since this model does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
+The graphs do differ among sites if we calculate a different $\omega_r$ for each site $r$ in the GY94 model using the amino-acid preferences~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
+However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
+The ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
+}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.9\textwidth]{fig3.pdf}}
+\caption{\label{fig:simulations}
+\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
+We simulated alignments along a HA phylogenetic tree using an ExpCM parameterized by the actual site-specific amino-acid preferences for an H1 HA~\citep{doud2016accurate}.
+We then inferred the branch lengths of this tree from the simulated alignments.
+The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
+We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
+The grey dashed line at $y=x$ represents what would be seen if the inferred branch lengths exactly matched those used in the simulations. 
+}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+\centerline{\includegraphics[width=\textwidth]{fig4.pdf}}
+\caption{\label{fig:empirical_trees}
+\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
+The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
+The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}.
+The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
+Circle denotes the H1 clade and triangle denotes the H3 clade.
+The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
+This figure enables qualitative visualization of the trees; for a quantitative comparison of branch lengths optimized by different models, see \ref{fig:empirical_scatter}. 
+}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+\centerline{\includegraphics[width=\textwidth]{fig5.pdf}}
+\caption{\label{fig:empirical_scatter}
+\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments extends branch lengths, especially for branches leading to the HA used in the experiment.}
+The points indicate the total length of branches separating all pairs of tips on the HA phylogenetic tree when the tree is optimized under the indicated model. 
+Blue and orange denote branches that lead to the H1 and H3 HAs used in the deep mutational scanning. 
+The amino-acid preference set defining the ExpCM is labeled above each each plot.
+(A) ExpCMs defined by amino-acid preferences from any of the deep mutational scanning experiments estimate generally longer branches than the GY94 model, with the increase particularly pronounced for branches leading to the HA used in the experiment.
+(B) The addition of $\Gamma\omega$ rate variation further extends branch lengths, without any apparent bias towards the HAs used in the experiment.
+Note that this figure shows the same data as \ref{fig:empirical_trees} in a different form.
+}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.75\textwidth]{fig6.pdf}}
+\caption{\label{fig:compete}
+\textbf{The congruence between natural selection and the deep mutational scanning measurements decreases with sequence divergence.} 
+We fit an ExpCM informed by the H1 or H3 deep mutational scanning experiments to trees spanning sequences with low, intermediate, and high divergence from the sequence used in the experiment. 
+The ExpCM stringency parameter ($\beta$) is a measure of the congruence between natural selection and the experimental measurements~\citep{bloom2014informed,hilton2017phydms}. 
+Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with greater stringency. 
+As divergence increases between the HA used in the experiment and the other sequences in the tree, the $\beta$ value decreases and the amino-acid preference ``flatten."
+Therefore, the preferences measured in each experiment are progressively less congruent with natural selection as we include increasingly diverged sequences. 
+}
+\end{figure}
+\newpage
+\begin{figure}[h!]
+\centerline{\includegraphics[width=0.99\textwidth]{fig7.pdf}}
+\caption{\label{fig:phylobayes}
+\textbf{Models inferred from natural sequences have similar stationary states to models defined by experimental preferences and estimate similar branch lengths.}
+We fit an ExpCM(H1+H3 avg)+$\Gamma\omega$ and a pbMutSel to the full HA tree in \ref{fig:empirical_trees}. 
+The pbMutSel amino-acid preferences are inferred from the natural HA sequences, while the ExpCM amino-acid preferences are experimentally measured and then rescaled by the stringency parameter in \ref{tab:empirical_data}.
+(A) The pbMutSel preferences are more correlated with the re-scaled average of the H1 and H3 deep mutational scanning preferences than the individual re-scaled H1 and H3 deep mutational scanning preferences are to each other (Pearson's $r$: 0.74 versus 0.52).
+(B) The ExpCM(H1+H3 avg)+$\Gamma\omega$ and pbMutSel models estimated similar branch lengths when fit to the entire HA tree. 
+Points denote branch lengths between all pairs of tips on the tree. 
+Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
+The \texttt{phydms} program implementing ExpCMs and the \texttt{PhyloBayes-MPI} program implementing pbMutSel models give branch lengths in different units, so to facilitate direct comparison between the models, we have normalized all branch lengths returned by each program by the length of the branches separating the earliest (A/South Carolina/1918) and latest (A/Solomon Islands/2006) seasonal human H1 sequences on the tree.
+}
+\end{figure}
+
+%% tables
+\newpage
+\begin{table}[t!]
+\caption{\label{tab:empirical_data}
+{\bf Fitting of substitution models to the HA phylogenetic tree.}
+All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
+The models fit here are the same ones in~\ref{fig:empirical_trees}. 
+The $\omega$ value for each of the $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation. 
+All ExpCMs fit a stringency parameter $>1$.
+} 
+     \begin{tabular}{cccccccccc}
+        \hline
+         Model & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & $\omega$ & {\shortstack{Stringency\\ parameter ($\beta$)}}\\ \hline
+       	ExpCM  (H1+H3 avg) + $\Gamma\omega$  & 0 & -51083 & 0.19,  0.50,  0.91,  1.86 &  1.69\\
+	ExpCM (H1+H3 avg)  &  1063 & -51616 & 0.14 & 1.77\\
+	ExpCM  (H1) + $\Gamma\omega$  & 1321 & -51744  & 0.12 ,  0.42,  0.89,  2.13 & 1.11\\
+	ExpCM (H3) + $\Gamma\omega$ & 1777 & -51972 & 0.10,  0.36,  0.76,  1.84 & 1.28\\
+	ExpCM (H1) & 2670 & -52419 &  0.12 & 1.21\\
+	ExpCM (H3) &  3377 & -52773 & 0.12 & 1.43\\
+	GY94 + $\Gamma\omega$  & 4817 & -53487 & 0.00,  0.03,  0.08,  0.24 & - \\
+	GY94 & 7892 & -55025  & 0.07 & -\\
+      \end{tabular}
+\end{table}
+
+\clearpage
+\newpage
 \setcounter{page}{1}
@@
 \begin{suppfig}[H]
-\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
+\centerline{\includegraphics[width=\textwidth]{suppfig1.pdf}}
 \caption{\label{suppfig:prefs_doud}
@@
 \begin{suppfig}[H]
-\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
+\centerline{\includegraphics[width=\textwidth]{suppfig2.pdf}}
 \caption{\label{suppfig:prefs_lee}
@@
 \begin{suppfig}[H]
-\centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
+\centerline{\includegraphics[width=\textwidth]{suppfig3.pdf}}
 \caption{\label{suppfig:prefs_average}
@@
 \begin{suppfig}[H]
-\centerline{\includegraphics[width=\textwidth]{figures/prefs_mutSel}}
+\centerline{\includegraphics[width=\textwidth]{suppfig4.pdf}}
 \caption{\label{suppfig:prefs_mutSel}
@@
 \begin{suppfig}[H]
-\centerline{\includegraphics[width=0.50\textwidth]{figures/divergence_distances.pdf}}
+\centerline{\includegraphics[width=0.50\textwidth]{suppfig5.pdf}}
 \caption{\label{suppfig:subalignments}
